### 2022-06-30 submission/manuscript.Rmd
Create initial outline of manuscript

---
output:
  word_document: default
  pdf_document:
    keep_tex: true
geometry: margin=1.0in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \linenumbers
  - \raggedright
editor_options:
  chunk_output_type: console
bibliography: references.bib
csl: asm.cls
---

```{r, echo=FALSE}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(here))

opts_chunk$set(echo = FALSE)

inline_hook <- function(x){

	if(is.numeric(x)){
		if(abs(x - round(x)) < .Machine$double.eps){
			# treat as integer
			formatted <- format(x, digits=0, big.mark = ",")

		} else {
			# treat as floating point number
			formatted <- format(x, digits=2, nsmall=2, big.mark = ",")
		}
	} else {
		formatted <-x
	}

  formatted

}

knit_hooks$set(inline=inline_hook)

format_pct <- function(x, my_digits=1){

	format(x, digits=my_digits, nsmall=my_digits)

}
```

# Rarefy your data

\vspace{20mm}

**Running title:** Rarefy your data

\vspace{20mm}

Patrick D. Schloss${^\dagger}$

\vspace{40mm}

${\dagger}$ To whom corresponsdence should be addressed:

\href{mailto:pschloss@umich.edu}{pschloss@umich.edu}

Department of Microbiology & Immunology

University of Michigan

Ann Arbor, MI 48109

\vspace{20mm}

**Research article**

\newpage

## Abstract

## Importance

\newpage

## Introduction

* What is rarefaction? History, reason for rarefaction
  * Repeated down sampling of datasets to a common number of observations to calculate the average value to ascertain the expected value of a metric for the metric under study; typically richness
  * Control for unneven sampling effort
  * Methods vary in their sensitivity to uneven sampling
* Reasons behind "rarefaction is inadmissable"
  * Weird simulation
* Alternative approaches and claims
  * sampling invariance
* Goal of this study

## Results

### Datasets
* Represented diverse communities
* High quality V4 sequence data that consisted of paired 250 nt reads
* Wide distribution in the number of reads within each dataset (Figure S1, Table 1)

### Null models
* Null models were samplings, without replacement, of all samples from a study that have been pooled together and sampled to the same depth of sampling
* Created 100 random sets of samples per dataset

* Would expect a zero correlation between alpha diversity metric and number of sequences in the community
* Rarefaction was the only approach that resulted in a near-zero correlation (Figure 1)

* Would expect a zero correlation between beta diversity metrics and the difference in the number of sequences in the samples going into the distance calculation
* Rarefaction was the only approach that resulted in a near-zero correlation (Figure 2)

### Result when sampling effort is confounded with treatment group
* Used Null model
* Created two treatment groups for each study by dividing samples based on whether they were above or below the median number of sequences per sample

* Used Wilcoxon test to detect differences between treatment groups for alpha diversity, would expect ~5% of random sets to be significant
* Rarefaction the only approach that reliably resulted in the expected Type I error rate

* Used AMOVA test to detect differences between treatment groups for beta diversity, would expect ~5% of random sets to be significant
* Rarefaction the only approach that reliably resulted in the expected Type I error rate

### Result when known effect size imposed between two groups of samples
* For each dataset I divided samples into two groups. The probability distribution for the first group was determined by the null model and the second was determined by xxxxxxxxxxx
* These parameters were selected because they produced results were some of the simulations for the rarefied data did not result in a significant test
* Fraction of tests that were significant would be a measure of statistical power

* Used Wilcoxon test to detect differences between treatment groups for alpha diversity
* All tests resulted in reduced power to detect differences relative to the power of using rarefaction

* Used AMOVA test to detect differences between treatment groups for beta diversity
* All tests resulted in reduced power to detect differences relative to the power of using rarefaction

* This simulation would work well for measures based on the structure of the community, but lacked resolution for those based on the richness or membership of commmunities. For an additional simulation, the first treatment group followed a null distribution for the second, XX% of OTUs were removed from the distribution
* Tests for richness and membership (i.e. Jaccard) resutled in reduced power relative to rarefaction

### Effect of sampling depth on metrics
* Simulated rarefying null model datasets to retain varying percentage of samples
* High sampling depth resulted in less variation between samples for alpha and beta diversity

* Good's coverage of observed data (take from raw data files)

### Re-analysis of previously published datasets
* Human study: Did not alter effect size or significance of alpha or beta diversity, but did result in reduced effect sizes for measures of richness and non-parametric estimators of richness; breakaway detected a difference

## Discussion

* Rarefy your data
* Problems with recommended methods...
  * Many recommended methods are borrowed from gene expression analysis
  * Meaning of zeroes in data - structural vs. below limit of detection
* Factors that determine what number of sequences to rarefy to
* Need better methods of pooling libraries that result in more even distribution of sequences across samples
* Rarefy your data

## Materials and Methods

**Data availability.**

**Reproducible data analysis.**

\vspace{10mm}

**Acknowledgements.**

\newpage

## References

\setlength{\parindent}{-0.25in}
\setlength{\leftskip}{0.25in}
\noindent

<div id="refs"></div>

\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}

\newpage

**Figure 1.**

\newpage

**Figure S1.**

### 2022-06-30 submission/manuscript.tex
Create initial outline of manuscript

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1.0in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{helvet}
\renewcommand*\familydefault{\sfdefault}
\usepackage{setspace}
\doublespacing
\usepackage[left]{lineno}
\linenumbers
\raggedright
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\author{}
\date{\vspace{-2.5em}}

\begin{document}

\hypertarget{rarefy-your-data}{%
\section{Rarefy your data}\label{rarefy-your-data}}

\vspace{20mm}

\textbf{Running title:} Rarefy your data

\vspace{20mm}

Patrick D. Schloss\({^\dagger}\)

\vspace{40mm}

\({\dagger}\) To whom corresponsdence should be addressed:

\href{mailto:pschloss@umich.edu}{pschloss@umich.edu}

Department of Microbiology \& Immunology

University of Michigan

Ann Arbor, MI 48109

\vspace{20mm}

\textbf{Research article}

\newpage

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

\hypertarget{importance}{%
\subsection{Importance}\label{importance}}

\newpage

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

\begin{itemize}
\tightlist
\item
  What is rarefaction? History, reason for rarefaction

  \begin{itemize}
  \tightlist
  \item
    Repeated down sampling of datasets to a common number of
    observations to calculate the average value to ascertain the
    expected value of a metric for the metric under study; typically
    richness
  \item
    Control for unneven sampling effort
  \item
    Methods vary in their sensitivity to uneven sampling
  \end{itemize}
\item
  Reasons behind ``rarefaction is inadmissable''

  \begin{itemize}
  \tightlist
  \item
    Weird simulation
  \end{itemize}
\item
  Alternative approaches and claims

  \begin{itemize}
  \tightlist
  \item
    sampling invariance
  \end{itemize}
\item
  Goal of this study
\end{itemize}

\hypertarget{results}{%
\subsection{Results}\label{results}}

\hypertarget{datasets}{%
\subsubsection{Datasets}\label{datasets}}

\begin{itemize}
\tightlist
\item
  Represented diverse communities
\item
  High quality V4 sequence data that consisted of paired 250 nt reads
\item
  Wide distribution in the number of reads within each dataset (Figure
  S1, Table 1)
\end{itemize}

\hypertarget{null-models}{%
\subsubsection{Null models}\label{null-models}}

\begin{itemize}
\item
  Null models were samplings, without replacement, of all samples from a
  study that have been pooled together and sampled to the same depth of
  sampling
\item
  Created 100 random sets of samples per dataset
\item
  Would expect a zero correlation between alpha diversity metric and
  number of sequences in the community
\item
  Rarefaction was the only approach that resulted in a near-zero
  correlation (Figure 1)
\item
  Would expect a zero correlation between beta diversity metrics and the
  difference in the number of sequences in the samples going into the
  distance calculation
\item
  Rarefaction was the only approach that resulted in a near-zero
  correlation (Figure 2)
\end{itemize}

\hypertarget{result-when-sampling-effort-is-confounded-with-treatment-group}{%
\subsubsection{Result when sampling effort is confounded with treatment
group}\label{result-when-sampling-effort-is-confounded-with-treatment-group}}

\begin{itemize}
\item
  Used Null model
\item
  Created two treatment groups for each study by dividing samples based
  on whether they were above or below the median number of sequences per
  sample
\item
  Used Wilcoxon test to detect differences between treatment groups for
  alpha diversity, would expect \textasciitilde5\% of random sets to be
  significant
\item
  Rarefaction the only approach that reliably resulted in the expected
  Type I error rate
\item
  Used AMOVA test to detect differences between treatment groups for
  beta diversity, would expect \textasciitilde5\% of random sets to be
  significant
\item
  Rarefaction the only approach that reliably resulted in the expected
  Type I error rate
\end{itemize}

\hypertarget{result-when-known-effect-size-imposed-between-two-groups-of-samples}{%
\subsubsection{Result when known effect size imposed between two groups
of
samples}\label{result-when-known-effect-size-imposed-between-two-groups-of-samples}}

\begin{itemize}
\item
  For each dataset I divided samples into two groups. The probability
  distribution for the first group was determined by the null model and
  the second was determined by xxxxxxxxxxx
\item
  These parameters were selected because they produced results were some
  of the simulations for the rarefied data did not result in a
  significant test
\item
  Fraction of tests that were significant would be a measure of
  statistical power
\item
  Used Wilcoxon test to detect differences between treatment groups for
  alpha diversity
\item
  All tests resulted in reduced power to detect differences relative to
  the power of using rarefaction
\item
  Used AMOVA test to detect differences between treatment groups for
  beta diversity
\item
  All tests resulted in reduced power to detect differences relative to
  the power of using rarefaction
\item
  This simulation would work well for measures based on the structure of
  the community, but lacked resolution for those based on the richness
  or membership of commmunities. For an additional simulation, the first
  treatment group followed a null distribution for the second, XX\% of
  OTUs were removed from the distribution
\item
  Tests for richness and membership (i.e.~Jaccard) resutled in reduced
  power relative to rarefaction
\end{itemize}

\hypertarget{effect-of-sampling-depth-on-metrics}{%
\subsubsection{Effect of sampling depth on
metrics}\label{effect-of-sampling-depth-on-metrics}}

\begin{itemize}
\tightlist
\item
  Simulated rarefying null model datasets to retain varying percentage
  of samples
\item
  High sampling depth resulted in less variation between samples for
  alpha and beta diversity
\end{itemize}

\hypertarget{re-analysis-of-previously-published-datasets}{%
\subsubsection{Re-analysis of previously published
datasets}\label{re-analysis-of-previously-published-datasets}}

\begin{itemize}
\tightlist
\item
  Human study: Did not alter effect size or significance of alpha or
  beta diversity, but did result in reduced effect sizes for measures of
  richness and non-parametric estimators of richness; breakaway detected
  a difference
\end{itemize}

\hypertarget{discussion}{%
\subsection{Discussion}\label{discussion}}

\begin{itemize}
\tightlist
\item
  Rarefy your data
\item
  Problems with recommended methods\ldots{}

  \begin{itemize}
  \tightlist
  \item
    Many recommended methods are borrowed from gene expression analysis
  \item
    Meaning of zeroes in data - structural vs.~below limit of detection
  \end{itemize}
\item
  Factors that determine what number of sequences to rarefy to
\item
  Need better methods of pooling libraries that result in more even
  distribution of sequences across samples
\item
  Rarefy your data
\end{itemize}

\hypertarget{materials-and-methods}{%
\subsection{Materials and Methods}\label{materials-and-methods}}

\textbf{Data availability.}

\textbf{Reproducible data analysis.}

\vspace{10mm}

\textbf{Acknowledgements.}

\newpage

\hypertarget{references}{%
\subsection{References}\label{references}}

\setlength{\parindent}{-0.25in}
\setlength{\leftskip}{0.25in}

\noindent

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\end{CSLReferences}

\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}

\newpage

\textbf{Figure 1.}

\newpage

\textbf{Figure S1.}

\end{document}

### 2022-07-05 code/get_observed_coverage.R
Analyze influence of sampling depth on rarefying alpha and beta diversity metrics

@@ -1,0 +1,34 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(glue)
+
+input <- commandArgs(trailingOnly = TRUE)
+
+shared_file <- input[1] # shared_file <- "data/marine/data.otu.shared"
+output_file <- str_replace(shared_file, "shared", "obs_coverage")
+
+temp_shared_file <- str_replace(shared_file, "shared", "temp_coverage.shared")
+
+if (file.exists(temp_shared_file)) {
+  unlink(temp_shared_file)
+}
+
+file.copy(shared_file, temp_shared_file, overwrite = TRUE)
+
+mothur_command <- glue("mothur '#summary.single(shared={temp_shared_file},
+                                                calc=nseqs-coverage)'")
+
+system(mothur_command)
+
+
+str_replace(temp_shared_file, "shared", "groups.summary") %>%
+  read_tsv() %>%
+  select(-label) %>%
+  write_tsv(output_file)
+  
+  
+temp_files <- list.files(path = dirname(temp_shared_file),
+                         pattern = ".*temp_coverage.*",
+                         full.names = TRUE)
+unlink(temp_files)

### 2022-07-05 code/get_observed_depth_alpha.R

@@ -1,0 +1,45 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(glue)
+
+input <- commandArgs(trailingOnly = TRUE)
+
+shared_file <- input[1] # shared_file <- "data/human/data.otu.shared"
+depth <- as.numeric(input[2]) # depth <- 1000
+
+mothur_output_file <-
+      str_replace(shared_file,
+                  "shared",
+                  glue("temp_obs_{depth}_alpha.groups.ave-std.summary"))
+
+script_output_file <- str_replace(shared_file,
+                                  "shared",
+                                  glue("obs_{depth}_alpha.summary"))
+
+temp_shared_file <- str_replace(shared_file, "shared",
+                                glue("temp_obs_{depth}_alpha.shared"))
+
+if (file.exists(temp_shared_file)) {
+  unlink(temp_shared_file)
+}
+
+file.copy(shared_file, temp_shared_file, overwrite = TRUE)
+
+
+
+
+mothur <- glue("mothur '#summary.single(shared = {temp_shared_file},
+                        calc=sobs-shannon-simpson-coverage,
+                        subsample={depth})'")
+
+system(mothur)
+
+file.rename(mothur_output_file, script_output_file)
+
+temp_files <- list.files(path = dirname(temp_shared_file),
+                         pattern = str_replace(basename(temp_shared_file),
+                                               ".shared", ""),
+                         full.names = TRUE)
+
+unlink(temp_files)

### 2022-07-05 code/get_observed_depth_alpha_kw.R

@@ -1,0 +1,71 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(glue)
+library(effectsize)
+library(broom)
+
+get_rarefy_results <- function(rfile, design) {
+
+  alpha_data <- read_tsv(rfile, col_types = cols(group = col_character(),
+                                                 method = col_character(),
+                                                 .default = col_double()))
+
+  inner_join(alpha_data,
+             design,
+             by = "group") %>% 
+    filter(method == "ave") %>%
+    select(Group = group, treatment, sobs, shannon, simpson, coverage) %>%
+    pivot_longer(cols = -c(Group, treatment), names_to = "metric") %>%
+    nest(data = -metric) %>%
+    mutate(kw = map(data, ~tidy(kruskal.test(.x$value, g = .x$treatment))),
+           biserial = map(data, ~rank_biserial(.x$value~as.factor(.x$treatment),
+                                                    ci = NULL)[[1]][[1]]),
+           median = map(data,
+                        ~group_by(.x, treatment) %>%
+                          summarize(median = median(value),
+                                    n = n())
+                          )) %>%
+    unnest(c(kw, median, biserial)) %>%
+    select(metric, p.value, biserial, treatment, median, n) %>%
+    pivot_wider(names_from = "treatment", values_from = c("median", "n"))
+
+}
+
+
+run_kw_tests <- function(alpha_file, treatment) {
+
+  nseqs <- as.numeric(str_replace(alpha_file, ".*_(\\d*)_.*", "\\1"))
+
+  design_file <- "data/human/data.odesign"
+
+  design <- read_tsv(design_file,
+                     col_types = cols(.default = col_character())) %>%
+    select(group = Group, treatment = all_of(treatment))
+
+  get_rarefy_results(alpha_file, design) %>%
+    mutate("treatment" = treatment) %>%
+    mutate(nseqs = nseqs, .after = "metric")
+
+}
+
+#########
+
+design_file <- "data/human/data.odesign"
+output_file <- "data/human/data.otu.obs_depth_alpha_kw.sumary"
+
+summary_files <- list.files(path = "data/human",
+                            pattern = "data.otu.obs_\\d*_alpha.summary",
+                            full.names = TRUE)
+
+metadata <- c("srn", "obese", "female", "smoker", "nsaid", "prev_crc",
+              "prev_polyps")
+
+
[…]

### 2022-07-05 code/get_observed_depth_beta.R

@@ -1,0 +1,26 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(glue)
+library(vegan)
+library(data.table)
+
+
+input <- commandArgs(trailingOnly = TRUE)
+
+shared_file <- input[1] # shared_file <- "data/human/data.otu.shared"
+depth <- as.numeric(input[2]) # depth <- 1000
+distance <- input[3] # distance <- "bray"
+
+dist_file <- str_replace(shared_file,
+                         "shared",
+                         glue("obs_{depth}_{distance}.dist"))
+
+
+fread(shared_file, colClasses = c(Group = "character")) %>%
+  select(Group, starts_with("Otu")) %>%
+  column_to_rownames("Group") %>%
+  avgdist(sample = depth, dmethod = distance)  %>%
+      as.matrix() %>%
+      as_tibble(rownames = "Group") %>%
+      write_tsv(dist_file)

### 2022-07-05 code/get_observed_depth_beta_amova.R

@@ -24,7 +24,10 @@ run_amova <- function(distance_file, treatment) {
     nrow()
 
   result <- tibble(p.value = NA_real_,
-                   r_squared = NA_real_)
+                   r_squared = NA_real_,
+                   n_FALSE = NA_real_,
+                   n_TRUE = NA_real_
+                   )
 
   if (w_na == wo_na) {
     distance_matrix <- distance_design %>%
@@ -34,26 +37,28 @@ run_amova <- function(distance_file, treatment) {
 
     test <- vegan::adonis2(distance_matrix~distance_design$treatment)
 
+    tf_count <- distance_design %>% count(treatment) %>%
+      pivot_wider(names_from = treatment, values_from = "n")
+
     result <- tibble(p.value = test$`Pr(>F)`[1],
-                     r_squared = test$R2[1])
+                     r_squared = test$R2[1],
+                     n_FALSE = tf_count$`FALSE`,
+                     n_TRUE = tf_count$`TRUE`)
   }
 
   return(result)
 
 }
 
-dataset <- "data/human"
-output_file <- "data/human/data.o_oamova"
-
-model <- str_replace(output_file, ".*\\.(.)_.amova", "\\1")
-design <- str_replace(output_file, ".*\\.._(.)amova", "\\1")
 
 design_file <- read_tsv("data/human/data.odesign")
 treatments <- colnames(design_file)[-1]
 
+output_file <- "data/human/data.otu.obs_depth_beta_amova.summary"
+
 distance_design <- expand_grid(
-    distance_file = list.files(path = dataset,
-                               pattern = glue(".*\\.{model}_.*.dist"),
+    distance_file = list.files(path = "data/human",
+                               pattern = "data.otu.obs_\\d*_\\w*.dist",
                                full.names = TRUE),
     treatment = treatments
     )
@@ -65,8 +70,8 @@ distance_design %>%
   unnest(result) %>%
   ungroup() %>%
   separate(distance_file,
-           into = c("model", "beta_process", "beta_calculator"),
+           into = c("dummy", "nseqs", "beta_calculator"),
            sep = "_") %>%
   mutate(beta_calculator = str_replace(beta_calculator, ".dist", "")) %>%
-  select(-model) %>%
+  select(-dummy) %>%
   write_tsv(output_file)

### 2022-07-05 code/get_observed_shared_design.R

@@ -16,12 +16,17 @@ removal <- scan(remove_file, quiet = TRUE, what = character()) %>%
 # some sample names appear twice, but their diagnoses are the same. this was
 # because some samples were sequenced twice
 read_tsv("data/human/sra_info.tsv",
-         col_type=cols(Sample_Name = col_character())) %>%
-  mutate(treatment = if_else(
-    diagnosis == "Cancer" | diagnosis == "adv Adenoma",
-    "srn", "healthy"))          %>%
-  count(Sample_Name, treatment) %>%
-  select(Group = Sample_Name, treatment) %>%
+         col_type = cols(Sample_Name = col_character())) %>% 
+  mutate(srn = diagnosis == "Cancer" | diagnosis == "adv Adenoma",
+         obese = BMI >= 30,
+         smoker = as.logical(Smoke),
+         nsaid = as.logical(NSAID),
+         prev_crc = as.logical(Hx_Prev),
+         prev_polyps = as.logical(Hx_of_Polyps),
+         female = Gender == 'female'
+         ) %>%
+  select(Group = Sample_Name, srn, obese, female, smoker, nsaid, 
+         prev_crc, prev_polyps) %>%
   filter(! Group %in% removal) %>%
   write_tsv(obs_design_file)
 

### 2022-07-05 code/pool_alpha_summary.R

@@ -1,0 +1,31 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+
+input <- commandArgs(trailingOnly = TRUE)
+
+summary_files <- input
+# summary_files <- list.files(path = "data/soil",
+#                             pattern = ".*r_rare_alpha.summary",
+#                             full.names = TRUE)
+
+output_file <- str_replace(summary_files[1],
+                           "data.otu.*",
+                           "data.otu.alpha_depth.summary")
+
+reader <- function(x) {
+
+  seed <- str_replace(x, ".*\\.(\\d*)\\.r_rare_.*.summary", "\\1")
+  dataset <- str_replace(x, "data/(.*)/data.otu.*", "\\1")
+  # dmethod <- str_replace(x, ".*\\.\\d*\\.r_rare_(.*).summary", "\\1")
+
+  read_tsv(x,
+           col_types = cols(metric = col_character(),
+                            .default = col_double())) %>%
+  mutate(seed = seed, dataset = dataset) %>%
+  select(seed, dataset, method = metric, n_seqs, mean, sd, n_samples)
+
+}
+
+map_dfr(summary_files, reader) %>%
+  write_tsv(output_file)

### 2022-07-05 code/pool_beta_summary.R

@@ -1,0 +1,31 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+
+input <- commandArgs(trailingOnly = TRUE)
+
+summary_files <- input
+# summary_files <- list.files(path = "data/mice",
+#                             pattern = ".*r_rare_.*.summary",
+#                             full.names = TRUE)
+
+output_file <- str_replace(summary_files[1],
+                           "data.otu.*",
+                           "data.otu.beta_depth.summary")
+
+reader <- function(x) {
+
+  seed <- str_replace(x, ".*\\.(\\d*)\\.r_rare_.*.summary", "\\1")
+  dataset <- str_replace(x, "data/(.*)/data.otu.*", "\\1")
+  dmethod <- str_replace(x, ".*\\.\\d*\\.r_rare_(.*).summary", "\\1")
+
+  read_tsv(x,
+           col_types = cols(.default = col_double())) %>%
+    mutate(seed = as.numeric(seed), dataset = dataset, method = dmethod,
+           .before = n_seqs)
+
+}
+
+map_dfr(summary_files, reader) %>%
+  arrange(seed, method, n_seqs) %>%
+  write_tsv(output_file)

### 2022-07-05 code/rarefy_beta_analysis.R

@@ -1,0 +1,63 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(data.table)
+library(vegan)
+library(glue)
+library(magrittr)
+
+
+input <- commandArgs(trailingOnly = TRUE)
+
+shared_file <- input[1] # shared_file <- "data/mice/data.otu.1.rshared"
+nseqs_file <- input[2] # nseqs_file <- "data/mice/data.group_count"
+remove_file <- input[3] # remove_file <- "data/mice/data.remove_accnos"
+output_file <- input[4] # output_file <- "data/mice/data.otu.1.1.r_rare_bray.summary"
+
+index <- str_replace(output_file,
+            "data/.*/data\\.otu\\.(\\d*)\\.\\d*\\.r_.*summary",
+            "\\1") %>%
+  as.numeric()
+
+to_remove <- read_tsv(remove_file, col_names="Group")
+
+max_nseqs <- read_tsv(nseqs_file,
+                      col_types = cols(Group = col_character()),
+                      col_names = c("Group", "nseqs")) %>%
+  anti_join(., to_remove) %>%
+  arrange(nseqs) %>% 
+  slice(n = nrow(.) - 4) %>% # insures there are at least 10 distances in 
+  pull(nseqs)                     # final class
+
+
+
+nseqs <- 10^seq(log10(1000), log10(max_nseqs), length.out = 100) %>%
+  as.integer() %>%
+  extract(index)
+
+distance <- str_replace(output_file, ".*r_rare_(.*)\\.summary", "\\1")
+distance <- case_when(distance == "bray" ~ "braycurtis",
+            distance == "jaccard" ~ "jclass")
+
+temp_shared <- str_replace(output_file, "summary", "shared")
+file.copy(shared_file, temp_shared, overwrite = TRUE)
+
+mothur_cmd <- glue("mothur '#dist.shared(shared={temp_shared}, subsample={nseqs},
+                                         calc={distance}, iters=100, processors=1, output=square)'")
+system(mothur_cmd)
+
+temp_dist <- str_replace(temp_shared, "shared", glue("{distance}.0.03.square.ave.dist"))
+
+read_tsv(temp_dist, skip = 1, col_names = FALSE) %>%
+  set_colnames(c("samples", all_of(.[["X1"]]))) %>%
+      pivot_longer(cols = -samples) %>%
+      filter(samples < name) %>%
+      summarize(mean = mean(value),
[…]

### 2022-07-05 code/run_oalpha_kw.R

@@ -4,11 +4,15 @@ library(tidyverse)
 library(glue)
 library(broom)
 
-get_rarefy_results <- function(rfile, dfile) {
+get_rarefy_results <- function(rfile, design) {
 
-  inner_join(read_tsv(rfile, show_col_types = FALSE),
-             read_tsv(dfile, show_col_types = FALSE),
-             by = c("group" = "Group")) %>%
+  alpha_data <- read_tsv(rfile, col_types = cols(group = col_character(),
+                                                 method = col_character(),
+                                                 .default = col_double()))
+
+  inner_join(alpha_data,
+             design,
+             by = "group") %>%
     filter(method == "ave") %>%
     select(Group = group, treatment, nseqs, sobs, shannon,
           simpson, chao, ace, npshannon, coverage) %>%
@@ -27,11 +31,16 @@ get_rarefy_results <- function(rfile, dfile) {
 
 }
 
-get_breakaway_results <- function(bafile, dfile) {
+get_breakaway_results <- function(bafile, design) {
 
-  inner_join(read_tsv(bafile, show_col_types = FALSE),
-             read_tsv(dfile, show_col_types = FALSE),
-             by = c("Group" = "Group")) %>%
+  alpha_data <- read_tsv(bafile, col_types = cols(Group = col_character(),
+                                                 ba_model = col_character(),
+                                                 p_model = col_character(),
+                                                 .default = col_double()))
+
+  inner_join(alpha_data,
+             design,
+             by = c("Group" = "group")) %>%
     select(Group, treatment, ba_est, ba_model, p_est, p_model) %>%
     pivot_longer(cols = -c(Group, treatment), names_to = c("set", ".value"),
                  names_pattern = "(.*)_(.*)") %>%
@@ -50,14 +59,17 @@ get_breakaway_results <- function(bafile, dfile) {
 
 }
 
-get_raw_results <- function(rawfile, dfile) {
+get_raw_results <- function(rawfile, design) {
 
-  inner_join(read_tsv(rawfile, show_col_types = FALSE),
-             read_tsv(dfile, show_col_types = FALSE),
-             by = c("group" = "Group")) %>%
-    select(Group = group, treatment, nseqs, sobs, shannon,
+  alpha_data <- read_tsv(rawfile, col_types = cols(group = col_character(),
[…]

### 2022-07-05 code/pool_beta_summary.R
Analyze influence of sampling depth on rarefying alpha and beta diversity metrics

@@ -1,0 +1,29 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+
+input <- commandArgs(trailingOnly = TRUE)
+
+summary_files <- input
+# summary_files <- list.files(path = "data/mice",
+#                             pattern = ".*r_rare_.*.summary",
+#                             full.names = TRUE)
+
+output_file <- str_replace(summary_files[1],
+                           "data.otu.*",
+                           "data.otu.beta_depth.summary")
+
+reader <- function(x) {
+
+  seed <- str_replace(x, ".*\\.(\\d*)\\.r_rare_.*.summary", "\\1")
+  dataset <- str_replace(x, "data/(.*)/data.otu.*", "\\1")
+  dmethod <- str_replace(x, ".*\\.\\d*\\.r_rare_(.*).summary", "\\1")
+
+  read_tsv(x,
+           col_types = cols(.default = col_double())) %>%
+  mutate(seed = seed, dataset = dataset, method = dmethod, .before = n_seqs)
+
+}
+
+map_dfr(summary_files, reader) %>%
+  write_tsv(output_file)

### 2022-07-05 code/rarefy_alpha_analysis.R

@@ -1,0 +1,61 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(glue)
+
+input <- commandArgs(trailingOnly = TRUE)
+
+shared_file <- input[1] # shared_file <- "data/soil/data.otu.24.rshared"
+output_file <- input[2] # output_file <- "data/soil/data.otu.24.r_rare_alpha.summary"
+
+
+temp_shared_file <- str_replace(shared_file, "rshared", "temp_alpha.rshared")
+
+if (file.exists(temp_shared_file)) {
+  unlink(temp_shared_file)
+}
+
+file.copy(shared_file, temp_shared_file, overwrite = TRUE)
+
+mothur <- glue("mothur '#rarefaction.single(shared = {temp_shared_file},
+                        calc=sobs-shannon-simpson-coverage,
+                        freq=1000,
+                        processors = 1)'")
+
+system(mothur)
+
+read_rarefaction_files <- function(x) {
+
+  m <- if_else(str_detect(x, "rarefaction"),
+               "sobs",
+               str_replace(x, ".*temp_alpha.groups.r_", "")
+               )
+  
+  read_tsv(x, col_types = cols(.default = col_double())) %>%
+    filter(numsampled == 1000 | (numsampled %% 2000 == 0)) %>%
+    select(n_seqs = numsampled, starts_with("0.03")) %>%
+    pivot_longer(-n_seqs, names_to = "group", values_to = "value") %>%
+    mutate(group = str_replace(group, "0\\.03-", ""),
+           metric = m)
+
+}
+
+extensions <- c("r_coverage", "r_shannon", "r_simpson", "rarefaction")
+file_stub <- str_replace(temp_shared_file, "rshared", "groups")
+files <- glue("{file_stub}.{extensions}")
+
+
+map_dfr(files, read_rarefaction_files) %>% 
+  drop_na() %>%
+  group_by(n_seqs, metric) %>%
+  summarize(mean = mean(value),
+            sd = sd(value),
+            n_samples = n(), .groups = "drop") %>%
+  write_tsv(output_file)
+
+temp_files <- list.files(path = dirname(temp_shared_file),
+                         pattern = str_replace(basename(temp_shared_file),
+                                               ".rshared", ""),
+                         full.names = TRUE)
+
+unlink(temp_files)

### 2022-07-05 code/rarefy_beta_analysis.R

@@ -1,0 +1,53 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(data.table)
+library(vegan)
+# library(furrr)
+
+# plan(multicore, workers = 8)
+
+input <- commandArgs(trailingOnly = TRUE)
+
+shared_file <- input[1] # shared_file <- "data/mice/data.otu.1.rshared"
+nseqs_file <- input[2] # nseqs_file <- "data/mice/data.group_count"
+output_file <- input[3] # output_file <- "data/mice/data.otu.1.r_rare_bray.summary"
+
+shared_df <- fread(shared_file, colClasses = c(Group = "character")) %>%
+  select(Group, starts_with("Otu")) %>%
+  column_to_rownames("Group")
+
+groups <- rownames(shared_df)
+
+max_nseqs <- read_tsv(nseqs_file,
+                  col_names = c("Group", "nseqs")) %>%
+  filter(Group %in% groups) %>%
+  arrange(nseqs) %>% 
+  slice(n = nrow(.) - 4) %>% # insures there are at least 10 distances in 
+  pull(nseqs)                     # final class
+
+
+nseqs <- c(1000, seq(2000, max_nseqs, 2000))
+
+distance <- str_replace(output_file, ".*r_rare_(.*)\\.summary", "\\1")
+
+run_rarefy_beta <- function(df, x, distance) {
+
+  avgdist(df, sample = x, dmethod = distance)  %>%
+      as.matrix() %>%
+      as_tibble(rownames = "samples") %>%
+      pivot_longer(cols = -samples) %>%
+      filter(samples < name) %>%
+      summarize(mean = mean(value),
+                sd = sd(value),
+                n_samples = (1 + sqrt(1 + n() * 8)) / 2) %>%
+      mutate(n_seqs = x, .before = mean)
+
+}
+
+#future_map_dfr(nseqs,
+map_dfr(nseqs,
+               ~run_rarefy_beta(df = shared_df,
+                                x = .x,
+                                distance = distance)) %>%
+  write_tsv(output_file)

### 2022-07-07 submission/manuscript.Rmd
First rough draft of Results

@@
 
+* Motivation
+  * Problem of uneven sampling effort
 * What is rarefaction? History, reason for rarefaction
@@
 
-### Datasets
-* Represented diverse communities
-* High quality V4 sequence data that consisted of paired 250 nt reads
-* Wide distribution in the number of reads within each dataset (Figure S1, Table 1)
+***Choice of datasets.*** My analysis incorporated community distribution data from 12 studies that represented bacterial communities from diverse environments.
+The specific studies were selected because their data was publicly accessible through the Sequence Read Archive, the original investigators sequenced the V4 region of the 16S rRNA gene using paired 250 nt reads, and my previous familiarity with the data.
+The use of paired 250 nt reads to sequence the V4 region resulted in a near complete two-fold overap of the V4 region resulting in high quality contigs with a low sequencing error rate.
+These data were processed through the standard sequence curation pipeline to generate operational taxonomic units (OTUs) using the mothur software package.
+The original studies generated the sequence data by pooling separate PCR products that were generated by amplifying the V4 region from the bacterial DNA in each sample.
+Because pooling equimolar quantities of DNA is frought with difficulties, it was common to observe wide variation in the number of sequences in each sample (Figure S1, Table 1).
 
-### Null models
-* Null models were samplings, without replacement, of all samples from a study that have been pooled together and sampled to the same depth of sampling
-* Created 100 random sets of samples per dataset
+***Without rarefaction, metrics of alpha diversity are sensitive to sampling effort.*** To test the sesitivity of various approaches of measuring alpha diversity to sampling effort, I generated null models for each study.
+Under a null model, each sample from the same study would be expected to have the same alpha diversity regardless of the sampling effort.
+I assessed richness without any correction, using normalized OTU counts, with estimates based on non-parametric and parametric approaches, and using rarefaction.
+For each study, all of the approaches, except for rarefaction, showed a strong correlation between richness and the number of sequences in the sample (Figure 1A).
+Next, I assessed diversity using the Shannon diversity index and the inverse Simpson diversity index without any correction, using normalized OTU counts, and rarefaction; I also used a non-parametric estimator of Shannon diversity.
+The correlation between sampling depth and the diversity metric was not as strong as it was for richness and the inverse Simpson diversity values were less sensitive than the Shannon diversity values; however, the correlation to the rarefied diversity metrics were the lowest for all of the metrics and studies (Figure 1B).
+The rarefied alpha-diversity metrics consistently demonstrated a lack of sensitivity to sampling depth.
 
-* Would expect a zero correlation between alpha diversity metric and number of sequences in the community
-* Rarefaction was the only approach that resulted in a near-zero correlation (Figure 1)
+***Without rarefaction, metrics of beta diversity are sensitive to sampling effort.*** To test the sesitivity of various approaches of measuring beta diversity to sampling effort, I used the same null models used for studying the sensitivity of alpha diversity.
+Under a null model, the ecological distance between any pair of samples would be the same regardless of the difference in the number of sequences observed in each sample.
+First, I analyzed the sensitivity of the Jaccard distance coefficient, which incorporates whether an OTU is present in each community and not their relative abundance.
+When calculating Jaccard distances using the uncorrected OTU counts, normalized OTU counts, relative abundances, and rarefaction only the rarefied data showed a lack of sensitivity to sampling effort (Figure 2A).
+Second, I analyzed the sensitivity of the Bray-Curtis distance coefficient, which is a popular metric that incorporates the abundance of each OTU.
+Similar to what I observed with the Jaccard coefficient, only the rarefied data showed a lack of sensitivity to sampling effort (Figure 2B).
+Third, I calcualted Aitchison distances on raw OTU counts where the central log-ratio (CLR) was calculated by ignoring OTUs in samples with zero counts (robust CLR), adding a pseudocount of 1 to all OTU counts prior to calculating the CLR (one CLR), XXXXX XXXXX XXXXX XXXXX XXXXX XXXXX XXXXX XXXXX XXXXX XXXXX XXXXXXXXXX (n CLR), and imputing the value of zero counts (z CLR).
+Regardless of the approach, the Aitchison distances were all strongly sensitive to sampling effort (Figure 2C).
+Finally, I used the cumulative sum scaling (CSS) normalization from metagenomeSeq and variance stabilization technique (VST) from DeSeq2 prior to calculating Euclidean distances.
+Both approachs revealed a strong sensitivity to sampling effort (Figure 2D).
+Although Euclidean distances are not typically used on raw or rarefied count data in ecology, rarefied Euclidean distances were not sensitive to sampling effort.
+Across each of the beta diversity metrics and approaches used to account for uneven sampling effort and sparsity, rarefaction was the least sensitive approach to differences in sampling effort.
 
-* Would expect a zero correlation between beta diversity metrics and the difference in the number of sequences in the samples going into the distance calculation
-* Rarefaction was the only approach that resulted in a near-zero correlation (Figure 2)
+***Rarefaction limits the detection of false positives when sampling effort and treatment group are confounded.*** Next, I investigated the impact of the various strategies and metrics on falsely detecting a significant diffeerence using the the same communities generated from the null model in the analysis of alpha and beta diversity metrics.
+To test for differences in alpha and beta diversity I used the non-parametric Wilcoxon test and non-parametric permutation-based multivariate analysis of variance (PERMANOVA).
+First, within each study, I randomly assigned each sample to one of two treatment groups.
+My expectation was that approximately 5 of the 100 (5%) random tests for each comparison would yield a significant test result.
+Indeed, for each study and alpha and beta diversity metric and strategy for accounting for uneven sampling, approximately 5% of the tests yielded a significant result (Figure 3).
+Second, within each study, I assigned samples with more than the median number of sequences per sample to one treatment group and the rest to another treatment group.
+If there is no sensitivity to sampling effort, I would again expect that 5% of the tests would yield a significant result.
+In fact, only the rarefied data consistently resulted in a 5% false positive rate for alpha and beta diversity metrics (Figure 4).
+These results align with the observed sensitivity of alpha and beta diversity metrics to sampling effort and underscore the value of rarefaction.
 
-### Result when sampling effort is confounded with treatment group
-* Used Null model
-* Created two treatment groups for each study by dividing samples based on whether they were above or below the median number of sequences per sample
+***Rarefaction preserves the statistical power to detect differences between treatment groups.*** To assess the impact of different approaches to control for uneven sampling effort I performed two additional simulations.
+In the first simulation, for each study samples were randomly assigned to one of two treatment groups.
+Samples in the first treatment group were generated by sampling from the null distribution.
+Samples in the second treatment group were generated by perturbing the null distribution by increasing the relative abundance of 10% of the OTUs by 5%.
+These values were determined after empirically searching for conditions that resulted in a large fraction of the randomizations yieleding a significant result across most of the studies.
+The fraction of tests that yielded a significant test was a measure of the statistical power for the test.
+Relative to the rarefied data, the power to detect differences in alpha and beta diversity was considerably lower for each of the strategies for handling uneven sampling effort.
+The power to detect differences in richness by all approaches was low (Figures 7 and 8).
+This was likely because the approach to generating the second community did not necessarily change the number of OTUs in each treatment group.
+To explore this further, in the second simulation the second treatment group was perturbed by removing 3% of the OTUs from the model.
+Again, the rarefied data had a considerably higher statistical power than the other approaches when measuring richness (Figure 9).
+Both simulations highlight the value of rarefaction for preserving the statistical power to detect differences between treatment groups for measures of alpha and beta diversity.
 
-* Used Wilcoxon test to detect differences between treatment groups for alpha diversity, would expect ~5% of random sets to be significant
-* Rarefaction the only approach that reliably resulted in the expected Type I error rate
+***Increased rarefaction depth reduces intra-sample variation in alpha and beta diversity.*** To assess the sensitivity of alpha and beta diversity metrics to rarefaction depth, I again used the dataset generated using the null models and rarefied them to varying depths.
+For the alpha diversity metrics, the value of the metrics plateaued and the coefficient of variation (i.e., the mean divided by the standard deviation) between samples remained constant as the rarefaction depth increased (Figure 10).
+For beta diversity metrics, the distance between samples and the coefficient of variation between samples decreased as sampling depth increased (Figure 11).
+These results confirm that greater sequencing depth provides a more robust estimate of the metrics.
 
-* Used AMOVA test to detect differences between treatment groups for beta diversity, would expect ~5% of random sets to be significant
-* Rarefaction the only approach that reliably resulted in the expected Type I error rate
+***Most studies have a high level of sequencing coverage.*** I calculated the Good's coverage for the observed data and found that each of the studies had a minimum coverage greater than 90% at their lowest sequencing depth (Figure 12A).
+This suggests that most studies have a high level of sequencing coverage.
+Next, I returned to the null models to ask how much sequencing effort was required to obtain higher levels of coverage.
+To obtain 95 and 99% coverage, an average of XX and XX-fold more sequence data was required, respectively (Figure 12B).
+Although it is clear that most researchers would desire greater coverage, the sequencing effort required to acheive that sequencing depth would likely limit the number of samples that could be sequenced when controlling for costs.
 
-### Result when known effect size imposed between two groups of samples
-* For each dataset I divided samples into two groups. The probability distribution for the first group was determined by the null model and the second was determined by xxxxxxxxxxx
-* These parameters were selected because they produced results were some of the simulations for the rarefied data did not result in a significant test
-* Fraction of tests that were significant would be a measure of statistical power
+***Evidence that not rarefying data can impact results in original studies.*** To assess the impact of using alternatives to rarefaction, I reassessed hypotheses from the human, mice, and peromyscus studies.
+These studies were selected because they originated from my research group and know the datasets well.
+First, the study that published the human dataset was interested in the difference in the fecal microbial communities of patients with and without colorectal cancer.
+Here, I divided the patients into peope with and without screen relevant neoplasia (SRN).
+Among the alpha diversity metrics, the test were non-significant with similar effect sizes; however, the richness estimates obtained using parametric appraoches detected statistically significant differences in richness.
+Although the tests using Bray-Curtis distances were significant regardless of whether rarefaction was used, the effect sizes were smaller for the distances calculated using the CLR, VST, CSS transformations.
 
-* Used Wilcoxon test to detect differences between treatment groups for alpha diversity
-* All tests resulted in reduced power to detect differences relative to the power of using rarefaction
+Mouse
 
-* Used AMOVA test to detect differences between treatment groups for beta diversity
-* All tests resulted in reduced power to detect differences relative to the power of using rarefaction
-
-* This simulation would work well for measures based on the structure of the community, but lacked resolution for those based on the richness or membership of commmunities.
-For an additional simulation, the first treatment group followed a null distribution for the second, XX% of OTUs were removed from the distribution
-* Tests for richness and membership (i.e. Jaccard) resutled in reduced power relative to rarefaction
-
-### Effect of sampling depth on metrics
-* Simulated rarefying null model datasets to retain varying percentage of samples
-* High sampling depth resulted in less variation between samples for alpha and beta diversity
-
-* Good's coverage of observed data (take from raw data files)
+Peromyscus
 
@@
 * Problems with recommended methods...
-  * Many recommended methods are borrowed from gene expression analysis
+  * Many recommended methods are borrowed from gened expression analysis
   * Meaning of zeroes in data - structural vs. below limit of detection
@@
 ## Materials and Methods
+
+**Null models.** Null models were generated by randomly assigning each sequence in the study to an OTU and sample while keeping constant the number of sequences per sample and the total number of sequences in each OTU.
+Because the construction of the null models was a stochastic process, 100 replicates were geneated for each study.
 

### 2022-10-06 submission/manuscript.Rmd
Bring in Table 1 from Singletons papper

@@
 ---
+bibliography: references.bib
 output:
-  word_document: default
   pdf_document:
     keep_tex: true
+csl: asm.csl
 geometry: margin=1.0in
-font-size: 11pt
 header-includes:
-  - \usepackage{helvet}
-  - \renewcommand*\familydefault{\sfdefault}
+ - \usepackage{booktabs}
+ - \usepackage{longtable}
+ - \usepackage{array}
+ - \usepackage{multirow}
+ - \usepackage{wrapfig}
+ - \usepackage{float}
+ - \usepackage{colortbl}
+ - \usepackage{pdflscape}
+ - \usepackage{tabu}
+ - \usepackage{threeparttable}
+ - \usepackage{threeparttablex}
+ - \usepackage[normalem]{ulem}
+ - \usepackage{makecell}
  - \usepackage{setspace}
@@
  - \linenumbers
-  - \raggedright
-editor_options:
-  chunk_output_type: console
-bibliography: references.bib
-csl: asm.cls
+ - \modulolinenumbers
+ - \usepackage{helvet} % Helvetica font
+ - \renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
+ - \usepackage[T1]{fontenc}
+ - \usepackage[shortcuts]{extdash}
 ---
@@
 ```{r, echo=FALSE}
+options(tidyverse.quiet = TRUE)
 suppressPackageStartupMessages(library(tidyverse))
@@
 suppressPackageStartupMessages(library(here))
+suppressPackageStartupMessages(library(kableExtra))
+suppressPackageStartupMessages(library(glue))
 
-opts_chunk$set(echo = FALSE)
+opts_chunk$set("tidy" = TRUE)
+opts_chunk$set("echo" = FALSE)
+opts_chunk$set("eval" = TRUE)
+opts_chunk$set("warning" = FALSE)
+opts_chunk$set("message" = FALSE)
+opts_chunk$set("cache" = FALSE)
 
-inline_hook <- function(x){
+inline_hook <- function(x, digits=2){
 
+  if(is.list(x)){
+    x <- unlist(x)
+  }
   if(is.numeric(x)){
-		if(abs(x - round(x)) < .Machine$double.eps){
-			# treat as integer
-			formatted <- format(x, digits=0, big.mark = ",")
+      paste(format(x,big.mark=',', digits=digits, nsmall=digits, scientific=FALSE))
+  } else {
+      paste(x)
+  }
+}
+knitr::knit_hooks$set(inline=inline_hook)
 
-		} else {
-			# treat as floating point number
-			formatted <- format(x, digits=2, nsmall=2, big.mark = ",")
-		}
-	} else {
-		formatted <-x
-	}
+package_version <- function(package){
 
-  formatted
+  paste(unlist(packageVersion(package)), collapse='.')
 
@@
 
-knit_hooks$set(inline=inline_hook)
+oxford_comma <- function(x, digits=2) {
 
-format_pct <- function(x, my_digits=1){
+  x <- map_chr(x, inline_hook, digits=digits)
 
-	format(x, digits=my_digits, nsmall=my_digits)
-
+  if(length(x) < 2){
+    x
+  } else if(length(x) == 2){
+    paste(x, collapse = " and ")
+  } else {
+    paste(paste(x[-length(x)], collapse=", "), x[length(x)], sep=", and ")
+  }
 }
@@
 
+```{r datasets}
+table_1 <- read_tsv(here('data/process/study_summary_statistics.tsv'))
+
+n_sample_range <- table_1 %>%
+  pull(n_samples) %>%
+  range() %>% as.integer()
+
+n_med_sequence_range <- table_1 %>%
+  pull(median) %>%
+  range() %>% as.integer()
+
+low_fold_samples <- table_1 %>%
+  filter(fold_difference < 2) %>%
+  pull(directory)
+
+high_fold_range <- table_1 %>%
+  filter(fold_difference >= 2) %>%
+  pull(fold_difference) %>%
+  range() %>%
+  round(digits=1)
+```
+
 ## Abstract
@@
 
-***Choice of datasets.*** My analysis incorporated community distribution data from 12 studies that represented bacterial communities from diverse environments.
+***Choice of datasets.*** I selected 16S rRNA gene sequence data from from 12 studies that characterized the variation in bacterial communities from diverse environments (Table 1).
 The specific studies were selected because their data was publicly accessible through the Sequence Read Archive, the original investigators sequenced the V4 region of the 16S rRNA gene using paired 250 nt reads, and my previous familiarity with the data.
-The use of paired 250 nt reads to sequence the V4 region resulted in a near complete two-fold overap of the V4 region resulting in high quality contigs with a low sequencing error rate.
-These data were processed through the standard sequence curation pipeline to generate operational taxonomic units (OTUs) using the mothur software package.
-The original studies generated the sequence data by pooling separate PCR products that were generated by amplifying the V4 region from the bacterial DNA in each sample.
-Because pooling equimolar quantities of DNA is frought with difficulties, it was common to observe wide variation in the number of sequences in each sample (Figure S1, Table 1).
+The use of paired 250 nt reads to sequence the V4 region resulted in a near complete two-fold overap of the V4 region resulting in high quality contigs with a low sequencing error rate [@Kozich2013].
+These data were processed through the standard sequence curation pipeline to generate operational taxonomic units (OTUs) using the mothur software package [@Kozich2013; @Schloss2009].
+The original studies generated the sequence data by pooling separate PCR products that were generated by amplifying the V4 region of the 16S rRNA gene from the bacterial DNA in multiple samples.
+Because pooling equimolar quantities of DNA is frought with difficulties, it was common to observe wide variation in the number of sequences in each sample (Figure S1).
 
@@
 ***Evidence that not rarefying data can impact results in original studies.*** To assess the impact of using alternatives to rarefaction, I reassessed hypotheses from the human, mice, and peromyscus studies.
-These studies were selected because they originated from my research group and know the datasets well.
+These studies were selected because they originated from my research group and I know the datasets well.
 First, the study that published the human dataset was interested in the difference in the fecal microbial communities of patients with and without colorectal cancer.
@@
 
+* Human study: Did not alter effect size or significance of alpha or beta diversity, but did result in reduced effect sizes for measures of richness and non-parametric estimators of richness; breakaway detected a difference
+
 Mouse
@@
 Peromyscus
-
-### Re-analysis of previously published datasets
-* Human study: Did not alter effect size or significance of alpha or beta diversity, but did result in reduced effect sizes for measures of richness and non-parametric estimators of richness; breakaway detected a difference
 
@@
 <div id="refs"></div>
-
+\bibliography{ref}
 \setlength{\parindent}{0in}
 \setlength{\leftskip}{0in}
+
+\newpage
+
+**Table 1.
+Summary of studies used in the analysis.** For all studies, the number of sequences used from each study was rarefied to the smallest sample size.
+A graphical represenation of the distribution of sample sizes for each study and the samples that were removed from each study are provided in Figure S1.
+
+\small
+
+```{r results_f}
+table_1 %>%
+  arrange(nice_name) %>%
+  mutate(
+    nice_name = glue("{nice_name}\\ {reference}"),
+    total_seqs=format(total_seqs, big.mark=",", trim=TRUE),
+    median=format(as.integer(median), big.mark=",", trim=TRUE),
+    min=format(min, big.mark=",", trim=TRUE),
+    max=format(max, big.mark=",", trim=TRUE),
+    range = glue("{min}\\Hyphdash*{max}"),
+    n_samples = n_samples) %>%
+  select(nice_name, n_samples, total_seqs, median, range, sra_study) %>%
+  kableExtra::kable(format="markdown", booktabs=TRUE, escape=F, align="lrrrrr",
+    linesep="",
+    col.names = linebreak(
+      c(
+        "\\textbf{Study\\nobreakspace{}(Ref)}",
+        "\\textbf{Samples}",
+        "\\textbf{Total}\n\\textbf{sequences}",
+        "\\textbf{Median}\n\\textbf{sequences}",
+        "\\textbf{Range of}\n\\textbf{sequences}",
+        "\\textbf{SRA study}\n\\textbf{accession}"
+      ),
+      align="c")
+  )
+```
+\normalsize
 

### 2022-10-06 submission/manuscript.tex
Bring in Table 1 from Singletons papper

@@
 \usepackage[margin=1.0in]{geometry}
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
 \usepackage{graphicx}
@@
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-\usepackage{helvet}
-\renewcommand*\familydefault{\sfdefault}
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
 \usepackage{setspace}
@@
 \linenumbers
-\raggedright
+\modulolinenumbers
+\usepackage{helvet} % Helvetica font
+\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
+\usepackage[T1]{fontenc}
+\usepackage[shortcuts]{extdash}
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 \ifLuaTeX
@@
 \item
+  Motivation
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Problem of uneven sampling effort
+  \end{itemize}
+\item
   What is rarefaction? History, reason for rarefaction
@@
 
-\hypertarget{datasets}{%
-\subsubsection{Datasets}\label{datasets}}
+\textbf{\emph{Choice of datasets.}} I selected 16S rRNA gene sequence
+data from from 12 studies that characterized the variation in bacterial
+communities from diverse environments (Table 1). The specific studies
+were selected because their data was publicly accessible through the
+Sequence Read Archive, the original investigators sequenced the V4
+region of the 16S rRNA gene using paired 250 nt reads, and my previous
+familiarity with the data. The use of paired 250 nt reads to sequence
+the V4 region resulted in a near complete two-fold overap of the V4
+region resulting in high quality contigs with a low sequencing error
+rate (1). These data were processed through the standard sequence
+curation pipeline to generate operational taxonomic units (OTUs) using
+the mothur software package (1, 2). The original studies generated the
+sequence data by pooling separate PCR products that were generated by
+amplifying the V4 region of the 16S rRNA gene from the bacterial DNA in
+multiple samples. Because pooling equimolar quantities of DNA is frought
+with difficulties, it was common to observe wide variation in the number
+of sequences in each sample (Figure S1).
 
-\begin{itemize}
-\tightlist
-\item
-  Represented diverse communities
-\item
-  High quality V4 sequence data that consisted of paired 250 nt reads
-\item
-  Wide distribution in the number of reads within each dataset (Figure
-  S1, Table 1)
-\end{itemize}
+\textbf{\emph{Without rarefaction, metrics of alpha diversity are
+sensitive to sampling effort.}} To test the sesitivity of various
+approaches of measuring alpha diversity to sampling effort, I generated
+null models for each study. Under a null model, each sample from the
+same study would be expected to have the same alpha diversity regardless
+of the sampling effort. I assessed richness without any correction,
+using normalized OTU counts, with estimates based on non-parametric and
+parametric approaches, and using rarefaction. For each study, all of the
+approaches, except for rarefaction, showed a strong correlation between
+richness and the number of sequences in the sample (Figure 1A). Next, I
+assessed diversity using the Shannon diversity index and the inverse
+Simpson diversity index without any correction, using normalized OTU
+counts, and rarefaction; I also used a non-parametric estimator of
+Shannon diversity. The correlation between sampling depth and the
+diversity metric was not as strong as it was for richness and the
+inverse Simpson diversity values were less sensitive than the Shannon
+diversity values; however, the correlation to the rarefied diversity
+metrics were the lowest for all of the metrics and studies (Figure 1B).
+The rarefied alpha-diversity metrics consistently demonstrated a lack of
+sensitivity to sampling depth.
 
-\hypertarget{null-models}{%
-\subsubsection{Null models}\label{null-models}}
+\textbf{\emph{Without rarefaction, metrics of beta diversity are
+sensitive to sampling effort.}} To test the sesitivity of various
+approaches of measuring beta diversity to sampling effort, I used the
+same null models used for studying the sensitivity of alpha diversity.
+Under a null model, the ecological distance between any pair of samples
+would be the same regardless of the difference in the number of
+sequences observed in each sample. First, I analyzed the sensitivity of
+the Jaccard distance coefficient, which incorporates whether an OTU is
+present in each community and not their relative abundance. When
+calculating Jaccard distances using the uncorrected OTU counts,
+normalized OTU counts, relative abundances, and rarefaction only the
+rarefied data showed a lack of sensitivity to sampling effort (Figure
+2A). Second, I analyzed the sensitivity of the Bray-Curtis distance
+coefficient, which is a popular metric that incorporates the abundance
+of each OTU. Similar to what I observed with the Jaccard coefficient,
+only the rarefied data showed a lack of sensitivity to sampling effort
+(Figure 2B). Third, I calcualted Aitchison distances on raw OTU counts
+where the central log-ratio (CLR) was calculated by ignoring OTUs in
+samples with zero counts (robust CLR), adding a pseudocount of 1 to all
+OTU counts prior to calculating the CLR (one CLR), XXXXX XXXXX XXXXX
+XXXXX XXXXX XXXXX XXXXX XXXXX XXXXX XXXXX XXXXXXXXXX (n CLR), and
+imputing the value of zero counts (z CLR). Regardless of the approach,
+the Aitchison distances were all strongly sensitive to sampling effort
+(Figure 2C). Finally, I used the cumulative sum scaling (CSS)
+normalization from metagenomeSeq and variance stabilization technique
+(VST) from DeSeq2 prior to calculating Euclidean distances. Both
+approachs revealed a strong sensitivity to sampling effort (Figure 2D).
+Although Euclidean distances are not typically used on raw or rarefied
+count data in ecology, rarefied Euclidean distances were not sensitive
+to sampling effort. Across each of the beta diversity metrics and
+approaches used to account for uneven sampling effort and sparsity,
+rarefaction was the least sensitive approach to differences in sampling
+effort.
 
-\begin{itemize}
-\item
-  Null models were samplings, without replacement, of all samples from a
-  study that have been pooled together and sampled to the same depth of
-  sampling
-\item
-  Created 100 random sets of samples per dataset
-\item
-  Would expect a zero correlation between alpha diversity metric and
-  number of sequences in the community
-\item
-  Rarefaction was the only approach that resulted in a near-zero
-  correlation (Figure 1)
-\item
-  Would expect a zero correlation between beta diversity metrics and the
-  difference in the number of sequences in the samples going into the
-  distance calculation
-\item
-  Rarefaction was the only approach that resulted in a near-zero
-  correlation (Figure 2)
-\end{itemize}
+\textbf{\emph{Rarefaction limits the detection of false positives when
+sampling effort and treatment group are confounded.}} Next, I
+investigated the impact of the various strategies and metrics on falsely
+detecting a significant diffeerence using the the same communities
+generated from the null model in the analysis of alpha and beta
+diversity metrics. To test for differences in alpha and beta diversity I
+used the non-parametric Wilcoxon test and non-parametric
+permutation-based multivariate analysis of variance (PERMANOVA). First,
+within each study, I randomly assigned each sample to one of two
+treatment groups. My expectation was that approximately 5 of the 100
+(5\%) random tests for each comparison would yield a significant test
+result. Indeed, for each study and alpha and beta diversity metric and
+strategy for accounting for uneven sampling, approximately 5\% of the
+tests yielded a significant result (Figure 3). Second, within each
+study, I assigned samples with more than the median number of sequences
+per sample to one treatment group and the rest to another treatment
+group. If there is no sensitivity to sampling effort, I would again
+expect that 5\% of the tests would yield a significant result. In fact,
+only the rarefied data consistently resulted in a 5\% false positive
+rate for alpha and beta diversity metrics (Figure 4). These results
+align with the observed sensitivity of alpha and beta diversity metrics
+to sampling effort and underscore the value of rarefaction.
 
-\hypertarget{result-when-sampling-effort-is-confounded-with-treatment-group}{%
-\subsubsection{Result when sampling effort is confounded with treatment
-group}\label{result-when-sampling-effort-is-confounded-with-treatment-group}}
+\textbf{\emph{Rarefaction preserves the statistical power to detect
+differences between treatment groups.}} To assess the impact of
+different approaches to control for uneven sampling effort I performed
+two additional simulations. In the first simulation, for each study
+samples were randomly assigned to one of two treatment groups. Samples
+in the first treatment group were generated by sampling from the null
+distribution. Samples in the second treatment group were generated by
+perturbing the null distribution by increasing the relative abundance of
+10\% of the OTUs by 5\%. These values were determined after empirically
+searching for conditions that resulted in a large fraction of the
+randomizations yieleding a significant result across most of the
+studies. The fraction of tests that yielded a significant test was a
+measure of the statistical power for the test. Relative to the rarefied
+data, the power to detect differences in alpha and beta diversity was
+considerably lower for each of the strategies for handling uneven
+sampling effort. The power to detect differences in richness by all
+approaches was low (Figures 7 and 8). This was likely because the
+approach to generating the second community did not necessarily change
+the number of OTUs in each treatment group. To explore this further, in
+the second simulation the second treatment group was perturbed by
+removing 3\% of the OTUs from the model. Again, the rarefied data had a
+considerably higher statistical power than the other approaches when
+measuring richness (Figure 9). Both simulations highlight the value of
+rarefaction for preserving the statistical power to detect differences
+between treatment groups for measures of alpha and beta diversity.
 
-\begin{itemize}
-\item
-  Used Null model
-\item
-  Created two treatment groups for each study by dividing samples based
-  on whether they were above or below the median number of sequences per
-  sample
-\item
-  Used Wilcoxon test to detect differences between treatment groups for
-  alpha diversity, would expect \textasciitilde5\% of random sets to be
-  significant
-\item
-  Rarefaction the only approach that reliably resulted in the expected
-  Type I error rate
-\item
-  Used AMOVA test to detect differences between treatment groups for
-  beta diversity, would expect \textasciitilde5\% of random sets to be
-  significant
-\item
-  Rarefaction the only approach that reliably resulted in the expected
-  Type I error rate
-\end{itemize}
+\textbf{\emph{Increased rarefaction depth reduces intra-sample variation
+in alpha and beta diversity.}} To assess the sensitivity of alpha and
+beta diversity metrics to rarefaction depth, I again used the dataset
+generated using the null models and rarefied them to varying depths. For
+the alpha diversity metrics, the value of the metrics plateaued and the
+coefficient of variation (i.e., the mean divided by the standard
+deviation) between samples remained constant as the rarefaction depth
+increased (Figure 10). For beta diversity metrics, the distance between
+samples and the coefficient of variation between samples decreased as
+sampling depth increased (Figure 11). These results confirm that greater
+sequencing depth provides a more robust estimate of the metrics.
 
-\hypertarget{result-when-known-effect-size-imposed-between-two-groups-of-samples}{%
-\subsubsection{Result when known effect size imposed between two groups
-of
-samples}\label{result-when-known-effect-size-imposed-between-two-groups-of-samples}}
+\textbf{\emph{Most studies have a high level of sequencing coverage.}} I
+calculated the Good's coverage for the observed data and found that each
+of the studies had a minimum coverage greater than 90\% at their lowest
+sequencing depth (Figure 12A). This suggests that most studies have a
+high level of sequencing coverage. Next, I returned to the null models
+to ask how much sequencing effort was required to obtain higher levels
+of coverage. To obtain 95 and 99\% coverage, an average of XX and
+XX-fold more sequence data was required, respectively (Figure 12B).
+Although it is clear that most researchers would desire greater
+coverage, the sequencing effort required to acheive that sequencing
+depth would likely limit the number of samples that could be sequenced
+when controlling for costs.
 
-\begin{itemize}
-\item
-  For each dataset I divided samples into two groups. The probability
-  distribution for the first group was determined by the null model and
-  the second was determined by xxxxxxxxxxx
-\item
-  These parameters were selected because they produced results were some
-  of the simulations for the rarefied data did not result in a
-  significant test
-\item
-  Fraction of tests that were significant would be a measure of
-  statistical power
-\item
-  Used Wilcoxon test to detect differences between treatment groups for
-  alpha diversity
-\item
-  All tests resulted in reduced power to detect differences relative to
-  the power of using rarefaction
-\item
-  Used AMOVA test to detect differences between treatment groups for
-  beta diversity
-\item
-  All tests resulted in reduced power to detect differences relative to
-  the power of using rarefaction
-\item
-  This simulation would work well for measures based on the structure of
-  the community, but lacked resolution for those based on the richness
-  or membership of commmunities. For an additional simulation, the first
-  treatment group followed a null distribution for the second, XX\% of
-  OTUs were removed from the distribution
-\item
-  Tests for richness and membership (i.e.~Jaccard) resutled in reduced
-  power relative to rarefaction
-\end{itemize}
-
-\hypertarget{effect-of-sampling-depth-on-metrics}{%
-\subsubsection{Effect of sampling depth on
-metrics}\label{effect-of-sampling-depth-on-metrics}}
-
-\begin{itemize}
-\tightlist
-\item
-  Simulated rarefying null model datasets to retain varying percentage
-  of samples
-\item
-  High sampling depth resulted in less variation between samples for
-  alpha and beta diversity
-\end{itemize}
-
-\hypertarget{re-analysis-of-previously-published-datasets}{%
-\subsubsection{Re-analysis of previously published
-datasets}\label{re-analysis-of-previously-published-datasets}}
+\textbf{\emph{Evidence that not rarefying data can impact results in
+original studies.}} To assess the impact of using alternatives to
+rarefaction, I reassessed hypotheses from the human, mice, and
+peromyscus studies. These studies were selected because they originated
+from my research group and I know the datasets well. First, the study
+that published the human dataset was interested in the difference in the
+fecal microbial communities of patients with and without colorectal
+cancer. Here, I divided the patients into peope with and without screen
+relevant neoplasia (SRN). Among the alpha diversity metrics, the test
+were non-significant with similar effect sizes; however, the richness
+estimates obtained using parametric appraoches detected statistically
+significant differences in richness. Although the tests using
+Bray-Curtis distances were significant regardless of whether rarefaction
+was used, the effect sizes were smaller for the distances calculated
+using the CLR, VST, CSS transformations.
 
@@
 \end{itemize}
+
+Mouse
+
+Peromyscus
 
@@
   \item
-    Many recommended methods are borrowed from gene expression analysis
+    Many recommended methods are borrowed from gened expression analysis
   \item
@@
 \subsection{Materials and Methods}\label{materials-and-methods}}
+
+\textbf{Null models.} Null models were generated by randomly assigning
+each sequence in the study to an OTU and sample while keeping constant
+the number of sequences per sample and the total number of sequences in
+each OTU. Because the construction of the null models was a stochastic
+process, 100 replicates were geneated for each study.
 
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
+\begin{CSLReferences}{0}{1}
+\leavevmode\vadjust pre{\hypertarget{ref-Kozich2013}{}}%
+\CSLLeftMargin{1. }%
+\CSLRightInline{\textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter
+NT}, \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. {Development of
+a dual-index sequencing strategy and curation pipeline for analyzing
+amplicon sequence data on the MiSeq Illumina sequencing platform}.
+Applied and environmental microbiology \textbf{79}:5112--5120.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Schloss2009}{}}%
+\CSLLeftMargin{2. }%
+\CSLRightInline{\textbf{Schloss PD}, \textbf{Westcott SL},
+\textbf{Ryabin T}, \textbf{Hall JR}, \textbf{Hartmann M},
+\textbf{Hollister EB}, \textbf{Lesniewski RA}, \textbf{Oakley BB},
+\textbf{Parks DH}, \textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres
+B}, \textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
+Introducing mothur: Open-source, platform-independent,
+community-supported software for describing and comparing microbial
+communities. Applied and Environmental Microbiology
+\textbf{75}:7537--7541.
+doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Li2015}{}}%
+\CSLLeftMargin{3. }%
+\CSLRightInline{\textbf{Li Q}, \textbf{Heist EP}, \textbf{Moe LA}. 2015.
+Bacterial community structure and dynamics during corn-based bioethanol
+fermentation. Microbial Ecology \textbf{71}:409--421.
+doi:\href{https://doi.org/10.1007/s00248-015-0673-9}{10.1007/s00248-015-0673-9}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Baxter2016}{}}%
+\CSLLeftMargin{4. }%
+\CSLRightInline{\textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers
+MAM}, \textbf{Schloss PD}. 2016. Microbiota-based model improves the
+sensitivity of fecal immunochemical test for detecting colonic lesions.
+Genome Medicine \textbf{8}.
+doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Beall2015}{}}%
+\CSLLeftMargin{5. }%
+\CSLRightInline{\textbf{Beall BFN}, \textbf{Twiss MR}, \textbf{Smith
+DE}, \textbf{Oyserman BO}, \textbf{Rozmarynowycz MJ}, \textbf{Binding
+CE}, \textbf{Bourbonniere RA}, \textbf{Bullerjahn GS}, \textbf{Palmer
+ME}, \textbf{Reavie ED}, \textbf{Waters LMK}, \textbf{Woityra LWC},
+\textbf{McKay RML}. 2015. Ice cover extent drives phytoplankton and
+bacterial community structure in a large north-temperate lake:
+Implications for a warming climate. Environmental Microbiology
+\textbf{18}:1704--1719.
+doi:\href{https://doi.org/10.1111/1462-2920.12819}{10.1111/1462-2920.12819}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Henson2016}{}}%
+\CSLLeftMargin{6. }%
+\CSLRightInline{\textbf{Henson MW}, \textbf{Pitre DM}, \textbf{Weckhorst
+JL}, \textbf{Lanclos VC}, \textbf{Webber AT}, \textbf{Thrash JC}. 2016.
+Artificial seawater media facilitate cultivating members of the
+microbial majority from the gulf of mexico. {mSphere} \textbf{1}.
+doi:\href{https://doi.org/10.1128/msphere.00028-16}{10.1128/msphere.00028-16}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Baxter2014}{}}%
+\CSLLeftMargin{7. }%
+\CSLRightInline{\textbf{Baxter NT}, \textbf{Wan JJ}, \textbf{Schubert
+AM}, \textbf{Jenior ML}, \textbf{Myers P}, \textbf{Schloss PD}. 2014.
+Intra- and interindividual variations mask interspecies variation in the
+microbiota of sympatric peromyscus populations. Applied and
+Environmental Microbiology \textbf{81}:396--404.
+doi:\href{https://doi.org/10.1128/aem.02303-14}{10.1128/aem.02303-14}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-LevyBooth2018}{}}%
+\CSLLeftMargin{8. }%
+\CSLRightInline{\textbf{Levy-Booth DJ}, \textbf{Giesbrecht IJW},
+\textbf{Kellogg CTE}, \textbf{Heger TJ}, \textbf{D'Amore DV},
+\textbf{Keeling PJ}, \textbf{Hallam SJ}, \textbf{Mohn WW}. 2018.
+Seasonal and ecohydrological regulation of active microbial populations
+involved in {DOC}, {CO}2, and {CH}4 fluxes in temperate rainforest soil.
+The {ISME} Journal \textbf{13}:950--963.
+doi:\href{https://doi.org/10.1038/s41396-018-0334-3}{10.1038/s41396-018-0334-3}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Edwards2015}{}}%
+\CSLLeftMargin{9. }%
+\CSLRightInline{\textbf{Edwards J}, \textbf{Johnson C},
+\textbf{Santos-Medellín C}, \textbf{Lurie E}, \textbf{Podishetty NK},
+\textbf{Bhatnagar S}, \textbf{Eisen JA}, \textbf{Sundaresan V}. 2015.
+Structure, variation, and assembly of the root-associated microbiomes of
+rice. Proceedings of the National Academy of Sciences
+\textbf{112}:E911--E920.
+doi:\href{https://doi.org/10.1073/pnas.1414592112}{10.1073/pnas.1414592112}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Ettinger2017}{}}%
+\CSLLeftMargin{10. }%
+\CSLRightInline{\textbf{Ettinger CL}, \textbf{Williams SL},
+\textbf{Abbott JM}, \textbf{Stachowicz JJ}, \textbf{Eisen JA}. 2017.
+Microbiome succession during ammonification in eelgrass bed sediments.
+{PeerJ} \textbf{5}:e3674.
+doi:\href{https://doi.org/10.7717/peerj.3674}{10.7717/peerj.3674}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Graw2018}{}}%
+\CSLLeftMargin{11. }%
+\CSLRightInline{\textbf{Graw MF}, \textbf{DAngelo G}, \textbf{Borchers
+M}, \textbf{Thurber AR}, \textbf{Johnson JE}, \textbf{Zhang C},
+\textbf{Liu H}, \textbf{Colwell FS}. 2018. Energy gradients structure
+microbial communities across sediment horizons in deep marine sediments
+of the south china sea. Frontiers in Microbiology \textbf{9}.
+doi:\href{https://doi.org/10.3389/fmicb.2018.00729}{10.3389/fmicb.2018.00729}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Johnston2016}{}}%
+\CSLLeftMargin{12. }%
+\CSLRightInline{\textbf{Johnston ER}, \textbf{Rodriguez-R LM},
+\textbf{Luo C}, \textbf{Yuan MM}, \textbf{Wu L}, \textbf{He Z},
+\textbf{Schuur EAG}, \textbf{Luo Y}, \textbf{Tiedje JM}, \textbf{Zhou
+J}, \textbf{Konstantinidis KT}. 2016. Metagenomics reveals pervasive
+bacterial populations and reduced community diversity across the alaska
+tundra ecosystem. Frontiers in Microbiology \textbf{7}.
+doi:\href{https://doi.org/10.3389/fmicb.2016.00579}{10.3389/fmicb.2016.00579}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Hassell2018}{}}%
+\CSLLeftMargin{13. }%
+\CSLRightInline{\textbf{Hassell N}, \textbf{Tinker KA}, \textbf{Moore
+T}, \textbf{Ottesen EA}. 2018. Temporal and spatial dynamics in
+microbial community composition within a temperate stream network.
+Environmental Microbiology \textbf{20}:3560--3572.
+doi:\href{https://doi.org/10.1111/1462-2920.14311}{10.1111/1462-2920.14311}.}
+
 \end{CSLReferences}
 
+\bibliography{ref}
 \setlength{\parindent}{0in}
 \setlength{\leftskip}{0in}
+
+\newpage
+
+\textbf{Table 1. Summary of studies used in the analysis.} For all
+studies, the number of sequences used from each study was rarefied to
+the smallest sample size. A graphical represenation of the distribution
+of sample sizes for each study and the samples that were removed from
+each study are provided in Figure S1.
+
+\small
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1367}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0664}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1914}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1953}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2031}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2070}}@{}}
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+\textbf{Study\nobreakspace{}(Ref)}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+\textbf{Samples}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+\makecell[c]{\textbf{Total}\\\textbf{sequences}}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+\makecell[c]{\textbf{Median}\\\textbf{sequences}}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+\makecell[c]{\textbf{Range of}\\\textbf{sequences}}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+\makecell[c]{\textbf{SRA study}\\\textbf{accession}}
+\end{minipage} \\
+\midrule()
+\endhead
+Bioethanol~(3) & 95 & 3,970,972 & 16,014 & 3,690\Hyphdash*356,027 &
+SRP055545 \\
+Human~(4) & 490 & 20,828,275 & 32,452 & 10,439\Hyphdash*422,904 &
+SRP062005 \\
+Lake~(5) & 52 & 3,145,486 & 69,205 & 15,135\Hyphdash*110,993 &
+SRP050963 \\
+Marine~(6) & 7 & 1,484,068 & 213,091 & 132,895\Hyphdash*256,758 &
+SRP068101 \\
+Mice~(1) & 348 & 2,785,641 & 6,426 & 1,804\Hyphdash*30,311 &
+SRP192323 \\
+Peromyscus~(7) & 111 & 1,545,288 & 12,393 & 4,454\Hyphdash*33,502 &
+SRP044050 \\
+Rainforest~(8) & 69 & 936,666 & 11,464 & 4,880\Hyphdash*37,403 &
+ERP023747 \\
+Rice~(9) & 490 & 22,623,937 & 43,399 & 2,777\Hyphdash*192,200 &
+SRP044745 \\
+Seagrass~(10) & 286 & 4,135,440 & 13,538 & 1,830\Hyphdash*45,076 &
+SRP092441 \\
+Sediment~(11) & 58 & 1,151,389 & 17,606 & 7,686\Hyphdash*67,763 &
+SRP097192 \\
+Soil~(12) & 18 & 932,563 & 50,487 & 46,622\Hyphdash*58,935 &
+ERP012016 \\
+Stream~(13) & 201 & 21,017,610 & 90,621 & 8,931\Hyphdash*394,419 &
+SRP075852 \\
+\bottomrule()
+\end{longtable}
+
+\normalsize
 

### 2022-10-06 code/get_sample_summary_statistics.R

@@ -1,0 +1,56 @@
+#!/bin/env Rscript
+
+library(tidyverse)
+
+summarize_data <- function(environment) {
+
+	removed_samples <- paste0("data/", environment, "/data.remove_accnos") %>%
+		read_lines(.) %>%
+		str_split("-") %>%
+		unlist() %>%
+		enframe(., name=NULL, value="sample") %>%
+		mutate(sample = as.character(sample))
+
+	sra_study <- paste0("data/", environment, "/sra_info.tsv") %>%
+		read_tsv() %>%
+		pull("SRA_Study") %>%
+		unique()
+
+
+	paste0("data/", environment, "/data.group_count") %>%
+		read_tsv(col_names=c("sample", "n_seqs"),
+						 col_types=cols(sample=col_character(), n_seqs=col_integer())) %>%
+		anti_join(., removed_samples, by="sample") %>%
+		summarize(
+			n_samples=n(), total_seqs=sum(n_seqs), mean=mean(n_seqs),
+			min=min(n_seqs), l_quartile = quantile(n_seqs, 0.25),
+			median=median(n_seqs), u_quartile = quantile(n_seqs, 0.75),
+			max=max(n_seqs),
+			n_removed = removed_samples %>% filter(sample != "") %>% nrow(),
+			fold_difference = max/min) %>%
+		mutate(sra_study = sra_study)
+}
+
+
+tibble(
+		directory=c(
+			"bioethanol", "human", "lake", "marine",
+			"mice", "peromyscus", "rainforest",	"rice",
+			"seagrass", "sediment", "soil", "stream"
+		),
+		nice_name=c(
+			"Bioethanol", "Human", "Lake", "Marine",
+			"Mice", "Peromyscus", "Rainforest", "Rice",
+			"Seagrass", "Sediment", "Soil", "Stream"
+		),
+		reference=c(
+			"[@Li2015]", "[@Baxter2016]", "[@Beall2015]", "[@Henson2016]",
+			"[@Kozich2013]", "[@Baxter2014]", "[@LevyBooth2018]", "[@Edwards2015]",
+			"[@Ettinger2017]", "[@Graw2018]", "[@Johnston2016]", "[@Hassell2018]"
+		)
+	) %>%
+	group_by(directory) %>%
+	mutate(summary_data = map(directory, summarize_data)) %>%
+	unnest(summary_data) %>%
+	ungroup() %>%
+	write_tsv("data/process/study_summary_statistics.tsv")

### 2022-10-11 code/plot_coverage.R
Add rice data and rarefied coverage values

@@ -26,7 +26,7 @@ obs_cor <- map_dfr(obs_coverage_files, read_tsv, .id = "dataset",
                          .default = col_double()))
 
 
-datasets <- datasets[datasets != "rice" ]
+# datasets <- datasets[datasets != "rice" ]
 rare_coverage_files <- glue("data/{datasets}/data.otu.rarefy_coverage")
 names(rare_coverage_files) <- datasets
 
@@ -39,10 +39,14 @@ cor_plot <- obs_cor %>%
   ggplot(aes(x = norare_nseqs, y = norare_coverage,
              color = dataset, shape = dataset)) +
   geom_point(fill = "white") +
+  geom_boxplot(aes(x=325, y=rare_coverage),
+               outlier.shape = NULL, outlier.size = 0.75,
+               width = 0.2) +
   geom_line(data = cor_line, aes(x = nseqs, y = mean), color = "black") +
-  scale_x_log10(breaks = c(1e3, 1e4, 1e5, 1e6),
-                labels = c("10^3", "10^4", "10^5", "10^6"),
-                limits = c(1e3, 1e6)) +
+  geom_vline(xintercept = 800, color = "darkgray") +
+  scale_x_log10(breaks = c(325, 1e3, 1e4, 1e5, 1e6),
+                labels = c("R", "10^3", "10^4", "10^5", "10^6"),
+                limits = c(2e2, 1e6)) +
   scale_color_manual(name = NULL,
                     values = rep(c('#1b9e77','#d95f02','#7570b3'), 4),
                     breaks = pretty_datasets$plain,

### 2022-10-11 submission/manuscript.Rmd
Editing results section

@@
 }
+
+capitalize <- function(string) {
+  substr(string, 1, 1) <- toupper(substr(string, 1, 1))
+  string
+}
+
+datasets <- c("bioethanol", "human", "lake", "marine", "mice", "peromyscus",
+              "rainforest", "rice", "seagrass", "sediment", "soil", "stream")
+
+pretty_datasets <- tibble(
+  plain = datasets,
+  pretty = capitalize(plain)
+)
+
 ```
@@
   * Methods vary in their sensitivity to uneven sampling
+  * Compositional data analysis
 * Reasons behind "rarefaction is inadmissable"
@@
 
+This analysis included 16S rRNA gene sequence data from from 12 studies that characterized the variation in bacterial communities from diverse environments (Table 1).
+The original studies generated the sequence data by pooling separate PCR products that were generated by amplifying the V4 region of the 16S rRNA gene from the bacterial DNA in multiple samples.
+Because pooling equimolar quantities of DNA is frought with difficulties, it was common to observe wide variation in the number of sequences in each sample (Table 1 and Figure S1).
+
 ## Results
 
-***Choice of datasets.*** I selected 16S rRNA gene sequence data from from 12 studies that characterized the variation in bacterial communities from diverse environments (Table 1).
-The specific studies were selected because their data was publicly accessible through the Sequence Read Archive, the original investigators sequenced the V4 region of the 16S rRNA gene using paired 250 nt reads, and my previous familiarity with the data.
-The use of paired 250 nt reads to sequence the V4 region resulted in a near complete two-fold overap of the V4 region resulting in high quality contigs with a low sequencing error rate [@Kozich2013].
-These data were processed through the standard sequence curation pipeline to generate operational taxonomic units (OTUs) using the mothur software package [@Kozich2013; @Schloss2009].
-The original studies generated the sequence data by pooling separate PCR products that were generated by amplifying the V4 region of the 16S rRNA gene from the bacterial DNA in multiple samples.
-Because pooling equimolar quantities of DNA is frought with difficulties, it was common to observe wide variation in the number of sequences in each sample (Figure S1).
-
 ***Without rarefaction, metrics of alpha diversity are sensitive to sampling effort.*** To test the sesitivity of various approaches of measuring alpha diversity to sampling effort, I generated null models for each study.
-Under a null model, each sample from the same study would be expected to have the same alpha diversity regardless of the sampling effort.
-I assessed richness without any correction, using normalized OTU counts, with estimates based on non-parametric and parametric approaches, and using rarefaction.
+Under a null model, each community from the same study would be expected to have the same alpha diversity regardless of the sampling effort.
+I measured the richness of the communities in each study without any correction, using scaled ranked subsampling (SRS) normalized OTU counts, with estimates based on non-parametric and parametric approaches, and using rarefaction (e.g.
+Figure S2).
 For each study, all of the approaches, except for rarefaction, showed a strong correlation between richness and the number of sequences in the sample (Figure 1A).
 Next, I assessed diversity using the Shannon diversity index and the inverse Simpson diversity index without any correction, using normalized OTU counts, and rarefaction; I also used a non-parametric estimator of Shannon diversity.
-The correlation between sampling depth and the diversity metric was not as strong as it was for richness and the inverse Simpson diversity values were less sensitive than the Shannon diversity values; however, the correlation to the rarefied diversity metrics were the lowest for all of the metrics and studies (Figure 1B).
+The correlation between sampling depth and the diversity metric was not as strong as it was for richness and the inverse Simpson diversity values were less sensitive than the Shannon diversity values; however, the correlation to the rarefied diversity metrics were the lowest for all of the metrics and studies (Figure 1A).
 The rarefied alpha-diversity metrics consistently demonstrated a lack of sensitivity to sampling depth.
@@
 ***Without rarefaction, metrics of beta diversity are sensitive to sampling effort.*** To test the sesitivity of various approaches of measuring beta diversity to sampling effort, I used the same null models used for studying the sensitivity of alpha diversity.
-Under a null model, the ecological distance between any pair of samples would be the same regardless of the difference in the number of sequences observed in each sample.
-First, I analyzed the sensitivity of the Jaccard distance coefficient, which incorporates whether an OTU is present in each community and not their relative abundance.
-When calculating Jaccard distances using the uncorrected OTU counts, normalized OTU counts, relative abundances, and rarefaction only the rarefied data showed a lack of sensitivity to sampling effort (Figure 2A).
+Under a null model, the ecological distance between any pair of samples would be the same regardless of the difference in the number of sequences observed in each sample (e.g., Figure S3).
+First, I calculated the Jaccard distance coefficient between all pairs of communities within a study.
+The Jaccard distance coefficient is the fraction of OTUs that are unique to either community and does not account for the abundance of the OTUs.
+Jaccard distances were calcualted using the uncorrected OTU counts, with rarefaction, relative abundances, and following normalization using cumulative sum scaling (CSS) and SRS.
+Only the rarefied distances showed a lack of sensitivity to sampling effort (Figure 1B).
 Second, I analyzed the sensitivity of the Bray-Curtis distance coefficient, which is a popular metric that incorporates the abundance of each OTU.
-Similar to what I observed with the Jaccard coefficient, only the rarefied data showed a lack of sensitivity to sampling effort (Figure 2B).
-Third, I calcualted Aitchison distances on raw OTU counts where the central log-ratio (CLR) was calculated by ignoring OTUs in samples with zero counts (robust CLR), adding a pseudocount of 1 to all OTU counts prior to calculating the CLR (one CLR), XXXXX XXXXX XXXXX XXXXX XXXXX XXXXX XXXXX XXXXX XXXXX XXXXX XXXXXXXXXX (n CLR), and imputing the value of zero counts (z CLR).
-Regardless of the approach, the Aitchison distances were all strongly sensitive to sampling effort (Figure 2C).
-Finally, I used the cumulative sum scaling (CSS) normalization from metagenomeSeq and variance stabilization technique (VST) from DeSeq2 prior to calculating Euclidean distances.
-Both approachs revealed a strong sensitivity to sampling effort (Figure 2D).
+Similar to what I observed with the Jaccard coefficient, only the rarefied data showed a lack of sensitivity to sampling effort (Figure 1B).
+Third, I calcualted the Euclidean distance on raw OTU counts where the central log-ratio (CLR) was calculated (i.e., Aitchison distances) by ignoring OTUs in samples with zero counts (Robust CLR), adding a pseudocount of one to all OTU counts prior to calculating the CLR (One CLR), adding a pseudocount of one divided by the total number of sequences obtained for the community (Nudge CLR), and imputing the value of zero counts (Zero CLR).
+The Aitchison distances were all strongly sensitive to sampling effort (Figure 1B).
+Finally, I used the variance stabilization technique (VST) from DeSeq2 prior to calculating Euclidean distances.
+Again, there was a strong sensitivity to sampling effort (Figure 1B).
 Although Euclidean distances are not typically used on raw or rarefied count data in ecology, rarefied Euclidean distances were not sensitive to sampling effort.
@@
 
-***Rarefaction limits the detection of false positives when sampling effort and treatment group are confounded.*** Next, I investigated the impact of the various strategies and metrics on falsely detecting a significant diffeerence using the the same communities generated from the null model in the analysis of alpha and beta diversity metrics.
+***Rarefaction limits the detection of false positives when sampling effort and treatment group are confounded.*** Next, I investigated the impact of the various strategies and metrics on falsely detecting a significant difference using the the same communities generated from the null model in the analysis of alpha and beta diversity metrics.
 To test for differences in alpha and beta diversity I used the non-parametric Wilcoxon test and non-parametric permutation-based multivariate analysis of variance (PERMANOVA).
@@
 My expectation was that approximately 5 of the 100 (5%) random tests for each comparison would yield a significant test result.
-Indeed, for each study and alpha and beta diversity metric and strategy for accounting for uneven sampling, approximately 5% of the tests yielded a significant result (Figure 3).
+Indeed, for each study and alpha and beta diversity metric and strategy for accounting for uneven sampling, approximately 5% of the tests yielded a significant result (Figure 2).
 Second, within each study, I assigned samples with more than the median number of sequences per sample to one treatment group and the rest to another treatment group.
-If there is no sensitivity to sampling effort, I would again expect that 5% of the tests would yield a significant result.
-In fact, only the rarefied data consistently resulted in a 5% false positive rate for alpha and beta diversity metrics (Figure 4).
+If there was no sensitivity to sampling effort, I would have expected that 5% of the tests would yield a significant result.
+In fact, only the rarefied data consistently resulted in a 5% false positive rate for alpha and beta diversity metrics (Figure 2).
 These results align with the observed sensitivity of alpha and beta diversity metrics to sampling effort and underscore the value of rarefaction.
@@
 ***Rarefaction preserves the statistical power to detect differences between treatment groups.*** To assess the impact of different approaches to control for uneven sampling effort I performed two additional simulations.
-In the first simulation, for each study samples were randomly assigned to one of two treatment groups.
-Samples in the first treatment group were generated by sampling from the null distribution.
+In the first simulation, for each study communities were randomly assigned to one of two treatment groups.
+Communities in the first treatment group were generated by sampling from the null distribution.
 Samples in the second treatment group were generated by perturbing the null distribution by increasing the relative abundance of 10% of the OTUs by 5%.
@@
 The fraction of tests that yielded a significant test was a measure of the statistical power for the test.
-Relative to the rarefied data, the power to detect differences in alpha and beta diversity was considerably lower for each of the strategies for handling uneven sampling effort.
-The power to detect differences in richness by all approaches was low (Figures 7 and 8).
-This was likely because the approach to generating the second community did not necessarily change the number of OTUs in each treatment group.
-To explore this further, in the second simulation the second treatment group was perturbed by removing 3% of the OTUs from the model.
-Again, the rarefied data had a considerably higher statistical power than the other approaches when measuring richness (Figure 9).
-Both simulations highlight the value of rarefaction for preserving the statistical power to detect differences between treatment groups for measures of alpha and beta diversity.
+The power to detect differences in richness by all approaches was low (Figure 4A).
+This was likely because the approach for generating the perturbed community did not necessarily change the number of OTUs in each treatment group.
+Regardless, the simulations testing differencse in richness using the Rice and Stream datasets had the greatest power when the richness data were rarefied.
+To explore this further, in a second simulation the second treatment group was perturbed by removing 3% of the OTUs from the model.
+As suggested by the first simulation, the rarefied richness data had a higher statistical power than the other approaches when measuring richness (Figure 5).
+The simulations testing the power to detect differences in Shannon diversity also showed that rarefied data performed other methods (Figure 4A).
+When testing for differences in the Inverse Simpson diversity index the the difference between rarefaction and the other methods was negligible (Figure 4A).
+For tests of beta diversity I found that rarefaction was the most reliable approach to maintain statistical power to detect differences between two communities.
+Among the tests using the Jaccard and Bray-Curtis metrics, raw count data and CSS normalized data had little power relative to rarefied, relative abundance, and SRS normalized data.
+The differences in power between rarefied, relative abundance, and SRS normalized data was small, but if there were differences, the power obtained using rarefied data was greater than the other methods.
+Among the tests using Euclidean distances, using raw counts and CLR and DeSeq2 transformed data had little power relative to the distances calcualted using rarefied and relative abundance data.
+This power-based analysis of the simulated communities using different methods of handling uneven sample sizes demonstrated the value of rarefaction for preserving the statistical power to detect differences between treatment groups for measures of alpha and beta diversity.
 
-***Increased rarefaction depth reduces intra-sample variation in alpha and beta diversity.*** To assess the sensitivity of alpha and beta diversity metrics to rarefaction depth, I again used the dataset generated using the null models and rarefied them to varying depths.
-For the alpha diversity metrics, the value of the metrics plateaued and the coefficient of variation (i.e., the mean divided by the standard deviation) between samples remained constant as the rarefaction depth increased (Figure 10).
-For beta diversity metrics, the distance between samples and the coefficient of variation between samples decreased as sampling depth increased (Figure 11).
-These results confirm that greater sequencing depth provides a more robust estimate of the metrics.
+```{r coverage}
+loss <- table_1 %>%
+  select(nice_name, fold_difference) %>%
+  mutate(loss = round(100* (1 - 1/fold_difference), 1)) %>%
+  slice_max(loss)
 
-***Most studies have a high level of sequencing coverage.*** I calculated the Good's coverage for the observed data and found that each of the studies had a minimum coverage greater than 90% at their lowest sequencing depth (Figure 12A).
-This suggests that most studies have a high level of sequencing coverage.
-Next, I returned to the null models to ask how much sequencing effort was required to obtain higher levels of coverage.
-To obtain 95 and 99% coverage, an average of XX and XX-fold more sequence data was required, respectively (Figure 12B).
-Although it is clear that most researchers would desire greater coverage, the sequencing effort required to acheive that sequencing depth would likely limit the number of samples that could be sequenced when controlling for costs.
+obs_coverage_files <- glue("../data/{datasets}/data.otu.obs_coverage")
+names(obs_coverage_files) <- datasets
 
-***Evidence that not rarefying data can impact results in original studies.*** To assess the impact of using alternatives to rarefaction, I reassessed hypotheses from the human, mice, and peromyscus studies.
-These studies were selected because they originated from my research group and I know the datasets well.
-First, the study that published the human dataset was interested in the difference in the fecal microbial communities of patients with and without colorectal cancer.
-Here, I divided the patients into peope with and without screen relevant neoplasia (SRN).
-Among the alpha diversity metrics, the test were non-significant with similar effect sizes; however, the richness estimates obtained using parametric appraoches detected statistically significant differences in richness.
-Although the tests using Bray-Curtis distances were significant regardless of whether rarefaction was used, the effect sizes were smaller for the distances calculated using the CLR, VST, CSS transformations.
+obs_cor <- map_dfr(obs_coverage_files, read_tsv, .id = "dataset",
+        col_types = cols(group = col_character(),
+                         .default = col_double()))
 
-* Human study: Did not alter effect size or significance of alpha or beta diversity, but did result in reduced effect sizes for measures of richness and non-parametric estimators of richness; breakaway detected a difference
+obs_median <- obs_cor %>%
+  group_by(dataset) %>%
+  summarize(med = round(100 * median(norare_coverage), 1)) %>%
+  arrange(med) %>%
+  slice(c(1, n())) %>%
+  mutate(med = format(med, 1))
 
-Mouse
+rare_median <- obs_cor %>%
+  group_by(dataset) %>%
+  summarize(med = round(100 * median(rare_coverage), 1)) %>%
+  arrange(med) %>%
+  filter(med < 90)
 
-Peromyscus
+rare_coverage_files <- glue("../data/{datasets}/data.otu.rarefy_coverage")
+names(rare_coverage_files) <- datasets
+
+cor_line <- map_dfr(rare_coverage_files, read_tsv, .id = "dataset",
+        col_types = cols(.default = col_double()))
+
+extrapolate <- cor_line %>%
+  mutate(ninety = abs(mean - 0.90),
+         ninetyfive = abs(mean - 0.95),
+         ninetynine = abs(mean - 0.99)) %>%
+  group_by(dataset) %>%
+  summarize(ninety = nseqs[which.min(ninety)],
+            ninetyfive = nseqs[which.min(ninetyfive)],
+            ninetynine = nseqs[which.min(ninetynine)]) %>%
+  mutate(ninetyfive = ninetyfive / ninety - 1,
+         ninetynine = ninetynine / ninety - 1) %>%
+  summarize(m = round(mean(ninetyfive), 1),
+            n = round(mean(ninetynine), 1)
+            )
+```
+
+***Increased rarefaction depth reduces intra-sample variation in alpha and beta diversity.*** Once concern with rarefying communities is the perceived loss of sequencing information when more a large fraction of data appears to be removed when the community with the greatest sequencing depth is rarefied to the size of the community with the least (e.g., `r paste0(loss$loss, "% with the ", loss$nice_name, " dataset")`).
+To assess the sensitivity of alpha and beta diversity metrics to rarefaction depth, I again used the dataset generated using the null models, but rarefied each community to varying sampling depths (Figure 6).
+The richness values increased with sampling effort as rare OTUs would continue be detected.
+In contrast, the Shannon diversity and Bray-Curtis values plateaued with increased sampling effort.
+This result was expected since increased sampling would lead to increased precision in the measured abundance of OTUs.
+Next, I measured the coefficient of variation (i.e., the mean divided by the standard deviation) between samples for richness, Shannon diversity, and Bray-Curtis distances.
+Although the richness values appeared to increased unbounded with smapling effort, the coefficient of variation for each dataset was relatively stable.
+In general, the coefficient of variation increased slightly with sampling depth only to decline once smaller samples were removed from the analysis at higher sampling depths.
+Interestingly, the coefficient of variation between Shannon diveristy values decreased towards zero with increased sampling effort and the coefficient of variation between Bray-Curtis distances tended to increased.
+Regardless, the coefficients of variation were relatively small.
+
+***Most studies have a high level of sequencing coverage.*** To explore the concern over loss of sequencing depth further, I calculated the Good's coverage for the observed data.
+The median coverage for each dataset ranged between `r oxford_comma(obs_median$med)`% for the `r oxford_comma(capitalize(obs_median$dataset))` datasets, respectively (Figure 7).
+When I rarefied each dataset to the size of the smallest community in the dataset, with the exception of the `r oxford_comma(capitalize(rare_median$dataset))` datasets, the median coverage for the rarefied communities was still greater than 90%.
+These results suggest that most studies had a level of sequencing coverage that aligned with the diversity of the communities.
+Next, I used the null model for each dataset to ask how much sequencing effort was required to obtain higher levels of coverage.
+To obtain 95 and 99% coverage, an average of `r extrapolate$m` and `r extrapolate$n`-fold more sequence data was estimated to be required than was required to obtain 90% coverage, respectively (Figure 7).
+As suggested by the simulated coverages curve in Figure 7, these estimates are conservative.
+Regardless, the sequencing effort required to acheive higher sequencing depth would likely limit the number of samples that could be sequenced when controlling for costs.
+Although it may be disconcerting to rarefy to a sequencing depth that is considerably lower than that obtained for the best sequenced community in a study, sequencing coverage for many environments is probably adequate even at the lower sequencing depth.
+Of course, the results above have demonstrated that rarefaction is necessary to avoid problems with making inferences.
 
@@
 
-**Null models.** Null models were generated by randomly assigning each sequence in the study to an OTU and sample while keeping constant the number of sequences per sample and the total number of sequences in each OTU.
+**Choice of datasets.** The specific studies were selected because their data was publicly accessible through the Sequence Read Archive, the original investigators sequenced the V4 region of the 16S rRNA gene using paired 250 nt reads, and my previous familiarity with the data.
+The use of paired 250 nt reads to sequence the V4 region resulted in a near complete two-fold overap of the V4 region resulting in high quality contigs with a low sequencing error rate [@Kozich2013].
+These data were processed through the standard sequence curation pipeline to generate operational taxonomic units (OTUs) using the mothur software package [@Kozich2013; @Schloss2009].
+
+**Null community model.** Null models were generated by randomly assigning each sequence in the study to an OTU and sample while keeping constant the number of sequences per sample and the total number of sequences in each OTU.
 Because the construction of the null models was a stochastic process, 100 replicates were geneated for each study.
+
+**10%/5% model**
+
+**Remove model**
+
+**Null treatment model**
+
+**Size-based treatment model**
 
@@
 
+***doi: 10.1038/nmeth.2658.*** Should CSS be used in alpha diversity analysis?
+No.
+The richness and shannon diversity values were no different from those obtained with the raw abundances.
+From the paper, "The relative proportion of the features is unaffected by the normalization".
+CSS paper refers to relative abundance approach as total-sum normalization (TSN)
+
+Robust CLR - remove zero counts and calculate CLR
+One CLR - add a pseudocount of 1 to all observations
+Nudge CLR - add a pseudocount of 1/total number of sequences in sample
+Zero CLR - Impute the value of zeroes using zCompositions package
+
+***https://edepot.wur.nl/547087.*** "Log- ratio PCA is designed to give results that are library size- independent.
+However, as we demonstrated mathematically and with examples based on simulated and real data, log- ratio PCA becomes library size- dependent, if there are many infrequent taxa (many zeroes) and library sizes differ largely.
+In this situation, the row centring used in log- ratio PCA brings an effect of r (the row mean of the log- transformed counts) in the clr- transformed matrix.
+Note that this effect is irrespective of whether or not these infrequent taxa are genuine or due to sequencing noise or allocation error.
+This library size dependence is unexpected in the sense that, after applying the clr, the transformed matrix is free of the effect of the row totals for strictly positive data (yij>0 for all i and j).
+We additionally demonstrate that library size variability causes a loss in power in detecting an effect of x with log- ratio RDA.
+If there is additionally a correlation between treatment and the library size, the type 1 error for detecting the effect of x can be seriously inflated."
+
+***https://www.nature.com/articles/s41522-020-00160-w.*** "An important characteristic of a feature table is that it is typically sparse, sometimes as many as ~90% are zero entries21, which creates a challenge for analyzing rare taxa.
+A quick and simple strategy to deal with excess zeros is to add a small positive constant (e.g.
+1) called pseudo-count14,22 to each cell of the feature table.
+The addition of a pseudo-count becomes necessary when using methods of analysis that require log transformation of the observed counts.
+Even though adding a pseudo-count is simple and widely used, the choice of the pseudo-count is ad hoc.
+Studies have shown that differential abundance or clustering results could be sensitive to the choice of pseudo count23,24.
+Although different values of pseudo counts have been discussed in the literature23,24,25,26, to the best of our knowledge, there is no consensus on how to choose the optimal value.
+Other strategies involve modeling zero counts by some probability models21,27."
+
+https://academic.oup.com/bioinformatics/article/34/16/2870/4956011
+https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6755255/
 \vspace{10mm}

### 2022-10-11 submission/manuscript.tex
Editing results section

@@
     Methods vary in their sensitivity to uneven sampling
+  \item
+    Compositional data analysis
   \end{itemize}
@@
 
+This analysis included 16S rRNA gene sequence data from from 12 studies
+that characterized the variation in bacterial communities from diverse
+environments (Table 1). The original studies generated the sequence data
+by pooling separate PCR products that were generated by amplifying the
+V4 region of the 16S rRNA gene from the bacterial DNA in multiple
+samples. Because pooling equimolar quantities of DNA is frought with
+difficulties, it was common to observe wide variation in the number of
+sequences in each sample (Table 1 and Figure S1).
+
 \hypertarget{results}{%
 \subsection{Results}\label{results}}
-
-\textbf{\emph{Choice of datasets.}} I selected 16S rRNA gene sequence
-data from from 12 studies that characterized the variation in bacterial
-communities from diverse environments (Table 1). The specific studies
-were selected because their data was publicly accessible through the
-Sequence Read Archive, the original investigators sequenced the V4
-region of the 16S rRNA gene using paired 250 nt reads, and my previous
-familiarity with the data. The use of paired 250 nt reads to sequence
-the V4 region resulted in a near complete two-fold overap of the V4
-region resulting in high quality contigs with a low sequencing error
-rate (1). These data were processed through the standard sequence
-curation pipeline to generate operational taxonomic units (OTUs) using
-the mothur software package (1, 2). The original studies generated the
-sequence data by pooling separate PCR products that were generated by
-amplifying the V4 region of the 16S rRNA gene from the bacterial DNA in
-multiple samples. Because pooling equimolar quantities of DNA is frought
-with difficulties, it was common to observe wide variation in the number
-of sequences in each sample (Figure S1).
 
@@
 approaches of measuring alpha diversity to sampling effort, I generated
-null models for each study. Under a null model, each sample from the
+null models for each study. Under a null model, each community from the
 same study would be expected to have the same alpha diversity regardless
-of the sampling effort. I assessed richness without any correction,
-using normalized OTU counts, with estimates based on non-parametric and
-parametric approaches, and using rarefaction. For each study, all of the
-approaches, except for rarefaction, showed a strong correlation between
-richness and the number of sequences in the sample (Figure 1A). Next, I
-assessed diversity using the Shannon diversity index and the inverse
-Simpson diversity index without any correction, using normalized OTU
-counts, and rarefaction; I also used a non-parametric estimator of
-Shannon diversity. The correlation between sampling depth and the
-diversity metric was not as strong as it was for richness and the
-inverse Simpson diversity values were less sensitive than the Shannon
-diversity values; however, the correlation to the rarefied diversity
-metrics were the lowest for all of the metrics and studies (Figure 1B).
-The rarefied alpha-diversity metrics consistently demonstrated a lack of
-sensitivity to sampling depth.
+of the sampling effort. I measured the richness of the communities in
+each study without any correction, using scaled ranked subsampling (SRS)
+normalized OTU counts, with estimates based on non-parametric and
+parametric approaches, and using rarefaction (e.g.~Figure S2). For each
+study, all of the approaches, except for rarefaction, showed a strong
+correlation between richness and the number of sequences in the sample
+(Figure 1A). Next, I assessed diversity using the Shannon diversity
+index and the inverse Simpson diversity index without any correction,
+using normalized OTU counts, and rarefaction; I also used a
+non-parametric estimator of Shannon diversity. The correlation between
+sampling depth and the diversity metric was not as strong as it was for
+richness and the inverse Simpson diversity values were less sensitive
+than the Shannon diversity values; however, the correlation to the
+rarefied diversity metrics were the lowest for all of the metrics and
+studies (Figure 1A). The rarefied alpha-diversity metrics consistently
+demonstrated a lack of sensitivity to sampling depth.
 
@@
 would be the same regardless of the difference in the number of
-sequences observed in each sample. First, I analyzed the sensitivity of
-the Jaccard distance coefficient, which incorporates whether an OTU is
-present in each community and not their relative abundance. When
-calculating Jaccard distances using the uncorrected OTU counts,
-normalized OTU counts, relative abundances, and rarefaction only the
-rarefied data showed a lack of sensitivity to sampling effort (Figure
-2A). Second, I analyzed the sensitivity of the Bray-Curtis distance
-coefficient, which is a popular metric that incorporates the abundance
-of each OTU. Similar to what I observed with the Jaccard coefficient,
-only the rarefied data showed a lack of sensitivity to sampling effort
-(Figure 2B). Third, I calcualted Aitchison distances on raw OTU counts
-where the central log-ratio (CLR) was calculated by ignoring OTUs in
-samples with zero counts (robust CLR), adding a pseudocount of 1 to all
-OTU counts prior to calculating the CLR (one CLR), XXXXX XXXXX XXXXX
-XXXXX XXXXX XXXXX XXXXX XXXXX XXXXX XXXXX XXXXXXXXXX (n CLR), and
-imputing the value of zero counts (z CLR). Regardless of the approach,
-the Aitchison distances were all strongly sensitive to sampling effort
-(Figure 2C). Finally, I used the cumulative sum scaling (CSS)
-normalization from metagenomeSeq and variance stabilization technique
-(VST) from DeSeq2 prior to calculating Euclidean distances. Both
-approachs revealed a strong sensitivity to sampling effort (Figure 2D).
-Although Euclidean distances are not typically used on raw or rarefied
-count data in ecology, rarefied Euclidean distances were not sensitive
-to sampling effort. Across each of the beta diversity metrics and
-approaches used to account for uneven sampling effort and sparsity,
-rarefaction was the least sensitive approach to differences in sampling
-effort.
+sequences observed in each sample (e.g., Figure S3). First, I calculated
+the Jaccard distance coefficient between all pairs of communities within
+a study. The Jaccard distance coefficient is the fraction of OTUs that
+are unique to either community and does not account for the abundance of
+the OTUs. Jaccard distances were calcualted using the uncorrected OTU
+counts, with rarefaction, relative abundances, and following
+normalization using cumulative sum scaling (CSS) and SRS. Only the
+rarefied distances showed a lack of sensitivity to sampling effort
+(Figure 1B). Second, I analyzed the sensitivity of the Bray-Curtis
+distance coefficient, which is a popular metric that incorporates the
+abundance of each OTU. Similar to what I observed with the Jaccard
+coefficient, only the rarefied data showed a lack of sensitivity to
+sampling effort (Figure 1B). Third, I calcualted the Euclidean distance
+on raw OTU counts where the central log-ratio (CLR) was calculated
+(i.e., Aitchison distances) by ignoring OTUs in samples with zero counts
+(Robust CLR), adding a pseudocount of one to all OTU counts prior to
+calculating the CLR (One CLR), adding a pseudocount of one divided by
+the total number of sequences obtained for the community (Nudge CLR),
+and imputing the value of zero counts (Zero CLR). The Aitchison
+distances were all strongly sensitive to sampling effort (Figure 1B).
+Finally, I used the variance stabilization technique (VST) from DeSeq2
+prior to calculating Euclidean distances. Again, there was a strong
+sensitivity to sampling effort (Figure 1B). Although Euclidean distances
+are not typically used on raw or rarefied count data in ecology,
+rarefied Euclidean distances were not sensitive to sampling effort.
+Across each of the beta diversity metrics and approaches used to account
+for uneven sampling effort and sparsity, rarefaction was the least
+sensitive approach to differences in sampling effort.
 
@@
 investigated the impact of the various strategies and metrics on falsely
-detecting a significant diffeerence using the the same communities
+detecting a significant difference using the the same communities
 generated from the null model in the analysis of alpha and beta
@@
 strategy for accounting for uneven sampling, approximately 5\% of the
-tests yielded a significant result (Figure 3). Second, within each
+tests yielded a significant result (Figure 2). Second, within each
 study, I assigned samples with more than the median number of sequences
 per sample to one treatment group and the rest to another treatment
-group. If there is no sensitivity to sampling effort, I would again
-expect that 5\% of the tests would yield a significant result. In fact,
-only the rarefied data consistently resulted in a 5\% false positive
-rate for alpha and beta diversity metrics (Figure 4). These results
-align with the observed sensitivity of alpha and beta diversity metrics
-to sampling effort and underscore the value of rarefaction.
+group. If there was no sensitivity to sampling effort, I would have
+expected that 5\% of the tests would yield a significant result. In
+fact, only the rarefied data consistently resulted in a 5\% false
+positive rate for alpha and beta diversity metrics (Figure 2). These
+results align with the observed sensitivity of alpha and beta diversity
+metrics to sampling effort and underscore the value of rarefaction.
 
@@
 two additional simulations. In the first simulation, for each study
-samples were randomly assigned to one of two treatment groups. Samples
-in the first treatment group were generated by sampling from the null
-distribution. Samples in the second treatment group were generated by
-perturbing the null distribution by increasing the relative abundance of
-10\% of the OTUs by 5\%. These values were determined after empirically
-searching for conditions that resulted in a large fraction of the
-randomizations yieleding a significant result across most of the
+communities were randomly assigned to one of two treatment groups.
+Communities in the first treatment group were generated by sampling from
+the null distribution. Samples in the second treatment group were
+generated by perturbing the null distribution by increasing the relative
+abundance of 10\% of the OTUs by 5\%. These values were determined after
+empirically searching for conditions that resulted in a large fraction
+of the randomizations yieleding a significant result across most of the
 studies. The fraction of tests that yielded a significant test was a
-measure of the statistical power for the test. Relative to the rarefied
-data, the power to detect differences in alpha and beta diversity was
-considerably lower for each of the strategies for handling uneven
-sampling effort. The power to detect differences in richness by all
-approaches was low (Figures 7 and 8). This was likely because the
-approach to generating the second community did not necessarily change
-the number of OTUs in each treatment group. To explore this further, in
-the second simulation the second treatment group was perturbed by
-removing 3\% of the OTUs from the model. Again, the rarefied data had a
-considerably higher statistical power than the other approaches when
-measuring richness (Figure 9). Both simulations highlight the value of
-rarefaction for preserving the statistical power to detect differences
-between treatment groups for measures of alpha and beta diversity.
+measure of the statistical power for the test. The power to detect
+differences in richness by all approaches was low (Figure 4A). This was
+likely because the approach for generating the perturbed community did
+not necessarily change the number of OTUs in each treatment group.
+Regardless, the simulations testing differencse in richness using the
+Rice and Stream datasets had the greatest power when the richness data
+were rarefied. To explore this further, in a second simulation the
+second treatment group was perturbed by removing 3\% of the OTUs from
+the model. As suggested by the first simulation, the rarefied richness
+data had a higher statistical power than the other approaches when
+measuring richness (Figure 5). The simulations testing the power to
+detect differences in Shannon diversity also showed that rarefied data
+performed other methods (Figure 4A). When testing for differences in the
+Inverse Simpson diversity index the the difference between rarefaction
+and the other methods was negligible (Figure 4A). For tests of beta
+diversity I found that rarefaction was the most reliable approach to
+maintain statistical power to detect differences between two
+communities. Among the tests using the Jaccard and Bray-Curtis metrics,
+raw count data and CSS normalized data had little power relative to
+rarefied, relative abundance, and SRS normalized data. The differences
+in power between rarefied, relative abundance, and SRS normalized data
+was small, but if there were differences, the power obtained using
+rarefied data was greater than the other methods. Among the tests using
+Euclidean distances, using raw counts and CLR and DeSeq2 transformed
+data had little power relative to the distances calcualted using
+rarefied and relative abundance data. This power-based analysis of the
+simulated communities using different methods of handling uneven sample
+sizes demonstrated the value of rarefaction for preserving the
+statistical power to detect differences between treatment groups for
+measures of alpha and beta diversity.
 
 \textbf{\emph{Increased rarefaction depth reduces intra-sample variation
-in alpha and beta diversity.}} To assess the sensitivity of alpha and
-beta diversity metrics to rarefaction depth, I again used the dataset
-generated using the null models and rarefied them to varying depths. For
-the alpha diversity metrics, the value of the metrics plateaued and the
-coefficient of variation (i.e., the mean divided by the standard
-deviation) between samples remained constant as the rarefaction depth
-increased (Figure 10). For beta diversity metrics, the distance between
-samples and the coefficient of variation between samples decreased as
-sampling depth increased (Figure 11). These results confirm that greater
-sequencing depth provides a more robust estimate of the metrics.
+in alpha and beta diversity.}} Once concern with rarefying communities
+is the perceived loss of sequencing information when more a large
+fraction of data appears to be removed when the community with the
+greatest sequencing depth is rarefied to the size of the community with
+the least (e.g., 99\% with the Bioethanol dataset). To assess the
+sensitivity of alpha and beta diversity metrics to rarefaction depth, I
+again used the dataset generated using the null models, but rarefied
+each community to varying sampling depths (Figure 6). The richness
+values increased with sampling effort as rare OTUs would continue be
+detected. In contrast, the Shannon diversity and Bray-Curtis values
+plateaued with increased sampling effort. This result was expected since
+increased sampling would lead to increased precision in the measured
+abundance of OTUs. Next, I measured the coefficient of variation (i.e.,
+the mean divided by the standard deviation) between samples for
+richness, Shannon diversity, and Bray-Curtis distances. Although the
+richness values appeared to increased unbounded with smapling effort,
+the coefficient of variation for each dataset was relatively stable. In
+general, the coefficient of variation increased slightly with sampling
+depth only to decline once smaller samples were removed from the
+analysis at higher sampling depths. Interestingly, the coefficient of
+variation between Shannon diveristy values decreased towards zero with
+increased sampling effort and the coefficient of variation between
+Bray-Curtis distances tended to increased. Regardless, the coefficients
+of variation were relatively small.
 
-\textbf{\emph{Most studies have a high level of sequencing coverage.}} I
-calculated the Good's coverage for the observed data and found that each
-of the studies had a minimum coverage greater than 90\% at their lowest
-sequencing depth (Figure 12A). This suggests that most studies have a
-high level of sequencing coverage. Next, I returned to the null models
-to ask how much sequencing effort was required to obtain higher levels
-of coverage. To obtain 95 and 99\% coverage, an average of XX and
-XX-fold more sequence data was required, respectively (Figure 12B).
-Although it is clear that most researchers would desire greater
-coverage, the sequencing effort required to acheive that sequencing
-depth would likely limit the number of samples that could be sequenced
-when controlling for costs.
-
-\textbf{\emph{Evidence that not rarefying data can impact results in
-original studies.}} To assess the impact of using alternatives to
-rarefaction, I reassessed hypotheses from the human, mice, and
-peromyscus studies. These studies were selected because they originated
-from my research group and I know the datasets well. First, the study
-that published the human dataset was interested in the difference in the
-fecal microbial communities of patients with and without colorectal
-cancer. Here, I divided the patients into peope with and without screen
-relevant neoplasia (SRN). Among the alpha diversity metrics, the test
-were non-significant with similar effect sizes; however, the richness
-estimates obtained using parametric appraoches detected statistically
-significant differences in richness. Although the tests using
-Bray-Curtis distances were significant regardless of whether rarefaction
-was used, the effect sizes were smaller for the distances calculated
-using the CLR, VST, CSS transformations.
-
-\begin{itemize}
-\tightlist
-\item
-  Human study: Did not alter effect size or significance of alpha or
-  beta diversity, but did result in reduced effect sizes for measures of
-  richness and non-parametric estimators of richness; breakaway detected
-  a difference
-\end{itemize}
-
-Mouse
-
-Peromyscus
+\textbf{\emph{Most studies have a high level of sequencing coverage.}}
+To explore the concern over loss of sequencing depth further, I
+calculated the Good's coverage for the observed data. The median
+coverage for each dataset ranged between 89.4 and 99.8\% for the
+Seagrass and Human datasets, respectively (Figure 7). When I rarefied
+each dataset to the size of the smallest community in the dataset, with
+the exception of the Seagrass, Rice, and Stream datasets, the median
+coverage for the rarefied communities was still greater than 90\%. These
+results suggest that most studies had a level of sequencing coverage
+that aligned with the diversity of the communities. Next, I used the
+null model for each dataset to ask how much sequencing effort was
+required to obtain higher levels of coverage. To obtain 95 and 99\%
+coverage, an average of 2.70 and 101.20-fold more sequence data was
+estimated to be required than was required to obtain 90\% coverage,
+respectively (Figure 7). As suggested by the simulated coverages curve
+in Figure 7, these estimates are conservative. Regardless, the
+sequencing effort required to acheive higher sequencing depth would
+likely limit the number of samples that could be sequenced when
+controlling for costs. Although it may be disconcerting to rarefy to a
+sequencing depth that is considerably lower than that obtained for the
+best sequenced community in a study, sequencing coverage for many
+environments is probably adequate even at the lower sequencing depth. Of
+course, the results above have demonstrated that rarefaction is
+necessary to avoid problems with making inferences.
 
@@
 
-\textbf{Null models.} Null models were generated by randomly assigning
-each sequence in the study to an OTU and sample while keeping constant
-the number of sequences per sample and the total number of sequences in
-each OTU. Because the construction of the null models was a stochastic
-process, 100 replicates were geneated for each study.
+\textbf{Choice of datasets.} The specific studies were selected because
+their data was publicly accessible through the Sequence Read Archive,
+the original investigators sequenced the V4 region of the 16S rRNA gene
+using paired 250 nt reads, and my previous familiarity with the data.
+The use of paired 250 nt reads to sequence the V4 region resulted in a
+near complete two-fold overap of the V4 region resulting in high quality
+contigs with a low sequencing error rate (1). These data were processed
+through the standard sequence curation pipeline to generate operational
+taxonomic units (OTUs) using the mothur software package (1, 2).
+
+\textbf{Null community model.} Null models were generated by randomly
+assigning each sequence in the study to an OTU and sample while keeping
+constant the number of sequences per sample and the total number of
+sequences in each OTU. Because the construction of the null models was a
+stochastic process, 100 replicates were geneated for each study.
+
+\textbf{10\%/5\% model}
+
+\textbf{Remove model}
+
+\textbf{Null treatment model}
+
+\textbf{Size-based treatment model}
 
@@
 
+\textbf{\emph{doi: 10.1038/nmeth.2658.}} Should CSS be used in alpha
+diversity analysis? No.~The richness and shannon diversity values were
+no different from those obtained with the raw abundances. From the
+paper, ``The relative proportion of the features is unaffected by the
+normalization''. CSS paper refers to relative abundance approach as
+total-sum normalization (TSN)
+
+Robust CLR - remove zero counts and calculate CLR One CLR - add a
+pseudocount of 1 to all observations Nudge CLR - add a pseudocount of
+1/total number of sequences in sample Zero CLR - Impute the value of
+zeroes using zCompositions package
+
+\textbf{\emph{\url{https://edepot.wur.nl/547087}.}} ``Log- ratio PCA is
+designed to give results that are library size- independent. However, as
+we demonstrated mathematically and with examples based on simulated and
+real data, log- ratio PCA becomes library size- dependent, if there are
+many infrequent taxa (many zeroes) and library sizes differ largely. In
+this situation, the row centring used in log- ratio PCA brings an effect
+of r (the row mean of the log- transformed counts) in the clr-
+transformed matrix. Note that this effect is irrespective of whether or
+not these infrequent taxa are genuine or due to sequencing noise or
+allocation error. This library size dependence is unexpected in the
+sense that, after applying the clr, the transformed matrix is free of
+the effect of the row totals for strictly positive data
+(yij\textgreater0 for all i and j). We additionally demonstrate that
+library size variability causes a loss in power in detecting an effect
+of x with log- ratio RDA. If there is additionally a correlation between
+treatment and the library size, the type 1 error for detecting the
+effect of x can be seriously inflated.''
+
+\textbf{\emph{\url{https://www.nature.com/articles/s41522-020-00160-w}.}}
+``An important characteristic of a feature table is that it is typically
+sparse, sometimes as many as \textasciitilde90\% are zero entries21,
+which creates a challenge for analyzing rare taxa. A quick and simple
+strategy to deal with excess zeros is to add a small positive constant
+(e.g.~1) called pseudo-count14,22 to each cell of the feature table. The
+addition of a pseudo-count becomes necessary when using methods of
+analysis that require log transformation of the observed counts. Even
+though adding a pseudo-count is simple and widely used, the choice of
+the pseudo-count is ad hoc. Studies have shown that differential
+abundance or clustering results could be sensitive to the choice of
+pseudo count23,24. Although different values of pseudo counts have been
+discussed in the literature23,24,25,26, to the best of our knowledge,
+there is no consensus on how to choose the optimal value. Other
+strategies involve modeling zero counts by some probability
+models21,27.''
+
+\url{https://academic.oup.com/bioinformatics/article/34/16/2870/4956011}
+\url{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6755255/}
 \vspace{10mm}

### 2022-10-11 code/plot_alpha_beta_depth_correlation.R

@@ -21,21 +21,21 @@ pretty_datasets <- tibble(
 ## Alpha diversity panel...
 
 alpha_metrics <- c(
-  "raw_sobs",
   "rare_sobs",
+  "raw_sobs",
   "srs_sobs",
   "raw_ace",
   "raw_chao",
   "ba_default",
   "ba_poisson",
   
-  "raw_shannon",
   "rare_shannon",
+  "raw_shannon",
   "srs_shannon",
   "raw_npshannon",
   
-  "raw_simpson",
   "rare_simpson",
+  "raw_simpson",
   "srs_simpson"
   
   # "ba_sobs", "rare_ace", "rare_chao", "rare_coverage", "rare_npshannon",
@@ -64,21 +64,21 @@ alpha_classes <- c(
 )
 
 alpha_labels <- c(
-  "Raw",
   "Rarefied",
+  "Raw",
   "SRS Normalized",
   "ACE Estimate",
   "Chao1 Estimate",
   "BA Default",
   "BA Poisson",
   
-  "Raw",
   "Rarefied",
+  "Raw",
   "SRS Normalized",
   "Estimate",
   
-  "Raw",
   "Rarefied",
+  "Raw",
   "SRS Normalized"
 )
 
@@ -132,28 +132,26 @@ alpha <- alpha_composite %>%
 ## Beta diversity panel...
 
 beta_metrics <- c(
-  "raw_bray",
-  "rare_bray",
-  "relabund_bray",
-  "srs_bray",
-  "metagenomeseq_bray",
-
-  "raw_jaccard",
   "rare_jaccard",
+  "raw_jaccard",
   "relabund_jaccard",
   "srs_jaccard",
   "metagenomeseq_jaccard",
 
-  "raw_euclidean",
+  "rare_bray",
+  "raw_bray",
+  "relabund_bray",
+  "srs_bray",
+  "metagenomeseq_bray",
+
   "rare_euclidean",
+  "raw_euclidean",
   "relabund_euclidean",
-  "nclr_euclidean",
-  "oclr_euclidean",
   "rclr_euclidean",
+  "oclr_euclidean",
+  "nclr_euclidean",
   "zclr_euclidean",
   "deseq2_euclidean"
-  # "srs_euclidean",
-  # "metagenomeseq_euclidean"
 )
 
 # beta_classes <- c(
[…]

### 2022-10-11 code/plot_example_alpha_cor.R

@@ -67,7 +67,7 @@ pretty_labels <- c(
 
 aplot <- tidy_composite %>%
   mutate(metric = factor(metric,
-                         levels = c("raw", "rare", "srs", "ba_default"))) %>%
+                         levels = c("rare", "raw", "srs", "ba_default"))) %>%
   filter(!(metric == "ba_default" & value > 4000)) %>%
   ggplot(aes(x = nseqs, y = value)) +
   geom_point() +

### 2022-10-11 code/plot_example_beta_cor.R

@@ -8,9 +8,9 @@ library(patchwork)
 
 dist_files <- c(
     "data/human/data.otu.1.r_deseq2_euclidean.dist",
-    "data/human/data.otu.1.r_nclr_euclidean.dist",
-    "data/human/data.otu.1.r_oclr_euclidean.dist",
     "data/human/data.otu.1.r_rclr_euclidean.dist",
+    "data/human/data.otu.1.r_oclr_euclidean.dist",
+    "data/human/data.otu.1.r_nclr_euclidean.dist",
     "data/human/data.otu.1.r_zclr_euclidean.dist",
 
     "data/human/data.otu.1.r_rare_bray.dist",
@@ -57,8 +57,8 @@ get_dists <- function(x) {
 
 distances <- map_dfr(dist_files, get_dists) %>%
   mutate(process = factor(process,
-                          levels = c("raw", "rare", "relabund", "srs",
-                                     "metagenomeseq", "nclr", "oclr", "rclr",
+                          levels = c("rare", "raw", "relabund", "srs",
+                                     "metagenomeseq", "rclr", "oclr", "nclr",
                                      "zclr", "deseq2")))
 
 correlations <- distances %>%
@@ -92,26 +92,26 @@ get_rho <- function(p, m, df = correlations) {
 
 
 j_beta_labels <- c(
-  raw = glue("Raw\n(\u03C1 = {get_rho('raw', 'jaccard')})"),
   rare = glue("Rarefied\n(\u03C1 = {get_rho('rare', 'jaccard')})"),
+  raw = glue("Raw\n(\u03C1 = {get_rho('raw', 'jaccard')})"),
   relabund = glue("Rel. abundance\n(\u03C1 = {get_rho('relabund', 'jaccard')})"),
   srs = glue("SRS Normalized\n(\u03C1 = {get_rho('srs', 'jaccard')})"),
   metagenomeseq = glue("CSS Normalized\n(\u03C1 = {get_rho('metagenomeseq', 'jaccard')})")
 )
 
 b_beta_labels <- c(
-  raw = glue("Raw\n(\u03C1 = {get_rho('raw', 'bray')})"),
   rare = glue("Rarefied\n(\u03C1 = {get_rho('rare', 'bray')})"),
[…]

### 2022-10-11 code/plot_false_positive_null_model.R

@@ -21,26 +21,23 @@ pretty_datasets <- tibble(
 ## Alpha diversity panel...
 
 alpha_metrics <- c(
-  "sobs_raw",
   "sobs_rarefy",
+  "sobs_raw",
   "sobs_srs",
   "ace_raw",
   "chao_raw",
   "default_breakaway",
   "poisson_breakaway",
 
-  "shannon_raw",
   "shannon_rarefy",
+  "shannon_raw",
   "shannon_srs",
   "npshannon_raw",
 
-  "simpson_raw",
   "simpson_rarefy",
+  "simpson_raw",
   "simpson_srs"
 
-#"ace_rarefy", "chao_rarefy", "coverage_rarefy", "npshannon_rarefy",
-#"nseqs_rarefy", "npshannon_srs", "coverage_raw", "coverage_srs", "nseqs_srs",
-#"nseqs_raw", "ace_srs", "chao_srs",
 )
 
 alpha_classes <- c(
@@ -63,21 +60,21 @@ alpha_classes <- c(
 )
 
 alpha_labels <- c(
-  "Raw",
   "Rarefied",
+  "Raw",
   "SRS Normalized",
   "ACE Estimate",
   "Chao1 Estimate",
   "BA Default",
   "BA Poisson",
 
-  "Raw",
   "Rarefied",
+  "Raw",
   "SRS Normalized",
   "Estimate",
 
-  "Raw",
   "Rarefied",
+  "Raw",
   "SRS Normalized"
 )
 
@@ -134,24 +131,24 @@ alpha <- alpha_composite %>%
 ## Beta diversity panel...
 
 beta_metrics <- c(
-  "raw_bray",
-  "rare_bray",
-  "relabund_bray",
-  "srs_bray",
-  "metagenomeseq_bray",
-
-  "raw_jaccard",
   "rare_jaccard",
+  "raw_jaccard",
   "relabund_jaccard",
   "srs_jaccard",
   "metagenomeseq_jaccard",
 
-  "raw_euclidean",
+  "rare_bray",
+  "raw_bray",
+  "relabund_bray",
+  "srs_bray",
+  "metagenomeseq_bray",
+
   "rare_euclidean",
+  "raw_euclidean",
   "relabund_euclidean",
-  "nclr_euclidean",
-  "oclr_euclidean",
   "rclr_euclidean",
+  "oclr_euclidean",
+  "nclr_euclidean",
[…]

### 2022-10-11 code/plot_false_positive_null_model_size.R

@@ -21,21 +21,21 @@ pretty_datasets <- tibble(
 ## Alpha diversity panel...
 
 alpha_metrics <- c(
-  "sobs_raw",
   "sobs_rarefy",
+  "sobs_raw",
   "sobs_srs",
   "ace_raw",
   "chao_raw",
   "default_breakaway",
   "poisson_breakaway",
 
-  "shannon_raw",
   "shannon_rarefy",
+  "shannon_raw",
   "shannon_srs",
   "npshannon_raw",
 
-  "simpson_raw",
   "simpson_rarefy",
+  "simpson_raw",
   "simpson_srs"
 
 #"ace_rarefy", "chao_rarefy", "coverage_rarefy", "npshannon_rarefy",
@@ -63,21 +63,21 @@ alpha_classes <- c(
 )
 
 alpha_labels <- c(
-  "Raw",
   "Rarefied",
+  "Raw",
   "SRS Normalized",
   "ACE Estimate",
   "Chao1 Estimate",
   "BA Default",
   "BA Poisson",
 
-  "Raw",
   "Rarefied",
+  "Raw",
   "SRS Normalized",
   "Estimate",
 
-  "Raw",
   "Rarefied",
+  "Raw",
   "SRS Normalized"
 )
 
@@ -135,28 +135,26 @@ alpha <- alpha_composite %>%
 ## Beta diversity panel...
 
 beta_metrics <- c(
-  "raw_bray",
   "rare_bray",
+  "raw_bray",
   "relabund_bray",
   "srs_bray",
   "metagenomeseq_bray",
 
-  "raw_jaccard",
   "rare_jaccard",
+  "raw_jaccard",
   "relabund_jaccard",
   "srs_jaccard",
   "metagenomeseq_jaccard",
 
-  "raw_euclidean",
   "rare_euclidean",
+  "raw_euclidean",
   "relabund_euclidean",
-  "nclr_euclidean",
-  "oclr_euclidean",
   "rclr_euclidean",
+  "oclr_euclidean",
+  "nclr_euclidean",
   "zclr_euclidean",
   "deseq2_euclidean"
-  # "srs_euclidean",
-  # "metagenomeseq_euclidean"
 )
 
 # beta_classes <- c(
@@ -164,26 +162,27 @@ beta_metrics <- c(
 # )
 
 beta_labels <- c(
-  "Raw",
   "Rarefied",
+  "Raw",
[…]

### 2022-10-11 code/plot_intrasample_variation.R

@@ -70,16 +70,16 @@ alpha_composite <- map_dfr(alpha_summary_files, read_tsv, .id = "dataset") %>%
 
 
 datasets <- c("bioethanol", "human", "lake", "marine", "mice", "peromyscus",
-              "rainforest", #"rice",
-              "seagrass", "sediment", "soil"#,
-              #"stream"
+              "rainforest", "rice",
+              "seagrass", "sediment", "soil",
+              "stream"
               )
 
 beta_summary_files <- glue("data/{datasets}/data.otu.beta_depth.summary")
 names(beta_summary_files) <- datasets
 
 beta_composite <- map_dfr(beta_summary_files, read_tsv, .id = "dataset") %>%
-  filter(method == "bray" | method == "jaccard") %>%
+  filter(method == "bray") %>%
   group_by(dataset, method, n_seqs) %>%
   summarize(mean = mean(mean),
             cov = 100 * mean(sd / mean),
@@ -91,7 +91,6 @@ beta_composite <- map_dfr(beta_summary_files, read_tsv, .id = "dataset") %>%
                names_to = "statistic",
                values_to = "value") %>%
   mutate(dataset = factor(dataset, level = datasets),
-         method = factor(method, levels = c("bray", "jaccard")),
          statistic = factor(statistic, levels = c("mean", "cov"))
          )
 
@@ -103,7 +102,13 @@ plot_method_statistic <- function(m, s, data = composite,
     filter(method == m & statistic == s)
 
   filtered_point <- filtered_line %>%
-    inner_join(., smallest, by = c("dataset", "n_seqs"))
+    inner_join(., smallest, by = c("dataset")) %>%
+    mutate(diff = abs(n_seqs.x - n_seqs.y)) %>%
+    group_by(dataset) %>%
+    slice_min(diff) %>%
+    select(dataset, n_seqs = n_seqs.y, value) %>%
+    ungroup()
+    
 
   filtered_line %>%
     ggplot(aes(x = n_seqs, y = value,
@@ -134,15 +139,7 @@ plot_method_statistic <- function(m, s, data = composite,
     )
 }
 
-composite <- bind_rows(alpha_composite, beta_composite) %>%
-  bind_rows(tibble(dataset = c(rep("rice", 8), rep("stream", 8)),
-                   method = rep(rep(c("bray", "jaccard"), each = 4), 2),
-                   n_seqs = 1000,
-                   sd = 0,
[…]

### 2022-10-11 code/plot_power_cffect_model.R

@@ -21,26 +21,22 @@ pretty_datasets <- tibble(
 ## Alpha diversity panel...
 
 alpha_metrics <- c(
-  "sobs_raw",
   "sobs_rarefy",
+  "sobs_raw",
   "sobs_srs",
   "ace_raw",
   "chao_raw",
   "default_breakaway",
   "poisson_breakaway",
 
-  "shannon_raw",
   "shannon_rarefy",
+  "shannon_raw",
   "shannon_srs",
   "npshannon_raw",
 
-  "simpson_raw",
   "simpson_rarefy",
+  "simpson_raw",
   "simpson_srs"
-
-#"ace_rarefy", "chao_rarefy", "coverage_rarefy", "npshannon_rarefy",
-#"nseqs_rarefy", "npshannon_srs", "coverage_raw", "coverage_srs", "nseqs_srs",
-#"nseqs_raw", "ace_srs", "chao_srs",
 )
 
 alpha_classes <- c(
@@ -63,21 +59,21 @@ alpha_classes <- c(
 )
 
 alpha_labels <- c(
-  "Raw",
   "Rarefied",
+  "Raw",
   "SRS Normalized",
   "ACE Estimate",
   "Chao1 Estimate",
   "BA Default",
   "BA Poisson",
 
-  "Raw",
   "Rarefied",
+  "Raw",
   "SRS Normalized",
   "Estimate",
 
-  "Raw",
   "Rarefied",
+  "Raw",
   "SRS Normalized"
 )
 

### 2022-10-11 code/plot_power_effect_model.R

@@ -21,26 +21,22 @@ pretty_datasets <- tibble(
 ## Alpha diversity panel...
 
 alpha_metrics <- c(
-  "sobs_raw",
   "sobs_rarefy",
+  "sobs_raw",
   "sobs_srs",
   "ace_raw",
   "chao_raw",
   "default_breakaway",
   "poisson_breakaway",
 
-  "shannon_raw",
   "shannon_rarefy",
+  "shannon_raw",
   "shannon_srs",
   "npshannon_raw",
 
   "simpson_raw",
   "simpson_rarefy",
   "simpson_srs"
-
-#"ace_rarefy", "chao_rarefy", "coverage_rarefy", "npshannon_rarefy",
-#"nseqs_rarefy", "npshannon_srs", "coverage_raw", "coverage_srs", "nseqs_srs",
-#"nseqs_raw", "ace_srs", "chao_srs",
 )
 
 alpha_classes <- c(
@@ -63,21 +59,21 @@ alpha_classes <- c(
 )
 
 alpha_labels <- c(
-  "Raw",
   "Rarefied",
+  "Raw",
   "SRS Normalized",
   "ACE Estimate",
   "Chao1 Estimate",
   "BA Default",
   "BA Poisson",
 
-  "Raw",
   "Rarefied",
+  "Raw",
   "SRS Normalized",
   "Estimate",
 
-  "Raw",
   "Rarefied",
+  "Raw",
   "SRS Normalized"
 )
 
@@ -137,28 +133,27 @@ alpha <- alpha_composite %>%
 ## Beta diversity panel...
 
 beta_metrics <- c(
-  "raw_bray",
+
   "rare_bray",
+  "raw_bray",
   "relabund_bray",
   "srs_bray",
   "metagenomeseq_bray",
 
-  "raw_jaccard",
   "rare_jaccard",
+  "raw_jaccard",
   "relabund_jaccard",
   "srs_jaccard",
   "metagenomeseq_jaccard",
 
-  "raw_euclidean",
   "rare_euclidean",
+  "raw_euclidean",
   "relabund_euclidean",
-  "nclr_euclidean",
-  "oclr_euclidean",
   "rclr_euclidean",
+  "oclr_euclidean",
+  "nclr_euclidean",
   "zclr_euclidean",
   "deseq2_euclidean"
-  # "srs_euclidean",
-  # "metagenomeseq_euclidean"
 )
 
[…]

### 2022-11-01 submission/manuscript.Rmd
Incorporate figures and legends into manuscript

@@
 header-includes:
+ - \usepackage{upgreek}
  - \usepackage{booktabs}
  - \usepackage{longtable}
+ - \usepackage{graphicx}
  - \usepackage{array}
@@
 
-**Figure 1.**
+## Figures
 
@@
 
-**Figure S1.**
+\includegraphics[height=17cm]{figure_1.png}
+
+**Figure 1.
+Rarefaction eliminates the correlation between sequencing depth and alpha diversity (A) and between differences in sampling depth and beta (B) diversity metrics when using null community models.** Examples of the relationship between different metrics and methods for controlling for uneven sequencing effort are provided in Figures S2 and S3 for alpha and beta diversity metrics, respectively.
+Each point represents the mean of 100 random null community models; the standard deviation was smaller than the size of the plotting symbol.
+
+\newpage
+
+\includegraphics[height=17cm]{figure_2.png}
+
+**Figure 2.
+The risk of falsely detecting a difference between treatment groups drawn from a null model does not meaningfully vary from 5%, regardless of approach for controlling for uneven sequencing depth**.
+Samples were randomly assigned to different treatment groups.
+To calculate the false detection rate, datasets were regenerated 100 times and differences in alpha diversity were tested using a Wilcoxon test (A) and differences in beta diversity were tested using PERMANOVA (B) at a 5% threshold.
+The false positive rate was the number of times a dataset yeilded a significant result.
+
+\newpage
+
+\includegraphics[height=17cm]{figure_3.png}
+
+**Figure 3.The risk of falsely detecting a difference between treatment groups drawn from a null model does not meaningfully vary from 5% when data are rarefied when sequencing depth is confounded with treatement group**.
+Samples were assigned to different treatment groups based on whether they were above the median number of sequences for each dataset.
+To calculate the false detection rate, datasets were regenerated 100 times and differences in alpha diversity were tested using a Wilcoxon test (A) and differences in beta diversity were tested using PERMANOVA (B) at a 5% threshold.
+The false positive rate was the number of times a dataset yeilded a significant result.
+
+\newpage
+
+\includegraphics[height=17cm]{figure_4.png}
+
+**Figure 4.
+The ability to detect true differences in treatment groups for alpha (A) and beta (B) diversity metrics is greatest when communities differing in the relative abundance of their OTUs are rarefied.** For each dataset samples were randomly assigned to one of two community distributions where the abundance of OTUs differed.
+To calculate the power for each study, datasets were regenerated 100 times and differences in alpha diversity were tested using a Wilcoxon test (A) and differences in beta diversity were tested using PERMANOVA (B) at a 5% threshold.
+The power was the number of times a dataset yielded a significant result.
+
+\newpage
+
+\includegraphics{figure_5.png}
+
+**Figure 5.
+The ability to detect true differences in treatment groups for alpha diversity metrics is greatest when communities differing in richness are rarefied.** For each dataset samples were randomly assigned to one of two community distributions where one distribution contained a subset of OTUs found in the other.
+To calculate the power for each study, datasets were regenerated 100 times and differences in alpha diversity were tested using a Wilcoxon test (A) and differences in beta diversity were tested using PERMANOVA (B) at a 5% threshold.
+The power was the number of times a dataset yielded a significant result.
+
+\newpage
+
+\includegraphics{figure_6.png}
+
+**Figure 6.
+The mean and coefficient of variation for rarefied richness, shannon diversity, and Bray-Curtis dissimilarity vary with sequencing depth.** For each dataset, a null community distribution was created and samples were created to have the same sequencing depth as they did originally.
+The placement of the plotting symbol indicates the size of the smallest sample.
+Results are only shown for sequencing depths where a dataset had 5 or more samples.
+
+\newpage
+
+\includegraphics{figure_7.png}
+
+**Figure 7.
+Most datasets are sequenced to a level that provides a high level of coverage.** Each plotting symbol represents the observed Good's coverage for a different sample in each dataset.
+The smoothed line indicates the simulated coverage for varying levels of sampling effort when a null community is generated from the observed data.
+The box and whisker plot indicates the range of coverage values when the observed commmunity data were rarefied to the size of the least sequenced sample.
+
+\newpage
+
+\includegraphics{figure_s1.png}
+
+**Figure S1.
+The number of sequences observed in each sample for each dataset included in this analysis generally varied by 10 to 100-fold.** The threshold for specifying the number of sequences per sample varied by dataset and was determined based on identifying natural breaks in the data.
+
+\newpage
+
+\includegraphics{figure_s2.png}
+
+**Figure S2.
+Examples of the richness in each of the 490 samples that were generated for one randomization of the null model using the human dataset.** The x-axis indicates the number of sequences in each of the samples prior to each method's appraoch of controlling for uneven sampling effort.
+The Spearman correlation coefficient ($\uprho$) and test of whether the coefficient was significantly different from zero are indicated for each panel.
+
+\newpage
+
+\includegraphics{figure_s3.png}
+
+**Figure S3.
+Examples of differences in beta diversity in each of the 490 samples that were generated for one randomization of the null model using the human dataset.** The x-axis indicates the difference in the number of sequences in each of the samples that went into calcualting the pairwise distance prior to each method's appraoch of controlling for uneven sampling effort.
+The Spearman correlation coefficient ($\uprho$) and test of whether the coefficient was significantly different from zero are indicated for each panel.

### 2022-11-01 submission/manuscript.tex
Incorporate figures and legends into manuscript

@@
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+\usepackage{upgreek}
 \usepackage{booktabs}
 \usepackage{longtable}
+\usepackage{graphicx}
 \usepackage{array}
@@
 
-\textbf{Figure 1.}
+\hypertarget{figures}{%
+\subsection{Figures}\label{figures}}
 
@@
 
-\textbf{Figure S1.}
+\includegraphics[height=17cm]{figure_1.png}
+
+\textbf{Figure 1. Rarefaction eliminates the correlation between
+sequencing depth and alpha diversity (A) and between differences in
+sampling depth and beta (B) diversity metrics when using null community
+models.} Examples of the relationship between different metrics and
+methods for controlling for uneven sequencing effort are provided in
+Figures S2 and S3 for alpha and beta diversity metrics, respectively.
+Each point represents the mean of 100 random null community models; the
+standard deviation was smaller than the size of the plotting symbol.
+
+\newpage
+
+\includegraphics[height=17cm]{figure_2.png}
+
+\textbf{Figure 2. The risk of falsely detecting a difference between
+treatment groups drawn from a null model does not meaningfully vary from
+5\%, regardless of approach for controlling for uneven sequencing
+depth}. Samples were randomly assigned to different treatment groups. To
+calculate the false detection rate, datasets were regenerated 100 times
+and differences in alpha diversity were tested using a Wilcoxon test (A)
+and differences in beta diversity were tested using PERMANOVA (B) at a
+5\% threshold. The false positive rate was the number of times a dataset
+yeilded a significant result.
+
+\newpage
+
+\includegraphics[height=17cm]{figure_3.png}
+
+\textbf{Figure 3.The risk of falsely detecting a difference between
+treatment groups drawn from a null model does not meaningfully vary from
+5\% when data are rarefied when sequencing depth is confounded with
+treatement group}. Samples were assigned to different treatment groups
+based on whether they were above the median number of sequences for each
+dataset. To calculate the false detection rate, datasets were
+regenerated 100 times and differences in alpha diversity were tested
+using a Wilcoxon test (A) and differences in beta diversity were tested
+using PERMANOVA (B) at a 5\% threshold. The false positive rate was the
+number of times a dataset yeilded a significant result.
+
+\newpage
+
+\includegraphics[height=17cm]{figure_4.png}
+
+\textbf{Figure 4. The ability to detect true differences in treatment
+groups for alpha (A) and beta (B) diversity metrics is greatest when
+communities differing in the relative abundance of their OTUs are
+rarefied.} For each dataset samples were randomly assigned to one of two
+community distributions where the abundance of OTUs differed. To
+calculate the power for each study, datasets were regenerated 100 times
+and differences in alpha diversity were tested using a Wilcoxon test (A)
+and differences in beta diversity were tested using PERMANOVA (B) at a
+5\% threshold. The power was the number of times a dataset yielded a
+significant result.
+
+\newpage
+
+\includegraphics{figure_5.png}
+
+\textbf{Figure 5. The ability to detect true differences in treatment
+groups for alpha diversity metrics is greatest when communities
+differing in richness are rarefied.} For each dataset samples were
+randomly assigned to one of two community distributions where one
+distribution contained a subset of OTUs found in the other. To calculate
+the power for each study, datasets were regenerated 100 times and
+differences in alpha diversity were tested using a Wilcoxon test (A) and
+differences in beta diversity were tested using PERMANOVA (B) at a 5\%
+threshold. The power was the number of times a dataset yielded a
+significant result.
+
+\newpage
+
+\includegraphics{figure_6.png}
+
+\textbf{Figure 6. The mean and coefficient of variation for rarefied
+richness, shannon diversity, and Bray-Curtis dissimilarity vary with
+sequencing depth.} For each dataset, a null community distribution was
+created and samples were created to have the same sequencing depth as
+they did originally. The placement of the plotting symbol indicates the
+size of the smallest sample. Results are only shown for sequencing
+depths where a dataset had 5 or more samples.
+
+\newpage
+
+\includegraphics{figure_7.png}
+
+\textbf{Figure 7. Most datasets are sequenced to a level that provides a
+high level of coverage.} Each plotting symbol represents the observed
+Good's coverage for a different sample in each dataset. The smoothed
+line indicates the simulated coverage for varying levels of sampling
+effort when a null community is generated from the observed data. The
+box and whisker plot indicates the range of coverage values when the
+observed commmunity data were rarefied to the size of the least
+sequenced sample.
+
+\newpage
+
+\includegraphics{figure_s1.png}
+
+\textbf{Figure S1. The number of sequences observed in each sample for
+each dataset included in this analysis generally varied by 10 to
+100-fold.} The threshold for specifying the number of sequences per
+sample varied by dataset and was determined based on identifying natural
+breaks in the data.
+
+\newpage
+
+\includegraphics{figure_s2.png}
+
+\textbf{Figure S2. Examples of the richness in each of the 490 samples
+that were generated for one randomization of the null model using the
+human dataset.} The x-axis indicates the number of sequences in each of
+the samples prior to each method's appraoch of controlling for uneven
+sampling effort. The Spearman correlation coefficient (\(\uprho\)) and
+test of whether the coefficient was significantly different from zero
+are indicated for each panel.
+
+\newpage
+
+\includegraphics{figure_s3.png}
+
+\textbf{Figure S3. Examples of differences in beta diversity in each of
+the 490 samples that were generated for one randomization of the null
+model using the human dataset.} The x-axis indicates the difference in
+the number of sequences in each of the samples that went into
+calcualting the pairwise distance prior to each method's appraoch of
+controlling for uneven sampling effort. The Spearman correlation
+coefficient (\(\uprho\)) and test of whether the coefficient was
+significantly different from zero are indicated for each panel.
 

### 2022-11-01 code/render_manuscript.R

@@ -1,0 +1,11 @@
+#!/usr/bin/env Rscript
+
+library('rmarkdown')
+
+if(!tinytex::is_tinytex()) {
+  tinytex::install_tinytex()
+} 
+
+render('submission/manuscript.Rmd', clean=FALSE)
+
+file.rename("submission/manuscript.knit.md", "submission/manuscript.md")

### 2022-11-02 submission/manuscript.Rmd
Writing M&M section

@@
 
-***Without rarefaction, metrics of alpha diversity are sensitive to sampling effort.*** To test the sesitivity of various approaches of measuring alpha diversity to sampling effort, I generated null models for each study.
-Under a null model, each community from the same study would be expected to have the same alpha diversity regardless of the sampling effort.
-I measured the richness of the communities in each study without any correction, using scaled ranked subsampling (SRS) normalized OTU counts, with estimates based on non-parametric and parametric approaches, and using rarefaction (e.g.
+***Without rarefaction, metrics of alpha diversity are sensitive to sampling effort.*** To test the sesitivity of various approaches of measuring alpha diversity to sampling effort, I generated null models for each dataset.
+Under a null model, each community from the same dataset would be expected to have the same alpha diversity regardless of the sampling effort.
+I measured the richness of the communities in each dataset without any correction, using scaled ranked subsampling (SRS) normalized OTU counts, with estimates based on non-parametric and parametric approaches, and using rarefaction (e.g.
 Figure S2).
-For each study, all of the approaches, except for rarefaction, showed a strong correlation between richness and the number of sequences in the sample (Figure 1A).
+For each dataset, all of the approaches, except for rarefaction, showed a strong correlation between richness and the number of sequences in the sample (Figure 1A).
 Next, I assessed diversity using the Shannon diversity index and the inverse Simpson diversity index without any correction, using normalized OTU counts, and rarefaction; I also used a non-parametric estimator of Shannon diversity.
@@
 Under a null model, the ecological distance between any pair of samples would be the same regardless of the difference in the number of sequences observed in each sample (e.g., Figure S3).
-First, I calculated the Jaccard distance coefficient between all pairs of communities within a study.
+First, I calculated the Jaccard distance coefficient between all pairs of communities within a dataset.
 The Jaccard distance coefficient is the fraction of OTUs that are unique to either community and does not account for the abundance of the OTUs.
@@
 To test for differences in alpha and beta diversity I used the non-parametric Wilcoxon test and non-parametric permutation-based multivariate analysis of variance (PERMANOVA).
-First, within each study, I randomly assigned each sample to one of two treatment groups.
-My expectation was that approximately 5 of the 100 (5%) random tests for each comparison would yield a significant test result.
-Indeed, for each study and alpha and beta diversity metric and strategy for accounting for uneven sampling, approximately 5% of the tests yielded a significant result (Figure 2).
-Second, within each study, I assigned samples with more than the median number of sequences per sample to one treatment group and the rest to another treatment group.
-If there was no sensitivity to sampling effort, I would have expected that 5% of the tests would yield a significant result.
-In fact, only the rarefied data consistently resulted in a 5% false positive rate for alpha and beta diversity metrics (Figure 2).
-These results align with the observed sensitivity of alpha and beta diversity metrics to sampling effort and underscore the value of rarefaction.
+First, I employed an unbiased null treatment model to measure the false detection rate, which should not have meaningfully differed from 5%.
+Indeed, for each dataset and alpha and beta diversity metric and strategy for accounting for uneven sampling, approximately 5% of the tests yielded a significant result (Figure 2).
+Second, I employed a biased null treatment model where the treatment group was completely confounded with the number of sequences in each sample.
+Under this model, only the rarefied data consistently resulted in a 5% false positive rate for alpha and beta diversity metrics (Figure 2).
+These results aligned with the observed sensitivity of alpha and beta diversity metrics to sampling effort and underscore the value of rarefaction.
 
 ***Rarefaction preserves the statistical power to detect differences between treatment groups.*** To assess the impact of different approaches to control for uneven sampling effort I performed two additional simulations.
-In the first simulation, for each study communities were randomly assigned to one of two treatment groups.
-Communities in the first treatment group were generated by sampling from the null distribution.
-Samples in the second treatment group were generated by perturbing the null distribution by increasing the relative abundance of 10% of the OTUs by 5%.
-These values were determined after empirically searching for conditions that resulted in a large fraction of the randomizations yieleding a significant result across most of the studies.
-The fraction of tests that yielded a significant test was a measure of the statistical power for the test.
-The power to detect differences in richness by all approaches was low (Figure 4A).
+In the first simulation, I implemented a skewed abundance distribution model to create two treatment groups for each datasets that were each populated with half of the samples each with the same number of sequences as the samples had in the observed data.
+The power to detect differences in richness between the two simulated treatment groups by all approaches was low (Figure 4A).
 This was likely because the approach for generating the perturbed community did not necessarily change the number of OTUs in each treatment group.
 Regardless, the simulations testing differencse in richness using the Rice and Stream datasets had the greatest power when the richness data were rarefied.
-To explore this further, in a second simulation the second treatment group was perturbed by removing 3% of the OTUs from the model.
+To explore this further, a richness-adjusted community model was created by removing 3% of the OTUs from a null model model.
 As suggested by the first simulation, the rarefied richness data had a higher statistical power than the other approaches when measuring richness (Figure 5).
@@
 Regardless, the sequencing effort required to acheive higher sequencing depth would likely limit the number of samples that could be sequenced when controlling for costs.
-Although it may be disconcerting to rarefy to a sequencing depth that is considerably lower than that obtained for the best sequenced community in a study, sequencing coverage for many environments is probably adequate even at the lower sequencing depth.
+Although it may be disconcerting to rarefy to a sequencing depth that is considerably lower than that obtained for the best sequenced community in a dataset, sequencing coverage for many environments is probably adequate even at the lower sequencing depth.
 Of course, the results above have demonstrated that rarefaction is necessary to avoid problems with making inferences.
@@
 
-**Choice of datasets.** The specific studies were selected because their data was publicly accessible through the Sequence Read Archive, the original investigators sequenced the V4 region of the 16S rRNA gene using paired 250 nt reads, and my previous familiarity with the data.
+**Choice of datasets.** The specific studies used in this study were selected because their data was publicly accessible through the Sequence Read Archive, the original investigators sequenced the V4 region of the 16S rRNA gene using paired 250 nt reads, and my previous familiarity with the data.
 The use of paired 250 nt reads to sequence the V4 region resulted in a near complete two-fold overap of the V4 region resulting in high quality contigs with a low sequencing error rate [@Kozich2013].
-These data were processed through the standard sequence curation pipeline to generate operational taxonomic units (OTUs) using the mothur software package [@Kozich2013; @Schloss2009].
+These data were processed through a standardized sequence curation pipeline to generate operational taxonomic units (OTUs) using the mothur software package [@Kozich2013; @Schloss2009].
+OTUs were identified using the OptiClust algorithm to cluster sequences together that were not more than 3% different from each other [@Westcott2017].
 
-**Null community model.** Null models were generated by randomly assigning each sequence in the study to an OTU and sample while keeping constant the number of sequences per sample and the total number of sequences in each OTU.
-Because the construction of the null models was a stochastic process, 100 replicates were geneated for each study.
+**Null community model.** Null community models were generated such that within a dataset the number of sequences per sample and the number of sequences per OTU across all samples within the dataset were the same as was observed in original.
+This model effectively generated statistical samples of a single community so that there should not have been a statistical difference between the samples.
+This model implemented by randomly assigning each sequence in the dataset to an OTU and sample while keeping constant the number of sequences per sample and the total number of sequences in each OTU.
+This is a similar approach to that of the IS algorithm described by Ulrich and Gotelli [@Ulrich2010].
+Because the construction of the null models was a stochastic process, 100 replicates were geneated for each dataset.
 
-**10%/5% model**
+**Null treatment model.** I created an unbiased and biased treatment model.
+In the unbiased model, samples were randomly assigned to one of two treatment groups.
+In the biased treatment model, samples that had more than the median number of seqeunces for a dataset were assigned to one treatment group and the rest were assigned to a second treatment group.
+Comparison of any diversity metric across the two treatment groups should have only yielded a significant result in 5% of the simulations when testing under a Type I error (i.e., $\upalpha$) of 0.05.
 
-**Remove model**
+**Skewed abundance community model.** In the skewed abundance community model, communities were randomly assigned to one of two simulated treatment groups.
+Communities in the first treatment group were generated by calculating the relative abundance of each OTU across all samples and using those values as the probability of sampling each OTU.
+This probability distribution was sampled until each sample had the same number of sequences that it did in the observed data.
+Samples in the second treatment group were generated by adjusting the relative abundances of the OTUs in the firs treatment group by increasing the relative abundance of 10% of the OTUs by 5%.
+These values were determined after empirically searching for conditions that resulted in a large fraction of the randomizations yieleding a significant result across most of the studies.
+Sequences were sampled from the skewed community community until each sample had the same number of sequences that it did in the observed data.
+Under the skewed abundance community model each sample represented a statistical sampling of two communities such that there should not have been a statistically significant difference within a treatment group, but there was between the treatment groups.
+Because the construction of the skewed abundance community model was a stochastic process, 100 replicates were geneated for each dataset.
 
-**Null treatment model**
+**Richness-adjusted community model.** In the richness-adjusted community model, communities were randomly assigned to one of two simulated treatment groups.
+Communities in the first treatment group were generated by calculating the relative abundance of each OTU across all samples and using those values as the probability of sampling each OTU.
+This probability distribution was sampled until each sample had the same number of sequences that it did in the observed data.
+Samples in the second treatment group were generated by removing 3% of the OTUs from the dataset and recalculating the relative abundance of the remaining OTUs.
+This percentage was determined after empirically searching for a value that resulted in a large fraction of the randomizations yieleding a significant result across most of the studies.
+Sequences were sampled from the richness-adjusted community distrubtion until each sample had the same number of sequences that it did in the observed data.
+Under the richness-adjusted community model each sample represented a statistical sampling of two communities such that there should not have been a statistically significant difference within a treatment group, but there was between the treatment groups.
+Because the construction of the richness-adjusted community model was a stochastic process, 100 replicates were geneated for each dataset.
 
-**Size-based treatment model**
+**Alpha diversity calculations.** Various strategies for handling uneven sampling effort were evaluated to identify the best approach for calculating community richness and Shannon and inverse Simpson diversity indeices.
+Raw OTU counts were used as input to calculate sample richness and Shannon and inverse Simpson diversity using mothur [@Schloss2009; @Magurran2004].
+Shannon diversity was calculated as
 
-**Data availability.**
+$$H_{shannon} = - \sum_{i=1}^{S_{obs}} \frac{n_i}{N} ln \frac{n_i}{N}$$
+
+The Simpson diversity was calculated as
+
+$$D_{simpson} = \frac {\sum_{i=1}^{S_{obs}} {n_i \left ( n_i - 1 \right )}}{N \left( N-1 \right )}$$
+
+The inverse Simpson diversity was calculated as $1/D_{simpson}$.
+In both formulae, $n_i$ is the number of sequences in OTU $i$ and $N$ is the number of sequences in the sample.
+Rarefaction of richness, Shannon diversity and Inverse Simpson diversity values were carried out in mothur.
+Briefly, mothur calculates each value on a random draw of the same number of sequences from each sample and obtains a mean value based on 1,000 random draws.
+Scaled ranked subsampling (SRS) was used to normalize OTU counts to the size of the smallest sample in each dataset using the SRS R package (v.`r package_version("SRS")`)[@Beule2020].
+Normalized OTU counts were used to calculate sample richness and Shannon and inverse Simpson diversity values using mothur.
+The non-parametric bias-corrected Chao1 and ACE richness estimators [@Chao2016] and a non-parametric estimator of the Shannon diversity [@Chao2003] were calculated using raw OTU counts with mothur.
+Parametric estimates of sample richness were calculated using the breakaway (BA) R package (v.`r package_version("breakaway")`)[@Willis2015].
+The current analysis reports both the results from running default model selection procedure and the Poisson model.
+The default model selection returned either the Kemp, Negative Binomial, or Poisson models.
+
+**Beta diversity calculations.** Similar to the alpha diversity calculations, multiple approaches were used to control for uneven sampling effort and calculate beta diversity.
+Raw and OTU counts were used for input to calculate the Jaccard, Bray-Curtis, and Euclidean dissimilarity indices using the vegdist function from the vegan R package (v.`r package_version("vegan")`)[@Oksanen2022].
+The Jaccard index was calculated as
+
+$$D_{Jaccard}=1-\frac{S_{AB}}{S_A+S_B-S_{AB}}$$
+
+where $S_A$ and $S_B$ were the number of OTUs in samples $A$ and $B$ and $S_{AB}$ was the number of OTUs shared between the two samples. The Bray-Curtis index was calculated as
+
+$$D_{Bray-Curtis}=1-\frac{\sum_{i=1}^{S_T} \left| n_{A,i} - n_{B,i}\right| }{ N_A + N_B}$$
+
+where $n_{A,i}$ and $n_{B,i}$ are the number of sequences observed in OTU $i$ from samples $A$ and $B$, respectively.
+$N_A$ and $N_B$ are the total number of sequences in samples $A$ and $B$, respectively.
+$S_T$ is the total number of OTUs observed between the two samples.
+The Euclidean distance was calculated as
+
+$$D_{Euclidean}=1-\sqrt{\sum_{i=1}^{S_T}\left(n_{A,i} - n_{B,i}\right)^2}$$
+
+These metrics were calculated using the relative abundance of each OTU using the vegdist function from vegan.
+The relative abundance was calculated as the number of sequences in the OTU (e.g., $n_{A,i}$) divided by the total number of sequences in the sample (e.g., $N_A$).
+Rarefied beta-diversity values were calculated using the avgdist function in vegan.
+Briefly, vegan's avgdist function calculates each pairwise dissimilarity index after obtaining a random draw of the same number of sequences from each sample.
+After obtaining 100 random draws it returns the mean value.
+
+SRS normalized - Jaccard/BC
+CSS normalized - Jaccard/BC
+DeSeq2
+Center-log ratio
+
+**Power analysis.**
+ The fraction of tests that yielded a significant test was a measure of the statistical power for the test.
+
+**Test of statistical significance.**
+Wilcox, PERMANOVA
 
 **Reproducible data analysis.** 
+mothur - 1.47.0
 
@@
 **Table 1.
-Summary of studies used in the analysis.** For all studies, the number of sequences used from each study was rarefied to the smallest sample size.
-A graphical represenation of the distribution of sample sizes for each study and the samples that were removed from each study are provided in Figure S1.
+Summary of studies used in the analysis.** For all studies, the number of sequences used from each dataset was rarefied to the smallest sample size.
+A graphical represenation of the distribution of sample sizes for each dataset and the samples that were removed from each dataset are provided in Figure S1.
 
@@
       c(
-        "\\textbf{Study\\nobreakspace{}(Ref)}",
+        "\\textbf{Dataset\\nobreakspace{}(Ref)}",
         "\\textbf{Samples}",
@@
 The ability to detect true differences in treatment groups for alpha (A) and beta (B) diversity metrics is greatest when communities differing in the relative abundance of their OTUs are rarefied.** For each dataset samples were randomly assigned to one of two community distributions where the abundance of OTUs differed.
-To calculate the power for each study, datasets were regenerated 100 times and differences in alpha diversity were tested using a Wilcoxon test (A) and differences in beta diversity were tested using PERMANOVA (B) at a 5% threshold.
+To calculate the power for each datasets, datasets were regenerated 100 times and differences in alpha diversity were tested using a Wilcoxon test (A) and differences in beta diversity were tested using PERMANOVA (B) at a 5% threshold.
 The power was the number of times a dataset yielded a significant result.
@@
 The ability to detect true differences in treatment groups for alpha diversity metrics is greatest when communities differing in richness are rarefied.** For each dataset samples were randomly assigned to one of two community distributions where one distribution contained a subset of OTUs found in the other.
-To calculate the power for each study, datasets were regenerated 100 times and differences in alpha diversity were tested using a Wilcoxon test (A) and differences in beta diversity were tested using PERMANOVA (B) at a 5% threshold.
+To calculate the power for each dataset, datasets were regenerated 100 times and differences in alpha diversity were tested using a Wilcoxon test (A) and differences in beta diversity were tested using PERMANOVA (B) at a 5% threshold.
 The power was the number of times a dataset yielded a significant result.

### 2022-11-02 submission/manuscript.tex
Writing M&M section

@@
 approaches of measuring alpha diversity to sampling effort, I generated
-null models for each study. Under a null model, each community from the
-same study would be expected to have the same alpha diversity regardless
-of the sampling effort. I measured the richness of the communities in
-each study without any correction, using scaled ranked subsampling (SRS)
-normalized OTU counts, with estimates based on non-parametric and
-parametric approaches, and using rarefaction (e.g.~Figure S2). For each
-study, all of the approaches, except for rarefaction, showed a strong
-correlation between richness and the number of sequences in the sample
-(Figure 1A). Next, I assessed diversity using the Shannon diversity
-index and the inverse Simpson diversity index without any correction,
-using normalized OTU counts, and rarefaction; I also used a
-non-parametric estimator of Shannon diversity. The correlation between
-sampling depth and the diversity metric was not as strong as it was for
-richness and the inverse Simpson diversity values were less sensitive
-than the Shannon diversity values; however, the correlation to the
-rarefied diversity metrics were the lowest for all of the metrics and
-studies (Figure 1A). The rarefied alpha-diversity metrics consistently
-demonstrated a lack of sensitivity to sampling depth.
+null models for each dataset. Under a null model, each community from
+the same dataset would be expected to have the same alpha diversity
+regardless of the sampling effort. I measured the richness of the
+communities in each dataset without any correction, using scaled ranked
+subsampling (SRS) normalized OTU counts, with estimates based on
+non-parametric and parametric approaches, and using rarefaction
+(e.g.~Figure S2). For each dataset, all of the approaches, except for
+rarefaction, showed a strong correlation between richness and the number
+of sequences in the sample (Figure 1A). Next, I assessed diversity using
+the Shannon diversity index and the inverse Simpson diversity index
+without any correction, using normalized OTU counts, and rarefaction; I
+also used a non-parametric estimator of Shannon diversity. The
+correlation between sampling depth and the diversity metric was not as
+strong as it was for richness and the inverse Simpson diversity values
+were less sensitive than the Shannon diversity values; however, the
+correlation to the rarefied diversity metrics were the lowest for all of
+the metrics and studies (Figure 1A). The rarefied alpha-diversity
+metrics consistently demonstrated a lack of sensitivity to sampling
+depth.
 
@@
 the Jaccard distance coefficient between all pairs of communities within
-a study. The Jaccard distance coefficient is the fraction of OTUs that
+a dataset. The Jaccard distance coefficient is the fraction of OTUs that
 are unique to either community and does not account for the abundance of
@@
 permutation-based multivariate analysis of variance (PERMANOVA). First,
-within each study, I randomly assigned each sample to one of two
-treatment groups. My expectation was that approximately 5 of the 100
-(5\%) random tests for each comparison would yield a significant test
-result. Indeed, for each study and alpha and beta diversity metric and
+I employed an unbiased null treatment model to measure the false
+detection rate, which should not have meaningfully differed from 5\%.
+Indeed, for each dataset and alpha and beta diversity metric and
 strategy for accounting for uneven sampling, approximately 5\% of the
-tests yielded a significant result (Figure 2). Second, within each
-study, I assigned samples with more than the median number of sequences
-per sample to one treatment group and the rest to another treatment
-group. If there was no sensitivity to sampling effort, I would have
-expected that 5\% of the tests would yield a significant result. In
-fact, only the rarefied data consistently resulted in a 5\% false
+tests yielded a significant result (Figure 2). Second, I employed a
+biased null treatment model where the treatment group was completely
+confounded with the number of sequences in each sample. Under this
+model, only the rarefied data consistently resulted in a 5\% false
 positive rate for alpha and beta diversity metrics (Figure 2). These
-results align with the observed sensitivity of alpha and beta diversity
-metrics to sampling effort and underscore the value of rarefaction.
+results aligned with the observed sensitivity of alpha and beta
+diversity metrics to sampling effort and underscore the value of
+rarefaction.
 
@@
 different approaches to control for uneven sampling effort I performed
-two additional simulations. In the first simulation, for each study
-communities were randomly assigned to one of two treatment groups.
-Communities in the first treatment group were generated by sampling from
-the null distribution. Samples in the second treatment group were
-generated by perturbing the null distribution by increasing the relative
-abundance of 10\% of the OTUs by 5\%. These values were determined after
-empirically searching for conditions that resulted in a large fraction
-of the randomizations yieleding a significant result across most of the
-studies. The fraction of tests that yielded a significant test was a
-measure of the statistical power for the test. The power to detect
-differences in richness by all approaches was low (Figure 4A). This was
-likely because the approach for generating the perturbed community did
-not necessarily change the number of OTUs in each treatment group.
+two additional simulations. In the first simulation, I implemented a
+skewed abundance distribution model to create two treatment groups for
+each datasets that were each populated with half of the samples each
+with the same number of sequences as the samples had in the observed
+data. The power to detect differences in richness between the two
+simulated treatment groups by all approaches was low (Figure 4A). This
+was likely because the approach for generating the perturbed community
+did not necessarily change the number of OTUs in each treatment group.
 Regardless, the simulations testing differencse in richness using the
 Rice and Stream datasets had the greatest power when the richness data
-were rarefied. To explore this further, in a second simulation the
-second treatment group was perturbed by removing 3\% of the OTUs from
-the model. As suggested by the first simulation, the rarefied richness
-data had a higher statistical power than the other approaches when
-measuring richness (Figure 5). The simulations testing the power to
-detect differences in Shannon diversity also showed that rarefied data
+were rarefied. To explore this further, a richness-adjusted community
+model was created by removing 3\% of the OTUs from a null model model.
+As suggested by the first simulation, the rarefied richness data had a
+higher statistical power than the other approaches when measuring
+richness (Figure 5). The simulations testing the power to detect
+differences in Shannon diversity also showed that rarefied data
 performed other methods (Figure 4A). When testing for differences in the
@@
 sequencing depth that is considerably lower than that obtained for the
-best sequenced community in a study, sequencing coverage for many
+best sequenced community in a dataset, sequencing coverage for many
 environments is probably adequate even at the lower sequencing depth. Of
@@
 
-\textbf{Choice of datasets.} The specific studies were selected because
-their data was publicly accessible through the Sequence Read Archive,
-the original investigators sequenced the V4 region of the 16S rRNA gene
-using paired 250 nt reads, and my previous familiarity with the data.
-The use of paired 250 nt reads to sequence the V4 region resulted in a
-near complete two-fold overap of the V4 region resulting in high quality
-contigs with a low sequencing error rate (1). These data were processed
-through the standard sequence curation pipeline to generate operational
-taxonomic units (OTUs) using the mothur software package (1, 2).
+\textbf{Choice of datasets.} The specific studies used in this study
+were selected because their data was publicly accessible through the
+Sequence Read Archive, the original investigators sequenced the V4
+region of the 16S rRNA gene using paired 250 nt reads, and my previous
+familiarity with the data. The use of paired 250 nt reads to sequence
+the V4 region resulted in a near complete two-fold overap of the V4
+region resulting in high quality contigs with a low sequencing error
+rate (1). These data were processed through a standardized sequence
+curation pipeline to generate operational taxonomic units (OTUs) using
+the mothur software package (1, 2). OTUs were identified using the
+OptiClust algorithm to cluster sequences together that were not more
+than 3\% different from each other (3).
 
-\textbf{Null community model.} Null models were generated by randomly
-assigning each sequence in the study to an OTU and sample while keeping
-constant the number of sequences per sample and the total number of
-sequences in each OTU. Because the construction of the null models was a
-stochastic process, 100 replicates were geneated for each study.
+\textbf{Null community model.} Null community models were generated such
+that within a dataset the number of sequences per sample and the number
+of sequences per OTU across all samples within the dataset were the same
+as was observed in original. This model effectively generated
+statistical samples of a single community so that there should not have
+been a statistical difference between the samples. This model
+implemented by randomly assigning each sequence in the dataset to an OTU
+and sample while keeping constant the number of sequences per sample and
+the total number of sequences in each OTU. This is a similar approach to
+that of the IS algorithm described by Ulrich and Gotelli (4). Because
+the construction of the null models was a stochastic process, 100
+replicates were geneated for each dataset.
 
-\textbf{10\%/5\% model}
+\textbf{Null treatment model.} I created an unbiased and biased
+treatment model. In the unbiased model, samples were randomly assigned
+to one of two treatment groups. In the biased treatment model, samples
+that had more than the median number of seqeunces for a dataset were
+assigned to one treatment group and the rest were assigned to a second
+treatment group. Comparison of any diversity metric across the two
+treatment groups should have only yielded a significant result in 5\% of
+the simulations when testing under a Type I error (i.e., \(\upalpha\))
+of 0.05.
 
-\textbf{Remove model}
+\textbf{Skewed abundance community model.} In the skewed abundance
+community model, communities were randomly assigned to one of two
+simulated treatment groups. Communities in the first treatment group
+were generated by calculating the relative abundance of each OTU across
+all samples and using those values as the probability of sampling each
+OTU. This probability distribution was sampled until each sample had the
+same number of sequences that it did in the observed data. Samples in
+the second treatment group were generated by adjusting the relative
+abundances of the OTUs in the firs treatment group by increasing the
+relative abundance of 10\% of the OTUs by 5\%. These values were
+determined after empirically searching for conditions that resulted in a
+large fraction of the randomizations yieleding a significant result
+across most of the studies. Sequences were sampled from the skewed
+community community until each sample had the same number of sequences
+that it did in the observed data. Under the skewed abundance community
+model each sample represented a statistical sampling of two communities
+such that there should not have been a statistically significant
+difference within a treatment group, but there was between the treatment
+groups. Because the construction of the skewed abundance community model
+was a stochastic process, 100 replicates were geneated for each dataset.
 
-\textbf{Null treatment model}
+\textbf{Richness-adjusted community model.} In the richness-adjusted
+community model, communities were randomly assigned to one of two
+simulated treatment groups. Communities in the first treatment group
+were generated by calculating the relative abundance of each OTU across
+all samples and using those values as the probability of sampling each
+OTU. This probability distribution was sampled until each sample had the
+same number of sequences that it did in the observed data. Samples in
+the second treatment group were generated by removing 3\% of the OTUs
+from the dataset and recalculating the relative abundance of the
+remaining OTUs. This percentage was determined after empirically
+searching for a value that resulted in a large fraction of the
+randomizations yieleding a significant result across most of the
+studies. Sequences were sampled from the richness-adjusted community
+distrubtion until each sample had the same number of sequences that it
+did in the observed data. Under the richness-adjusted community model
+each sample represented a statistical sampling of two communities such
+that there should not have been a statistically significant difference
+within a treatment group, but there was between the treatment groups.
+Because the construction of the richness-adjusted community model was a
+stochastic process, 100 replicates were geneated for each dataset.
 
-\textbf{Size-based treatment model}
+\textbf{Alpha diversity calculations.} Various strategies for handling
+uneven sampling effort were evaluated to identify the best approach for
+calculating community richness and Shannon and inverse Simpson diversity
+indeices. Raw OTU counts were used as input to calculate sample richness
+and Shannon and inverse Simpson diversity using mothur (2, 5). Shannon
+diversity was calculated as
 
-\textbf{Data availability.}
+\[H_{shannon} = - \sum_{i=1}^{S_{obs}} \frac{n_i}{N} ln \frac{n_i}{N}\]
 
-\textbf{Reproducible data analysis.}
+The Simpson diversity was calculated as
+
+\[D_{simpson} = \frac {\sum_{i=1}^{S_{obs}} {n_i \left ( n_i - 1 \right )}}{N \left( N-1 \right )}\]
+
+The inverse Simpson diversity was calculated as \(1/D_{simpson}\). In
+both formulae, \(n_i\) is the number of sequences in OTU \(i\) and \(N\)
+is the number of sequences in the sample. Rarefaction of richness,
+Shannon diversity and Inverse Simpson diversity values were carried out
+in mothur. Briefly, mothur calculates each value on a random draw of the
+same number of sequences from each sample and obtains a mean value based
+on 1,000 random draws. Scaled ranked subsampling (SRS) was used to
+normalize OTU counts to the size of the smallest sample in each dataset
+using the SRS R package (v.0.2.3)(6). Normalized OTU counts were used to
+calculate sample richness and Shannon and inverse Simpson diversity
+values using mothur. The non-parametric bias-corrected Chao1 and ACE
+richness estimators (7) and a non-parametric estimator of the Shannon
+diversity (8) were calculated using raw OTU counts with mothur.
+Parametric estimates of sample richness were calculated using the
+breakaway (BA) R package (v.4.7.9)(9). The current analysis reports both
+the results from running default model selection procedure and the
+Poisson model. The default model selection returned either the Kemp,
+Negative Binomial, or Poisson models.
+
+\textbf{Beta diversity calculations.} Similar to the alpha diversity
+calculations, multiple approaches were used to control for uneven
+sampling effort and calculate beta diversity. Raw and OTU counts were
+used for input to calculate the Jaccard, Bray-Curtis, and Euclidean
+dissimilarity indices using the vegdist function from the vegan R
+package (v.2.6.2)(10). The Jaccard index was calculated as
+
+\[D_{Jaccard}=1-\frac{S_{AB}}{S_A+S_B-S_{AB}}\]
+
+where \(S_A\) and \(S_B\) were the number of OTUs in samples \(A\) and
+\(B\) and \(S_{AB}\) was the number of OTUs shared between the two
+samples. The Bray-Curtis index was calculated as
+
+\[D_{Bray-Curtis}=1-\frac{\sum_{i=1}^{S_T} \left| n_{A,i} - n_{B,i}\right| }{ N_A + N_B}\]
+
+where \(n_{A,i}\) and \(n_{B,i}\) are the number of sequences observed
+in OTU \(i\) from samples \(A\) and \(B\), respectively. \(N_A\) and
+\(N_B\) are the total number of sequences in samples \(A\) and \(B\),
+respectively. \(S_T\) is the total number of OTUs observed between the
+two samples. The Euclidean distance was calculated as
+
+\[D_{Euclidean}=1-\sqrt{\sum_{i=1}^{S_T}\left(n_{A,i} - n_{B,i}\right)^2}\]
+
+These metrics were calculated using the relative abundance of each OTU
+using the vegdist function from vegan. The relative abundance was
+calculated as the number of sequences in the OTU (e.g., \(n_{A,i}\))
+divided by the total number of sequences in the sample (e.g., \(N_A\)).
+Rarefied beta-diversity values were calculated using the avgdist
+function in vegan. Briefly, vegan's avgdist function calculates each
+pairwise dissimilarity index after obtaining a random draw of the same
+number of sequences from each sample. After obtaining 100 random draws
+it returns the mean value.
+
+SRS normalized - Jaccard/BC CSS normalized - Jaccard/BC DeSeq2
+Center-log ratio
+
+\textbf{Power analysis.} The fraction of tests that yielded a
+significant test was a measure of the statistical power for the test.
+
+\textbf{Test of statistical significance.} Wilcox, PERMANOVA
+
+\textbf{Reproducible data analysis.} mothur - 1.47.0
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-Westcott2017}{}}%
+\CSLLeftMargin{3. }%
+\CSLRightInline{\textbf{Westcott SL}, \textbf{Schloss PD}. 2017.
+{OptiClust}, an improved method for assigning amplicon-based sequence
+data to operational taxonomic units. {mSphere} \textbf{2}.
+doi:\href{https://doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Ulrich2010}{}}%
+\CSLLeftMargin{4. }%
+\CSLRightInline{\textbf{Ulrich W}, \textbf{Gotelli NJ}. 2010. Null model
+analysis of species associations using abundance data. Ecology
+\textbf{91}:3384--3397.
+doi:\href{https://doi.org/10.1890/09-2157.1}{10.1890/09-2157.1}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Magurran2004}{}}%
+\CSLLeftMargin{5. }%
+\CSLRightInline{\textbf{Magurran AE}. 2004.
+\href{https://books.google.com/books?id=tUqzLSUzXxcC}{Measuring
+biological diversity}. Wiley.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Beule2020}{}}%
+\CSLLeftMargin{6. }%
+\CSLRightInline{\textbf{Beule L}, \textbf{Karlovsky P}. 2020.
+\href{https://doi.org/10.7717/peerj.9593}{Improved normalization of
+species count data in ecology by scaling with ranked subsampling (SRS):
+Application to microbial communities}. PeerJ \textbf{8}:e9593.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Chao2016}{}}%
+\CSLLeftMargin{7. }%
+\CSLRightInline{\textbf{Chao A}, \textbf{Chiu C-H}. 2016. Species
+richness: Estimation and comparison. Wiley StatsRef: statistics
+reference online \textbf{1}:26.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Chao2003}{}}%
+\CSLLeftMargin{8. }%
+\CSLRightInline{\textbf{Chao A}, \textbf{Shen T-J}. 2003. Nonparametric
+estimation of shannon's index of diversity when there are unseen species
+in sample. Environmental and ecological statistics
+\textbf{10}:429--443.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Willis2015}{}}%
+\CSLLeftMargin{9. }%
+\CSLRightInline{\textbf{Willis A}, \textbf{Bunge J}. 2015. Estimating
+diversity via frequency ratios. Biometrics \textbf{71}:1042--1049.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Oksanen2022}{}}%
+\CSLLeftMargin{10. }%
+\CSLRightInline{\textbf{Oksanen J}, \textbf{Simpson GL},
+\textbf{Blanchet FG}, \textbf{Kindt R}, \textbf{Legendre P},
+\textbf{Minchin PR}, \textbf{O'Hara RB}, \textbf{Solymos P},
+\textbf{Stevens MHH}, \textbf{Szoecs E}, \textbf{Wagner H},
+\textbf{Barbour M}, \textbf{Bedward M}, \textbf{Bolker B},
+\textbf{Borcard D}, \textbf{Carvalho G}, \textbf{Chirico M}, \textbf{De
+Caceres M}, \textbf{Durand S}, \textbf{Evangelista HBA},
+\textbf{FitzJohn R}, \textbf{Friendly M}, \textbf{Furneaux B},
+\textbf{Hannigan G}, \textbf{Hill MO}, \textbf{Lahti L}, \textbf{McGlinn
+D}, \textbf{Ouellette M-H}, \textbf{Ribeiro Cunha E}, \textbf{Smith T},
+\textbf{Stier A}, \textbf{Ter Braak CJF}, \textbf{Weedon J}. 2022.
+\href{https://CRAN.R-project.org/package=vegan}{Vegan: Community ecology
+package}.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-Li2015}{}}%
-\CSLLeftMargin{3. }%
+\CSLLeftMargin{11. }%
 \CSLRightInline{\textbf{Li Q}, \textbf{Heist EP}, \textbf{Moe LA}. 2015.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Baxter2016}{}}%
-\CSLLeftMargin{4. }%
+\CSLLeftMargin{12. }%
 \CSLRightInline{\textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Beall2015}{}}%
-\CSLLeftMargin{5. }%
+\CSLLeftMargin{13. }%
 \CSLRightInline{\textbf{Beall BFN}, \textbf{Twiss MR}, \textbf{Smith
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Henson2016}{}}%
-\CSLLeftMargin{6. }%
+\CSLLeftMargin{14. }%
 \CSLRightInline{\textbf{Henson MW}, \textbf{Pitre DM}, \textbf{Weckhorst
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Baxter2014}{}}%
-\CSLLeftMargin{7. }%
+\CSLLeftMargin{15. }%
 \CSLRightInline{\textbf{Baxter NT}, \textbf{Wan JJ}, \textbf{Schubert
@@
 \leavevmode\vadjust pre{\hypertarget{ref-LevyBooth2018}{}}%
-\CSLLeftMargin{8. }%
+\CSLLeftMargin{16. }%
 \CSLRightInline{\textbf{Levy-Booth DJ}, \textbf{Giesbrecht IJW},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Edwards2015}{}}%
-\CSLLeftMargin{9. }%
+\CSLLeftMargin{17. }%
 \CSLRightInline{\textbf{Edwards J}, \textbf{Johnson C},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Ettinger2017}{}}%
-\CSLLeftMargin{10. }%
+\CSLLeftMargin{18. }%
 \CSLRightInline{\textbf{Ettinger CL}, \textbf{Williams SL},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Graw2018}{}}%
-\CSLLeftMargin{11. }%
+\CSLLeftMargin{19. }%
 \CSLRightInline{\textbf{Graw MF}, \textbf{DAngelo G}, \textbf{Borchers
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Johnston2016}{}}%
-\CSLLeftMargin{12. }%
+\CSLLeftMargin{20. }%
 \CSLRightInline{\textbf{Johnston ER}, \textbf{Rodriguez-R LM},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Hassell2018}{}}%
-\CSLLeftMargin{13. }%
+\CSLLeftMargin{21. }%
 \CSLRightInline{\textbf{Hassell N}, \textbf{Tinker KA}, \textbf{Moore
@@
 \textbf{Table 1. Summary of studies used in the analysis.} For all
-studies, the number of sequences used from each study was rarefied to
+studies, the number of sequences used from each dataset was rarefied to
 the smallest sample size. A graphical represenation of the distribution
-of sample sizes for each study and the samples that were removed from
-each study are provided in Figure S1.
+of sample sizes for each dataset and the samples that were removed from
+each dataset are provided in Figure S1.
 
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1367}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0664}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1914}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1953}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2031}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2070}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1434}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0659}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1899}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1938}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2016}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2054}}@{}}
 \toprule()
 \begin{minipage}[b]{\linewidth}\raggedright
-\textbf{Study\nobreakspace{}(Ref)}
+\textbf{Dataset\nobreakspace{}(Ref)}
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
@@
 \endhead
-Bioethanol~(3) & 95 & 3,970,972 & 16,014 & 3,690\Hyphdash*356,027 &
+Bioethanol~(11) & 95 & 3,970,972 & 16,014 & 3,690\Hyphdash*356,027 &
 SRP055545 \\
-Human~(4) & 490 & 20,828,275 & 32,452 & 10,439\Hyphdash*422,904 &
+Human~(12) & 490 & 20,828,275 & 32,452 & 10,439\Hyphdash*422,904 &
 SRP062005 \\
-Lake~(5) & 52 & 3,145,486 & 69,205 & 15,135\Hyphdash*110,993 &
+Lake~(13) & 52 & 3,145,486 & 69,205 & 15,135\Hyphdash*110,993 &
 SRP050963 \\
-Marine~(6) & 7 & 1,484,068 & 213,091 & 132,895\Hyphdash*256,758 &
+Marine~(14) & 7 & 1,484,068 & 213,091 & 132,895\Hyphdash*256,758 &
 SRP068101 \\
@@
 SRP192323 \\
-Peromyscus~(7) & 111 & 1,545,288 & 12,393 & 4,454\Hyphdash*33,502 &
+Peromyscus~(15) & 111 & 1,545,288 & 12,393 & 4,454\Hyphdash*33,502 &
 SRP044050 \\
-Rainforest~(8) & 69 & 936,666 & 11,464 & 4,880\Hyphdash*37,403 &
+Rainforest~(16) & 69 & 936,666 & 11,464 & 4,880\Hyphdash*37,403 &
 ERP023747 \\
-Rice~(9) & 490 & 22,623,937 & 43,399 & 2,777\Hyphdash*192,200 &
+Rice~(17) & 490 & 22,623,937 & 43,399 & 2,777\Hyphdash*192,200 &
 SRP044745 \\
-Seagrass~(10) & 286 & 4,135,440 & 13,538 & 1,830\Hyphdash*45,076 &
+Seagrass~(18) & 286 & 4,135,440 & 13,538 & 1,830\Hyphdash*45,076 &
 SRP092441 \\
-Sediment~(11) & 58 & 1,151,389 & 17,606 & 7,686\Hyphdash*67,763 &
+Sediment~(19) & 58 & 1,151,389 & 17,606 & 7,686\Hyphdash*67,763 &
 SRP097192 \\
-Soil~(12) & 18 & 932,563 & 50,487 & 46,622\Hyphdash*58,935 &
+Soil~(20) & 18 & 932,563 & 50,487 & 46,622\Hyphdash*58,935 &
 ERP012016 \\
-Stream~(13) & 201 & 21,017,610 & 90,621 & 8,931\Hyphdash*394,419 &
+Stream~(21) & 201 & 21,017,610 & 90,621 & 8,931\Hyphdash*394,419 &
 SRP075852 \\
@@
 community distributions where the abundance of OTUs differed. To
-calculate the power for each study, datasets were regenerated 100 times
-and differences in alpha diversity were tested using a Wilcoxon test (A)
-and differences in beta diversity were tested using PERMANOVA (B) at a
-5\% threshold. The power was the number of times a dataset yielded a
-significant result.
+calculate the power for each datasets, datasets were regenerated 100
+times and differences in alpha diversity were tested using a Wilcoxon
+test (A) and differences in beta diversity were tested using PERMANOVA
+(B) at a 5\% threshold. The power was the number of times a dataset
+yielded a significant result.
 
@@
 distribution contained a subset of OTUs found in the other. To calculate
-the power for each study, datasets were regenerated 100 times and
+the power for each dataset, datasets were regenerated 100 times and
 differences in alpha diversity were tested using a Wilcoxon test (A) and

### 2022-11-03 code/plot_coverage.R
Re-run with merged R scripts

@@ -45,7 +45,7 @@ cor_plot <- obs_cor %>%
   geom_line(data = cor_line, aes(x = nseqs, y = mean), color = "black") +
   geom_vline(xintercept = 800, color = "darkgray") +
   scale_x_log10(breaks = c(325, 1e3, 1e4, 1e5, 1e6),
-                labels = c("R", "10^3", "10^4", "10^5", "10^6"),
+                labels = c("R", "10^3^", "10^4^", "10^5^", "10^6^"),
                 limits = c(2e2, 1e6)) +
   scale_color_manual(name = NULL,
                     values = rep(c('#1b9e77','#d95f02','#7570b3'), 4),

### 2022-11-03 code/pool_beta_summary.R

@@ -21,18 +21,11 @@ reader <- function(x) {
 
   read_tsv(x,
            col_types = cols(.default = col_double())) %>%
-<<<<<<< HEAD
     mutate(seed = as.numeric(seed), dataset = dataset, method = dmethod,
            .before = n_seqs)
-=======
-  mutate(seed = seed, dataset = dataset, method = dmethod, .before = n_seqs)
->>>>>>> paper
 
 }
 
 map_dfr(summary_files, reader) %>%
-<<<<<<< HEAD
   arrange(seed, method, n_seqs) %>%
-=======
->>>>>>> paper
   write_tsv(output_file)

### 2022-11-03 code/rarefy_beta_analysis.R

@@ -3,21 +3,14 @@
 library(tidyverse)
 library(data.table)
 library(vegan)
-<<<<<<< HEAD
 library(glue)
 library(magrittr)
 
-=======
-# library(furrr)
-
-# plan(multicore, workers = 8)
->>>>>>> paper
 
 input <- commandArgs(trailingOnly = TRUE)
 
 shared_file <- input[1] # shared_file <- "data/mice/data.otu.1.rshared"
 nseqs_file <- input[2] # nseqs_file <- "data/mice/data.group_count"
-<<<<<<< HEAD
 remove_file <- input[3] # remove_file <- "data/mice/data.remove_accnos"
 output_file <- input[4] # output_file <- "data/mice/data.otu.1.1.r_rare_bray.summary"
 
@@ -32,25 +25,11 @@ max_nseqs <- read_tsv(nseqs_file,
                       col_types = cols(Group = col_character()),
                       col_names = c("Group", "nseqs")) %>%
   anti_join(., to_remove) %>%
-=======
-output_file <- input[3] # output_file <- "data/mice/data.otu.1.r_rare_bray.summary"
-
-shared_df <- fread(shared_file, colClasses = c(Group = "character")) %>%
-  select(Group, starts_with("Otu")) %>%
-  column_to_rownames("Group")
-
-groups <- rownames(shared_df)
-
-max_nseqs <- read_tsv(nseqs_file,
-                  col_names = c("Group", "nseqs")) %>%
-  filter(Group %in% groups) %>%
->>>>>>> paper
   arrange(nseqs) %>% 
   slice(n = nrow(.) - 4) %>% # insures there are at least 10 distances in 
   pull(nseqs)                     # final class
 
 
-<<<<<<< HEAD
 
 nseqs <- 10^seq(log10(1000), log10(max_nseqs), length.out = 100) %>%
   as.integer() %>%
@@ -71,38 +50,14 @@ temp_dist <- str_replace(temp_shared, "shared", glue("{distance}.0.03.square.ave
 
 read_tsv(temp_dist, skip = 1, col_names = FALSE) %>%
   set_colnames(c("samples", all_of(.[["X1"]]))) %>%
-=======
-nseqs <- c(1000, seq(2000, max_nseqs, 2000))
-
-distance <- str_replace(output_file, ".*r_rare_(.*)\\.summary", "\\1")
-
-run_rarefy_beta <- function(df, x, distance) {
-
-  avgdist(df, sample = x, dmethod = distance)  %>%
-      as.matrix() %>%
[…]

### 2022-11-08 submission/manuscript.Rmd
First full draft of Materials and Methods

@@
 
+***https://edepot.wur.nl/547087.*** "Log- ratio PCA is designed to give results that are library size- independent.
+However, as we demonstrated mathematically and with examples based on simulated and real data, log- ratio PCA becomes library size- dependent, if there are many infrequent taxa (many zeroes) and library sizes differ largely.
+In this situation, the row centring used in log- ratio PCA brings an effect of r (the row mean of the log- transformed counts) in the clr- transformed matrix.
+Note that this effect is irrespective of whether or not these infrequent taxa are genuine or due to sequencing noise or allocation error.
+This library size dependence is unexpected in the sense that, after applying the clr, the transformed matrix is free of the effect of the row totals for strictly positive data (yij>0 for all i and j).
+We additionally demonstrate that library size variability causes a loss in power in dextecting an effect of x with log- ratio RDA.
+If there is additionally a correlation between treatment and the library size, the type 1 error for detecting the effect of x can be seriously inflated."
+
+***https://www.nature.com/articles/s41522-020-00160-w.*** "An important characteristic of a feature table is that it is typically sparse, sometimes as many as ~90% are zero entries21, which creates a challenge for analyzing rare taxa.
+A quick and simple strategy to deal with excess zeros is to add a small positive constant (e.g.
+1) called pseudo-count14,22 to each cell of the feature table.
+The addition of a pseudo-count becomes necessary when using methods of analysis that require log transformation of the observed counts.
+Even though adding a pseudo-count is simple and widely used, the choice of the pseudo-count is ad hoc.
+Studies have shown that differential abundance or clustering results could be sensitive to the choice of pseudo count23,24.
+Although different values of pseudo counts have been discussed in the literature23,24,25,26, to the best of our knowledge, there is no consensus on how to choose the optimal value.
+Other strategies involve modeling zero counts by some probability models21,27."
+
+https://academic.oup.com/bioinformatics/article/34/16/2870/4956011
+https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6755255/
+
 * Motivation
@@
 The Aitchison distances were all strongly sensitive to sampling effort (Figure 1B).
-Finally, I used the variance stabilization technique (VST) from DeSeq2 prior to calculating Euclidean distances.
+Finally, I used the variance-stabilization transformation (VST) from DeSeq2 prior to calculating Euclidean distances.
 Again, there was a strong sensitivity to sampling effort (Figure 1B).
@@
 Samples in the second treatment group were generated by removing 3% of the OTUs from the dataset and recalculating the relative abundance of the remaining OTUs.
-This percentage was determined after empirically searching for a value that resulted in a large fraction of the randomizations yieleding a significant result across most of the studies.
 Sequences were sampled from the richness-adjusted community distrubtion until each sample had the same number of sequences that it did in the observed data.
@@
 Because the construction of the richness-adjusted community model was a stochastic process, 100 replicates were geneated for each dataset.
+
+**Test of statistical significance.** Statistical comparisons of alpha diversity metrics across the simulated treatment groups were performed using the non-parametric two-sample Wilcoxon test as implemented in R.
+This test was selected because the alpha-diversity metrics tended to not be normally distributed and each dataset required a different transformation to normalize the data.
+Comparisons of beta diversity metrics were performed using the adonis2 function from the vegan (v.`r package_version("vegan")`) R package [@Oksanen2022].
+The adonis2 function implements a non-parametric multivariate analysis of variance using distances matrices as described by McArdle and Anderson [@McArdle2001].
+Throughout this study I used 0.05 as the threshold for asssessing statistical significance.
+
+**Power analysis.** The parameters used to design the skewed abundance and richness-adjusted community models were set to impose a known effect size when using rarefied community data.
+The statistical power to detect these differences was determined by calculating the p-value for each of 100 replicate simulated set of samples from each dataset using the various alpha and beta diversity metrics.
+The percentage of tests that yielded a significant P-value was considered the statistical power (i.e., 1 minus the Type II error) to detect the difference.
 
@@
 Normalized OTU counts were used to calculate sample richness and Shannon and inverse Simpson diversity values using mothur.
+Data normalized by cumulative sum scaling (CSS) were not reported for alpha-diversity values since the relative abundances of the features do not change with the normalization procedure [@Paulson2013].
 The non-parametric bias-corrected Chao1 and ACE richness estimators [@Chao2016] and a non-parametric estimator of the Shannon diversity [@Chao2003] were calculated using raw OTU counts with mothur.
@@
 The default model selection returned either the Kemp, Negative Binomial, or Poisson models.
+Relative abundance data were not used to calculate alpha diversity metrics since the richness and evenness does not change from the raw data when dividing each sample by the total number of sequences in the sample.
 
@@
 The relative abundance was calculated as the number of sequences in the OTU (e.g., $n_{A,i}$) divided by the total number of sequences in the sample (e.g., $N_A$).
+
 Rarefied beta-diversity values were calculated using the avgdist function in vegan.
@@
 
-SRS normalized - Jaccard/BC
-CSS normalized - Jaccard/BC
-DeSeq2
-Center-log ratio
+Three approaches were taken to normalize the number of sequences across samples within a dataset.
+Scaled ranked subsampling (SRS) and cumulative sum scaling (CSS) were used to normalize raw OTU counts using the SRS (v.`r package_version("SRS")`) and metagenomeSeq (v.`r package_version("metagenomeSeq")`) R packages, respectively [@Beule2020; @Paulson2013].
+The normalized counts were then used to calculate Jaccard and Bray-Curtis dissimilarity indices.
+Finally, the variance-stabilization transformation (VST) as implemented in the DESeq2 (v.`r package_version("DESeq2")`) R package was used to normalize the data as described by McMurdie and Holmes [@Love2014; @McMurdie2014].
+Because the VST approach generates negative values, which are incompatible with calculating Jaccard and Bray-Curtis dissimilarity values, Euclidean distances were calcualted instead.
 
-**Power analysis.**
- The fraction of tests that yielded a significant test was a measure of the statistical power for the test.
+Raw OTU counts were used to calculate centered log ratio values for each OTU, which were then used to calcualte Euclidean distances; such distances are commonly refered to as Aitchison distances.
+Centered log-ratio (CLR) abundances are calculated as:
 
-**Test of statistical significance.**
-Wilcox, PERMANOVA
+$$
+clr\left(n_j\right) = \left[ \ln\frac{x_{ij}}{g(x_j)}, ..., \ln\frac{x_{S_Tj}}{g(x_j)}\right]
+$$
 
-**Reproducible data analysis.** 
-mothur - 1.47.0
+where $x_{ij} was the number of sequences observed for OTU $i$ in sample $j$ and $g()$ was the geometric mean $x_j$ was the counts of the $S_T$ OTUs in sample $j$.
+Because the geometric mean is zero if any OTU is absent from a sample, the CLR is undefined when there are unobserved OTUs in a sample.
+To overcome this problem, I attempted a four approaches.
+The first, Zero CLR, imputed the value of the zeroes based on the observed data using the zCompositions (v.`r package_version("zCompositions")`) R package [@Palarea2015].
+The second, One CLR, added a pseudocount of 1 to the abundance of all OTUs [@Paulson2013; @Lin2020].
+The third, Nudge CLR, added a pseudocount of 1 divided by the total number of sequences in a sample to each OTU in the sample [@Costea2014; @Lin2020].
+The final approach, Robust CLR, removed unobserved OTUs prior to calcualting the CLR [@Martino2019].
 
-***doi: 10.1038/nmeth.2658.*** Should CSS be used in alpha diversity analysis?
-No.
-The richness and shannon diversity values were no different from those obtained with the raw abundances.
-From the paper, "The relative proportion of the features is unaffected by the normalization".
-CSS paper refers to relative abundance approach as total-sum normalization (TSN)
+**Analysis of sequencing coverage.** To assess the level of sequencing coverage I calcualted Good's coverage ($C_{Good}$) using mothur:
 
-Robust CLR - remove zero counts and calculate CLR
-One CLR - add a pseudocount of 1 to all observations
-Nudge CLR - add a pseudocount of 1/total number of sequences in sample
-Zero CLR - Impute the value of zeroes using zCompositions package
+$$C_{Good} = 100\% \times \left(1-\frac{n_1}{N_T} \right)$$
 
-***https://edepot.wur.nl/547087.*** "Log- ratio PCA is designed to give results that are library size- independent.
-However, as we demonstrated mathematically and with examples based on simulated and real data, log- ratio PCA becomes library size- dependent, if there are many infrequent taxa (many zeroes) and library sizes differ largely.
-In this situation, the row centring used in log- ratio PCA brings an effect of r (the row mean of the log- transformed counts) in the clr- transformed matrix.
-Note that this effect is irrespective of whether or not these infrequent taxa are genuine or due to sequencing noise or allocation error.
-This library size dependence is unexpected in the sense that, after applying the clr, the transformed matrix is free of the effect of the row totals for strictly positive data (yij>0 for all i and j).
-We additionally demonstrate that library size variability causes a loss in power in detecting an effect of x with log- ratio RDA.
-If there is additionally a correlation between treatment and the library size, the type 1 error for detecting the effect of x can be seriously inflated."
+where $n_1$ is the number of OTUs with only one sequence in the sample and $N_T$ is the total number of sequences in the sample.
+Good's coverage was calculated using (i) the observed OTU counts for each sample and dataset, (ii) following rarefaction (1,000 iterations) of the observed OTU counts to the size of the smallest sample in each dataset, and (iii) after rarefying (1,000 iterations) the null community distribution.
 
-***https://www.nature.com/articles/s41522-020-00160-w.*** "An important characteristic of a feature table is that it is typically sparse, sometimes as many as ~90% are zero entries21, which creates a challenge for analyzing rare taxa.
-A quick and simple strategy to deal with excess zeros is to add a small positive constant (e.g.
-1) called pseudo-count14,22 to each cell of the feature table.
-The addition of a pseudo-count becomes necessary when using methods of analysis that require log transformation of the observed counts.
-Even though adding a pseudo-count is simple and widely used, the choice of the pseudo-count is ad hoc.
-Studies have shown that differential abundance or clustering results could be sensitive to the choice of pseudo count23,24.
-Although different values of pseudo counts have been discussed in the literature23,24,25,26, to the best of our knowledge, there is no consensus on how to choose the optimal value.
-Other strategies involve modeling zero counts by some probability models21,27."
+**Reproducible data analysis.** A complete reproducible workflow written in Snakemake (v.7.15.2) and conda (v.4.12.0) computational environment can be obtained from the GitHub hosted git repository for this project (https://github.com/SchlossLab/Schloss_Rarefaction_XXXXX_2022).
+This paper was written in R markdown (v.`r package_version("rmarkdown")`) with the kableExtra (v.`r package_version("kableExtra")`) package.
+The mothur (v.1.47.0) and R (`r glue("{version$major}.{version$minor}")`) software packages were used for all analyses with extensive use of functions in the tidyverse metapackage (v.`r package_version("tidyverse")`).
+A preliminary version of this analysis was presented as the Rarefaction video series on the Riffomonas Project YouTube channel (https://www.youtube.com/playlist?list=PLmNrK_nkqBpJuhS93PYC-Xr5oqur7IIWf).
 
-https://academic.oup.com/bioinformatics/article/34/16/2870/4956011
-https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6755255/
 \vspace{10mm}
@@
 **Acknowledgements.**
+
+I am grateful to the researchers who generated the datasets used in this study.
+I also thank the individuals who asked questions and commented on the preliminary version of this project, which was released as a YouTube playlist on the Riffomonas channel.
+This work was supported in part by funding from the National Institutes of Health (U01AI124255, P30DK034933, R01CA215574).
 
@@
 \newpage
+## Tables
 
@@
 ## Figures
-
-\newpage
 

### 2022-11-08 submission/manuscript.tex
First full draft of Materials and Methods

@@
 
+\textbf{\emph{\url{https://edepot.wur.nl/547087}.}} ``Log- ratio PCA is
+designed to give results that are library size- independent. However, as
+we demonstrated mathematically and with examples based on simulated and
+real data, log- ratio PCA becomes library size- dependent, if there are
+many infrequent taxa (many zeroes) and library sizes differ largely. In
+this situation, the row centring used in log- ratio PCA brings an effect
+of r (the row mean of the log- transformed counts) in the clr-
+transformed matrix. Note that this effect is irrespective of whether or
+not these infrequent taxa are genuine or due to sequencing noise or
+allocation error. This library size dependence is unexpected in the
+sense that, after applying the clr, the transformed matrix is free of
+the effect of the row totals for strictly positive data
+(yij\textgreater0 for all i and j). We additionally demonstrate that
+library size variability causes a loss in power in dextecting an effect
+of x with log- ratio RDA. If there is additionally a correlation between
+treatment and the library size, the type 1 error for detecting the
+effect of x can be seriously inflated.''
+
+\textbf{\emph{\url{https://www.nature.com/articles/s41522-020-00160-w}.}}
+``An important characteristic of a feature table is that it is typically
+sparse, sometimes as many as \textasciitilde90\% are zero entries21,
+which creates a challenge for analyzing rare taxa. A quick and simple
+strategy to deal with excess zeros is to add a small positive constant
+(e.g.~1) called pseudo-count14,22 to each cell of the feature table. The
+addition of a pseudo-count becomes necessary when using methods of
+analysis that require log transformation of the observed counts. Even
+though adding a pseudo-count is simple and widely used, the choice of
+the pseudo-count is ad hoc. Studies have shown that differential
+abundance or clustering results could be sensitive to the choice of
+pseudo count23,24. Although different values of pseudo counts have been
+discussed in the literature23,24,25,26, to the best of our knowledge,
+there is no consensus on how to choose the optimal value. Other
+strategies involve modeling zero counts by some probability
+models21,27.''
+
+\url{https://academic.oup.com/bioinformatics/article/34/16/2870/4956011}
+\url{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6755255/}
+
 \begin{itemize}
@@
 distances were all strongly sensitive to sampling effort (Figure 1B).
-Finally, I used the variance stabilization technique (VST) from DeSeq2
-prior to calculating Euclidean distances. Again, there was a strong
-sensitivity to sampling effort (Figure 1B). Although Euclidean distances
-are not typically used on raw or rarefied count data in ecology,
-rarefied Euclidean distances were not sensitive to sampling effort.
-Across each of the beta diversity metrics and approaches used to account
-for uneven sampling effort and sparsity, rarefaction was the least
-sensitive approach to differences in sampling effort.
+Finally, I used the variance-stabilization transformation (VST) from
+DeSeq2 prior to calculating Euclidean distances. Again, there was a
+strong sensitivity to sampling effort (Figure 1B). Although Euclidean
+distances are not typically used on raw or rarefied count data in
+ecology, rarefied Euclidean distances were not sensitive to sampling
+effort. Across each of the beta diversity metrics and approaches used to
+account for uneven sampling effort and sparsity, rarefaction was the
+least sensitive approach to differences in sampling effort.
 
@@
 from the dataset and recalculating the relative abundance of the
-remaining OTUs. This percentage was determined after empirically
-searching for a value that resulted in a large fraction of the
-randomizations yieleding a significant result across most of the
-studies. Sequences were sampled from the richness-adjusted community
-distrubtion until each sample had the same number of sequences that it
-did in the observed data. Under the richness-adjusted community model
-each sample represented a statistical sampling of two communities such
-that there should not have been a statistically significant difference
-within a treatment group, but there was between the treatment groups.
-Because the construction of the richness-adjusted community model was a
-stochastic process, 100 replicates were geneated for each dataset.
+remaining OTUs. Sequences were sampled from the richness-adjusted
+community distrubtion until each sample had the same number of sequences
+that it did in the observed data. Under the richness-adjusted community
+model each sample represented a statistical sampling of two communities
+such that there should not have been a statistically significant
+difference within a treatment group, but there was between the treatment
+groups. Because the construction of the richness-adjusted community
+model was a stochastic process, 100 replicates were geneated for each
+dataset.
+
+\textbf{Test of statistical significance.} Statistical comparisons of
+alpha diversity metrics across the simulated treatment groups were
+performed using the non-parametric two-sample Wilcoxon test as
+implemented in R. This test was selected because the alpha-diversity
+metrics tended to not be normally distributed and each dataset required
+a different transformation to normalize the data. Comparisons of beta
+diversity metrics were performed using the adonis2 function from the
+vegan (v.2.6.2) R package (5). The adonis2 function implements a
+non-parametric multivariate analysis of variance using distances
+matrices as described by McArdle and Anderson (6). Throughout this study
+I used 0.05 as the threshold for asssessing statistical significance.
+
+\textbf{Power analysis.} The parameters used to design the skewed
+abundance and richness-adjusted community models were set to impose a
+known effect size when using rarefied community data. The statistical
+power to detect these differences was determined by calculating the
+p-value for each of 100 replicate simulated set of samples from each
+dataset using the various alpha and beta diversity metrics. The
+percentage of tests that yielded a significant P-value was considered
+the statistical power (i.e., 1 minus the Type II error) to detect the
+difference.
 
@@
 indeices. Raw OTU counts were used as input to calculate sample richness
-and Shannon and inverse Simpson diversity using mothur (2, 5). Shannon
+and Shannon and inverse Simpson diversity using mothur (2, 7). Shannon
 diversity was calculated as
@@
 normalize OTU counts to the size of the smallest sample in each dataset
-using the SRS R package (v.0.2.3)(6). Normalized OTU counts were used to
+using the SRS R package (v.0.2.3)(8). Normalized OTU counts were used to
 calculate sample richness and Shannon and inverse Simpson diversity
-values using mothur. The non-parametric bias-corrected Chao1 and ACE
-richness estimators (7) and a non-parametric estimator of the Shannon
-diversity (8) were calculated using raw OTU counts with mothur.
-Parametric estimates of sample richness were calculated using the
-breakaway (BA) R package (v.4.7.9)(9). The current analysis reports both
-the results from running default model selection procedure and the
-Poisson model. The default model selection returned either the Kemp,
-Negative Binomial, or Poisson models.
+values using mothur. Data normalized by cumulative sum scaling (CSS)
+were not reported for alpha-diversity values since the relative
+abundances of the features do not change with the normalization
+procedure (9). The non-parametric bias-corrected Chao1 and ACE richness
+estimators (10) and a non-parametric estimator of the Shannon diversity
+(11) were calculated using raw OTU counts with mothur. Parametric
+estimates of sample richness were calculated using the breakaway (BA) R
+package (v.4.7.9)(12). The current analysis reports both the results
+from running default model selection procedure and the Poisson model.
+The default model selection returned either the Kemp, Negative Binomial,
+or Poisson models. Relative abundance data were not used to calculate
+alpha diversity metrics since the richness and evenness does not change
+from the raw data when dividing each sample by the total number of
+sequences in the sample.
 
@@
 dissimilarity indices using the vegdist function from the vegan R
-package (v.2.6.2)(10). The Jaccard index was calculated as
+package (v.2.6.2)(5). The Jaccard index was calculated as
 
@@
 divided by the total number of sequences in the sample (e.g., \(N_A\)).
+
 Rarefied beta-diversity values were calculated using the avgdist
@@
 
-SRS normalized - Jaccard/BC CSS normalized - Jaccard/BC DeSeq2
-Center-log ratio
-
-\textbf{Power analysis.} The fraction of tests that yielded a
-significant test was a measure of the statistical power for the test.
-
-\textbf{Test of statistical significance.} Wilcox, PERMANOVA
-
-\textbf{Reproducible data analysis.} mothur - 1.47.0
-
-\textbf{\emph{doi: 10.1038/nmeth.2658.}} Should CSS be used in alpha
-diversity analysis? No.~The richness and shannon diversity values were
-no different from those obtained with the raw abundances. From the
-paper, ``The relative proportion of the features is unaffected by the
-normalization''. CSS paper refers to relative abundance approach as
-total-sum normalization (TSN)
-
-Robust CLR - remove zero counts and calculate CLR One CLR - add a
-pseudocount of 1 to all observations Nudge CLR - add a pseudocount of
-1/total number of sequences in sample Zero CLR - Impute the value of
-zeroes using zCompositions package
-
-\textbf{\emph{\url{https://edepot.wur.nl/547087}.}} ``Log- ratio PCA is
-designed to give results that are library size- independent. However, as
-we demonstrated mathematically and with examples based on simulated and
-real data, log- ratio PCA becomes library size- dependent, if there are
-many infrequent taxa (many zeroes) and library sizes differ largely. In
-this situation, the row centring used in log- ratio PCA brings an effect
-of r (the row mean of the log- transformed counts) in the clr-
-transformed matrix. Note that this effect is irrespective of whether or
-not these infrequent taxa are genuine or due to sequencing noise or
-allocation error. This library size dependence is unexpected in the
-sense that, after applying the clr, the transformed matrix is free of
-the effect of the row totals for strictly positive data
-(yij\textgreater0 for all i and j). We additionally demonstrate that
-library size variability causes a loss in power in detecting an effect
-of x with log- ratio RDA. If there is additionally a correlation between
-treatment and the library size, the type 1 error for detecting the
-effect of x can be seriously inflated.''
-
-\textbf{\emph{\url{https://www.nature.com/articles/s41522-020-00160-w}.}}
-``An important characteristic of a feature table is that it is typically
-sparse, sometimes as many as \textasciitilde90\% are zero entries21,
-which creates a challenge for analyzing rare taxa. A quick and simple
-strategy to deal with excess zeros is to add a small positive constant
-(e.g.~1) called pseudo-count14,22 to each cell of the feature table. The
-addition of a pseudo-count becomes necessary when using methods of
-analysis that require log transformation of the observed counts. Even
-though adding a pseudo-count is simple and widely used, the choice of
-the pseudo-count is ad hoc. Studies have shown that differential
-abundance or clustering results could be sensitive to the choice of
-pseudo count23,24. Although different values of pseudo counts have been
-discussed in the literature23,24,25,26, to the best of our knowledge,
-there is no consensus on how to choose the optimal value. Other
-strategies involve modeling zero counts by some probability
-models21,27.''
-
-\url{https://academic.oup.com/bioinformatics/article/34/16/2870/4956011}
-\url{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6755255/}
+Three approaches were taken to normalize the number of sequences across
+samples within a dataset. Scaled ranked subsampling (SRS) and cumulative
+sum scaling (CSS) were used to normalize raw OTU counts using the SRS
+(v.0.2.3) and metagenomeSeq (v.1.36.0) R packages, respectively (8, 9).
+The normalized counts were then used to calculate Jaccard and
+Bray-Curtis dissimilarity indices. Finally, the variance-stabilization
+transformation (VST) as implemented in the DESeq2 (v.1.34.0) R package
+was used to normalize the data as described by McMurdie and Holmes (13,
+14). Because the VST approach generates negative values, which are
+incompatible with calculating Jaccard and Bray-Curtis dissimilarity
+values, Euclidean distances were calcualted instead.
+
+Raw OTU counts were used to calculate centered log ratio values for each
+OTU, which were then used to calcualte Euclidean distances; such
+distances are commonly refered to as Aitchison distances. Centered
+log-ratio (CLR) abundances are calculated as:
+
+\[
+clr\left(n_j\right) = \left[ \ln\frac{x_{ij}}{g(x_j)}, ..., \ln\frac{x_{S_Tj}}{g(x_j)}\right]
+\]
+
+where \$x\_\{ij\} was the number of sequences observed for OTU \(i\) in
+sample \(j\) and \(g()\) was the geometric mean \(x_j\) was the counts
+of the \(S_T\) OTUs in sample \(j\). Because the geometric mean is zero
+if any OTU is absent from a sample, the CLR is undefined when there are
+unobserved OTUs in a sample. To overcome this problem, I attempted a
+four approaches. The first, Zero CLR, imputed the value of the zeroes
+based on the observed data using the zCompositions (v.1.4.0.1) R package
+(15). The second, One CLR, added a pseudocount of 1 to the abundance of
+all OTUs (9, 16). The third, Nudge CLR, added a pseudocount of 1 divided
+by the total number of sequences in a sample to each OTU in the sample
+(16, 17). The final approach, Robust CLR, removed unobserved OTUs prior
+to calcualting the CLR (18).
+
+\textbf{Analysis of sequencing coverage.} To assess the level of
+sequencing coverage I calcualted Good's coverage (\(C_{Good}\)) using
+mothur:
+
+\[C_{Good} = 100\% \times \left(1-\frac{n_1}{N_T} \right)\]
+
+where \(n_1\) is the number of OTUs with only one sequence in the sample
+and \(N_T\) is the total number of sequences in the sample. Good's
+coverage was calculated using (i) the observed OTU counts for each
+sample and dataset, (ii) following rarefaction (1,000 iterations) of the
+observed OTU counts to the size of the smallest sample in each dataset,
+and (iii) after rarefying (1,000 iterations) the null community
+distribution.
+
+\textbf{Reproducible data analysis.} A complete reproducible workflow
+written in Snakemake (v.7.15.2) and conda (v.4.12.0) computational
+environment can be obtained from the GitHub hosted git repository for
+this project
+(\url{https://github.com/SchlossLab/Schloss_Rarefaction_XXXXX_2022}).
+This paper was written in R markdown (v.2.16) with the kableExtra
+(v.1.3.4) package. The mothur (v.1.47.0) and R (4.1.3) software packages
+were used for all analyses with extensive use of functions in the
+tidyverse metapackage (v.1.3.1). A preliminary version of this analysis
+was presented as the Rarefaction video series on the Riffomonas Project
+YouTube channel
+(\url{https://www.youtube.com/playlist?list=PLmNrK_nkqBpJuhS93PYC-Xr5oqur7IIWf}).
+
 \vspace{10mm}
@@
 \textbf{Acknowledgements.}
+
+I am grateful to the researchers who generated the datasets used in this
+study. I also thank the individuals who asked questions and commented on
+the preliminary version of this project, which was released as a YouTube
+playlist on the Riffomonas channel. This work was supported in part by
+funding from the National Institutes of Health (U01AI124255,
+P30DK034933, R01CA215574).
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Magurran2004}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-Oksanen2022}{}}%
 \CSLLeftMargin{5. }%
-\CSLRightInline{\textbf{Magurran AE}. 2004.
-\href{https://books.google.com/books?id=tUqzLSUzXxcC}{Measuring
-biological diversity}. Wiley.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Beule2020}{}}%
-\CSLLeftMargin{6. }%
-\CSLRightInline{\textbf{Beule L}, \textbf{Karlovsky P}. 2020.
-\href{https://doi.org/10.7717/peerj.9593}{Improved normalization of
-species count data in ecology by scaling with ranked subsampling (SRS):
-Application to microbial communities}. PeerJ \textbf{8}:e9593.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Chao2016}{}}%
-\CSLLeftMargin{7. }%
-\CSLRightInline{\textbf{Chao A}, \textbf{Chiu C-H}. 2016. Species
-richness: Estimation and comparison. Wiley StatsRef: statistics
-reference online \textbf{1}:26.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Chao2003}{}}%
-\CSLLeftMargin{8. }%
-\CSLRightInline{\textbf{Chao A}, \textbf{Shen T-J}. 2003. Nonparametric
-estimation of shannon's index of diversity when there are unseen species
-in sample. Environmental and ecological statistics
-\textbf{10}:429--443.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Willis2015}{}}%
-\CSLLeftMargin{9. }%
-\CSLRightInline{\textbf{Willis A}, \textbf{Bunge J}. 2015. Estimating
-diversity via frequency ratios. Biometrics \textbf{71}:1042--1049.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Oksanen2022}{}}%
-\CSLLeftMargin{10. }%
 \CSLRightInline{\textbf{Oksanen J}, \textbf{Simpson GL},
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-McArdle2001}{}}%
+\CSLLeftMargin{6. }%
+\CSLRightInline{\textbf{McArdle BH}, \textbf{Anderson MJ}. 2001.
+{FITTING} {MULTIVARIATE} {MODELS} {TO} {COMMUNITY} {DATA}: A {COMMENT}
+{ON} {DISTANCE}-{BASED} {REDUNDANCY} {ANALYSIS}. Ecology
+\textbf{82}:290--297.
+doi:\href{https://doi.org/10.1890/0012-9658(2001)082\%5B0290:fmmtcd\%5D2.0.co;2}{10.1890/0012-9658(2001)082{[}0290:fmmtcd{]}2.0.co;2}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Magurran2004}{}}%
+\CSLLeftMargin{7. }%
+\CSLRightInline{\textbf{Magurran AE}. 2004.
+\href{https://books.google.com/books?id=tUqzLSUzXxcC}{Measuring
+biological diversity}. Wiley.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Beule2020}{}}%
+\CSLLeftMargin{8. }%
+\CSLRightInline{\textbf{Beule L}, \textbf{Karlovsky P}. 2020.
+\href{https://doi.org/10.7717/peerj.9593}{Improved normalization of
+species count data in ecology by scaling with ranked subsampling (SRS):
+Application to microbial communities}. PeerJ \textbf{8}:e9593.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Paulson2013}{}}%
+\CSLLeftMargin{9. }%
+\CSLRightInline{\textbf{Paulson JN}, \textbf{Stine OC}, \textbf{Bravo
+HC}, \textbf{Pop M}. 2013. Differential abundance analysis for microbial
+marker-gene surveys. Nature Methods \textbf{10}:1200--1202.
+doi:\href{https://doi.org/10.1038/nmeth.2658}{10.1038/nmeth.2658}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Chao2016}{}}%
+\CSLLeftMargin{10. }%
+\CSLRightInline{\textbf{Chao A}, \textbf{Chiu C-H}. 2016. Species
+richness: Estimation and comparison. Wiley StatsRef: statistics
+reference online \textbf{1}:26.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Chao2003}{}}%
+\CSLLeftMargin{11. }%
+\CSLRightInline{\textbf{Chao A}, \textbf{Shen T-J}. 2003. Nonparametric
+estimation of shannon's index of diversity when there are unseen species
+in sample. Environmental and ecological statistics
+\textbf{10}:429--443.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Willis2015}{}}%
+\CSLLeftMargin{12. }%
+\CSLRightInline{\textbf{Willis A}, \textbf{Bunge J}. 2015. Estimating
+diversity via frequency ratios. Biometrics \textbf{71}:1042--1049.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Love2014}{}}%
+\CSLLeftMargin{13. }%
+\CSLRightInline{\textbf{Love MI}, \textbf{Huber W}, \textbf{Anders S}.
+2014. Moderated estimation of fold change and dispersion for {RNA}-seq
+data with {DESeq}2. Genome Biology \textbf{15}.
+doi:\href{https://doi.org/10.1186/s13059-014-0550-8}{10.1186/s13059-014-0550-8}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-McMurdie2014}{}}%
+\CSLLeftMargin{14. }%
+\CSLRightInline{\textbf{McMurdie PJ}, \textbf{Holmes S}. 2014. Waste
+not, want not: Why rarefying microbiome data is inadmissible. {PLoS}
+Computational Biology \textbf{10}:e1003531.
+doi:\href{https://doi.org/10.1371/journal.pcbi.1003531}{10.1371/journal.pcbi.1003531}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Palarea2015}{}}%
+\CSLLeftMargin{15. }%
+\CSLRightInline{\textbf{Palarea-Albaladejo J}, \textbf{Martin-Fernandez
+J}. 2015. zCompositions -- {R} package for multivariate imputation of
+left-censored data under a compositional approach. Chemometrics and
+Intelligent Laboratory Systems \textbf{143}:85--96.
+doi:\href{https://doi.org/10.1016/j.chemolab.2015.02.019}{10.1016/j.chemolab.2015.02.019}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Lin2020}{}}%
+\CSLLeftMargin{16. }%
+\CSLRightInline{\textbf{Lin H}, \textbf{Peddada SD}. 2020. Analysis of
+microbial compositions: A review of normalization and differential
+abundance analysis. npj Biofilms and Microbiomes \textbf{6}.
+doi:\href{https://doi.org/10.1038/s41522-020-00160-w}{10.1038/s41522-020-00160-w}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Costea2014}{}}%
+\CSLLeftMargin{17. }%
+\CSLRightInline{\textbf{Costea PI}, \textbf{Zeller G}, \textbf{Sunagawa
+S}, \textbf{Bork P}. 2014. A fair comparison. Nature Methods
+\textbf{11}:359--359.
+doi:\href{https://doi.org/10.1038/nmeth.2897}{10.1038/nmeth.2897}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Martino2019}{}}%
+\CSLLeftMargin{18. }%
+\CSLRightInline{\textbf{Martino C}, \textbf{Morton JT}, \textbf{Marotz
+CA}, \textbf{Thompson LR}, \textbf{Tripathi A}, \textbf{Knight R},
+\textbf{Zengler K}. 2019. A novel sparse compositional technique reveals
+microbial perturbations. {mSystems} \textbf{4}.
+doi:\href{https://doi.org/10.1128/msystems.00016-19}{10.1128/msystems.00016-19}.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-Li2015}{}}%
-\CSLLeftMargin{11. }%
+\CSLLeftMargin{19. }%
 \CSLRightInline{\textbf{Li Q}, \textbf{Heist EP}, \textbf{Moe LA}. 2015.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Baxter2016}{}}%
-\CSLLeftMargin{12. }%
+\CSLLeftMargin{20. }%
 \CSLRightInline{\textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Beall2015}{}}%
-\CSLLeftMargin{13. }%
+\CSLLeftMargin{21. }%
 \CSLRightInline{\textbf{Beall BFN}, \textbf{Twiss MR}, \textbf{Smith
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Henson2016}{}}%
-\CSLLeftMargin{14. }%
+\CSLLeftMargin{22. }%
 \CSLRightInline{\textbf{Henson MW}, \textbf{Pitre DM}, \textbf{Weckhorst
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Baxter2014}{}}%
-\CSLLeftMargin{15. }%
+\CSLLeftMargin{23. }%
 \CSLRightInline{\textbf{Baxter NT}, \textbf{Wan JJ}, \textbf{Schubert
@@
 \leavevmode\vadjust pre{\hypertarget{ref-LevyBooth2018}{}}%
-\CSLLeftMargin{16. }%
+\CSLLeftMargin{24. }%
 \CSLRightInline{\textbf{Levy-Booth DJ}, \textbf{Giesbrecht IJW},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Edwards2015}{}}%
-\CSLLeftMargin{17. }%
+\CSLLeftMargin{25. }%
 \CSLRightInline{\textbf{Edwards J}, \textbf{Johnson C},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Ettinger2017}{}}%
-\CSLLeftMargin{18. }%
+\CSLLeftMargin{26. }%
 \CSLRightInline{\textbf{Ettinger CL}, \textbf{Williams SL},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Graw2018}{}}%
-\CSLLeftMargin{19. }%
+\CSLLeftMargin{27. }%
 \CSLRightInline{\textbf{Graw MF}, \textbf{DAngelo G}, \textbf{Borchers
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Johnston2016}{}}%
-\CSLLeftMargin{20. }%
+\CSLLeftMargin{28. }%
 \CSLRightInline{\textbf{Johnston ER}, \textbf{Rodriguez-R LM},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Hassell2018}{}}%
-\CSLLeftMargin{21. }%
+\CSLLeftMargin{29. }%
 \CSLRightInline{\textbf{Hassell N}, \textbf{Tinker KA}, \textbf{Moore
@@
 \newpage
+
+\hypertarget{tables}{%
+\subsection{Tables}\label{tables}}
 
@@
 \endhead
-Bioethanol~(11) & 95 & 3,970,972 & 16,014 & 3,690\Hyphdash*356,027 &
+Bioethanol~(19) & 95 & 3,970,972 & 16,014 & 3,690\Hyphdash*356,027 &
 SRP055545 \\
-Human~(12) & 490 & 20,828,275 & 32,452 & 10,439\Hyphdash*422,904 &
+Human~(20) & 490 & 20,828,275 & 32,452 & 10,439\Hyphdash*422,904 &
 SRP062005 \\
-Lake~(13) & 52 & 3,145,486 & 69,205 & 15,135\Hyphdash*110,993 &
+Lake~(21) & 52 & 3,145,486 & 69,205 & 15,135\Hyphdash*110,993 &
 SRP050963 \\
-Marine~(14) & 7 & 1,484,068 & 213,091 & 132,895\Hyphdash*256,758 &
+Marine~(22) & 7 & 1,484,068 & 213,091 & 132,895\Hyphdash*256,758 &
 SRP068101 \\
@@
 SRP192323 \\
-Peromyscus~(15) & 111 & 1,545,288 & 12,393 & 4,454\Hyphdash*33,502 &
+Peromyscus~(23) & 111 & 1,545,288 & 12,393 & 4,454\Hyphdash*33,502 &
 SRP044050 \\
-Rainforest~(16) & 69 & 936,666 & 11,464 & 4,880\Hyphdash*37,403 &
+Rainforest~(24) & 69 & 936,666 & 11,464 & 4,880\Hyphdash*37,403 &
 ERP023747 \\
-Rice~(17) & 490 & 22,623,937 & 43,399 & 2,777\Hyphdash*192,200 &
+Rice~(25) & 490 & 22,623,937 & 43,399 & 2,777\Hyphdash*192,200 &
 SRP044745 \\
-Seagrass~(18) & 286 & 4,135,440 & 13,538 & 1,830\Hyphdash*45,076 &
+Seagrass~(26) & 286 & 4,135,440 & 13,538 & 1,830\Hyphdash*45,076 &
 SRP092441 \\
-Sediment~(19) & 58 & 1,151,389 & 17,606 & 7,686\Hyphdash*67,763 &
+Sediment~(27) & 58 & 1,151,389 & 17,606 & 7,686\Hyphdash*67,763 &
 SRP097192 \\
-Soil~(20) & 18 & 932,563 & 50,487 & 46,622\Hyphdash*58,935 &
+Soil~(28) & 18 & 932,563 & 50,487 & 46,622\Hyphdash*58,935 &
 ERP012016 \\
-Stream~(21) & 201 & 21,017,610 & 90,621 & 8,931\Hyphdash*394,419 &
+Stream~(29) & 201 & 21,017,610 & 90,621 & 8,931\Hyphdash*394,419 &
 SRP075852 \\
@@
 \subsection{Figures}\label{figures}}
-
-\newpage
 

### 2022-11-10 code/plot_power_cffect_model.R
Remove lines from Figures 4 and 5

@@ -97,8 +97,8 @@ alpha_composite <- map_dfr(alpha_summary_files, read_tsv, .id = "dataset") %>%
 alpha <- alpha_composite %>%
   ggplot(aes(x = metric_method, y = frac, color = dataset, shape = dataset)) +
     # geom_hline(yintercept = 0.05, size = 0.25, color = "gray") +
-    geom_line(aes(group = dataset), position = position_dodge(width = 0.3),
-              color = "gray", size = 0.1) +
+    # geom_line(aes(group = dataset), position = position_dodge(width = 0.3),
+    #           color = "gray", size = 0.1) +
     geom_point(position = position_dodge(width = 0.3), fill = "white") +
     facet_grid(. ~ class,
                scales = "free_x", space = "free_x",

### 2022-11-10 submission/manuscript.Rmd
Change coverage from fraction to percent

@@
 
-* Rarefy your data
-* Problems with recommended methods...
-  * Many recommended methods are borrowed from gened expression analysis
-  * Meaning of zeroes in data - structural vs. below limit of detection
-* Factors that determine what number of sequences to rarefy to
+Over the past decade, the question of whether to rarefy microbial community sequence data has been controversial.
+The analyses I presented here strongly indicate that rarefaction is necessary to control for uneven sampling effort when comparing communities based on alpha and beta diversity indices.
+Compared to all other methods, rarefaction removed the correlation between sequencing depth and alpha or beta diversity metrics when the sequencing depth varied by as much as 97-fold across samples.
+I showed that this correlation could lead falsely detecting differences between treatment groups if sampling depth and sequencing effort are confounded [e.g., @Morris2013].
+The correlation with sampling effort leads to an artificial increase in the variation between samples and a reduced power to detect true differences in alpha and beta diversity.
+For these reasons, rarefaction is a valuable tool to control for uneven sampling effort until improved statistical procedures are developed or it becomes possible to more evenly distribute sequencing effort across samples.
+
+The primary alternative to rarefaction that many advise is to estimate alpha diversity has been to use non-parametric or parametric methods using raw data and to then compare the estimates [@Willis2019].
+My data demonstrates that such approaches are limited for several reasons.
+First, it has been widely observed that non-parametric richness esimates such as ACE and Chao1 are sensitive to sampling effort.
+Therefore, these strategies do not, in practice, remove the effects of sampling effort.
+Second, parametric approaches, such as those implemented in the breakaway R package, generate confidence intervals that are likely to include the true richness and theoretically shrink with increased sampling effort.
+For most samples, the confidence intervals around the estimates are too wide to be meaningful, again leading to an inability to remvoe the effects of sampling effort.
+Third, it has become an increasingly common practice for researchers to remove sequences that are rare in a sample (e.g., those that appear once).
+Although that approach was not taken in this study, removing rare sequences would skew the distribtion of sequences and OTUs leading to a distortion of the measurement of alpha diversity [@Schloss2020].
+The effects of removing rare sequences would vary across samples depending on the number of sequences in each sample.
+One interesting result of this analysis was the demonstration that as metrics that depend less on rare taxa are used, the effect of uneven sampling effort was reduced.
+For example, richness counts a sequence appearing once as much as sequence appearing 1,000 times, while Shannon diversityindex places more emphasis on the more abundant sequence, and the inverse Simpson index even more.
+Although normalizing communities to a common number of sequences is also suggested (e.g., SRS normalization) to control for uneven sampling effort, the current analysis indicates that its performance does not meet that of rarefaction.
+For analysis of alpha diversity metrics, rarefaction is the most effective and consistent approach to control fo uneven sampling effort.
+
+Use of relative abundances, normalized counts, variance stabilizing transformations, and centered log ratios have each been recommended as more robust alternatives to rarefaction.
+Again, the only approach that consistently removed the effects of uneven sampling effort was rarefaction.
+Most of the recommendations borrow techniques from methods for identifying differential gene expression.
+Unfortunately, there is an important but underappreciated difference between gene expression and community data.
+This is the interpretation of unobserved data.
+For gene expression analysis in a single organism the lack of any sequencing data for a gene would indicate that its expression was below the limit of detection.
+Sequencing the same organism under multiple conditions would not lead to a seemingly unbouned number of genes in the organims.
+Rather, the number of genes has a definite limit that is knowable from the genome sequence.
+With microbiome data, an unobserved sequence could mean that the organism was present, but below the limit of detection or that the organism was missing.
+Because we have yet to exhaustively sample any community in the same way we have sequenced a single genome it is unreliable to impute the presence of all organisms.
+Yet, this is exactly what the variance stabilization transformation and most CLR techniques do.
+This analysis has demonstrated a clear correlation between distances calculated by these methods and sequencing effort.
+This result is at odds with the claims by others that the distances are scale invariant [@Martino2019].
+Again, rarefaction is the most effective and consistent approach to control for uneven sampling effort when calculating beta diversity metrics.
+
+Two common critique of rarefaction is that the approach ommits valid data and that the selection of the value to rarefy the sequence data to is arbitrary [@McMurdie2014].
+I disagree with both critiques.
+All of the data are used to calculate the mean value of the rarefied metrics; none of it is excluded.
+When the dataset is subsampled, every sequence has a random chance of being included in the calculated metric.
+When that subsampling is repeated a large number of times (e.g., 100 or 1000) the risk of oversampling rare taxa is mititgated.
+As for the second critism, I would resist the claim that the selection of the rarefaction threshold is arbitrary.
+In practice, there is a tradeoff between sampling breadth and depth.
+Greater breadth will increase the statistical power to compare treatment groups and greater depth will increase the resolution to describe the communities.
+My process for picking a break involves looking for a natural break in the distribution of the number of sequences.
+For example, the Lake dataset used in this study had a clear break around 10,000 sequences.
+I would also consider what samples are below any break that I select.
+If there were critical samples below the break I would either reduce the break point or I would generate more sequences from those samples.
+Furthermore, regardless of the break point one selects, they could repeat an analysis at multiple thresholds to assess the tradeoff between statistical power and resolution.
+As shown in my analysis of Good's coverage values, most studies obtain an ample level of coverage and would need to increase their sampling depth by 10 fold to increase the coverage by several percentage points.
+I contend that for most studies increased sequencing effort should be applied to improve sampling breadth and statistica power over depth and resolution.
+
 * Need better methods of pooling libraries that result in more even distribution of sequences across samples
+* We don't have true absolute numbers
 * Rarefy your data

### 2022-11-10 submission/manuscript.tex
Change coverage from fraction to percent

@@
 
+Over the past decade, the question of whether to rarefy microbial
+community sequence data has been controversial. The analyses I presented
+here strongly indicate that rarefaction is necessary to control for
+uneven sampling effort when comparing communities based on alpha and
+beta diversity indices. Compared to all other methods, rarefaction
+removed the correlation between sequencing depth and alpha or beta
+diversity metrics when the sequencing depth varied by as much as 97-fold
+across samples. I showed that this correlation could lead falsely
+detecting differences between treatment groups if sampling depth and
+sequencing effort are confounded (e.g., 1). The correlation with
+sampling effort leads to an artificial increase in the variation between
+samples and a reduced power to detect true differences in alpha and beta
+diversity. For these reasons, rarefaction is a valuable tool to control
+for uneven sampling effort until improved statistical procedures are
+developed or it becomes possible to more evenly distribute sequencing
+effort across samples.
+
+The primary alternative to rarefaction that many advise is to estimate
+alpha diversity has been to use non-parametric or parametric methods
+using raw data and to then compare the estimates (2). My data
+demonstrates that such approaches are limited for several reasons.
+First, it has been widely observed that non-parametric richness esimates
+such as ACE and Chao1 are sensitive to sampling effort. Therefore, these
+strategies do not, in practice, remove the effects of sampling effort.
+Second, parametric approaches, such as those implemented in the
+breakaway R package, generate confidence intervals that are likely to
+include the true richness and theoretically shrink with increased
+sampling effort. For most samples, the confidence intervals around the
+estimates are too wide to be meaningful, again leading to an inability
+to remvoe the effects of sampling effort. Third, it has become an
+increasingly common practice for researchers to remove sequences that
+are rare in a sample (e.g., those that appear once). Although that
+approach was not taken in this study, removing rare sequences would skew
+the distribtion of sequences and OTUs leading to a distortion of the
+measurement of alpha diversity (3). The effects of removing rare
+sequences would vary across samples depending on the number of sequences
+in each sample. One interesting result of this analysis was the
+demonstration that as metrics that depend less on rare taxa are used,
+the effect of uneven sampling effort was reduced. For example, richness
+counts a sequence appearing once as much as sequence appearing 1,000
+times, while Shannon diversityindex places more emphasis on the more
+abundant sequence, and the inverse Simpson index even more. Although
+normalizing communities to a common number of sequences is also
+suggested (e.g., SRS normalization) to control for uneven sampling
+effort, the current analysis indicates that its performance does not
+meet that of rarefaction. For analysis of alpha diversity metrics,
+rarefaction is the most effective and consistent approach to control fo
+uneven sampling effort.
+
+Use of relative abundances, normalized counts, variance stabilizing
+transformations, and centered log ratios have each been recommended as
+more robust alternatives to rarefaction. Again, the only approach that
+consistently removed the effects of uneven sampling effort was
+rarefaction. Most of the recommendations borrow techniques from methods
+for identifying differential gene expression. Unfortunately, there is an
+important but underappreciated difference between gene expression and
+community data. This is the interpretation of unobserved data. For gene
+expression analysis in a single organism the lack of any sequencing data
+for a gene would indicate that its expression was below the limit of
+detection. Sequencing the same organism under multiple conditions would
+not lead to a seemingly unbouned number of genes in the organims.
+Rather, the number of genes has a definite limit that is knowable from
+the genome sequence. With microbiome data, an unobserved sequence could
+mean that the organism was present, but below the limit of detection or
+that the organism was missing. Because we have yet to exhaustively
+sample any community in the same way we have sequenced a single genome
+it is unreliable to impute the presence of all organisms. Yet, this is
+exactly what the variance stabilization transformation and most CLR
+techniques do. This analysis has demonstrated a clear correlation
+between distances calculated by these methods and sequencing effort.
+This result is at odds with the claims by others that the distances are
+scale invariant (4). Again, rarefaction is the most effective and
+consistent approach to control for uneven sampling effort when
+calculating beta diversity metrics.
+
+Two common critique of rarefaction is that the approach ommits valid
+data and that the selection of the value to rarefy the sequence data to
+is arbitrary (5). I disagree with both critiques. All of the data are
+used to calculate the mean value of the rarefied metrics; none of it is
+excluded. When the dataset is subsampled, every sequence has a random
+chance of being included in the calculated metric. When that subsampling
+is repeated a large number of times (e.g., 100 or 1000) the risk of
+oversampling rare taxa is mititgated. As for the second critism, I would
+resist the claim that the selection of the rarefaction threshold is
+arbitrary. In practice, there is a tradeoff between sampling breadth and
+depth. Greater breadth will increase the statistical power to compare
+treatment groups and greater depth will increase the resolution to
+describe the communities. My process for picking a break involves
+looking for a natural break in the distribution of the number of
+sequences. For example, the Lake dataset used in this study had a clear
+break around 10,000 sequences. I would also consider what samples are
+below any break that I select. If there were critical samples below the
+break I would either reduce the break point or I would generate more
+sequences from those samples. Furthermore, regardless of the break point
+one selects, they could repeat an analysis at multiple thresholds to
+assess the tradeoff between statistical power and resolution. As shown
+in my analysis of Good's coverage values, most studies obtain an ample
+level of coverage and would need to increase their sampling depth by 10
+fold to increase the coverage by several percentage points. I contend
+that for most studies increased sequencing effort should be applied to
+improve sampling breadth and statistica power over depth and resolution.
+
 \begin{itemize}
@@
 \item
-  Rarefy your data
-\item
-  Problems with recommended methods\ldots{}
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Many recommended methods are borrowed from gened expression analysis
-  \item
-    Meaning of zeroes in data - structural vs.~below limit of detection
-  \end{itemize}
-\item
-  Factors that determine what number of sequences to rarefy to
-\item
   Need better methods of pooling libraries that result in more even
   distribution of sequences across samples
+\item
+  We don't have true absolute numbers
 \item
@@
 region resulting in high quality contigs with a low sequencing error
-rate (1). These data were processed through a standardized sequence
+rate (6). These data were processed through a standardized sequence
 curation pipeline to generate operational taxonomic units (OTUs) using
-the mothur software package (1, 2). OTUs were identified using the
+the mothur software package (6, 7). OTUs were identified using the
 OptiClust algorithm to cluster sequences together that were not more
-than 3\% different from each other (3).
+than 3\% different from each other (8).
 
@@
 the total number of sequences in each OTU. This is a similar approach to
-that of the IS algorithm described by Ulrich and Gotelli (4). Because
+that of the IS algorithm described by Ulrich and Gotelli (9). Because
 the construction of the null models was a stochastic process, 100
@@
 diversity metrics were performed using the adonis2 function from the
-vegan (v.2.6.2) R package (5). The adonis2 function implements a
+vegan (v.2.6.2) R package (10). The adonis2 function implements a
 non-parametric multivariate analysis of variance using distances
-matrices as described by McArdle and Anderson (6). Throughout this study
-I used 0.05 as the threshold for asssessing statistical significance.
+matrices as described by McArdle and Anderson (11). Throughout this
+study I used 0.05 as the threshold for asssessing statistical
+significance.
 
@@
 indeices. Raw OTU counts were used as input to calculate sample richness
-and Shannon and inverse Simpson diversity using mothur (2, 7). Shannon
+and Shannon and inverse Simpson diversity using mothur (7, 12). Shannon
 diversity was calculated as
@@
 normalize OTU counts to the size of the smallest sample in each dataset
-using the SRS R package (v.0.2.3)(8). Normalized OTU counts were used to
-calculate sample richness and Shannon and inverse Simpson diversity
+using the SRS R package (v.0.2.3)(13). Normalized OTU counts were used
+to calculate sample richness and Shannon and inverse Simpson diversity
 values using mothur. Data normalized by cumulative sum scaling (CSS)
@@
 abundances of the features do not change with the normalization
-procedure (9). The non-parametric bias-corrected Chao1 and ACE richness
-estimators (10) and a non-parametric estimator of the Shannon diversity
-(11) were calculated using raw OTU counts with mothur. Parametric
+procedure (14). The non-parametric bias-corrected Chao1 and ACE richness
+estimators (15) and a non-parametric estimator of the Shannon diversity
+(16) were calculated using raw OTU counts with mothur. Parametric
 estimates of sample richness were calculated using the breakaway (BA) R
-package (v.4.7.9)(12). The current analysis reports both the results
+package (v.4.7.9)(17). The current analysis reports both the results
 from running default model selection procedure and the Poisson model.
@@
 dissimilarity indices using the vegdist function from the vegan R
-package (v.2.6.2)(5). The Jaccard index was calculated as
+package (v.2.6.2)(10). The Jaccard index was calculated as
 
@@
 sum scaling (CSS) were used to normalize raw OTU counts using the SRS
-(v.0.2.3) and metagenomeSeq (v.1.36.0) R packages, respectively (8, 9).
-The normalized counts were then used to calculate Jaccard and
+(v.0.2.3) and metagenomeSeq (v.1.36.0) R packages, respectively (13,
+14). The normalized counts were then used to calculate Jaccard and
 Bray-Curtis dissimilarity indices. Finally, the variance-stabilization
 transformation (VST) as implemented in the DESeq2 (v.1.34.0) R package
-was used to normalize the data as described by McMurdie and Holmes (13,
-14). Because the VST approach generates negative values, which are
+was used to normalize the data as described by McMurdie and Holmes (5,
+18). Because the VST approach generates negative values, which are
 incompatible with calculating Jaccard and Bray-Curtis dissimilarity
@@
 based on the observed data using the zCompositions (v.1.4.0.1) R package
-(15). The second, One CLR, added a pseudocount of 1 to the abundance of
-all OTUs (9, 16). The third, Nudge CLR, added a pseudocount of 1 divided
-by the total number of sequences in a sample to each OTU in the sample
-(16, 17). The final approach, Robust CLR, removed unobserved OTUs prior
-to calcualting the CLR (18).
+(19). The second, One CLR, added a pseudocount of 1 to the abundance of
+all OTUs (14, 20). The third, Nudge CLR, added a pseudocount of 1
+divided by the total number of sequences in a sample to each OTU in the
+sample (20, 21). The final approach, Robust CLR, removed unobserved OTUs
+prior to calcualting the CLR (4).
 
@@
 \begin{CSLReferences}{0}{1}
+\leavevmode\vadjust pre{\hypertarget{ref-Morris2013}{}}%
+\CSLLeftMargin{1. }%
+\CSLRightInline{\textbf{Morris A}, \textbf{Beck JM}, \textbf{Schloss
+PD}, \textbf{Campbell TB}, \textbf{Crothers K}, \textbf{Curtis JL},
+\textbf{Flores SC}, \textbf{Fontenot AP}, \textbf{Ghedin E},
+\textbf{Huang L}, \textbf{Jablonski K}, \textbf{Kleerup E},
+\textbf{Lynch SV}, \textbf{Sodergren E}, \textbf{Twigg H}, \textbf{Young
+VB}, \textbf{Bassis CM}, \textbf{Venkataraman A}, \textbf{Schmidt TM},
+\textbf{Weinstock GM}. 2013. Comparison of the respiratory microbiome in
+healthy nonsmokers and smokers. American Journal of Respiratory and
+Critical Care Medicine \textbf{187}:1067--1075.
+doi:\href{https://doi.org/10.1164/rccm.201210-1913oc}{10.1164/rccm.201210-1913oc}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Willis2019}{}}%
+\CSLLeftMargin{2. }%
+\CSLRightInline{\textbf{Willis AD}. 2019. Rarefaction, alpha diversity,
+and statistics. Frontiers in Microbiology \textbf{10}.
+doi:\href{https://doi.org/10.3389/fmicb.2019.02407}{10.3389/fmicb.2019.02407}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Schloss2020}{}}%
+\CSLLeftMargin{3. }%
+\CSLRightInline{\textbf{Schloss PD}. 2020. Removal of rare amplicon
+sequence variants from 16S {rRNA} gene sequence surveys biases the
+interpretation of community structure data.
+doi:\href{https://doi.org/10.1101/2020.12.11.422279}{10.1101/2020.12.11.422279}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Martino2019}{}}%
+\CSLLeftMargin{4. }%
+\CSLRightInline{\textbf{Martino C}, \textbf{Morton JT}, \textbf{Marotz
+CA}, \textbf{Thompson LR}, \textbf{Tripathi A}, \textbf{Knight R},
+\textbf{Zengler K}. 2019. A novel sparse compositional technique reveals
+microbial perturbations. {mSystems} \textbf{4}.
+doi:\href{https://doi.org/10.1128/msystems.00016-19}{10.1128/msystems.00016-19}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-McMurdie2014}{}}%
+\CSLLeftMargin{5. }%
+\CSLRightInline{\textbf{McMurdie PJ}, \textbf{Holmes S}. 2014. Waste
+not, want not: Why rarefying microbiome data is inadmissible. {PLoS}
+Computational Biology \textbf{10}:e1003531.
+doi:\href{https://doi.org/10.1371/journal.pcbi.1003531}{10.1371/journal.pcbi.1003531}.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-Kozich2013}{}}%
-\CSLLeftMargin{1. }%
+\CSLLeftMargin{6. }%
 \CSLRightInline{\textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Schloss2009}{}}%
-\CSLLeftMargin{2. }%
+\CSLLeftMargin{7. }%
 \CSLRightInline{\textbf{Schloss PD}, \textbf{Westcott SL},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Westcott2017}{}}%
-\CSLLeftMargin{3. }%
+\CSLLeftMargin{8. }%
 \CSLRightInline{\textbf{Westcott SL}, \textbf{Schloss PD}. 2017.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Ulrich2010}{}}%
-\CSLLeftMargin{4. }%
+\CSLLeftMargin{9. }%
 \CSLRightInline{\textbf{Ulrich W}, \textbf{Gotelli NJ}. 2010. Null model
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Oksanen2022}{}}%
-\CSLLeftMargin{5. }%
+\CSLLeftMargin{10. }%
 \CSLRightInline{\textbf{Oksanen J}, \textbf{Simpson GL},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-McArdle2001}{}}%
-\CSLLeftMargin{6. }%
+\CSLLeftMargin{11. }%
 \CSLRightInline{\textbf{McArdle BH}, \textbf{Anderson MJ}. 2001.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Magurran2004}{}}%
-\CSLLeftMargin{7. }%
+\CSLLeftMargin{12. }%
 \CSLRightInline{\textbf{Magurran AE}. 2004.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Beule2020}{}}%
-\CSLLeftMargin{8. }%
+\CSLLeftMargin{13. }%
 \CSLRightInline{\textbf{Beule L}, \textbf{Karlovsky P}. 2020.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Paulson2013}{}}%
-\CSLLeftMargin{9. }%
+\CSLLeftMargin{14. }%
 \CSLRightInline{\textbf{Paulson JN}, \textbf{Stine OC}, \textbf{Bravo
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Chao2016}{}}%
-\CSLLeftMargin{10. }%
+\CSLLeftMargin{15. }%
 \CSLRightInline{\textbf{Chao A}, \textbf{Chiu C-H}. 2016. Species
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Chao2003}{}}%
-\CSLLeftMargin{11. }%
+\CSLLeftMargin{16. }%
 \CSLRightInline{\textbf{Chao A}, \textbf{Shen T-J}. 2003. Nonparametric
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Willis2015}{}}%
-\CSLLeftMargin{12. }%
+\CSLLeftMargin{17. }%
 \CSLRightInline{\textbf{Willis A}, \textbf{Bunge J}. 2015. Estimating
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Love2014}{}}%
-\CSLLeftMargin{13. }%
+\CSLLeftMargin{18. }%
 \CSLRightInline{\textbf{Love MI}, \textbf{Huber W}, \textbf{Anders S}.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-McMurdie2014}{}}%
-\CSLLeftMargin{14. }%
-\CSLRightInline{\textbf{McMurdie PJ}, \textbf{Holmes S}. 2014. Waste
-not, want not: Why rarefying microbiome data is inadmissible. {PLoS}
-Computational Biology \textbf{10}:e1003531.
-doi:\href{https://doi.org/10.1371/journal.pcbi.1003531}{10.1371/journal.pcbi.1003531}.}
-
 \leavevmode\vadjust pre{\hypertarget{ref-Palarea2015}{}}%
-\CSLLeftMargin{15. }%
+\CSLLeftMargin{19. }%
 \CSLRightInline{\textbf{Palarea-Albaladejo J}, \textbf{Martin-Fernandez
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Lin2020}{}}%
-\CSLLeftMargin{16. }%
+\CSLLeftMargin{20. }%
 \CSLRightInline{\textbf{Lin H}, \textbf{Peddada SD}. 2020. Analysis of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Costea2014}{}}%
-\CSLLeftMargin{17. }%
+\CSLLeftMargin{21. }%
 \CSLRightInline{\textbf{Costea PI}, \textbf{Zeller G}, \textbf{Sunagawa
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Martino2019}{}}%
-\CSLLeftMargin{18. }%
-\CSLRightInline{\textbf{Martino C}, \textbf{Morton JT}, \textbf{Marotz
-CA}, \textbf{Thompson LR}, \textbf{Tripathi A}, \textbf{Knight R},
-\textbf{Zengler K}. 2019. A novel sparse compositional technique reveals
-microbial perturbations. {mSystems} \textbf{4}.
-doi:\href{https://doi.org/10.1128/msystems.00016-19}{10.1128/msystems.00016-19}.}
-
 \leavevmode\vadjust pre{\hypertarget{ref-Li2015}{}}%
-\CSLLeftMargin{19. }%
+\CSLLeftMargin{22. }%
 \CSLRightInline{\textbf{Li Q}, \textbf{Heist EP}, \textbf{Moe LA}. 2015.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Baxter2016}{}}%
-\CSLLeftMargin{20. }%
+\CSLLeftMargin{23. }%
 \CSLRightInline{\textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Beall2015}{}}%
-\CSLLeftMargin{21. }%
+\CSLLeftMargin{24. }%
 \CSLRightInline{\textbf{Beall BFN}, \textbf{Twiss MR}, \textbf{Smith
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Henson2016}{}}%
-\CSLLeftMargin{22. }%
+\CSLLeftMargin{25. }%
 \CSLRightInline{\textbf{Henson MW}, \textbf{Pitre DM}, \textbf{Weckhorst
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Baxter2014}{}}%
-\CSLLeftMargin{23. }%
+\CSLLeftMargin{26. }%
 \CSLRightInline{\textbf{Baxter NT}, \textbf{Wan JJ}, \textbf{Schubert
@@
 \leavevmode\vadjust pre{\hypertarget{ref-LevyBooth2018}{}}%
-\CSLLeftMargin{24. }%
+\CSLLeftMargin{27. }%
 \CSLRightInline{\textbf{Levy-Booth DJ}, \textbf{Giesbrecht IJW},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Edwards2015}{}}%
-\CSLLeftMargin{25. }%
+\CSLLeftMargin{28. }%
 \CSLRightInline{\textbf{Edwards J}, \textbf{Johnson C},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Ettinger2017}{}}%
-\CSLLeftMargin{26. }%
+\CSLLeftMargin{29. }%
 \CSLRightInline{\textbf{Ettinger CL}, \textbf{Williams SL},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Graw2018}{}}%
-\CSLLeftMargin{27. }%
+\CSLLeftMargin{30. }%
 \CSLRightInline{\textbf{Graw MF}, \textbf{DAngelo G}, \textbf{Borchers
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Johnston2016}{}}%
-\CSLLeftMargin{28. }%
+\CSLLeftMargin{31. }%
 \CSLRightInline{\textbf{Johnston ER}, \textbf{Rodriguez-R LM},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Hassell2018}{}}%
-\CSLLeftMargin{29. }%
+\CSLLeftMargin{32. }%
 \CSLRightInline{\textbf{Hassell N}, \textbf{Tinker KA}, \textbf{Moore
@@
 \endhead
-Bioethanol~(19) & 95 & 3,970,972 & 16,014 & 3,690\Hyphdash*356,027 &
+Bioethanol~(22) & 95 & 3,970,972 & 16,014 & 3,690\Hyphdash*356,027 &
 SRP055545 \\
-Human~(20) & 490 & 20,828,275 & 32,452 & 10,439\Hyphdash*422,904 &
+Human~(23) & 490 & 20,828,275 & 32,452 & 10,439\Hyphdash*422,904 &
 SRP062005 \\
-Lake~(21) & 52 & 3,145,486 & 69,205 & 15,135\Hyphdash*110,993 &
+Lake~(24) & 52 & 3,145,486 & 69,205 & 15,135\Hyphdash*110,993 &
 SRP050963 \\
-Marine~(22) & 7 & 1,484,068 & 213,091 & 132,895\Hyphdash*256,758 &
+Marine~(25) & 7 & 1,484,068 & 213,091 & 132,895\Hyphdash*256,758 &
 SRP068101 \\
-Mice~(1) & 348 & 2,785,641 & 6,426 & 1,804\Hyphdash*30,311 &
+Mice~(6) & 348 & 2,785,641 & 6,426 & 1,804\Hyphdash*30,311 &
 SRP192323 \\
-Peromyscus~(23) & 111 & 1,545,288 & 12,393 & 4,454\Hyphdash*33,502 &
+Peromyscus~(26) & 111 & 1,545,288 & 12,393 & 4,454\Hyphdash*33,502 &
 SRP044050 \\
-Rainforest~(24) & 69 & 936,666 & 11,464 & 4,880\Hyphdash*37,403 &
+Rainforest~(27) & 69 & 936,666 & 11,464 & 4,880\Hyphdash*37,403 &
 ERP023747 \\
-Rice~(25) & 490 & 22,623,937 & 43,399 & 2,777\Hyphdash*192,200 &
+Rice~(28) & 490 & 22,623,937 & 43,399 & 2,777\Hyphdash*192,200 &
 SRP044745 \\
-Seagrass~(26) & 286 & 4,135,440 & 13,538 & 1,830\Hyphdash*45,076 &
+Seagrass~(29) & 286 & 4,135,440 & 13,538 & 1,830\Hyphdash*45,076 &
 SRP092441 \\
-Sediment~(27) & 58 & 1,151,389 & 17,606 & 7,686\Hyphdash*67,763 &
+Sediment~(30) & 58 & 1,151,389 & 17,606 & 7,686\Hyphdash*67,763 &
 SRP097192 \\
-Soil~(28) & 18 & 932,563 & 50,487 & 46,622\Hyphdash*58,935 &
+Soil~(31) & 18 & 932,563 & 50,487 & 46,622\Hyphdash*58,935 &
 ERP012016 \\
-Stream~(29) & 201 & 21,017,610 & 90,621 & 8,931\Hyphdash*394,419 &
+Stream~(32) & 201 & 21,017,610 & 90,621 & 8,931\Hyphdash*394,419 &
 SRP075852 \\

### 2022-11-10 code/plot_coverage.R

@@ -47,6 +47,7 @@ cor_plot <- obs_cor %>%
   scale_x_log10(breaks = c(325, 1e3, 1e4, 1e5, 1e6),
                 labels = c("R", "10^3^", "10^4^", "10^5^", "10^6^"),
                 limits = c(2e2, 1e6)) +
+  scale_y_continuous(breaks = seq(0, 1, 0.2), labels = seq(0, 100, 20)) +
   scale_color_manual(name = NULL,
                     values = rep(c('#1b9e77','#d95f02','#7570b3'), 4),
                     breaks = pretty_datasets$plain,
@@ -61,7 +62,7 @@ cor_plot <- obs_cor %>%
                   ) +
 
   facet_wrap(~dataset, labeller = labeller(.default = capitalize)) +
-  labs(y = "Good's coverage",
+  labs(y = "Good's coverage (%)",
        x = "Number of sequences per sample") +
   theme(
     axis.text.x = element_markdown(),

### 2022-11-11 submission/manuscript.Rmd
First full version of Discussion

@@
 
-Over the past decade, the question of whether to rarefy microbial community sequence data has been controversial.
+Over the past decade, the question of whether to rarefy microbial community sequence data has become controversial.
 The analyses I presented here strongly indicate that rarefaction is necessary to control for uneven sampling effort when comparing communities based on alpha and beta diversity indices.
@@
 When the dataset is subsampled, every sequence has a random chance of being included in the calculated metric.
-When that subsampling is repeated a large number of times (e.g., 100 or 1000) the risk of oversampling rare taxa is mititgated.
+When that subsampling is repeated a large number of times (e.g., 100 or 1000) the risks of ignoring or oversampling rare taxa are mititgated.
+Furthermore, it is curious that the study making the original critique removed sequences that did not appear in at least three samples or had an abundance of one in any sample.
+A parallel analysis to this study has demonstrated that many of these sequences are likely valid and that removal of rare sequences can bias alpha and beta diversity metrics and reduce statistical power [@Schloss2020].
 As for the second critism, I would resist the claim that the selection of the rarefaction threshold is arbitrary.
@@
 As shown in my analysis of Good's coverage values, most studies obtain an ample level of coverage and would need to increase their sampling depth by 10 fold to increase the coverage by several percentage points.
-I contend that for most studies increased sequencing effort should be applied to improve sampling breadth and statistica power over depth and resolution.
+I contend that for most studies increased sequencing effort should be applied to improve sampling breadth and statistical power over depth and resolution.
 
-* Need better methods of pooling libraries that result in more even distribution of sequences across samples
-* We don't have true absolute numbers
-* Rarefy your data
+The up to 100-fold difference in sample sizes is an unfortunate byproduct of how seqencing libraries are constructed.
+Researchers perform separate PCRs for each sample with unique index (aka, barcode) sequencs that all them to later identify which sample each sequence belongs.
+When the PCRs are pooled efforts are taken to pool the fragments in equimolar ratios.
+Researchers use one of two approaches.
+First, they often will quantify the concentration of DNA from each PCR and then pool DNA in the desired amounts.
+Alternatively, they may use normalization plates where each well can hybridize a uniform amount of DNA that is then eluted and pooled.
+Clearly, both approaches have limitations that reduce the ability to truly achieve equimolar mixture.
+In addition, for some samples it is common to co-amplify non-specific DNAs which may add to the challenges of achieving equimolar mixtures of the desired amplicons [@Morris2013].
+Regardless, it is clear that better strategies are needed to reduce the variation in the number of sequences generated for each sample.
+
+All simulations have weaknesses and should be interpreted with caution.
+However, the simulated communites generated and analyzed in this study had the advantage of being designed with known properties including the alpha and beta diversity and the their differences between treatment groups.
+For now, is is perfectly admissible and proper to rarefy microbial community data or risk reaching unwarrented conclusions.
 
@@
 
-**Null treatment model.** I created an unbiased and biased treatment model.
+**Null treatment models.** I created an unbiased and biased treatment model.
 In the unbiased model, samples were randomly assigned to one of two treatment groups.

### 2022-11-11 submission/manuscript.tex
First full version of Discussion

@@
 Over the past decade, the question of whether to rarefy microbial
-community sequence data has been controversial. The analyses I presented
-here strongly indicate that rarefaction is necessary to control for
-uneven sampling effort when comparing communities based on alpha and
-beta diversity indices. Compared to all other methods, rarefaction
-removed the correlation between sequencing depth and alpha or beta
-diversity metrics when the sequencing depth varied by as much as 97-fold
-across samples. I showed that this correlation could lead falsely
-detecting differences between treatment groups if sampling depth and
-sequencing effort are confounded (e.g., 1). The correlation with
+community sequence data has become controversial. The analyses I
+presented here strongly indicate that rarefaction is necessary to
+control for uneven sampling effort when comparing communities based on
+alpha and beta diversity indices. Compared to all other methods,
+rarefaction removed the correlation between sequencing depth and alpha
+or beta diversity metrics when the sequencing depth varied by as much as
+97-fold across samples. I showed that this correlation could lead
+falsely detecting differences between treatment groups if sampling depth
+and sequencing effort are confounded (e.g., 1). The correlation with
 sampling effort leads to an artificial increase in the variation between
@@
 chance of being included in the calculated metric. When that subsampling
-is repeated a large number of times (e.g., 100 or 1000) the risk of
-oversampling rare taxa is mititgated. As for the second critism, I would
-resist the claim that the selection of the rarefaction threshold is
-arbitrary. In practice, there is a tradeoff between sampling breadth and
-depth. Greater breadth will increase the statistical power to compare
-treatment groups and greater depth will increase the resolution to
-describe the communities. My process for picking a break involves
-looking for a natural break in the distribution of the number of
-sequences. For example, the Lake dataset used in this study had a clear
-break around 10,000 sequences. I would also consider what samples are
-below any break that I select. If there were critical samples below the
-break I would either reduce the break point or I would generate more
-sequences from those samples. Furthermore, regardless of the break point
-one selects, they could repeat an analysis at multiple thresholds to
-assess the tradeoff between statistical power and resolution. As shown
-in my analysis of Good's coverage values, most studies obtain an ample
-level of coverage and would need to increase their sampling depth by 10
-fold to increase the coverage by several percentage points. I contend
-that for most studies increased sequencing effort should be applied to
-improve sampling breadth and statistica power over depth and resolution.
-
-\begin{itemize}
-\tightlist
-\item
-  Need better methods of pooling libraries that result in more even
-  distribution of sequences across samples
-\item
-  We don't have true absolute numbers
-\item
-  Rarefy your data
-\end{itemize}
+is repeated a large number of times (e.g., 100 or 1000) the risks of
+ignoring or oversampling rare taxa are mititgated. Furthermore, it is
+curious that the study making the original critique removed sequences
+that did not appear in at least three samples or had an abundance of one
+in any sample. A parallel analysis to this study has demonstrated that
+many of these sequences are likely valid and that removal of rare
+sequences can bias alpha and beta diversity metrics and reduce
+statistical power (3). As for the second critism, I would resist the
+claim that the selection of the rarefaction threshold is arbitrary. In
+practice, there is a tradeoff between sampling breadth and depth.
+Greater breadth will increase the statistical power to compare treatment
+groups and greater depth will increase the resolution to describe the
+communities. My process for picking a break involves looking for a
+natural break in the distribution of the number of sequences. For
+example, the Lake dataset used in this study had a clear break around
+10,000 sequences. I would also consider what samples are below any break
+that I select. If there were critical samples below the break I would
+either reduce the break point or I would generate more sequences from
+those samples. Furthermore, regardless of the break point one selects,
+they could repeat an analysis at multiple thresholds to assess the
+tradeoff between statistical power and resolution. As shown in my
+analysis of Good's coverage values, most studies obtain an ample level
+of coverage and would need to increase their sampling depth by 10 fold
+to increase the coverage by several percentage points. I contend that
+for most studies increased sequencing effort should be applied to
+improve sampling breadth and statistical power over depth and
+resolution.
+
+The up to 100-fold difference in sample sizes is an unfortunate
+byproduct of how seqencing libraries are constructed. Researchers
+perform separate PCRs for each sample with unique index (aka, barcode)
+sequencs that all them to later identify which sample each sequence
+belongs. When the PCRs are pooled efforts are taken to pool the
+fragments in equimolar ratios. Researchers use one of two approaches.
+First, they often will quantify the concentration of DNA from each PCR
+and then pool DNA in the desired amounts. Alternatively, they may use
+normalization plates where each well can hybridize a uniform amount of
+DNA that is then eluted and pooled. Clearly, both approaches have
+limitations that reduce the ability to truly achieve equimolar mixture.
+In addition, for some samples it is common to co-amplify non-specific
+DNAs which may add to the challenges of achieving equimolar mixtures of
+the desired amplicons (1). Regardless, it is clear that better
+strategies are needed to reduce the variation in the number of sequences
+generated for each sample.
+
+All simulations have weaknesses and should be interpreted with caution.
+However, the simulated communites generated and analyzed in this study
+had the advantage of being designed with known properties including the
+alpha and beta diversity and the their differences between treatment
+groups. For now, is is perfectly admissible and proper to rarefy
+microbial community data or risk reaching unwarrented conclusions.
 
@@
 
-\textbf{Null treatment model.} I created an unbiased and biased
+\textbf{Null treatment models.} I created an unbiased and biased
 treatment model. In the unbiased model, samples were randomly assigned

### 2022-11-11 code/plot_cluster_analysis.R
Replicate cluster analysis

@@ -1,0 +1,138 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(tidyverse)
+library(glue)
+library(patchwork)
+
+capitalize <- function(string) {
+  substr(string, 1, 1) <- toupper(substr(string, 1, 1))
+  string
+}
+
+
+datasets <- c("bioethanol", "human", "lake", "marine", "mice", "peromyscus",
+              "rainforest", "rice", "seagrass", "sediment", "soil", "stream")
+
+pretty_datasets <- tibble(
+  plain = datasets,
+  pretty = capitalize(plain)
+)
+
+beta_metrics <- c(
+  "rare_jaccard",
+  "raw_jaccard",
+  "relabund_jaccard",
+  "srs_jaccard",
+  "metagenomeseq_jaccard",
+
+  "rare_bray",
+  "raw_bray",
+  "relabund_bray",
+  "srs_bray",
+  "metagenomeseq_bray",
+
+  "rare_euclidean",
+  "raw_euclidean",
+  "relabund_euclidean",
+  "rclr_euclidean",
+  "oclr_euclidean",
+  "nclr_euclidean",
+  "zclr_euclidean",
+  "deseq2_euclidean"
+)
+
+beta_labels <- c(
+  "Rarefied",
+  "Raw",
+  "Rel. abundance",
+  "SRS Normalized",
+  "CSS Normalized",
+
+  "Rarefied",
+  "Raw",
+  "Rel. abundance",
+  "SRS Normalized",
+  "CSS Normalized",
+
+  "Rarefied",
+  "Raw",
+  "Rel. abundance",
+  "Robust CLR",
+  "One CLR",
+  "Nudge CLR",
+  "Zero CLR",
+  "DeSeq2"
+)
+
+pretty_beta_calcs <- c(
+  jaccard = "Jaccard",
+  bray = "Bray-Curtis",
+  euclidean = "Euclidean"
+)
+
+pretty_simulations <- c(
+  null = "Unbiased null model",
+  size = "Biased null model",
+  effect = "Skewed abundance model"
+)
+
+cluster_summary_files <- glue("data/{datasets}/cluster_analysis.tsv")
+names(cluster_summary_files) <- datasets
+
+beta_composite <- map_dfr(cluster_summary_files, read_tsv, .id = "dataset") %>%
+  mutate(process_calc = glue("{corrections}_{distances}")) %>%
+  inner_join(., tibble(process_calc = beta_metrics),
+             by = "process_calc") %>%
[…]

### 2022-11-11 code/run_cluster_analysis.R

@@ -1,0 +1,64 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(cluster)
+library(glue)
+
+get_accuracy <- function(seed, model, design, dataset, correct, dist) {
+
+  data <- read_tsv(glue("data/{dataset}/data.otu.{seed}.{model}_{correct}_{dist}.dist"),
+          col_types = cols(Group = col_character())) %>%
+    column_to_rownames("Group")
+  
+  if(is.na(data[1,2])) {
+    return(NA_real_)
+  }
+  
+  data %>%
+    pam(k = 2, diss = T) %>%
+    pluck("clustering") %>%
+    as_tibble(rownames = "Group") %>%
+    rename(cluster = value) %>%
+    inner_join(., read_tsv(glue("data/{dataset}/data.{seed}.{design}design"),
+                          col_types = cols(Group = col_character())),
+              by = "Group") %>%
+    count(cluster, treatment) %>%
+    pivot_wider(names_from = "treatment", values_from = "n", values_fill = 0) %>%
+    summarize(a = .[["A"]][1] + .[["B"]][2], b = .[["A"]][2] + .[["B"]][1]) %>%
+    pivot_longer(cols = c("a", "b")) %>%
+    mutate(accuracy = value / sum(value)) %>%
+    slice_max(accuracy, with_ties = FALSE) %>%
+    pull(accuracy)
+
+}
+
+input <- commandArgs(trailingOnly = TRUE)
+dataset <- input[1]
+
+output_file <- glue("data/{dataset}/cluster_analysis.tsv")
+
+corrections <- c("raw", "relabund", "rare",
+                 "deseq2", "metagenomeseq", "srs",
+                 "nclr", "oclr", "rclr", "zclr")
+
+distances <- c("bray",
+               "euclidean",
+               "jaccard")
+
+model <- c("r", "e")
+
+design <- c("r", "s", "e")
+
+seeds <- 1:100
+
+expand_grid(model, design, corrections, distances, seeds) %>% 
+  filter((model == "r" & design != "e") | (model == "e" & design == "e")) %>%
+  mutate(
+      dataset = dataset
+  ) %>%
+  rowwise() %>%
+  mutate(
+    accuracies = get_accuracy(seeds, model, design, dataset, corrections, distances)
+  ) %>%
+  write_tsv(output_file)
+

### 2022-11-11 submission/manuscript.Rmd
First full draft of Introduction

@@
 
-n_med_sequence_range <- table_1 %>%
-  pull(median) %>%
-  range() %>% as.integer()
+#n_med_sequence_range <- table_1 %>%
+#  pull(median) %>%
+#  range() %>% as.integer()
 
@@
 
-***https://edepot.wur.nl/547087.*** "Log- ratio PCA is designed to give results that are library size- independent.
-However, as we demonstrated mathematically and with examples based on simulated and real data, log- ratio PCA becomes library size- dependent, if there are many infrequent taxa (many zeroes) and library sizes differ largely.
-In this situation, the row centring used in log- ratio PCA brings an effect of r (the row mean of the log- transformed counts) in the clr- transformed matrix.
-Note that this effect is irrespective of whether or not these infrequent taxa are genuine or due to sequencing noise or allocation error.
-This library size dependence is unexpected in the sense that, after applying the clr, the transformed matrix is free of the effect of the row totals for strictly positive data (yij>0 for all i and j).
-We additionally demonstrate that library size variability causes a loss in power in dextecting an effect of x with log- ratio RDA.
-If there is additionally a correlation between treatment and the library size, the type 1 error for detecting the effect of x can be seriously inflated."
+The ability to generate millions sequence reads from 16S rRNA gene amplicons and metagenomic datasets has allowed researchers to multiplex multiple samples on the same sequencing run by pooling separate PCRs that can be deconvoluted later based on index (aka barcode) sequences that embeded into the primer sequence [@Kozich2013; @Caporaso2010].
+Unfortunately, it is common to observe variation in the number of sequence reads from each sample vary by as much as 100-fold (e.g, see Figure S1).
+This occurs because pooling of DNA from multiple PCRs is frought with numerous opportunities for technichal errors to compound leading to a skewed distribution.
+Aside from developing better methods of pooling DNA, the question of how to control for uneven sampling effort in microbial ecology studies has become controversial.
 
-***https://www.nature.com/articles/s41522-020-00160-w.*** "An important characteristic of a feature table is that it is typically sparse, sometimes as many as ~90% are zero entries21, which creates a challenge for analyzing rare taxa.
-A quick and simple strategy to deal with excess zeros is to add a small positive constant (e.g.
-1) called pseudo-count14,22 to each cell of the feature table.
-The addition of a pseudo-count becomes necessary when using methods of analysis that require log transformation of the observed counts.
-Even though adding a pseudo-count is simple and widely used, the choice of the pseudo-count is ad hoc.
-Studies have shown that differential abundance or clustering results could be sensitive to the choice of pseudo count23,24.
-Although different values of pseudo counts have been discussed in the literature23,24,25,26, to the best of our knowledge, there is no consensus on how to choose the optimal value.
-Other strategies involve modeling zero counts by some probability models21,27."
+The practice of rarefaction has been commonly used in ecology for more than 50 years as a tool to control for uneven sampling effort across experimental replicates [@Sanders1968; @Hurlbert1971].
+Microbial ecologists have used it to compare 16S rRNA gene sequence data for the past 25 years [@Dunbar1999; @Hughes2001; @Moyer1998].
+When data are rarefied, the investigator selects a desired threshold of sampling effort and removes any samples below that threshold.
+They then randomly select that many sequences, with replacement from each sample.
+Based on the obseved sequence counts, the researcher can then calcuate alpha diversity metrics including richness and diversity indices or beta divesity metrics such as a Jaccard or Bray-Curtis dissimilarity index.
+I refer to this single sampling as a subsample; this method is implemented as the sub.sample function in mothur [@Schloss2009] and the rrarefy function in the vegan R package [@Oksanen2022].
+Rarefaction repeats the subsampling a large number of times (e.g., 100 or 1,000) and calculates the mean of the alpha or beta diersity metric over those subsamplings; rarefaction is implemented in mothur using the summary.single and dist.shared functions [@Schloss2009] and with the vegan R package using the rarefy or avgdist fuctions [@Oksanen2022].
+Rarefaction effectively tells a researcher what an alpha or beta diversity metric would have been for a collection of samples if they were all evenly sampled.
+Although a closed form equation exists to calcualte the expected richness [@Hurlbert1971], it is computationally easier to empirically calculate rarefied richness and other alpha and beta diversity metrics.
 
-https://academic.oup.com/bioinformatics/article/34/16/2870/4956011
-https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6755255/
+In 2014, McMurdie and Holmes [@McMurdie2014] announced that rarefaction of microbial community data is "statistically inadmissible" because it omits valid data.
+In their simulations, they observed that the use of rarefaction reduced the statistical power to compare communities based on beta diversity metrics.
+The communities used in this study were simulated by mixing operational taxonomic unit (OTU) frequencies from marine and human fecal communities, which were then sampled to generate 40 replicate communities each sampled to different depths.
+The simulation with the greatest depth of coverage obtained an average of 10,000 sequences per sample.
+Analysis of the R code that they used indicates that the simulations were actually only based on a single subsample and not truly rarefied data.
+Furthermore, the authors removed any OTU from the simulated communities that was not observed in three or more samples.
+This is a frequently used approach for analyzing microbiome data [@Bokulich2013].
+Unfortunately, it removes valid data, skews the distribution of the community, and removes true patchiness from the representation of the communities [@Schloss2020].
+Finally, the analysis did not include analysis of alpha diersity metrics and chose to assess beta diversity metrics based on clustering quality of samples rather than using the more frequently used approach of employing non-parametric multivariate analysis of variance [@McArdle2001].
+Regardless of these caveats, McMurdie and Holmes recommended the use of variance stabilizing transformations such as those applied for gene expression analysis such as that found in the DeSeq R package.
 
-* Motivation
-  * Problem of uneven sampling effort
-* What is rarefaction? History, reason for rarefaction
-  * Repeated down sampling of datasets to a common number of observations to calculate the average value to ascertain the expected value of a metric for the metric under study; typically richness
-  * Control for unneven sampling effort
-  * Methods vary in their sensitivity to uneven sampling
-  * Compositional data analysis
-* Reasons behind "rarefaction is inadmissable"
-  * Weird simulation
-* Alternative approaches and claims
-  * sampling invariance
-* Goal of this study
+Beyond the critiques from McMurdie and Holmes others have developed alternative approaches to controlling for uneven sampling effort in amplicon sequencing studies.
+For alpha diversity metrics, Willis used toy datasets to demonstrate that one could estimate the richness for each sample in a dataset and use those values for statistical comparisons [@Willis2019].
+Non-parametric estimators of richness and diversity [@Chao2003; @Chao2016] and parametric estimators of richness [Willis2015] that have been used in microbial ecology studies.
+For beta diversity metrics at least four approaches have been pursued.
+First, one could use relative abundance values where the observed number of sequences in an OTU is divided by the total number of sequences in the sample [@Lin2020].
+Second, normalization strategies have been developed where the relative abundance is mutiplied by the size of the minimum desired sampling effort and fractional values are reapportioned to the OTUs to obtain integer values [@Beule2020; @Paulson2013].
+Third, a variety of center log-ratio methods have been developed where the compositional nature of the OTU counts is removed and used to calculated Euclidean distances (aka Aitchison distances) [@Quinn2019; @Paulson2013; @Lin2020; @Martino2019; @Costea2014.
+This strategy is purported to control for uneven sampling effort [@Martino2019; @Quinn2018]; however, some have noticed that this feature breaks down under certain conditions [@teBeest2021].
+Finally, as mentioned above, variance stabilization transformations have been recommended to generate values that can be used to calcualte Euclidean distances [@McMurdie2014].
 
-This analysis included 16S rRNA gene sequence data from from 12 studies that characterized the variation in bacterial communities from diverse environments (Table 1).
-The original studies generated the sequence data by pooling separate PCR products that were generated by amplifying the V4 region of the 16S rRNA gene from the bacterial DNA in multiple samples.
-Because pooling equimolar quantities of DNA is frought with difficulties, it was common to observe wide variation in the number of sequences in each sample (Table 1 and Figure S1).
+The ongoing controversy over the use of rarefaction and the development of numerous strategies to control for uneven sampling effort caused me to question how these methods compared to each other.
+My analysis included 16S rRNA gene sequence data from from 12 studies that characterized the variation in bacterial communities from diverse environments (Table 1 and Figure S1).
+The sequences were assigned to OTUs using a standard pipeline and their frequencies and the number of sequences found in each sample were used to generate simulated communities and treatment effects.
+For each dataset and simulation, 100 replicate datasets were generated and used as input to each of the strategies for controlling for uneven sampling effort.
+My overall conclusion was that rarefaction outperformed the alternative strategies.
 
@@
 This analysis has demonstrated a clear correlation between distances calculated by these methods and sequencing effort.
-This result is at odds with the claims by others that the distances are scale invariant [@Martino2019].
+This result is at odds with the claims by others that the distances are scale invariant [@Martino2019; @Quinn2018].
 Again, rarefaction is the most effective and consistent approach to control for uneven sampling effort when calculating beta diversity metrics.
@@
 
-\small
+\footnotesize
 
@@
     median=format(as.integer(median), big.mark=",", trim=TRUE),
+    mean=format(as.integer(mean), big.mark=",", trim=TRUE),
     min=format(min, big.mark=",", trim=TRUE),
@@
     n_samples = n_samples) %>%
-  select(nice_name, n_samples, total_seqs, median, range, sra_study) %>%
-  kableExtra::kable(format="markdown", booktabs=TRUE, escape=F, align="lrrrrr",
+  select(nice_name, n_samples, total_seqs, median, mean, range, sra_study) %>%
+  kbl(format="markdown", booktabs=TRUE, escape=F, align="lrrrrrr",
     linesep="",
@@
         "\\textbf{Total}\n\\textbf{sequences}",
-        "\\textbf{Median}\n\\textbf{sequences}",
-        "\\textbf{Range of}\n\\textbf{sequences}",
+        "\\textbf{Median}\n\\textbf{sample size}",
+        "\\textbf{Mean}\n\\textbf{sample size}",
+        "\\textbf{Range of}\n\\textbf{sample sizes}",
         "\\textbf{SRA study}\n\\textbf{accession}"

### 2022-11-11 submission/manuscript.tex
First full draft of Introduction

@@
 
-\textbf{\emph{\url{https://edepot.wur.nl/547087}.}} ``Log- ratio PCA is
-designed to give results that are library size- independent. However, as
-we demonstrated mathematically and with examples based on simulated and
-real data, log- ratio PCA becomes library size- dependent, if there are
-many infrequent taxa (many zeroes) and library sizes differ largely. In
-this situation, the row centring used in log- ratio PCA brings an effect
-of r (the row mean of the log- transformed counts) in the clr-
-transformed matrix. Note that this effect is irrespective of whether or
-not these infrequent taxa are genuine or due to sequencing noise or
-allocation error. This library size dependence is unexpected in the
-sense that, after applying the clr, the transformed matrix is free of
-the effect of the row totals for strictly positive data
-(yij\textgreater0 for all i and j). We additionally demonstrate that
-library size variability causes a loss in power in dextecting an effect
-of x with log- ratio RDA. If there is additionally a correlation between
-treatment and the library size, the type 1 error for detecting the
-effect of x can be seriously inflated.''
-
-\textbf{\emph{\url{https://www.nature.com/articles/s41522-020-00160-w}.}}
-``An important characteristic of a feature table is that it is typically
-sparse, sometimes as many as \textasciitilde90\% are zero entries21,
-which creates a challenge for analyzing rare taxa. A quick and simple
-strategy to deal with excess zeros is to add a small positive constant
-(e.g.~1) called pseudo-count14,22 to each cell of the feature table. The
-addition of a pseudo-count becomes necessary when using methods of
-analysis that require log transformation of the observed counts. Even
-though adding a pseudo-count is simple and widely used, the choice of
-the pseudo-count is ad hoc. Studies have shown that differential
-abundance or clustering results could be sensitive to the choice of
-pseudo count23,24. Although different values of pseudo counts have been
-discussed in the literature23,24,25,26, to the best of our knowledge,
-there is no consensus on how to choose the optimal value. Other
-strategies involve modeling zero counts by some probability
-models21,27.''
-
-\url{https://academic.oup.com/bioinformatics/article/34/16/2870/4956011}
-\url{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6755255/}
-
-\begin{itemize}
-\tightlist
-\item
-  Motivation
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Problem of uneven sampling effort
-  \end{itemize}
-\item
-  What is rarefaction? History, reason for rarefaction
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Repeated down sampling of datasets to a common number of
-    observations to calculate the average value to ascertain the
-    expected value of a metric for the metric under study; typically
-    richness
-  \item
-    Control for unneven sampling effort
-  \item
-    Methods vary in their sensitivity to uneven sampling
-  \item
-    Compositional data analysis
-  \end{itemize}
-\item
-  Reasons behind ``rarefaction is inadmissable''
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Weird simulation
-  \end{itemize}
-\item
-  Alternative approaches and claims
-
-  \begin{itemize}
-  \tightlist
-  \item
-    sampling invariance
-  \end{itemize}
-\item
-  Goal of this study
-\end{itemize}
-
-This analysis included 16S rRNA gene sequence data from from 12 studies
-that characterized the variation in bacterial communities from diverse
-environments (Table 1). The original studies generated the sequence data
-by pooling separate PCR products that were generated by amplifying the
-V4 region of the 16S rRNA gene from the bacterial DNA in multiple
-samples. Because pooling equimolar quantities of DNA is frought with
-difficulties, it was common to observe wide variation in the number of
-sequences in each sample (Table 1 and Figure S1).
+The ability to generate millions sequence reads from 16S rRNA gene
+amplicons and metagenomic datasets has allowed researchers to multiplex
+multiple samples on the same sequencing run by pooling separate PCRs
+that can be deconvoluted later based on index (aka barcode) sequences
+that embeded into the primer sequence (1, 2). Unfortunately, it is
+common to observe variation in the number of sequence reads from each
+sample vary by as much as 100-fold (e.g, see Figure S1). This occurs
+because pooling of DNA from multiple PCRs is frought with numerous
+opportunities for technichal errors to compound leading to a skewed
+distribution. Aside from developing better methods of pooling DNA, the
+question of how to control for uneven sampling effort in microbial
+ecology studies has become controversial.
+
+The practice of rarefaction has been commonly used in ecology for more
+than 50 years as a tool to control for uneven sampling effort across
+experimental replicates (3, 4). Microbial ecologists have used it to
+compare 16S rRNA gene sequence data for the past 25 years (5--7). When
+data are rarefied, the investigator selects a desired threshold of
+sampling effort and removes any samples below that threshold. They then
+randomly select that many sequences, with replacement from each sample.
+Based on the obseved sequence counts, the researcher can then calcuate
+alpha diversity metrics including richness and diversity indices or beta
+divesity metrics such as a Jaccard or Bray-Curtis dissimilarity index. I
+refer to this single sampling as a subsample; this method is implemented
+as the sub.sample function in mothur (8) and the rrarefy function in the
+vegan R package (9). Rarefaction repeats the subsampling a large number
+of times (e.g., 100 or 1,000) and calculates the mean of the alpha or
+beta diersity metric over those subsamplings; rarefaction is implemented
+in mothur using the summary.single and dist.shared functions (8) and
+with the vegan R package using the rarefy or avgdist fuctions (9).
+Rarefaction effectively tells a researcher what an alpha or beta
+diversity metric would have been for a collection of samples if they
+were all evenly sampled. Although a closed form equation exists to
+calcualte the expected richness (4), it is computationally easier to
+empirically calculate rarefied richness and other alpha and beta
+diversity metrics.
+
+In 2014, McMurdie and Holmes (10) announced that rarefaction of
+microbial community data is ``statistically inadmissible'' because it
+omits valid data. In their simulations, they observed that the use of
+rarefaction reduced the statistical power to compare communities based
+on beta diversity metrics. The communities used in this study were
+simulated by mixing operational taxonomic unit (OTU) frequencies from
+marine and human fecal communities, which were then sampled to generate
+40 replicate communities each sampled to different depths. The
+simulation with the greatest depth of coverage obtained an average of
+10,000 sequences per sample. Analysis of the R code that they used
+indicates that the simulations were actually only based on a single
+subsample and not truly rarefied data. Furthermore, the authors removed
+any OTU from the simulated communities that was not observed in three or
+more samples. This is a frequently used approach for analyzing
+microbiome data (11). Unfortunately, it removes valid data, skews the
+distribution of the community, and removes true patchiness from the
+representation of the communities (12). Finally, the analysis did not
+include analysis of alpha diersity metrics and chose to assess beta
+diversity metrics based on clustering quality of samples rather than
+using the more frequently used approach of employing non-parametric
+multivariate analysis of variance (13). Regardless of these caveats,
+McMurdie and Holmes recommended the use of variance stabilizing
+transformations such as those applied for gene expression analysis such
+as that found in the DeSeq R package.
+
+Beyond the critiques from McMurdie and Holmes others have developed
+alternative approaches to controlling for uneven sampling effort in
+amplicon sequencing studies. For alpha diversity metrics, Willis used
+toy datasets to demonstrate that one could estimate the richness for
+each sample in a dataset and use those values for statistical
+comparisons (14). Non-parametric estimators of richness and diversity
+(15, 16) and parametric estimators of richness {[}Willis2015{]} that
+have been used in microbial ecology studies. For beta diversity metrics
+at least four approaches have been pursued. First, one could use
+relative abundance values where the observed number of sequences in an
+OTU is divided by the total number of sequences in the sample (17).
+Second, normalization strategies have been developed where the relative
+abundance is mutiplied by the size of the minimum desired sampling
+effort and fractional values are reapportioned to the OTUs to obtain
+integer values (18, 19). Third, a variety of center log-ratio methods
+have been developed where the compositional nature of the OTU counts is
+removed and used to calculated Euclidean distances (aka Aitchison
+distances) {[}(20); (19); (17); (21); (22). This strategy is purported
+to control for uneven sampling effort (21, 23); however, some have
+noticed that this feature breaks down under certain conditions (24).
+Finally, as mentioned above, variance stabilization transformations have
+been recommended to generate values that can be used to calcualte
+Euclidean distances (10).
+
+The ongoing controversy over the use of rarefaction and the development
+of numerous strategies to control for uneven sampling effort caused me
+to question how these methods compared to each other. My analysis
+included 16S rRNA gene sequence data from from 12 studies that
+characterized the variation in bacterial communities from diverse
+environments (Table 1 and Figure S1). The sequences were assigned to
+OTUs using a standard pipeline and their frequencies and the number of
+sequences found in each sample were used to generate simulated
+communities and treatment effects. For each dataset and simulation, 100
+replicate datasets were generated and used as input to each of the
+strategies for controlling for uneven sampling effort. My overall
+conclusion was that rarefaction outperformed the alternative strategies.
 
@@
 falsely detecting differences between treatment groups if sampling depth
-and sequencing effort are confounded (e.g., 1). The correlation with
+and sequencing effort are confounded (e.g., 25). The correlation with
 sampling effort leads to an artificial increase in the variation between
@@
 alpha diversity has been to use non-parametric or parametric methods
-using raw data and to then compare the estimates (2). My data
+using raw data and to then compare the estimates (14). My data
 demonstrates that such approaches are limited for several reasons.
@@
 the distribtion of sequences and OTUs leading to a distortion of the
-measurement of alpha diversity (3). The effects of removing rare
+measurement of alpha diversity (12). The effects of removing rare
 sequences would vary across samples depending on the number of sequences
@@
 This result is at odds with the claims by others that the distances are
-scale invariant (4). Again, rarefaction is the most effective and
+scale invariant (21, 23). Again, rarefaction is the most effective and
 consistent approach to control for uneven sampling effort when
@@
 data and that the selection of the value to rarefy the sequence data to
-is arbitrary (5). I disagree with both critiques. All of the data are
+is arbitrary (10). I disagree with both critiques. All of the data are
 used to calculate the mean value of the rarefied metrics; none of it is
@@
 sequences can bias alpha and beta diversity metrics and reduce
-statistical power (3). As for the second critism, I would resist the
+statistical power (12). As for the second critism, I would resist the
 claim that the selection of the rarefaction threshold is arbitrary. In
@@
 DNAs which may add to the challenges of achieving equimolar mixtures of
-the desired amplicons (1). Regardless, it is clear that better
+the desired amplicons (25). Regardless, it is clear that better
 strategies are needed to reduce the variation in the number of sequences
@@
 region resulting in high quality contigs with a low sequencing error
-rate (6). These data were processed through a standardized sequence
+rate (1). These data were processed through a standardized sequence
 curation pipeline to generate operational taxonomic units (OTUs) using
-the mothur software package (6, 7). OTUs were identified using the
+the mothur software package (1, 8). OTUs were identified using the
 OptiClust algorithm to cluster sequences together that were not more
-than 3\% different from each other (8).
+than 3\% different from each other (26).
 
@@
 the total number of sequences in each OTU. This is a similar approach to
-that of the IS algorithm described by Ulrich and Gotelli (9). Because
+that of the IS algorithm described by Ulrich and Gotelli (27). Because
 the construction of the null models was a stochastic process, 100
@@
 diversity metrics were performed using the adonis2 function from the
-vegan (v.2.6.2) R package (10). The adonis2 function implements a
+vegan (v.2.6.2) R package (9). The adonis2 function implements a
 non-parametric multivariate analysis of variance using distances
-matrices as described by McArdle and Anderson (11). Throughout this
+matrices as described by McArdle and Anderson (13). Throughout this
 study I used 0.05 as the threshold for asssessing statistical
@@
 indeices. Raw OTU counts were used as input to calculate sample richness
-and Shannon and inverse Simpson diversity using mothur (7, 12). Shannon
+and Shannon and inverse Simpson diversity using mothur (8, 28). Shannon
 diversity was calculated as
@@
 normalize OTU counts to the size of the smallest sample in each dataset
-using the SRS R package (v.0.2.3)(13). Normalized OTU counts were used
+using the SRS R package (v.0.2.3)(18). Normalized OTU counts were used
 to calculate sample richness and Shannon and inverse Simpson diversity
@@
 abundances of the features do not change with the normalization
-procedure (14). The non-parametric bias-corrected Chao1 and ACE richness
-estimators (15) and a non-parametric estimator of the Shannon diversity
-(16) were calculated using raw OTU counts with mothur. Parametric
+procedure (19). The non-parametric bias-corrected Chao1 and ACE richness
+estimators (16) and a non-parametric estimator of the Shannon diversity
+(15) were calculated using raw OTU counts with mothur. Parametric
 estimates of sample richness were calculated using the breakaway (BA) R
-package (v.4.7.9)(17). The current analysis reports both the results
+package (v.4.7.9)(29). The current analysis reports both the results
 from running default model selection procedure and the Poisson model.
@@
 dissimilarity indices using the vegdist function from the vegan R
-package (v.2.6.2)(10). The Jaccard index was calculated as
+package (v.2.6.2)(9). The Jaccard index was calculated as
 
@@
 sum scaling (CSS) were used to normalize raw OTU counts using the SRS
-(v.0.2.3) and metagenomeSeq (v.1.36.0) R packages, respectively (13,
-14). The normalized counts were then used to calculate Jaccard and
+(v.0.2.3) and metagenomeSeq (v.1.36.0) R packages, respectively (18,
+19). The normalized counts were then used to calculate Jaccard and
 Bray-Curtis dissimilarity indices. Finally, the variance-stabilization
 transformation (VST) as implemented in the DESeq2 (v.1.34.0) R package
-was used to normalize the data as described by McMurdie and Holmes (5,
-18). Because the VST approach generates negative values, which are
+was used to normalize the data as described by McMurdie and Holmes (10,
+30). Because the VST approach generates negative values, which are
 incompatible with calculating Jaccard and Bray-Curtis dissimilarity
@@
 based on the observed data using the zCompositions (v.1.4.0.1) R package
-(19). The second, One CLR, added a pseudocount of 1 to the abundance of
-all OTUs (14, 20). The third, Nudge CLR, added a pseudocount of 1
+(31). The second, One CLR, added a pseudocount of 1 to the abundance of
+all OTUs (17, 19). The third, Nudge CLR, added a pseudocount of 1
 divided by the total number of sequences in a sample to each OTU in the
-sample (20, 21). The final approach, Robust CLR, removed unobserved OTUs
-prior to calcualting the CLR (4).
+sample (17, 22). The final approach, Robust CLR, removed unobserved OTUs
+prior to calcualting the CLR (21).
 
@@
 \begin{CSLReferences}{0}{1}
-\leavevmode\vadjust pre{\hypertarget{ref-Morris2013}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-Kozich2013}{}}%
 \CSLLeftMargin{1. }%
-\CSLRightInline{\textbf{Morris A}, \textbf{Beck JM}, \textbf{Schloss
-PD}, \textbf{Campbell TB}, \textbf{Crothers K}, \textbf{Curtis JL},
-\textbf{Flores SC}, \textbf{Fontenot AP}, \textbf{Ghedin E},
-\textbf{Huang L}, \textbf{Jablonski K}, \textbf{Kleerup E},
-\textbf{Lynch SV}, \textbf{Sodergren E}, \textbf{Twigg H}, \textbf{Young
-VB}, \textbf{Bassis CM}, \textbf{Venkataraman A}, \textbf{Schmidt TM},
-\textbf{Weinstock GM}. 2013. Comparison of the respiratory microbiome in
-healthy nonsmokers and smokers. American Journal of Respiratory and
-Critical Care Medicine \textbf{187}:1067--1075.
-doi:\href{https://doi.org/10.1164/rccm.201210-1913oc}{10.1164/rccm.201210-1913oc}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Willis2019}{}}%
-\CSLLeftMargin{2. }%
-\CSLRightInline{\textbf{Willis AD}. 2019. Rarefaction, alpha diversity,
-and statistics. Frontiers in Microbiology \textbf{10}.
-doi:\href{https://doi.org/10.3389/fmicb.2019.02407}{10.3389/fmicb.2019.02407}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Schloss2020}{}}%
-\CSLLeftMargin{3. }%
-\CSLRightInline{\textbf{Schloss PD}. 2020. Removal of rare amplicon
-sequence variants from 16S {rRNA} gene sequence surveys biases the
-interpretation of community structure data.
-doi:\href{https://doi.org/10.1101/2020.12.11.422279}{10.1101/2020.12.11.422279}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Martino2019}{}}%
-\CSLLeftMargin{4. }%
-\CSLRightInline{\textbf{Martino C}, \textbf{Morton JT}, \textbf{Marotz
-CA}, \textbf{Thompson LR}, \textbf{Tripathi A}, \textbf{Knight R},
-\textbf{Zengler K}. 2019. A novel sparse compositional technique reveals
-microbial perturbations. {mSystems} \textbf{4}.
-doi:\href{https://doi.org/10.1128/msystems.00016-19}{10.1128/msystems.00016-19}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-McMurdie2014}{}}%
-\CSLLeftMargin{5. }%
-\CSLRightInline{\textbf{McMurdie PJ}, \textbf{Holmes S}. 2014. Waste
-not, want not: Why rarefying microbiome data is inadmissible. {PLoS}
-Computational Biology \textbf{10}:e1003531.
-doi:\href{https://doi.org/10.1371/journal.pcbi.1003531}{10.1371/journal.pcbi.1003531}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Kozich2013}{}}%
-\CSLLeftMargin{6. }%
 \CSLRightInline{\textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-Caporaso2010}{}}%
+\CSLLeftMargin{2. }%
+\CSLRightInline{\textbf{Caporaso JG}, \textbf{Lauber CL},
+\textbf{Walters WA}, \textbf{Berg-Lyons D}, \textbf{Lozupone CA},
+\textbf{Turnbaugh PJ}, \textbf{Fierer N}, \textbf{Knight R}. 2010.
+Global patterns of 16S {rRNA} diversity at a depth of millions of
+sequences per sample. Proceedings of the National Academy of Sciences
+\textbf{108}:4516--4522.
+doi:\href{https://doi.org/10.1073/pnas.1000080107}{10.1073/pnas.1000080107}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Sanders1968}{}}%
+\CSLLeftMargin{3. }%
+\CSLRightInline{\textbf{Sanders HL}. 1968. Marine benthic diversity: A
+comparative study. The American Naturalist \textbf{102}:243--282.
+doi:\href{https://doi.org/10.1086/282541}{10.1086/282541}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Hurlbert1971}{}}%
+\CSLLeftMargin{4. }%
+\CSLRightInline{\textbf{Hurlbert SH}. 1971. The nonconcept of species
+diversity: A critique and alternative parameters. Ecology
+\textbf{52}:577--586.
+doi:\href{https://doi.org/10.2307/1934145}{10.2307/1934145}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Dunbar1999}{}}%
+\CSLLeftMargin{5. }%
+\CSLRightInline{\textbf{Dunbar J}, \textbf{Takala S}, \textbf{Barns SM},
+\textbf{Davis JA}, \textbf{Kuske CR}. 1999. Levels of bacterial
+community diversity in four arid soils compared by cultivation and 16S
+{rRNA} gene cloning. Applied and Environmental Microbiology
+\textbf{65}:1662--1669.
+doi:\href{https://doi.org/10.1128/aem.65.4.1662-1669.1999}{10.1128/aem.65.4.1662-1669.1999}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Hughes2001}{}}%
+\CSLLeftMargin{6. }%
+\CSLRightInline{\textbf{Hughes JB}, \textbf{Hellmann JJ},
+\textbf{Ricketts TH}, \textbf{Bohannan BJM}. 2001. Counting the
+uncountable: Statistical approaches to estimating microbial diversity.
+Applied and Environmental Microbiology \textbf{67}:4399--4406.
+doi:\href{https://doi.org/10.1128/aem.67.10.4399-4406.2001}{10.1128/aem.67.10.4399-4406.2001}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Moyer1998}{}}%
+\CSLLeftMargin{7. }%
+\CSLRightInline{\textbf{Moyer CL}, \textbf{Tiedje JM}, \textbf{Dobbs
+FC}, \textbf{Karl DM}. 1998. Diversity of deep-sea hydrothermal vent
+archaea from loihi seamount, hawaii. Deep Sea Research Part {II}:
+Topical Studies in Oceanography \textbf{45}:303--317.
+doi:\href{https://doi.org/10.1016/s0967-0645(97)00081-7}{10.1016/s0967-0645(97)00081-7}.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-Schloss2009}{}}%
-\CSLLeftMargin{7. }%
+\CSLLeftMargin{8. }%
 \CSLRightInline{\textbf{Schloss PD}, \textbf{Westcott SL},
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Westcott2017}{}}%
-\CSLLeftMargin{8. }%
-\CSLRightInline{\textbf{Westcott SL}, \textbf{Schloss PD}. 2017.
-{OptiClust}, an improved method for assigning amplicon-based sequence
-data to operational taxonomic units. {mSphere} \textbf{2}.
-doi:\href{https://doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Ulrich2010}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-Oksanen2022}{}}%
 \CSLLeftMargin{9. }%
-\CSLRightInline{\textbf{Ulrich W}, \textbf{Gotelli NJ}. 2010. Null model
-analysis of species associations using abundance data. Ecology
-\textbf{91}:3384--3397.
-doi:\href{https://doi.org/10.1890/09-2157.1}{10.1890/09-2157.1}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Oksanen2022}{}}%
-\CSLLeftMargin{10. }%
 \CSLRightInline{\textbf{Oksanen J}, \textbf{Simpson GL},
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-McMurdie2014}{}}%
+\CSLLeftMargin{10. }%
+\CSLRightInline{\textbf{McMurdie PJ}, \textbf{Holmes S}. 2014. Waste
+not, want not: Why rarefying microbiome data is inadmissible. {PLoS}
+Computational Biology \textbf{10}:e1003531.
+doi:\href{https://doi.org/10.1371/journal.pcbi.1003531}{10.1371/journal.pcbi.1003531}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Bokulich2013}{}}%
+\CSLLeftMargin{11. }%
+\CSLRightInline{\textbf{Bokulich NA}, \textbf{Subramanian S},
+\textbf{Faith JJ}, \textbf{Gevers D}, \textbf{Gordon JI}, \textbf{Knight
+R}, \textbf{Mills DA}, \textbf{Caporaso JG}. 2013. Quality-filtering
+vastly improves diversity estimates from illumina amplicon sequencing.
+Nature Methods \textbf{10}:57--59.
+doi:\href{https://doi.org/10.1038/nmeth.2276}{10.1038/nmeth.2276}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Schloss2020}{}}%
+\CSLLeftMargin{12. }%
+\CSLRightInline{\textbf{Schloss PD}. 2020. Removal of rare amplicon
+sequence variants from 16S {rRNA} gene sequence surveys biases the
+interpretation of community structure data.
+doi:\href{https://doi.org/10.1101/2020.12.11.422279}{10.1101/2020.12.11.422279}.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-McArdle2001}{}}%
-\CSLLeftMargin{11. }%
+\CSLLeftMargin{13. }%
 \CSLRightInline{\textbf{McArdle BH}, \textbf{Anderson MJ}. 2001.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Magurran2004}{}}%
-\CSLLeftMargin{12. }%
-\CSLRightInline{\textbf{Magurran AE}. 2004.
-\href{https://books.google.com/books?id=tUqzLSUzXxcC}{Measuring
-biological diversity}. Wiley.}
+\leavevmode\vadjust pre{\hypertarget{ref-Willis2019}{}}%
+\CSLLeftMargin{14. }%
+\CSLRightInline{\textbf{Willis AD}. 2019. Rarefaction, alpha diversity,
+and statistics. Frontiers in Microbiology \textbf{10}.
+doi:\href{https://doi.org/10.3389/fmicb.2019.02407}{10.3389/fmicb.2019.02407}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Chao2003}{}}%
+\CSLLeftMargin{15. }%
+\CSLRightInline{\textbf{Chao A}, \textbf{Shen T-J}. 2003. Nonparametric
+estimation of shannon's index of diversity when there are unseen species
+in sample. Environmental and ecological statistics
+\textbf{10}:429--443.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Chao2016}{}}%
+\CSLLeftMargin{16. }%
+\CSLRightInline{\textbf{Chao A}, \textbf{Chiu C-H}. 2016. Species
+richness: Estimation and comparison. Wiley StatsRef: statistics
+reference online \textbf{1}:26.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Lin2020}{}}%
+\CSLLeftMargin{17. }%
+\CSLRightInline{\textbf{Lin H}, \textbf{Peddada SD}. 2020. Analysis of
+microbial compositions: A review of normalization and differential
+abundance analysis. npj Biofilms and Microbiomes \textbf{6}.
+doi:\href{https://doi.org/10.1038/s41522-020-00160-w}{10.1038/s41522-020-00160-w}.}
 
 \leavevmode\vadjust pre{\hypertarget{ref-Beule2020}{}}%
-\CSLLeftMargin{13. }%
+\CSLLeftMargin{18. }%
 \CSLRightInline{\textbf{Beule L}, \textbf{Karlovsky P}. 2020.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Paulson2013}{}}%
-\CSLLeftMargin{14. }%
+\CSLLeftMargin{19. }%
 \CSLRightInline{\textbf{Paulson JN}, \textbf{Stine OC}, \textbf{Bravo
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Chao2016}{}}%
-\CSLLeftMargin{15. }%
-\CSLRightInline{\textbf{Chao A}, \textbf{Chiu C-H}. 2016. Species
-richness: Estimation and comparison. Wiley StatsRef: statistics
-reference online \textbf{1}:26.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Chao2003}{}}%
-\CSLLeftMargin{16. }%
-\CSLRightInline{\textbf{Chao A}, \textbf{Shen T-J}. 2003. Nonparametric
-estimation of shannon's index of diversity when there are unseen species
-in sample. Environmental and ecological statistics
-\textbf{10}:429--443.}
+\leavevmode\vadjust pre{\hypertarget{ref-Quinn2019}{}}%
+\CSLLeftMargin{20. }%
+\CSLRightInline{\textbf{Quinn TP}, \textbf{Erb I}, \textbf{Gloor G},
+\textbf{Notredame C}, \textbf{Richardson MF}, \textbf{Crowley TM}. 2019.
+A field guide for the compositional analysis of any-omics data.
+{GigaScience} \textbf{8}.
+doi:\href{https://doi.org/10.1093/gigascience/giz107}{10.1093/gigascience/giz107}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Martino2019}{}}%
+\CSLLeftMargin{21. }%
+\CSLRightInline{\textbf{Martino C}, \textbf{Morton JT}, \textbf{Marotz
+CA}, \textbf{Thompson LR}, \textbf{Tripathi A}, \textbf{Knight R},
+\textbf{Zengler K}. 2019. A novel sparse compositional technique reveals
+microbial perturbations. {mSystems} \textbf{4}.
+doi:\href{https://doi.org/10.1128/msystems.00016-19}{10.1128/msystems.00016-19}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Costea2014}{}}%
+\CSLLeftMargin{22. }%
+\CSLRightInline{\textbf{Costea PI}, \textbf{Zeller G}, \textbf{Sunagawa
+S}, \textbf{Bork P}. 2014. A fair comparison. Nature Methods
+\textbf{11}:359--359.
+doi:\href{https://doi.org/10.1038/nmeth.2897}{10.1038/nmeth.2897}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Quinn2018}{}}%
+\CSLLeftMargin{23. }%
+\CSLRightInline{\textbf{Quinn TP}, \textbf{Erb I}, \textbf{Richardson
+MF}, \textbf{Crowley TM}. 2018. Understanding sequencing data as
+compositions: An outlook and review. Bioinformatics
+\textbf{34}:2870--2878.
+doi:\href{https://doi.org/10.1093/bioinformatics/bty175}{10.1093/bioinformatics/bty175}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-teBeest2021}{}}%
+\CSLLeftMargin{24. }%
+\CSLRightInline{\textbf{Beest DE te}, \textbf{Nijhuis EH},
+\textbf{Möhlmann TWR}, \textbf{Braak CJF}. 2021. Log-ratio analysis of
+microbiome data with many zeroes is library size dependent. Molecular
+Ecology Resources \textbf{21}:1866--1874.
+doi:\href{https://doi.org/10.1111/1755-0998.13391}{10.1111/1755-0998.13391}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Morris2013}{}}%
+\CSLLeftMargin{25. }%
+\CSLRightInline{\textbf{Morris A}, \textbf{Beck JM}, \textbf{Schloss
+PD}, \textbf{Campbell TB}, \textbf{Crothers K}, \textbf{Curtis JL},
+\textbf{Flores SC}, \textbf{Fontenot AP}, \textbf{Ghedin E},
+\textbf{Huang L}, \textbf{Jablonski K}, \textbf{Kleerup E},
+\textbf{Lynch SV}, \textbf{Sodergren E}, \textbf{Twigg H}, \textbf{Young
+VB}, \textbf{Bassis CM}, \textbf{Venkataraman A}, \textbf{Schmidt TM},
+\textbf{Weinstock GM}. 2013. Comparison of the respiratory microbiome in
+healthy nonsmokers and smokers. American Journal of Respiratory and
+Critical Care Medicine \textbf{187}:1067--1075.
+doi:\href{https://doi.org/10.1164/rccm.201210-1913oc}{10.1164/rccm.201210-1913oc}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Westcott2017}{}}%
+\CSLLeftMargin{26. }%
+\CSLRightInline{\textbf{Westcott SL}, \textbf{Schloss PD}. 2017.
+{OptiClust}, an improved method for assigning amplicon-based sequence
+data to operational taxonomic units. {mSphere} \textbf{2}.
+doi:\href{https://doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Ulrich2010}{}}%
+\CSLLeftMargin{27. }%
+\CSLRightInline{\textbf{Ulrich W}, \textbf{Gotelli NJ}. 2010. Null model
+analysis of species associations using abundance data. Ecology
+\textbf{91}:3384--3397.
+doi:\href{https://doi.org/10.1890/09-2157.1}{10.1890/09-2157.1}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Magurran2004}{}}%
+\CSLLeftMargin{28. }%
+\CSLRightInline{\textbf{Magurran AE}. 2004.
+\href{https://books.google.com/books?id=tUqzLSUzXxcC}{Measuring
+biological diversity}. Wiley.}
 
 \leavevmode\vadjust pre{\hypertarget{ref-Willis2015}{}}%
-\CSLLeftMargin{17. }%
+\CSLLeftMargin{29. }%
 \CSLRightInline{\textbf{Willis A}, \textbf{Bunge J}. 2015. Estimating
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Love2014}{}}%
-\CSLLeftMargin{18. }%
+\CSLLeftMargin{30. }%
 \CSLRightInline{\textbf{Love MI}, \textbf{Huber W}, \textbf{Anders S}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Palarea2015}{}}%
-\CSLLeftMargin{19. }%
+\CSLLeftMargin{31. }%
 \CSLRightInline{\textbf{Palarea-Albaladejo J}, \textbf{Martin-Fernandez
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Lin2020}{}}%
-\CSLLeftMargin{20. }%
-\CSLRightInline{\textbf{Lin H}, \textbf{Peddada SD}. 2020. Analysis of
-microbial compositions: A review of normalization and differential
-abundance analysis. npj Biofilms and Microbiomes \textbf{6}.
-doi:\href{https://doi.org/10.1038/s41522-020-00160-w}{10.1038/s41522-020-00160-w}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Costea2014}{}}%
-\CSLLeftMargin{21. }%
-\CSLRightInline{\textbf{Costea PI}, \textbf{Zeller G}, \textbf{Sunagawa
-S}, \textbf{Bork P}. 2014. A fair comparison. Nature Methods
-\textbf{11}:359--359.
-doi:\href{https://doi.org/10.1038/nmeth.2897}{10.1038/nmeth.2897}.}
-
 \leavevmode\vadjust pre{\hypertarget{ref-Li2015}{}}%
-\CSLLeftMargin{22. }%
+\CSLLeftMargin{32. }%
 \CSLRightInline{\textbf{Li Q}, \textbf{Heist EP}, \textbf{Moe LA}. 2015.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Baxter2016}{}}%
-\CSLLeftMargin{23. }%
+\CSLLeftMargin{33. }%
 \CSLRightInline{\textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Beall2015}{}}%
-\CSLLeftMargin{24. }%
+\CSLLeftMargin{34. }%
 \CSLRightInline{\textbf{Beall BFN}, \textbf{Twiss MR}, \textbf{Smith
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Henson2016}{}}%
-\CSLLeftMargin{25. }%
+\CSLLeftMargin{35. }%
 \CSLRightInline{\textbf{Henson MW}, \textbf{Pitre DM}, \textbf{Weckhorst
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Baxter2014}{}}%
-\CSLLeftMargin{26. }%
+\CSLLeftMargin{36. }%
 \CSLRightInline{\textbf{Baxter NT}, \textbf{Wan JJ}, \textbf{Schubert
@@
 \leavevmode\vadjust pre{\hypertarget{ref-LevyBooth2018}{}}%
-\CSLLeftMargin{27. }%
+\CSLLeftMargin{37. }%
 \CSLRightInline{\textbf{Levy-Booth DJ}, \textbf{Giesbrecht IJW},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Edwards2015}{}}%
-\CSLLeftMargin{28. }%
+\CSLLeftMargin{38. }%
 \CSLRightInline{\textbf{Edwards J}, \textbf{Johnson C},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Ettinger2017}{}}%
-\CSLLeftMargin{29. }%
+\CSLLeftMargin{39. }%
 \CSLRightInline{\textbf{Ettinger CL}, \textbf{Williams SL},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Graw2018}{}}%
-\CSLLeftMargin{30. }%
+\CSLLeftMargin{40. }%
 \CSLRightInline{\textbf{Graw MF}, \textbf{DAngelo G}, \textbf{Borchers
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Johnston2016}{}}%
-\CSLLeftMargin{31. }%
+\CSLLeftMargin{41. }%
 \CSLRightInline{\textbf{Johnston ER}, \textbf{Rodriguez-R LM},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Hassell2018}{}}%
-\CSLLeftMargin{32. }%
+\CSLLeftMargin{42. }%
 \CSLRightInline{\textbf{Hassell N}, \textbf{Tinker KA}, \textbf{Moore
@@
 
-\small
+\footnotesize
 
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1434}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0659}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1899}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1938}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2016}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2054}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1182}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0543}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1565}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1661}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1597}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1757}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1693}}@{}}
 \toprule()
@@
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-\makecell[c]{\textbf{Median}\\\textbf{sequences}}
+\makecell[c]{\textbf{Median}\\\textbf{sample size}}
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-\makecell[c]{\textbf{Range of}\\\textbf{sequences}}
+\makecell[c]{\textbf{Mean}\\\textbf{sample size}}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+\makecell[c]{\textbf{Range of}\\\textbf{sample sizes}}
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
@@
 \endhead
-Bioethanol~(22) & 95 & 3,970,972 & 16,014 & 3,690\Hyphdash*356,027 &
-SRP055545 \\
-Human~(23) & 490 & 20,828,275 & 32,452 & 10,439\Hyphdash*422,904 &
-SRP062005 \\
-Lake~(24) & 52 & 3,145,486 & 69,205 & 15,135\Hyphdash*110,993 &
+Bioethanol~(32) & 95 & 3,970,972 & 16,014 & 41,799 &
+3,690\Hyphdash*356,027 & SRP055545 \\
+Human~(33) & 490 & 20,828,275 & 32,452 & 42,506 &
+10,439\Hyphdash*422,904 & SRP062005 \\
+Lake~(34) & 52 & 3,145,486 & 69,205 & 60,490 & 15,135\Hyphdash*110,993 &
 SRP050963 \\
-Marine~(25) & 7 & 1,484,068 & 213,091 & 132,895\Hyphdash*256,758 &
-SRP068101 \\
-Mice~(6) & 348 & 2,785,641 & 6,426 & 1,804\Hyphdash*30,311 &
+Marine~(35) & 7 & 1,484,068 & 213,091 & 212,009 &
+132,895\Hyphdash*256,758 & SRP068101 \\
+Mice~(1) & 348 & 2,785,641 & 6,426 & 8,004 & 1,804\Hyphdash*30,311 &
 SRP192323 \\
-Peromyscus~(26) & 111 & 1,545,288 & 12,393 & 4,454\Hyphdash*33,502 &
-SRP044050 \\
-Rainforest~(27) & 69 & 936,666 & 11,464 & 4,880\Hyphdash*37,403 &
-ERP023747 \\
-Rice~(28) & 490 & 22,623,937 & 43,399 & 2,777\Hyphdash*192,200 &
-SRP044745 \\
-Seagrass~(29) & 286 & 4,135,440 & 13,538 & 1,830\Hyphdash*45,076 &
-SRP092441 \\
-Sediment~(30) & 58 & 1,151,389 & 17,606 & 7,686\Hyphdash*67,763 &
-SRP097192 \\
-Soil~(31) & 18 & 932,563 & 50,487 & 46,622\Hyphdash*58,935 &
+Peromyscus~(36) & 111 & 1,545,288 & 12,393 & 13,921 &
+4,454\Hyphdash*33,502 & SRP044050 \\
+Rainforest~(37) & 69 & 936,666 & 11,464 & 13,574 & 4,880\Hyphdash*37,403
+& ERP023747 \\
+Rice~(38) & 490 & 22,623,937 & 43,399 & 46,171 & 2,777\Hyphdash*192,200
+& SRP044745 \\
+Seagrass~(39) & 286 & 4,135,440 & 13,538 & 14,459 &
+1,830\Hyphdash*45,076 & SRP092441 \\
+Sediment~(40) & 58 & 1,151,389 & 17,606 & 19,851 & 7,686\Hyphdash*67,763
+& SRP097192 \\
+Soil~(41) & 18 & 932,563 & 50,487 & 51,809 & 46,622\Hyphdash*58,935 &
 ERP012016 \\
-Stream~(32) & 201 & 21,017,610 & 90,621 & 8,931\Hyphdash*394,419 &
-SRP075852 \\
+Stream~(42) & 201 & 21,017,610 & 90,621 & 104,565 &
+8,931\Hyphdash*394,419 & SRP075852 \\
 \bottomrule()

### 2022-11-11 submission/manuscript.Rmd
First full version of Abstract and Importance

@@
 
-# Rarefy your data
+# Rarefaction is currently the best approach to control for uneven sampling effort in amplicon sequencing analyses
 
@@
 
+How to control for uneven sampling effort in amplicon sequencing analyses has become a contentious question in the microbial ecology field since it is common to find as much as 100-fold variation in the number of 16S rRNA gene sequences found across samples in a study.
+Some have argued that rarefaction is inadmissible because it omits valid data and reduces the statistical power to detect differences between treatment groups.
+A number of alternative approaches have been developed to normalize and rescale the data that purport to be invariant to the number of observations.
+I generated community distributions based on 12 published datasets where I was able to assess the sensitivity of different methods to uneven sampling effort.
+Rarefaction was the only method that could control for variation in uneven sampling effort when measuring commonly used alpha and beta diversity metrics.
+Next, I compared the false detection rate and power to detect true differences between simulated communities with a known effect size using various alpha and beta diversity metrics.
+Although all methods of controlling for uneven sampling effort had an acceptable false detection rate when samples were randomly assigned to two treatment groups, rarefaction was consistantly able to contorol for differences in sampling effort when sequencing depth was confounded with treatment group.
+Finally, the statistical power to detect differences in alpha and beta diversity metrics was consistently the highest when the data were rarefied.
+These simulations underscore the importance of rarefying the number of sequences across samples in amplicon sequencing analyses.
+
 ## Importance
+
+Sequencing of 16S rRNA gene sequences has become a fundamental tool for understanding the diversity of microbial communities and the factors that affect that diversity.
+Because of technical challenges, it is common to observe wide variation in the number of sequences that are collected from different samples.
+Because most of the diversity metrics used by microbial ecologists are affected by sampling effort, tools are needed to control for the uneven levels of sampling.
+This simulation-based analysis shows that despite ongoing controversy, rarefaction is the most robust approach to controlling for uneven sampling effort.
+I contend that the controversy was started because of confusion over how rarefaction is performed and because methods have been borrowed from other fields that make important assumptions that are not correct with microbial community data.
 

### 2022-11-11 submission/manuscript.tex
First full version of Abstract and Importance

@@
 
-\hypertarget{rarefy-your-data}{%
-\section{Rarefy your data}\label{rarefy-your-data}}
+\hypertarget{rarefaction-is-currently-the-best-approach-to-control-for-uneven-sampling-effort-in-amplicon-sequencing-analyses}{%
+\section{Rarefaction is currently the best approach to control for
+uneven sampling effort in amplicon sequencing
+analyses}\label{rarefaction-is-currently-the-best-approach-to-control-for-uneven-sampling-effort-in-amplicon-sequencing-analyses}}
 
@@
 
+How to control for uneven sampling effort in amplicon sequencing
+analyses has become a contentious question in the microbial ecology
+field since it is common to find as much as 100-fold variation in the
+number of 16S rRNA gene sequences found across samples in a study. Some
+have argued that rarefaction is inadmissible because it omits valid data
+and reduces the statistical power to detect differences between
+treatment groups. A number of alternative approaches have been developed
+to normalize and rescale the data that purport to be invariant to the
+number of observations. I generated community distributions based on 12
+published datasets where I was able to assess the sensitivity of
+different methods to uneven sampling effort. Rarefaction was the only
+method that could control for variation in uneven sampling effort when
+measuring commonly used alpha and beta diversity metrics. Next, I
+compared the false detection rate and power to detect true differences
+between simulated communities with a known effect size using various
+alpha and beta diversity metrics. Although all methods of controlling
+for uneven sampling effort had an acceptable false detection rate when
+samples were randomly assigned to two treatment groups, rarefaction was
+consistantly able to contorol for differences in sampling effort when
+sequencing depth was confounded with treatment group. Finally, the
+statistical power to detect differences in alpha and beta diversity
+metrics was consistently the highest when the data were rarefied. These
+simulations underscore the importance of rarefying the number of
+sequences across samples in amplicon sequencing analyses.
+
 \hypertarget{importance}{%
 \subsection{Importance}\label{importance}}
+
+Sequencing of 16S rRNA gene sequences has become a fundamental tool for
+understanding the diversity of microbial communities and the factors
+that affect that diversity. Because of technical challenges, it is
+common to observe wide variation in the number of sequences that are
+collected from different samples. Because most of the diversity metrics
+used by microbial ecologists are affected by sampling effort, tools are
+needed to control for the uneven levels of sampling. This
+simulation-based analysis shows that despite ongoing controversy,
+rarefaction is the most robust approach to controlling for uneven
+sampling effort. I contend that the controversy was started because of
+confusion over how rarefaction is performed and because methods have
+been borrowed from other fields that make important assumptions that are
+not correct with microbial community data.
 

### 2023-06-02 submission/manuscript.Rmd
Editing

--- 2022-11-11 submission/manuscript.Rmd
@@
  - \usepackage[shortcuts]{extdash}
+  - \usepackage[document]{ragged2e}
 ---
@@
 
-# Rarefy your data
+# Rarefaction is currently the best approach to control for uneven sequencing effort in amplicon sequence analyses
 
@@
 
-**Running title:** Rarefy your data
+**Running title:** Use rarefaction
 
@@
 
+Considering it is common to find as much as 100-fold variation in the number of 16S rRNA gene sequences found across samples in a study, researchers need to control for the effect of uneven sequencing effort.
+How to do this has become a contentious question.
+Some have argued that rarefaction is "inadmissible" because it omits valid data.
+A number of alternative approaches have been developed to normalize and rescale the data that purport to be invariant to the number of observations.
+I generated community distributions based on 12 published datasets where I was able to assess the ability of multiple methods to control for uneven sequencing effort.
+Rarefaction was the only method that could control for variation in uneven sequencing effort when measuring commonly used alpha and beta diversity metrics.
+Next, I compared the false detection rate and power to detect true differences between simulated communities with a known effect size using various alpha and beta diversity metrics.
+Although all methods of controlling for uneven sequencing effort had an acceptable false detection rate when samples were randomly assigned to two treatment groups, rarefaction was consistently able to control for differences in sequencing effort when sequencing depth was confounded with treatment group.
+Finally, the statistical power to detect differences in alpha and beta diversity metrics was consistently the highest when using rarefaction.
+These simulations underscore the importance of using rarefaction to normalize the number of sequences across samples in amplicon sequencing analyses.
+
 ## Importance
+
+Sequencing 16S rRNA gene fragments has become a fundamental tool for understanding the diversity of microbial communities and the factors that affect their diversity.
+Due to technical challenges, it is common to observe wide variation in the number of sequences that are collected from different samples within the same  study.
+However, the diversity metrics used by microbial ecologists are sensitive to differences in sequencing effort.
+Therefore, tools are needed to control for the uneven levels of sequencing.
+This simulation-based analysis shows that despite a longstanding controversy, rarefaction is the most robust approach to control for uneven sequencing effort.
+The controversy started because of confusion over the definition of rarefaction and violation of assumptions that are made by methods that have been borrowed from other fields.
+Microbial ecologists should use rarefaction.
 
@@
 
-The ability to generate millions sequence reads from 16S rRNA gene amplicons and metagenomic datasets has allowed researchers to multiplex multiple samples on the same sequencing run by pooling separate PCRs that can be deconvoluted later based on index (aka barcode) sequences that embeded into the primer sequence [@Kozich2013; @Caporaso2010].
+The ability to generate millions sequence reads from 16S rRNA gene amplicons and metagenomic datasets has allowed researchers to multiplex multiple samples on the same sequencing run by pooling separate PCRs that can be deconvoluted later based on index (aka barcode) sequences that are embedded into the sequence of the PCR primers [@Sogin2006; @Caporaso2010; @Kozich2013].
 Unfortunately, it is common to observe variation in the number of sequence reads from each sample vary by as much as 100-fold (e.g, see Figure S1).
-This occurs because pooling of DNA from multiple PCRs is frought with numerous opportunities for technichal errors to compound leading to a skewed distribution.
-Aside from developing better methods of pooling DNA, the question of how to control for uneven sampling effort in microbial ecology studies has become controversial.
+This occurs because pooling of DNA from multiple PCRs is fraught with numerous opportunities for technical errors to compound leading to a skewed distribution.
+Aside from developing better methods of pooling DNA, the question of how to control for uneven sequencing effort in microbial ecology studies has become controversial.
 
-The practice of rarefaction has been commonly used in ecology for more than 50 years as a tool to control for uneven sampling effort across experimental replicates [@Sanders1968; @Hurlbert1971].
+The practice of rarefaction has been commonly used in ecology for more than 50 years as a tool to control for uneven sequencing effort across experimental replicates [@Sanders1968; @Hurlbert1971].
 Microbial ecologists have used it to compare 16S rRNA gene sequence data for the past 25 years [@Dunbar1999; @Hughes2001; @Moyer1998].
-When data are rarefied, the investigator selects a desired threshold of sampling effort and removes any samples below that threshold.
+With rarefaction the investigator selects a desired threshold of sequencing effort and removes any samples below that threshold.
 They then randomly select that many sequences, with replacement from each sample.
-Based on the obseved sequence counts, the researcher can then calcuate alpha diversity metrics including richness and diversity indices or beta divesity metrics such as a Jaccard or Bray-Curtis dissimilarity index.
-I refer to this single sampling as a subsample; this method is implemented as the sub.sample function in mothur [@Schloss2009] and the rrarefy function in the vegan R package [@Oksanen2022].
-Rarefaction repeats the subsampling a large number of times (e.g., 100 or 1,000) and calculates the mean of the alpha or beta diersity metric over those subsamplings; rarefaction is implemented in mothur using the summary.single and dist.shared functions [@Schloss2009] and with the vegan R package using the rarefy or avgdist fuctions [@Oksanen2022].
-Rarefaction effectively tells a researcher what an alpha or beta diversity metric would have been for a collection of samples if they were all evenly sampled.
-Although a closed form equation exists to calcualte the expected richness [@Hurlbert1971], it is computationally easier to empirically calculate rarefied richness and other alpha and beta diversity metrics.
+Based on the observed sequence counts, the researcher can then calculate alpha diversity metrics including richness and diversity indices or beta diversity metrics such as a Jaccard or Bray-Curtis dissimilarity index.
+I refer to this single sampling as a subsample; this method is implemented as the `sub.sample` function in mothur [@Schloss2009] and the `rrarefy` function in the `vegan` R package [@Dixon2003].
+Rarefaction repeats the subsampling a large number of times (e.g., 100 or 1,000 times) and calculates the mean of the alpha or beta diversity metric over those subsamplings; rarefaction is implemented in mothur using the `summary.single` and `dist.shared` functions [@Schloss2009] and with the `vegan` R package using the `rarefy` or `avgdist` functions [@Dixon2003].
+Rarefaction effectively tells a researcher what an alpha or beta diversity metric would have been for a collection of samples if they were all sequenced to the same depth.
+Although a closed form equation exists to calculate the expected richness [@Hurlbert1971], it is computationally easier to empirically calculate richness and other alpha and beta diversity metrics by rarefaction.
 
-In 2014, McMurdie and Holmes [@McMurdie2014] announced that rarefaction of microbial community data is "statistically inadmissible" because it omits valid data.
+[@Schloss2023wnwn]
+Confusion between rarefy and rarefaction
+***In 2014, McMurdie and Holmes [@McMurdie2014] announced that rarefaction of microbial community data is "statistically inadmissible" because it omits valid data.
 In their simulations, they observed that the use of rarefaction reduced the statistical power to compare communities based on beta diversity metrics.
@@
 The simulation with the greatest depth of coverage obtained an average of 10,000 sequences per sample.
-Analysis of the R code that they used indicates that the simulations were actually only based on a single subsample and not truly rarefied data.
+Analysis of the R code that they used indicates that the simulations were actually only based on a single subsample and not traditional rarefaction.
 Furthermore, the authors removed any OTU from the simulated communities that was not observed in three or more samples.
 This is a frequently used approach for analyzing microbiome data [@Bokulich2013].
-Unfortunately, it removes valid data, skews the distribution of the community, and removes true patchiness from the representation of the communities [@Schloss2020].
-Finally, the analysis did not include analysis of alpha diersity metrics and chose to assess beta diversity metrics based on clustering quality of samples rather than using the more frequently used approach of employing non-parametric multivariate analysis of variance [@McArdle2001].
-Regardless of these caveats, McMurdie and Holmes recommended the use of variance stabilizing transformations such as those applied for gene expression analysis such as that found in the DeSeq R package.
+Unfortunately, it removes valid data, skews the distribution of the community, and removes true patchiness from the representation of the communities [@Schloss2023sing].
+Finally, the analysis did not include analysis of alpha diversity metrics and chose to assess beta diversity metrics based on clustering quality of samples rather than using the more frequently used approach of employing non-parametric multivariate analysis of variance [@McArdle2001].
+Regardless of these caveats, McMurdie and Holmes recommended the use of variance stabilizing transformations such as those applied for gene expression analysis such as that found in the DESeq and DESeq2 R packages.***
 
-Beyond the critiques from McMurdie and Holmes others have developed alternative approaches to controlling for uneven sampling effort in amplicon sequencing studies.
+**Since McMurdie and Holmes published their simulations, others have developed alternative approaches to controlling for uneven sequencing effort in amplicon sequencing studies.
 For alpha diversity metrics, Willis used toy datasets to demonstrate that one could estimate the richness for each sample in a dataset and use those values for statistical comparisons [@Willis2019].
-Non-parametric estimators of richness and diversity [@Chao2003; @Chao2016] and parametric estimators of richness [Willis2015] that have been used in microbial ecology studies.
+Non-parametric estimators of richness and diversity [@Chao2003; @Chao2016] and parametric estimators of richness [@Willis2015] that have been used in microbial ecology studies.
 For beta diversity metrics at least four approaches have been pursued.
 First, one could use relative abundance values where the observed number of sequences in an OTU is divided by the total number of sequences in the sample [@Lin2020].
-Second, normalization strategies have been developed where the relative abundance is mutiplied by the size of the minimum desired sampling effort and fractional values are reapportioned to the OTUs to obtain integer values [@Beule2020; @Paulson2013].
-Third, a variety of center log-ratio methods have been developed where the compositional nature of the OTU counts is removed and used to calculated Euclidean distances (aka Aitchison distances) [@Quinn2019; @Paulson2013; @Lin2020; @Martino2019; @Costea2014.
-This strategy is purported to control for uneven sampling effort [@Martino2019; @Quinn2018]; however, some have noticed that this feature breaks down under certain conditions [@teBeest2021].
-Finally, as mentioned above, variance stabilization transformations have been recommended to generate values that can be used to calcualte Euclidean distances [@McMurdie2014].
+Second, normalization strategies have been developed where the relative abundance is multiplied by the size of the minimum desired sequencing effort and fractional values are reapportioned to the OTUs to obtain integer values [@Beule2020; @Paulson2013].
+Third, a variety of center log-ratio methods have been developed where the compositional nature of the OTU counts is removed and used to calculated Euclidean distances (aka Aitchison distances) [@Quinn2019; @Paulson2013; @Lin2020; @Martino2019; @Costea2014].
+This strategy is purported to control for uneven sequencing effort [@Martino2019; @Quinn2018]; however, some have noticed that this feature breaks down under certain conditions [@teBeest2021].
+Finally, as mentioned above, variance stabilization transformations have been recommended to generate values that can be used to calculate Euclidean distances [@McMurdie2014].**
 
-The ongoing controversy over the use of rarefaction and the development of numerous strategies to control for uneven sampling effort caused me to question how these methods compared to each other.
+The ongoing controversy over the use of rarefaction and the recent development of alternative strategies to control for uneven sequencing effort caused me to question how these methods compared to each other using a simulation framework that overcame the issues with the original study.
 My analysis included 16S rRNA gene sequence data from from 12 studies that characterized the variation in bacterial communities from diverse environments (Table 1 and Figure S1).
 The sequences were assigned to OTUs using a standard pipeline and their frequencies and the number of sequences found in each sample were used to generate simulated communities and treatment effects.
-For each dataset and simulation, 100 replicate datasets were generated and used as input to each of the strategies for controlling for uneven sampling effort.
+For each dataset and simulation, 100 replicate datasets were generated and used as input to each of the strategies for controlling for uneven sequencing effort.
 My overall conclusion was that rarefaction outperformed the alternative strategies.
@@
 
-***Without rarefaction, metrics of alpha diversity are sensitive to sampling effort.*** To test the sesitivity of various approaches of measuring alpha diversity to sampling effort, I generated null models for each dataset.
-Under a null model, each community from the same dataset would be expected to have the same alpha diversity regardless of the sampling effort.
+***Without rarefaction, metrics of alpha diversity are sensitive to sequencing effort.*** To test the sensitivity of various approaches of measuring alpha diversity to sequencing effort, I generated null models using OTU counts and sequencing depths from 12 studies.
+Under a null model, each community from the same dataset would be expected to have the same alpha diversity regardless of the sequencing effort.
 I measured the richness of the communities in each dataset without any correction, using scaled ranked subsampling (SRS) normalized OTU counts, with estimates based on non-parametric and parametric approaches, and using rarefaction (e.g.
@@
 Next, I assessed diversity using the Shannon diversity index and the inverse Simpson diversity index without any correction, using normalized OTU counts, and rarefaction; I also used a non-parametric estimator of Shannon diversity.
-The correlation between sampling depth and the diversity metric was not as strong as it was for richness and the inverse Simpson diversity values were less sensitive than the Shannon diversity values; however, the correlation to the rarefied diversity metrics were the lowest for all of the metrics and studies (Figure 1A).
-The rarefied alpha-diversity metrics consistently demonstrated a lack of sensitivity to sampling depth.
+The correlation between sequencing depth and the diversity metric was not as strong as it was for richness and the inverse Simpson diversity values were less sensitive than the Shannon diversity values; however, the correlation to the diversity metrics calculated with rarefaction were the lowest for all of the metrics and studies (Figure 1A).
+The alpha-diversity metrics calculated with rarefaction consistently demonstrated a lack of sensitivity to sequencing depth.
 
-***Without rarefaction, metrics of beta diversity are sensitive to sampling effort.*** To test the sesitivity of various approaches of measuring beta diversity to sampling effort, I used the same null models used for studying the sensitivity of alpha diversity.
+***Without rarefaction, metrics of beta diversity are sensitive to sequencing effort.*** To test the sensitivity of various approaches of measuring beta diversity to sequencing effort, I used the same null models used for studying the sensitivity of alpha diversity.
 Under a null model, the ecological distance between any pair of samples would be the same regardless of the difference in the number of sequences observed in each sample (e.g., Figure S3).
@@
 The Jaccard distance coefficient is the fraction of OTUs that are unique to either community and does not account for the abundance of the OTUs.
-Jaccard distances were calcualted using the uncorrected OTU counts, with rarefaction, relative abundances, and following normalization using cumulative sum scaling (CSS) and SRS.
-Only the rarefied distances showed a lack of sensitivity to sampling effort (Figure 1B).
+Jaccard distances were calculated using the uncorrected OTU counts, with rarefaction, relative abundances, and following normalization using cumulative sum scaling (CSS) and SRS.
+Only the distances calculated with rarefaction showed a lack of sensitivity to sequencing effort (Figure 1B).
 Second, I analyzed the sensitivity of the Bray-Curtis distance coefficient, which is a popular metric that incorporates the abundance of each OTU.
-Similar to what I observed with the Jaccard coefficient, only the rarefied data showed a lack of sensitivity to sampling effort (Figure 1B).
-Third, I calcualted the Euclidean distance on raw OTU counts where the central log-ratio (CLR) was calculated (i.e., Aitchison distances) by ignoring OTUs in samples with zero counts (Robust CLR), adding a pseudocount of one to all OTU counts prior to calculating the CLR (One CLR), adding a pseudocount of one divided by the total number of sequences obtained for the community (Nudge CLR), and imputing the value of zero counts (Zero CLR).
-The Aitchison distances were all strongly sensitive to sampling effort (Figure 1B).
-Finally, I used the variance-stabilization transformation (VST) from DeSeq2 prior to calculating Euclidean distances.
-Again, there was a strong sensitivity to sampling effort (Figure 1B).
-Although Euclidean distances are not typically used on raw or rarefied count data in ecology, rarefied Euclidean distances were not sensitive to sampling effort.
-Across each of the beta diversity metrics and approaches used to account for uneven sampling effort and sparsity, rarefaction was the least sensitive approach to differences in sampling effort.
+Similar to what I observed with the Jaccard coefficient, only the data normalized with rarefaction showed a lack of sensitivity to sequencing effort (Figure 1B).
+Third, I calculated the Euclidean distance on raw OTU counts where the central log-ratio (CLR) was calculated (i.e., Aitchison distances) by ignoring OTUs in samples with zero counts (Robust CLR), adding a pseudo-count of one to all OTU counts prior to calculating the CLR (One CLR), adding a pseudo-count of one divided by the total number of sequences obtained for the community (Nudge CLR), and imputing the value of zero counts (Zero CLR).
+The Aitchison distances were all strongly sensitive to sequencing effort (Figure 1B).
+Finally, I used the variance-stabilization transformation (VST) from DESeq2 prior to calculating Euclidean distances.
+Again, there was a strong sensitivity to sequencing effort (Figure 1B).
+Although Euclidean distances are not typically used on raw or rarefaction-normalized count data in ecology [@Legendre2012], Euclidean distances calculated with rarefaction were not sensitive to sequencing effort.
+Across each of the beta diversity metrics and approaches used to account for uneven sequencing effort and sparsity, rarefaction was the least sensitive approach to differences in sequencing effort.
 
-***Rarefaction limits the detection of false positives when sampling effort and treatment group are confounded.*** Next, I investigated the impact of the various strategies and metrics on falsely detecting a significant difference using the the same communities generated from the null model in the analysis of alpha and beta diversity metrics.
+***Rarefaction limits the detection of false positives when sequencing effort and treatment group are confounded.*** Next, I investigated the impact of the various strategies and metrics on falsely detecting a significant difference using the the same communities generated from the null model in the analysis of alpha and beta diversity metrics.
 To test for differences in alpha and beta diversity I used the non-parametric Wilcoxon test and non-parametric permutation-based multivariate analysis of variance (PERMANOVA).
 First, I employed an unbiased null treatment model to measure the false detection rate, which should not have meaningfully differed from 5%.
-Indeed, for each dataset and alpha and beta diversity metric and strategy for accounting for uneven sampling, approximately 5% of the tests yielded a significant result (Figure 2).
+Indeed, for each dataset and alpha and beta diversity metric and strategy for accounting for uneven sequencing, approximately 5% of the tests yielded a significant result (Figure 2).
 Second, I employed a biased null treatment model where the treatment group was completely confounded with the number of sequences in each sample.
-Under this model, only the rarefied data consistently resulted in a 5% false positive rate for alpha and beta diversity metrics (Figure 2).
-These results aligned with the observed sensitivity of alpha and beta diversity metrics to sampling effort and underscore the value of rarefaction.
+Under this model, only the data normalized with rarefaction consistently resulted in a 5% false positive rate for alpha and beta diversity metrics (Figure 2).
+These results aligned with the observed sensitivity of alpha and beta diversity metrics to sequencing effort and underscore the value of rarefaction.
 
-***Rarefaction preserves the statistical power to detect differences between treatment groups.*** To assess the impact of different approaches to control for uneven sampling effort I performed two additional simulations.
-In the first simulation, I implemented a skewed abundance distribution model to create two treatment groups for each datasets that were each populated with half of the samples each with the same number of sequences as the samples had in the observed data.
+***Rarefaction preserves the statistical power to detect differences between treatment groups.*** To assess the impact of different approaches to control for uneven sequencing effort I performed two additional simulations.
+In the first simulation, I implemented a skewed abundance distribution model to create two treatment groups for each dataset that were each populated with half of the samples each with the same number of sequences as the samples had in the observed data.
+The two treatment groups varied in their structure such that one had the same abundances as the null distribution above and the other had 10% of its OTUs randomly selected to increase their counts by 5%.
 The power to detect differences in richness between the two simulated treatment groups by all approaches was low (Figure 4A).
 This was likely because the approach for generating the perturbed community did not necessarily change the number of OTUs in each treatment group.
-Regardless, the simulations testing differencse in richness using the Rice and Stream datasets had the greatest power when the richness data were rarefied.
+Regardless, the simulations testing differences in richness using the Rice and Stream datasets had the greatest power when the richness data were calculated with rarefaction.
 To explore this further, a richness-adjusted community model was created by removing 3% of the OTUs from a null model model.
-As suggested by the first simulation, the rarefied richness data had a higher statistical power than the other approaches when measuring richness (Figure 5).
-The simulations testing the power to detect differences in Shannon diversity also showed that rarefied data performed other methods (Figure 4A).
+As suggested by the first simulation, the richness data  calculated with rarefaction had a higher statistical power than the other approaches when measuring richness (Figure 5).
+The simulations testing the power to detect differences in Shannon diversity also showed that rarefaction performed better than the other methods (Figure 4A).
 When testing for differences in the Inverse Simpson diversity index the the difference between rarefaction and the other methods was negligible (Figure 4A).
 For tests of beta diversity I found that rarefaction was the most reliable approach to maintain statistical power to detect differences between two communities.
-Among the tests using the Jaccard and Bray-Curtis metrics, raw count data and CSS normalized data had little power relative to rarefied, relative abundance, and SRS normalized data.
-The differences in power between rarefied, relative abundance, and SRS normalized data was small, but if there were differences, the power obtained using rarefied data was greater than the other methods.
-Among the tests using Euclidean distances, using raw counts and CLR and DeSeq2 transformed data had little power relative to the distances calcualted using rarefied and relative abundance data.
+Among the tests using the Jaccard and Bray-Curtis metrics, raw count data and CSS normalized data had little power relative to using rarefaction, relative abundance, and SRS to normalize the uneven sequencing depths.
+The differences in power between counts normalized with rarefaction, relative abundance, and SRS data was small, but if there were differences, the power obtained using rarefaction was greater than the other methods.
+Among the tests using Euclidean distances, using raw counts and CLR and DESeq2 transformed data had little power relative to the distances calculated using rarefaction and relative abundances.
 This power-based analysis of the simulated communities using different methods of handling uneven sample sizes demonstrated the value of rarefaction for preserving the statistical power to detect differences between treatment groups for measures of alpha and beta diversity.
@@
   select(nice_name, fold_difference) %>%
-  mutate(loss = round(100* (1 - 1/fold_difference), 1)) %>%
+  mutate(loss = 100 - round(100* (1 - 1/fold_difference), 1)) %>%
   slice_max(loss)
@@
 
-***Increased rarefaction depth reduces intra-sample variation in alpha and beta diversity.*** Once concern with rarefying communities is the perceived loss of sequencing information when more a large fraction of data appears to be removed when the community with the greatest sequencing depth is rarefied to the size of the community with the least (e.g., `r paste0(loss$loss, "% with the ", loss$nice_name, " dataset")`).
-To assess the sensitivity of alpha and beta diversity metrics to rarefaction depth, I again used the dataset generated using the null models, but rarefied each community to varying sampling depths (Figure 6).
-The richness values increased with sampling effort as rare OTUs would continue be detected.
-In contrast, the Shannon diversity and Bray-Curtis values plateaued with increased sampling effort.
-This result was expected since increased sampling would lead to increased precision in the measured abundance of OTUs.
+***Increased rarefaction depth reduces intra-sample variation in alpha and beta diversity.*** Once concern with using rarefaction is the perceived loss of sequencing information when a large fraction of data appears to be removed when the community with the greatest sequencing depth is sampled to the size of the community with the least (e.g., the smallest sample in the `r loss$nice_name` dataset had `r loss$loss`% of sequences that were in the largest sample).
+To assess the sensitivity of alpha and beta diversity metrics to rarefaction depth, I again used the dataset generated using the null models, but used rarefaction with each community to varying sequencing depths (Figure 6).
+The richness values increased with sequencing effort as rare OTUs would continue be detected.
+In contrast, the Shannon diversity and Bray-Curtis values plateaued with increased sequencing effort.
+This result was expected since increased sequencing would lead to increased precision in the measured relative abundance of the OTUs.
 Next, I measured the coefficient of variation (i.e., the mean divided by the standard deviation) between samples for richness, Shannon diversity, and Bray-Curtis distances.
-Although the richness values appeared to increased unbounded with smapling effort, the coefficient of variation for each dataset was relatively stable.
-In general, the coefficient of variation increased slightly with sampling depth only to decline once smaller samples were removed from the analysis at higher sampling depths.
-Interestingly, the coefficient of variation between Shannon diveristy values decreased towards zero with increased sampling effort and the coefficient of variation between Bray-Curtis distances tended to increased.
+Although the mean richness appeared to increased unbounded with sequencing effort, the coefficient of variation for each dataset was relatively stable.
+In general, the coefficient of variation increased slightly with sequencing depth only to decline once smaller samples were removed from the analysis at higher sequencing depths.
+Interestingly, the coefficient of variation between Shannon diversity values decreased towards zero with increased sequencing effort and the coefficient of variation between Bray-Curtis distances tended to increase.
 Regardless, the coefficients of variation were relatively small.
+This analysis indicates that there are benefits to increased sequencing depths.
 
@@
 The median coverage for each dataset ranged between `r oxford_comma(obs_median$med)`% for the `r oxford_comma(capitalize(obs_median$dataset))` datasets, respectively (Figure 7).
-When I rarefied each dataset to the size of the smallest community in the dataset, with the exception of the `r oxford_comma(capitalize(rare_median$dataset))` datasets, the median coverage for the rarefied communities was still greater than 90%.
+When I used a rarefaction threshold with each dataset at the size of the smallest community in the dataset, with the exception of the `r oxford_comma(capitalize(rare_median$dataset))` datasets, the median coverage with rarefaction was still greater than 90%.
 These results suggest that most studies had a level of sequencing coverage that aligned with the diversity of the communities.
@@
 As suggested by the simulated coverages curve in Figure 7, these estimates are conservative.
-Regardless, the sequencing effort required to acheive higher sequencing depth would likely limit the number of samples that could be sequenced when controlling for costs.
-Although it may be disconcerting to rarefy to a sequencing depth that is considerably lower than that obtained for the best sequenced community in a dataset, sequencing coverage for many environments is probably adequate even at the lower sequencing depth.
-Of course, the results above have demonstrated that rarefaction is necessary to avoid problems with making inferences.
+Regardless, the sequencing effort required to achieve higher sequencing depth would likely limit the number of samples that could be sequenced when controlling for costs.
+Although it may be disconcerting to use rarefaction to normalize to a sequencing depth that is considerably lower than that obtained for the best sequenced community in a dataset, sequencing coverage for many environments is probably adequate even at the lower sequencing depth.
+Of course, regardless of the concerns surrounding the choice of the rarefaction depth, the results throughout this study demonstrate that rarefaction is necessary to avoid reaching incorrect inferences.
 
@@
 
-Over the past decade, the question of whether to rarefy microbial community sequence data has become controversial.
-The analyses I presented here strongly indicate that rarefaction is necessary to control for uneven sampling effort when comparing communities based on alpha and beta diversity indices.
+Over the past decade, the question of whether to use rarefaction with microbial community sequence data has become controversial.
+The analyses I presented here strongly indicate that rarefaction is necessary to control for uneven sequencing effort when comparing communities based on alpha and beta diversity indices.
 Compared to all other methods, rarefaction removed the correlation between sequencing depth and alpha or beta diversity metrics when the sequencing depth varied by as much as 97-fold across samples.
-I showed that this correlation could lead falsely detecting differences between treatment groups if sampling depth and sequencing effort are confounded [e.g., @Morris2013].
-The correlation with sampling effort leads to an artificial increase in the variation between samples and a reduced power to detect true differences in alpha and beta diversity.
-For these reasons, rarefaction is a valuable tool to control for uneven sampling effort until improved statistical procedures are developed or it becomes possible to more evenly distribute sequencing effort across samples.
+I showed that this correlation could lead falsely detecting differences between treatment groups if sequencing depth and sequencing effort are confounded.
+The correlation with sequencing effort leads to an artificial increase in the variation between samples and a reduced power to detect true differences in alpha and beta diversity.
+For these reasons, rarefaction is a valuable tool to control for uneven sequencing effort until improved statistical procedures are developed or it becomes possible to more evenly distribute sequencing effort across samples.
 
-The primary alternative to rarefaction that many advise is to estimate alpha diversity has been to use non-parametric or parametric methods using raw data and to then compare the estimates [@Willis2019].
-My data demonstrates that such approaches are limited for several reasons.
-First, it has been widely observed that non-parametric richness esimates such as ACE and Chao1 are sensitive to sampling effort.
-Therefore, these strategies do not, in practice, remove the effects of sampling effort.
-Second, parametric approaches, such as those implemented in the breakaway R package, generate confidence intervals that are likely to include the true richness and theoretically shrink with increased sampling effort.
-For most samples, the confidence intervals around the estimates are too wide to be meaningful, again leading to an inability to remvoe the effects of sampling effort.
+The primary alternative to rarefaction for measuring alpha diversity is to estimate the metric using non-parametric or parametric models with raw counts and to then compare the estimates [@Hughes2001; @Willis2019].
+My results demonstrate that such approaches are limited for several reasons.
+First, non-parametric richness estimates such as ACE and Chao1 are sensitive to sequencing effort.
+Therefore, these strategies do not, in practice, remove the effects of sequencing effort.
+Second, parametric approaches, such as those implemented in the `breakaway` R package, generate confidence intervals that are likely to include the true richness and theoretically shrink with increased sequencing effort.
+Yet for most samples, the confidence intervals around the estimates are too wide to be meaningful, again leading to an inability to remove the effects of sequencing effort.
 Third, it has become an increasingly common practice for researchers to remove sequences that are rare in a sample (e.g., those that appear once).
-Although that approach was not taken in this study, removing rare sequences would skew the distribtion of sequences and OTUs leading to a distortion of the measurement of alpha diversity [@Schloss2020].
+Although that approach was not taken in this study, removing rare sequences would skew the distribution of sequences and OTUs leading to a distortion of the measurement of alpha diversity [@Schloss2023sing].
 The effects of removing rare sequences would vary across samples depending on the number of sequences in each sample.
-One interesting result of this analysis was the demonstration that as metrics that depend less on rare taxa are used, the effect of uneven sampling effort was reduced.
-For example, richness counts a sequence appearing once as much as sequence appearing 1,000 times, while Shannon diversityindex places more emphasis on the more abundant sequence, and the inverse Simpson index even more.
-Although normalizing communities to a common number of sequences is also suggested (e.g., SRS normalization) to control for uneven sampling effort, the current analysis indicates that its performance does not meet that of rarefaction.
-For analysis of alpha diversity metrics, rarefaction is the most effective and consistent approach to control fo uneven sampling effort.
+One interesting result of this analysis was the demonstration that as metrics that depend less on rare taxa are used, the effect of uneven sequencing effort was reduced.
+For example, richness counts a sequence appearing once as much as sequence appearing 1,000 times, while the Shannon diversity index places more emphasis on the more abundant sequence, and the inverse Simpson index even more.
+Although normalizing communities to a common number of sequences is also suggested (e.g., SRS normalization) to control for uneven sequencing effort, the current analysis indicates that its performance does not meet that of rarefaction.
+For analysis of alpha diversity metrics, rarefaction is the most effective and consistent approach to control for uneven sequencing effort.
 
 Use of relative abundances, normalized counts, variance stabilizing transformations, and centered log ratios have each been recommended as more robust alternatives to rarefaction.
-Again, the only approach that consistently removed the effects of uneven sampling effort was rarefaction.
+Again, the only approach that consistently removed the effects of uneven sequencing effort was rarefaction.
 Most of the recommendations borrow techniques from methods for identifying differential gene expression.
-Unfortunately, there is an important but underappreciated difference between gene expression and community data.
+Unfortunately, there is an important but under appreciated difference between gene expression and community data.
 This is the interpretation of unobserved data.
-For gene expression analysis in a single organism the lack of any sequencing data for a gene would indicate that its expression was below the limit of detection.
-Sequencing the same organism under multiple conditions would not lead to a seemingly unbouned number of genes in the organims.
+For gene expression analysis in a single organism the lack of any sequencing data for a gene would indicate that although the gene was present, its expression was below the limit of detection.
+Sequencing the same organism under multiple conditions would not lead to a seemingly unbounded number of genes in the organisms.
 Rather, the number of genes has a definite limit that is knowable from the genome sequence.
 With microbiome data, an unobserved sequence could mean that the organism was present, but below the limit of detection or that the organism was missing.
-Because we have yet to exhaustively sample any community in the same way we have sequenced a single genome it is unreliable to impute the presence of all organisms.
+Because we have yet to exhaustively sample any community in the same way we have sequenced a single genome, it is unreliable to impute the presence of all organisms.
 Yet, this is exactly what the variance stabilization transformation and most CLR techniques do.
@@
 This result is at odds with the claims by others that the distances are scale invariant [@Martino2019; @Quinn2018].
-Again, rarefaction is the most effective and consistent approach to control for uneven sampling effort when calculating beta diversity metrics.
+Again, rarefaction is the most effective and consistent approach to control for uneven sequencing effort when calculating beta diversity metrics.
 
-Two common critique of rarefaction is that the approach ommits valid data and that the selection of the value to rarefy the sequence data to is arbitrary [@McMurdie2014].
+Two common critiques of rarefaction are that the approach omits valid data and that the selection of rarefaction threshold is arbitrary [@McMurdie2014].
 I disagree with both critiques.
-All of the data are used to calculate the mean value of the rarefied metrics; none of it is excluded.
+All of the data are used to calculate the mean value of the metrics after repeatedly subsampling the data.
 When the dataset is subsampled, every sequence has a random chance of being included in the calculated metric.
-When that subsampling is repeated a large number of times (e.g., 100 or 1000) the risks of ignoring or oversampling rare taxa are mititgated.
-Furthermore, it is curious that the study making the original critique removed sequences that did not appear in at least three samples or had an abundance of one in any sample.
-A parallel analysis to this study has demonstrated that many of these sequences are likely valid and that removal of rare sequences can bias alpha and beta diversity metrics and reduce statistical power [@Schloss2020].
-As for the second critism, I would resist the claim that the selection of the rarefaction threshold is arbitrary.
-In practice, there is a tradeoff between sampling breadth and depth.
+When that subsampling is repeated a large number of times (e.g., 100 or 1000) the risks of ignoring or oversampling rare taxa are mitigated.
+Furthermore, it is curious that the study making the original critique removed rare sequences.
+A parallel analysis to this study has demonstrated that many of these sequences are likely valid and that removal of rare sequences can bias alpha and beta diversity metrics and reduce statistical power [@Schloss2023sing].
+As for the second criticism, I would resist the claim that the selection of the rarefaction threshold is arbitrary.
+In practice, there is a tradeoff between sampling breadth and sequencing depth.
 Greater breadth will increase the statistical power to compare treatment groups and greater depth will increase the resolution to describe the communities.
-My process for picking a break involves looking for a natural break in the distribution of the number of sequences.
-For example, the Lake dataset used in this study had a clear break around 10,000 sequences.
+My personal process for rarefaction threshold involves looking for a natural break in the distribution of the number of sequences.
+For example, the Lake dataset used in this study had a clear break around 10,000 sequences (Figure S1).
 I would also consider what samples are below any break that I select.
-If there were critical samples below the break I would either reduce the break point or I would generate more sequences from those samples.
-Furthermore, regardless of the break point one selects, they could repeat an analysis at multiple thresholds to assess the tradeoff between statistical power and resolution.
-As shown in my analysis of Good's coverage values, most studies obtain an ample level of coverage and would need to increase their sampling depth by 10 fold to increase the coverage by several percentage points.
-I contend that for most studies increased sequencing effort should be applied to improve sampling breadth and statistical power over depth and resolution.
+If there were critical samples below the break I would either reduce the threshold or obtain more sequences from those samples.
+As shown in my analysis of Good's coverage values, most studies obtain an ample level of coverage and would need to increase their sequencing depth by 10-fold to increase the coverage by several percentage points.
+In past work, I have favored distributing the total number of sequences to increase sampling breadth rather than sequencing depth.
 
-The up to 100-fold difference in sample sizes is an unfortunate byproduct of how seqencing libraries are constructed.
-Researchers perform separate PCRs for each sample with unique index (aka, barcode) sequencs that all them to later identify which sample each sequence belongs.
-When the PCRs are pooled efforts are taken to pool the fragments in equimolar ratios.
+The up to 100-fold difference in sample sizes is an unfortunate byproduct of how sequencing libraries are constructed.
+Researchers perform separate PCRs for each sample with unique index (aka, barcode) sequences that allow them to later assign sequences back to the samples that they came from.
+When the PCR products are pooled, efforts are taken to pool the fragments in equimolar ratios.
 Researchers use one of two approaches.
@@
 Clearly, both approaches have limitations that reduce the ability to truly achieve equimolar mixture.
-In addition, for some samples it is common to co-amplify non-specific DNAs which may add to the challenges of achieving equimolar mixtures of the desired amplicons [@Morris2013].
+In addition, for some samples it is common to co-amplify non-specific DNAs which may add to the challenges of achieving equimolar mixtures of the desired amplicons since those sequences will be removed [@Morris2013].
 Regardless, it is clear that better strategies are needed to reduce the variation in the number of sequences generated for each sample.
@@
 All simulations have weaknesses and should be interpreted with caution.
-However, the simulated communites generated and analyzed in this study had the advantage of being designed with known properties including the alpha and beta diversity and the their differences between treatment groups.
-For now, is is perfectly admissible and proper to rarefy microbial community data or risk reaching unwarrented conclusions.
+However, the simulated communities generated and analyzed in this study had the advantage of being designed with known properties including the alpha and beta diversity and the their differences between treatment groups.
+Indeed, is is perfectly admissible and proper to use rarefaction with microbial community data.
+The alternatie is to risk reaching unwarranted conclusions.
 
@@
 **Choice of datasets.** The specific studies used in this study were selected because their data was publicly accessible through the Sequence Read Archive, the original investigators sequenced the V4 region of the 16S rRNA gene using paired 250 nt reads, and my previous familiarity with the data.
-The use of paired 250 nt reads to sequence the V4 region resulted in a near complete two-fold overap of the V4 region resulting in high quality contigs with a low sequencing error rate [@Kozich2013].
+The use of paired 250 nt reads to sequence the V4 region resulted in a near complete two-fold overlap of the V4 region resulting in high quality contigs with a low sequencing error rate [@Kozich2013].
 These data were processed through a standardized sequence curation pipeline to generate operational taxonomic units (OTUs) using the mothur software package [@Kozich2013; @Schloss2009].
@@
 This is a similar approach to that of the IS algorithm described by Ulrich and Gotelli [@Ulrich2010].
-Because the construction of the null models was a stochastic process, 100 replicates were geneated for each dataset.
+Because the construction of the null models was a stochastic process, 100 replicates were generated for each dataset.
 
@@
 In the unbiased model, samples were randomly assigned to one of two treatment groups.
-In the biased treatment model, samples that had more than the median number of seqeunces for a dataset were assigned to one treatment group and the rest were assigned to a second treatment group.
+In the biased treatment model, samples that had more than the median number of sequences for a dataset were assigned to one treatment group and the rest were assigned to a second treatment group.
 Comparison of any diversity metric across the two treatment groups should have only yielded a significant result in 5% of the simulations when testing under a Type I error (i.e., $\upalpha$) of 0.05.
@@
 Communities in the first treatment group were generated by calculating the relative abundance of each OTU across all samples and using those values as the probability of sampling each OTU.
-This probability distribution was sampled until each sample had the same number of sequences that it did in the observed data.
+This probability distribution was sampled, with replacement, until each sample had the same number of sequences that it did in the observed data.
 Samples in the second treatment group were generated by adjusting the relative abundances of the OTUs in the firs treatment group by increasing the relative abundance of 10% of the OTUs by 5%.
-These values were determined after empirically searching for conditions that resulted in a large fraction of the randomizations yieleding a significant result across most of the studies.
+These values were determined after empirically searching for conditions that resulted in a large fraction of the randomizations yielding a significant result across most of the studies.
 Sequences were sampled from the skewed community community until each sample had the same number of sequences that it did in the observed data.
-Under the skewed abundance community model each sample represented a statistical sampling of two communities such that there should not have been a statistically significant difference within a treatment group, but there was between the treatment groups.
-Because the construction of the skewed abundance community model was a stochastic process, 100 replicates were geneated for each dataset.
+Under the skewed abundance community model, each sample represented a statistical sampling of two communities such that there should not have been a statistically significant difference within a treatment group, but there was between the treatment groups.
+Because the construction of the skewed abundance community model was a stochastic process, 100 replicates were generated for each dataset.
 
@@
 Samples in the second treatment group were generated by removing 3% of the OTUs from the dataset and recalculating the relative abundance of the remaining OTUs.
-Sequences were sampled from the richness-adjusted community distrubtion until each sample had the same number of sequences that it did in the observed data.
-Under the richness-adjusted community model each sample represented a statistical sampling of two communities such that there should not have been a statistically significant difference within a treatment group, but there was between the treatment groups.
-Because the construction of the richness-adjusted community model was a stochastic process, 100 replicates were geneated for each dataset.
+Sequences were sampled from the richness-adjusted community distribtion, with replacement, until each sample had the same number of sequences that it did in the observed data.
+Under the richness-adjusted community model, each sample represented a statistical sampling of two communities such that there should not have been a statistically significant difference within a treatment group, but there was between the treatment groups.
+Because the construction of the richness-adjusted community model was a stochastic process, 100 replicates were generated for each dataset.
 
-**Test of statistical significance.** Statistical comparisons of alpha diversity metrics across the simulated treatment groups were performed using the non-parametric two-sample Wilcoxon test as implemented in R.
+**Test of statistical significance.** Statistical comparisons of alpha diversity metrics across the simulated treatment groups were performed using the non-parametric two-sample Wilcoxon test as implemented in `wilcoxon.test` in the `stats` base R package.
 This test was selected because the alpha-diversity metrics tended to not be normally distributed and each dataset required a different transformation to normalize the data.
-Comparisons of beta diversity metrics were performed using the adonis2 function from the vegan (v.`r package_version("vegan")`) R package [@Oksanen2022].
-The adonis2 function implements a non-parametric multivariate analysis of variance using distances matrices as described by McArdle and Anderson [@McArdle2001].
-Throughout this study I used 0.05 as the threshold for asssessing statistical significance.
+Comparisons of beta diversity metrics were performed using the `adonis2` function from the `vegan` (v.`r package_version("vegan")`) R package [@Dixon2003].
+The `adonis2` function implements a non-parametric multivariate analysis of variance using distances matrices [@Anderson2001].
+Throughout this study I used 0.05 as the threshold for assessing the statistical significance of any P-values.
 
-**Power analysis.** The parameters used to design the skewed abundance and richness-adjusted community models were set to impose a known effect size when using rarefied community data.
+**Power analysis.** The parameters used to design the skewed abundance and richness-adjusted community models were set to impose a known effect size when using community data normalized by rarefaction.
 The statistical power to detect these differences was determined by calculating the p-value for each of 100 replicate simulated set of samples from each dataset using the various alpha and beta diversity metrics.
@@
 
-**Alpha diversity calculations.** Various strategies for handling uneven sampling effort were evaluated to identify the best approach for calculating community richness and Shannon and inverse Simpson diversity indeices.
-Raw OTU counts were used as input to calculate sample richness and Shannon and inverse Simpson diversity using mothur [@Schloss2009; @Magurran2004].
+**Alpha diversity calculations.** Various strategies for handling uneven sequencing effort were evaluated to identify the best approach for calculating community richness and Shannon and inverse Simpson diversity indices.
+OTU counts were used as input to calculate sample richness and Shannon and inverse Simpson diversity using mothur [@Schloss2009; @Magurran2004].
 Shannon diversity was calculated as
@@
 The inverse Simpson diversity was calculated as $1/D_{simpson}$.
-In both formulae, $n_i$ is the number of sequences in OTU $i$ and $N$ is the number of sequences in the sample.
+In both formulae, $n_i$ was the number of sequences in OTU $i$ and $N$ iwass the number of sequences in the sample.
 Rarefaction of richness, Shannon diversity and Inverse Simpson diversity values were carried out in mothur.
@@
 The non-parametric bias-corrected Chao1 and ACE richness estimators [@Chao2016] and a non-parametric estimator of the Shannon diversity [@Chao2003] were calculated using raw OTU counts with mothur.
-Parametric estimates of sample richness were calculated using the breakaway (BA) R package (v.`r package_version("breakaway")`)[@Willis2015].
-The current analysis reports both the results from running default model selection procedure and the Poisson model.
+Parametric estimates of sample richness were calculated using the `breakaway` (BA) R package (v.`r package_version("breakaway")`)[@Willis2015].
+My analysis reports both the results from running default model selection procedure and the Poisson model.
 The default model selection returned either the Kemp, Negative Binomial, or Poisson models.
@@
 
-**Beta diversity calculations.** Similar to the alpha diversity calculations, multiple approaches were used to control for uneven sampling effort and calculate beta diversity.
-Raw and OTU counts were used for input to calculate the Jaccard, Bray-Curtis, and Euclidean dissimilarity indices using the vegdist function from the vegan R package (v.`r package_version("vegan")`)[@Oksanen2022].
+**Beta diversity calculations.** Similar to the alpha diversity calculations, multiple approaches were used to control for uneven sequencing effort and calculate beta diversity.
+Raw and OTU counts were used for input to calculate the Jaccard, Bray-Curtis, and Euclidean dissimilarity indices using the `vegdist` function from the `vegan` R package (v.`r package_version("vegan")`)[@Dixon2003].
 The Jaccard index was calculated as
@@
 
-where $n_{A,i}$ and $n_{B,i}$ are the number of sequences observed in OTU $i$ from samples $A$ and $B$, respectively.
-$N_A$ and $N_B$ are the total number of sequences in samples $A$ and $B$, respectively.
-$S_T$ is the total number of OTUs observed between the two samples.
+where $n_{A,i}$ and $n_{B,i}$ were the number of sequences observed in OTU $i$ from samples $A$ and $B$, respectively.
+$N_A$ and $N_B$ were the total number of sequences in samples $A$ and $B$, respectively.
+$S_T$ was the total number of OTUs observed between the two samples.
 The Euclidean distance was calculated as
@@
 
-These metrics were calculated using the relative abundance of each OTU using the vegdist function from vegan.
+These metrics were calculated using the relative abundance of each OTU using the `vegdist` function from the `vegan` R package.
 The relative abundance was calculated as the number of sequences in the OTU (e.g., $n_{A,i}$) divided by the total number of sequences in the sample (e.g., $N_A$).
 
-Rarefied beta-diversity values were calculated using the avgdist function in vegan.
-Briefly, vegan's avgdist function calculates each pairwise dissimilarity index after obtaining a random draw of the same number of sequences from each sample.
+Beta-diversity values generated with rarefaction were calculated using the `avgdist` function in vegan.
+Briefly, `vegan`'s `avgdist` function calculates each pairwise dissimilarity index after obtaining a random draw of the same number of sequences from each sample.
 After obtaining 100 random draws it returns the mean value.
@@
 Three approaches were taken to normalize the number of sequences across samples within a dataset.
-Scaled ranked subsampling (SRS) and cumulative sum scaling (CSS) were used to normalize raw OTU counts using the SRS (v.`r package_version("SRS")`) and metagenomeSeq (v.`r package_version("metagenomeSeq")`) R packages, respectively [@Beule2020; @Paulson2013].
+Scaled ranked subsampling (SRS) and cumulative sum scaling (CSS) were used to normalize raw OTU counts using the `SRS` (v.`r package_version("SRS")`) and `metagenomeSeq` (v.`r package_version("metagenomeSeq")`) R packages, respectively [@Beule2020; @Paulson2013].
 The normalized counts were then used to calculate Jaccard and Bray-Curtis dissimilarity indices.
-Finally, the variance-stabilization transformation (VST) as implemented in the DESeq2 (v.`r package_version("DESeq2")`) R package was used to normalize the data as described by McMurdie and Holmes [@Love2014; @McMurdie2014].
-Because the VST approach generates negative values, which are incompatible with calculating Jaccard and Bray-Curtis dissimilarity values, Euclidean distances were calcualted instead.
+Finally, the variance-stabilization transformation (VST) as implemented in the `DESeq2` (v.`r package_version("DESeq2")`) R package was used to normalize the data as described by McMurdie and Holmes [@Love2014; @McMurdie2014].
+Because the VST approach generated negative values, which are incompatible with calculating Jaccard and Bray-Curtis dissimilarity values, Euclidean distances were calculated instead.
 
-Raw OTU counts were used to calculate centered log ratio values for each OTU, which were then used to calcualte Euclidean distances; such distances are commonly refered to as Aitchison distances.
-Centered log-ratio (CLR) abundances are calculated as:
+Raw OTU counts were used to calculate centered log ratio (CLR) values for each OTU, which were then used to calculate Euclidean distances; such distances are commonly referred to as Aitchison distances.
+CLR abundances are calculated as:
 
 $$
-clr\left(n_j\right) = \left[ \ln\frac{x_{ij}}{g(x_j)}, ..., \ln\frac{x_{S_Tj}}{g(x_j)}\right]
+CLR\left(n_j\right) = \left[ \ln\frac{x_{ij}}{g(x_j)}, ..., \ln\frac{x_{S_Tj}}{g(x_j)}\right]
 $$
 
-where $x_{ij} was the number of sequences observed for OTU $i$ in sample $j$ and $g()$ was the geometric mean $x_j$ was the counts of the $S_T$ OTUs in sample $j$.
+where $x_{ij}$ was the number of sequences observed for OTU $i$ in sample $j$ and $g()$ was the geometric mean, $x_j$ was the count of the $S_T$ OTUs in sample $j$.
 Because the geometric mean is zero if any OTU is absent from a sample, the CLR is undefined when there are unobserved OTUs in a sample.
@@
 The first, Zero CLR, imputed the value of the zeroes based on the observed data using the zCompositions (v.`r package_version("zCompositions")`) R package [@Palarea2015].
-The second, One CLR, added a pseudocount of 1 to the abundance of all OTUs [@Paulson2013; @Lin2020].
-The third, Nudge CLR, added a pseudocount of 1 divided by the total number of sequences in a sample to each OTU in the sample [@Costea2014; @Lin2020].
-The final approach, Robust CLR, removed unobserved OTUs prior to calcualting the CLR [@Martino2019].
+The second, One CLR, added a pseudo-count of 1 to the abundance of all OTUs [@Paulson2013; @Lin2020].
+The third, Nudge CLR, added a pseudo-count of 1 divided by the total number of sequences in a sample to each OTU in the sample [@Costea2014; @Lin2020].
+The final approach, Robust CLR, removed unobserved OTUs prior to calculating the CLR [@Martino2019].
 
-**Analysis of sequencing coverage.** To assess the level of sequencing coverage I calcualted Good's coverage ($C_{Good}$) using mothur:
+**Analysis of sequencing coverage.** To assess the level of sequencing coverage I calculated Good's coverage ($C_{Good}$) using mothur:
 
@@
 
-where $n_1$ is the number of OTUs with only one sequence in the sample and $N_T$ is the total number of sequences in the sample.
-Good's coverage was calculated using (i) the observed OTU counts for each sample and dataset, (ii) following rarefaction (1,000 iterations) of the observed OTU counts to the size of the smallest sample in each dataset, and (iii) after rarefying (1,000 iterations) the null community distribution.
+where $n_1$ was the number of OTUs with only one sequence in the sample and $N_T$ was the total number of sequences in the sample.
+Good's coverage was calculated using (i) the observed OTU counts for each sample and dataset, (ii) following rarefaction (1,000 iterations) of the observed OTU counts to the size of the smallest sample in each dataset, and (iii) after rarefaction (1,000 iterations) of the null community distribution.
 
-**Reproducible data analysis.** A complete reproducible workflow written in Snakemake (v.7.15.2) and conda (v.4.12.0) computational environment can be obtained from the GitHub hosted git repository for this project (https://github.com/SchlossLab/Schloss_Rarefaction_XXXXX_2022).
-This paper was written in R markdown (v.`r package_version("rmarkdown")`) with the kableExtra (v.`r package_version("kableExtra")`) package.
+**Reproducible data analysis.** A complete reproducible workflow written in Snakemake (v.7.15.2) and Conda (v.4.12.0) computational environment can be obtained from the GitHub hosted git repository for this project (https://github.com/SchlossLab/Schloss_Rarefaction_XXXXX_2023).
+This paper was written in R-flavored markdown (v.`r package_version("rmarkdown")`) with the `kableExtra` (v.`r package_version("kableExtra")`) package.
 The mothur (v.1.47.0) and R (`r glue("{version$major}.{version$minor}")`) software packages were used for all analyses with extensive use of functions in the tidyverse metapackage (v.`r package_version("tidyverse")`).
@@
 **Table 1.
-Summary of studies used in the analysis.** For all studies, the number of sequences used from each dataset was rarefied to the smallest sample size.
-A graphical represenation of the distribution of sample sizes for each dataset and the samples that were removed from each dataset are provided in Figure S1.
+Summary of studies used in the analysis.** For all studies, when rarefaction was used the number of sequences used from each dataset was the size of the smallest sample.
+A graphical representation of the distribution of sample sizes for each dataset and the samples that were removed from each dataset are provided in Figure S1.
 
@@
 **Figure 1.
-Rarefaction eliminates the correlation between sequencing depth and alpha diversity (A) and between differences in sampling depth and beta (B) diversity metrics when using null community models.** Examples of the relationship between different metrics and methods for controlling for uneven sequencing effort are provided in Figures S2 and S3 for alpha and beta diversity metrics, respectively.
+Rarefaction eliminates the correlation between sequencing depth and alpha diversity (A) and between differences in squencing depth and beta (B) diversity metrics when using null community models.** Examples of the relationship between different metrics and methods for controlling for uneven sequencing effort are provided in Figures S2 and S3 for alpha and beta diversity metrics, respectively.
 Each point represents the mean of 100 random null community models; the standard deviation was smaller than the size of the plotting symbol.
@@
 
-**Figure 3.The risk of falsely detecting a difference between treatment groups drawn from a null model does not meaningfully vary from 5% when data are rarefied when sequencing depth is confounded with treatement group**.
+**Figure 3.The risk of falsely detecting a difference between treatment groups drawn from a null model does not meaningfully vary from 5% when data are normalized by rarefaction when sequencing depth is confounded with treatment group**.
 Samples were assigned to different treatment groups based on whether they were above the median number of sequences for each dataset.
@@
 **Figure 4.
-The ability to detect true differences in treatment groups for alpha (A) and beta (B) diversity metrics is greatest when communities differing in the relative abundance of their OTUs are rarefied.** For each dataset samples were randomly assigned to one of two community distributions where the abundance of OTUs differed.
+The ability to detect true differences in treatment groups for alpha (A) and beta (B) diversity metrics is greatest when communities differing in the relative abundance of their OTUs are normalized by rarefaction.** For each dataset samples were randomly assigned to one of two community distributions where the abundance of OTUs differed.
 To calculate the power for each datasets, datasets were regenerated 100 times and differences in alpha diversity were tested using a Wilcoxon test (A) and differences in beta diversity were tested using PERMANOVA (B) at a 5% threshold.
@@
 **Figure 5.
-The ability to detect true differences in treatment groups for alpha diversity metrics is greatest when communities differing in richness are rarefied.** For each dataset samples were randomly assigned to one of two community distributions where one distribution contained a subset of OTUs found in the other.
+The ability to detect true differences in treatment groups for alpha diversity metrics is greatest when sequencing depths from communities differing in richness are normalized by rarefaction.** For each dataset samples were randomly assigned to one of two community distributions where one distribution contained a subset of OTUs found in the other.
 To calculate the power for each dataset, datasets were regenerated 100 times and differences in alpha diversity were tested using a Wilcoxon test (A) and differences in beta diversity were tested using PERMANOVA (B) at a 5% threshold.
@@
 **Figure 6.
-The mean and coefficient of variation for rarefied richness, shannon diversity, and Bray-Curtis dissimilarity vary with sequencing depth.** For each dataset, a null community distribution was created and samples were created to have the same sequencing depth as they did originally.
+The mean and coefficient of variation for richness, Shannon diversity, and Bray-Curtis dissimilarity values calculated by rarefaction vary with sequencing depth.** For each dataset, a null community distribution was created and samples were created to have the same sequencing depth as they did originally.
 The placement of the plotting symbol indicates the size of the smallest sample.
@@
 Most datasets are sequenced to a level that provides a high level of coverage.** Each plotting symbol represents the observed Good's coverage for a different sample in each dataset.
-The smoothed line indicates the simulated coverage for varying levels of sampling effort when a null community is generated from the observed data.
-The box and whisker plot indicates the range of coverage values when the observed commmunity data were rarefied to the size of the least sequenced sample.
+The smoothed line indicates the simulated coverage for varying levels of sequencing effort when a null community is generated from the observed data.
+The box and whisker plot indicates the range of coverage values when the observed community data were normalized by rarefaction to the size of the least sequenced sample.
 
@@
 **Figure S2.
-Examples of the richness in each of the 490 samples that were generated for one randomization of the null model using the human dataset.** The x-axis indicates the number of sequences in each of the samples prior to each method's appraoch of controlling for uneven sampling effort.
+Examples of the richness in each of the 490 samples that were generated for one randomization of the null model using the human dataset.** The x-axis indicates the number of sequences in each of the samples prior to each method's approach of controlling for uneven sequencing effort.
 The Spearman correlation coefficient ($\uprho$) and test of whether the coefficient was significantly different from zero are indicated for each panel.
@@
 **Figure S3.
-Examples of differences in beta diversity in each of the 490 samples that were generated for one randomization of the null model using the human dataset.** The x-axis indicates the difference in the number of sequences in each of the samples that went into calcualting the pairwise distance prior to each method's appraoch of controlling for uneven sampling effort.
+Examples of differences in beta diversity in each of the 490 samples that were generated for one randomization of the null model using the human dataset.** The x-axis indicates the difference in the number of sequences in each of the samples that went into calculating the pairwise distance prior to each method's approach of controlling for uneven sequencing effort.
 The Spearman correlation coefficient ($\uprho$) and test of whether the coefficient was significantly different from zero are indicated for each panel.

### 2023-06-22 submission/manuscript.tex
Submitted

--- 2022-11-11 submission/manuscript.tex
@@
 \modulolinenumbers
-\usepackage{helvet} % Helvetica font
-\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
+\usepackage{helvet}
+\renewcommand*\familydefault{\sfdefault}
 \usepackage[T1]{fontenc}
 \usepackage[shortcuts]{extdash}
+\usepackage[document]{ragged2e}
 \usepackage{booktabs}
@@
 
-\hypertarget{rarefy-your-data}{%
-\section{Rarefy your data}\label{rarefy-your-data}}
+\raggedright
+
+\hypertarget{rarefaction-is-currently-the-best-approach-to-control-for-uneven-sequencing-effort-in-amplicon-sequence-analyses}{%
+\section{Rarefaction is currently the best approach to control for
+uneven sequencing effort in amplicon sequence
+analyses}\label{rarefaction-is-currently-the-best-approach-to-control-for-uneven-sequencing-effort-in-amplicon-sequence-analyses}}
 
@@
 
-\textbf{Running title:} Rarefy your data
+\textbf{Running title:} Use rarefaction
 
@@
 
-\({\dagger}\) To whom corresponsdence should be addressed:
+\({\dagger}\) To whom correspondence should be addressed:
 
@@
 
+Considering it is common to find as much as 100-fold variation in the
+number of 16S rRNA gene sequences found across samples in a study,
+researchers need to control for the effect of uneven sequencing effort.
+How to do this has become a contentious question. Some have argued that
+rarefying or rarefaction is ``inadmissible'' because it omits valid
+data. A number of alternative approaches have been developed to
+normalize and rescale the data that purport to be invariant to the
+number of observations. I generated community distributions based on 12
+published datasets where I was able to assess the ability of multiple
+methods to control for uneven sequencing effort. Rarefaction was the
+only method that could control for variation in uneven sequencing effort
+when measuring commonly used alpha and beta diversity metrics. Next, I
+compared the false detection rate and power to detect true differences
+between simulated communities with a known effect size using various
+alpha and beta diversity metrics. Although all methods of controlling
+for uneven sequencing effort had an acceptable false detection rate when
+samples were randomly assigned to two treatment groups, rarefaction was
+consistently able to control for differences in sequencing effort when
+sequencing depth was confounded with treatment group. Finally, the
+statistical power to detect differences in alpha and beta diversity
+metrics was consistently the highest when using rarefaction. These
+simulations underscore the importance of using rarefaction to normalize
+the number of sequences across samples in amplicon sequencing analyses.
+
 \hypertarget{importance}{%
@@
 
+Sequencing 16S rRNA gene fragments has become a fundamental tool for
+understanding the diversity of microbial communities and the factors
+that affect their diversity. Due to technical challenges, it is common
+to observe wide variation in the number of sequences that are collected
+from different samples within the same study. However, the diversity
+metrics used by microbial ecologists are sensitive to differences in
+sequencing effort. Therefore, tools are needed to control for the uneven
+levels of sequencing. This simulation-based analysis shows that despite
+a longstanding controversy, rarefaction is the most robust approach to
+control for uneven sequencing effort. The controversy started because of
+confusion over the definition of rarefaction and violation of
+assumptions that are made by methods that have been borrowed from other
+fields. Microbial ecologists should use rarefaction.
+
 \newpage
@@
 
-The ability to generate millions sequence reads from 16S rRNA gene
-amplicons and metagenomic datasets has allowed researchers to multiplex
-multiple samples on the same sequencing run by pooling separate PCRs
-that can be deconvoluted later based on index (aka barcode) sequences
-that embeded into the primer sequence (1, 2). Unfortunately, it is
-common to observe variation in the number of sequence reads from each
+The ability to generate millions of 16S rRNA gene amplicon and
+metagenomic sequence reads has allowed researchers to multiplex multiple
+samples on the same sequencing run by pooling separate PCRs that can be
+deconvoluted later based on index (aka barcode) sequences that are
+embedded into the sequence of the PCR primers (1--3). Unfortunately, it
+is common to observe variation in the number of sequence reads from each
 sample vary by as much as 100-fold (e.g, see Figure S1). This occurs
-because pooling of DNA from multiple PCRs is frought with numerous
-opportunities for technichal errors to compound leading to a skewed
+because pooling of DNA from multiple PCRs is fraught with numerous
+opportunities for technical errors to compound leading to a skewed
 distribution. Aside from developing better methods of pooling DNA, the
-question of how to control for uneven sampling effort in microbial
+question of how to control for uneven sequencing effort in microbial
 ecology studies has become controversial.
@@
 The practice of rarefaction has been commonly used in ecology for more
-than 50 years as a tool to control for uneven sampling effort across
-experimental replicates (3, 4). Microbial ecologists have used it to
-compare 16S rRNA gene sequence data for the past 25 years (5--7). When
-data are rarefied, the investigator selects a desired threshold of
-sampling effort and removes any samples below that threshold. They then
-randomly select that many sequences, with replacement from each sample.
-Based on the obseved sequence counts, the researcher can then calcuate
-alpha diversity metrics including richness and diversity indices or beta
-divesity metrics such as a Jaccard or Bray-Curtis dissimilarity index. I
-refer to this single sampling as a subsample; this method is implemented
-as the sub.sample function in mothur (8) and the rrarefy function in the
-vegan R package (9). Rarefaction repeats the subsampling a large number
-of times (e.g., 100 or 1,000) and calculates the mean of the alpha or
-beta diersity metric over those subsamplings; rarefaction is implemented
-in mothur using the summary.single and dist.shared functions (8) and
-with the vegan R package using the rarefy or avgdist fuctions (9).
-Rarefaction effectively tells a researcher what an alpha or beta
-diversity metric would have been for a collection of samples if they
-were all evenly sampled. Although a closed form equation exists to
-calcualte the expected richness (4), it is computationally easier to
-empirically calculate rarefied richness and other alpha and beta
-diversity metrics.
-
-In 2014, McMurdie and Holmes (10) announced that rarefaction of
-microbial community data is ``statistically inadmissible'' because it
-omits valid data. In their simulations, they observed that the use of
-rarefaction reduced the statistical power to compare communities based
-on beta diversity metrics. The communities used in this study were
-simulated by mixing operational taxonomic unit (OTU) frequencies from
-marine and human fecal communities, which were then sampled to generate
-40 replicate communities each sampled to different depths. The
-simulation with the greatest depth of coverage obtained an average of
-10,000 sequences per sample. Analysis of the R code that they used
-indicates that the simulations were actually only based on a single
-subsample and not truly rarefied data. Furthermore, the authors removed
-any OTU from the simulated communities that was not observed in three or
-more samples. This is a frequently used approach for analyzing
-microbiome data (11). Unfortunately, it removes valid data, skews the
-distribution of the community, and removes true patchiness from the
-representation of the communities (12). Finally, the analysis did not
-include analysis of alpha diersity metrics and chose to assess beta
-diversity metrics based on clustering quality of samples rather than
-using the more frequently used approach of employing non-parametric
-multivariate analysis of variance (13). Regardless of these caveats,
-McMurdie and Holmes recommended the use of variance stabilizing
-transformations such as those applied for gene expression analysis such
-as that found in the DeSeq R package.
-
-Beyond the critiques from McMurdie and Holmes others have developed
-alternative approaches to controlling for uneven sampling effort in
-amplicon sequencing studies. For alpha diversity metrics, Willis used
+than 50 years as a tool to control for uneven sequencing effort across
+experimental replicates (4, 5). Microbial ecologists have used it to
+compare 16S rRNA gene sequence data for the past 25 years (6--8). With
+rarefaction the investigator selects a desired threshold of sequencing
+effort and removes any samples below that threshold. They then randomly
+select that many sequences, with replacement from each sample. Based on
+the observed sequence counts, the researcher can then calculate alpha
+diversity metrics including richness and diversity indices or beta
+diversity metrics such as a Jaccard or Bray-Curtis dissimilarity index.
+I refer to this single sampling as a subsample; this method is
+implemented as the \texttt{sub.sample} function in mothur (9) and the
+\texttt{rrarefy} function in the \texttt{vegan} R package (10).
+Rarefaction repeats the subsampling a large number of times (e.g., 100
+or 1,000 times) and calculates the mean of the alpha or beta diversity
+metric over those subsamplings; rarefaction is implemented in mothur
+using the \texttt{summary.single} and \texttt{dist.shared} functions (9)
+and with the \texttt{vegan} R package using the \texttt{rarefy} or
+\texttt{avgdist} functions (10). Rarefaction effectively tells a
+researcher what an alpha or beta diversity metric would have been for a
+collection of samples if they were all sequenced to the same depth.
+Although a closed form equation exists to calculate the expected
+richness (5), it is computationally easier to empirically calculate
+richness and other alpha and beta diversity metrics by rarefaction.
+
+In 2014, McMurdie and Holmes (11) announced that ``rarefying'' of
+microbial community data was ``statistically inadmissible'' because it
+omits valid data. In their simulations, they observed that rarefying
+reduced the statistical power to correctly cluster samples into the same
+treatment groups based on beta diversity metrics. Although the detail
+was lost on many subsequent researchers, they did not describe
+rarefaction, but rarefying (12). According to McMurdie and Holmes,
+rarefying is a single subsampling of an OTU abundance table whereas
+rarefaction repeats the subsampling step many times. Furthermore, their
+simulations penalized the rarefied data by removing 15\% of the samples
+and evaluated clustering quality using a clustering algorithm that
+performed worse for rarefied data. According to my reanalysis of
+rarefaction using the original simulation framework, rarefaction
+outperformed other normalization method being used in 2014 for both
+alpha and beta diversity metrics (12). Others have also critiqued the
+original work (13). Regardless, there is a general sense in the
+community that rarefying and rarefaction should not be used with
+microbiome data.
+
+Since McMurdie and Holmes published their simulations, others have
+developed alternative approaches to control for uneven sequencing effort
+in amplicon sequencing studies. For alpha diversity metrics, Willis used
 toy datasets to demonstrate that one could estimate the richness for
@@
 comparisons (14). Non-parametric estimators of richness and diversity
-(15, 16) and parametric estimators of richness {[}Willis2015{]} that
-have been used in microbial ecology studies. For beta diversity metrics
-at least four approaches have been pursued. First, one could use
-relative abundance values where the observed number of sequences in an
-OTU is divided by the total number of sequences in the sample (17).
-Second, normalization strategies have been developed where the relative
-abundance is mutiplied by the size of the minimum desired sampling
-effort and fractional values are reapportioned to the OTUs to obtain
-integer values (18, 19). Third, a variety of center log-ratio methods
-have been developed where the compositional nature of the OTU counts is
-removed and used to calculated Euclidean distances (aka Aitchison
-distances) {[}(20); (19); (17); (21); (22). This strategy is purported
-to control for uneven sampling effort (21, 23); however, some have
-noticed that this feature breaks down under certain conditions (24).
-Finally, as mentioned above, variance stabilization transformations have
-been recommended to generate values that can be used to calcualte
-Euclidean distances (10).
-
-The ongoing controversy over the use of rarefaction and the development
-of numerous strategies to control for uneven sampling effort caused me
-to question how these methods compared to each other. My analysis
-included 16S rRNA gene sequence data from from 12 studies that
-characterized the variation in bacterial communities from diverse
-environments (Table 1 and Figure S1). The sequences were assigned to
-OTUs using a standard pipeline and their frequencies and the number of
-sequences found in each sample were used to generate simulated
-communities and treatment effects. For each dataset and simulation, 100
-replicate datasets were generated and used as input to each of the
-strategies for controlling for uneven sampling effort. My overall
-conclusion was that rarefaction outperformed the alternative strategies.
+(15, 16) and parametric estimators of richness (17) have been used in
+microbial ecology studies. For beta diversity metrics at least four
+approaches have been pursued. First, one could use relative abundance
+values where the observed number of sequences in an OTU is divided by
+the total number of sequences in the sample (18). Second, normalization
+strategies have been developed where the relative abundance is
+multiplied by the size of the minimum desired sequencing effort and
+fractional values are reapportioned to the OTUs to obtain integer values
+(19, 20). Third, a variety of center log-ratio methods have been
+developed where the compositional nature of the OTU counts is removed
+and used to calculated Euclidean distances (aka Aitchison distances)
+(18, 20--23). This strategy is purported to control for uneven
+sequencing effort (22, 24); however, some have noticed that this feature
+breaks down under certain conditions (25). Finally, as mentioned above,
+variance stabilization transformations have been recommended to generate
+values that can be used to calculate Euclidean distances (11).
+
+The ongoing controversy over the use of rarefaction and the recent
+development of alternative strategies to control for uneven sequencing
+effort caused me to question how these methods compared to each other
+using a simulation framework that overcame the issues with the McMurdie
+and Holmes study. My analysis included 16S rRNA gene sequence data from
+from 12 studies that characterized the variation in bacterial
+communities from diverse environments (Table 1 and Figure S1). The
+sequences were assigned to OTUs using a standard pipeline and their
+frequencies and the number of sequences found in each sample were used
+to generate simulated communities and treatment effects. For each
+dataset and simulation, 100 replicate datasets were generated and used
+as input to each of the strategies for controlling for uneven sequencing
+effort. My overall conclusion was that rarefaction outperformed the
+alternative strategies.
 
@@
 \textbf{\emph{Without rarefaction, metrics of alpha diversity are
-sensitive to sampling effort.}} To test the sesitivity of various
-approaches of measuring alpha diversity to sampling effort, I generated
-null models for each dataset. Under a null model, each community from
-the same dataset would be expected to have the same alpha diversity
-regardless of the sampling effort. I measured the richness of the
-communities in each dataset without any correction, using scaled ranked
-subsampling (SRS) normalized OTU counts, with estimates based on
-non-parametric and parametric approaches, and using rarefaction
-(e.g.~Figure S2). For each dataset, all of the approaches, except for
-rarefaction, showed a strong correlation between richness and the number
-of sequences in the sample (Figure 1A). Next, I assessed diversity using
-the Shannon diversity index and the inverse Simpson diversity index
-without any correction, using normalized OTU counts, and rarefaction; I
-also used a non-parametric estimator of Shannon diversity. The
-correlation between sampling depth and the diversity metric was not as
-strong as it was for richness and the inverse Simpson diversity values
-were less sensitive than the Shannon diversity values; however, the
-correlation to the rarefied diversity metrics were the lowest for all of
-the metrics and studies (Figure 1A). The rarefied alpha-diversity
-metrics consistently demonstrated a lack of sensitivity to sampling
-depth.
+sensitive to sequencing effort.}} To test the sensitivity of various
+approaches of measuring alpha diversity to sequencing effort, I
+generated null models using OTU counts and sequencing depths from 12
+studies. Under a null model, each community from the same dataset would
+be expected to have the same alpha diversity regardless of the
+sequencing effort. I measured the richness of the communities in each
+dataset without any correction, using scaled ranked subsampling (SRS)
+normalized OTU counts, with estimates based on non-parametric and
+parametric approaches, and using rarefaction (e.g.~Figure S2). For each
+dataset, all of the approaches, except for rarefaction, showed a strong
+correlation between richness and the number of sequences in the sample
+(Figure 1A). Next, I assessed diversity using the Shannon diversity
+index and the inverse Simpson diversity index without any correction,
+using normalized OTU counts, and rarefaction; I also used a
+non-parametric estimator of Shannon diversity. The correlation between
+sequencing depth and the diversity metric was not as strong as it was
+for richness and the inverse Simpson diversity values were less
+sensitive than the Shannon diversity values; however, the correlation to
+the diversity metrics calculated with rarefaction were the lowest for
+all of the metrics and studies (Figure 1A). The alpha-diversity metrics
+calculated with rarefaction consistently demonstrated a lack of
+sensitivity to sequencing depth.
 
 \textbf{\emph{Without rarefaction, metrics of beta diversity are
-sensitive to sampling effort.}} To test the sesitivity of various
-approaches of measuring beta diversity to sampling effort, I used the
+sensitive to sequencing effort.}} To test the sensitivity of various
+approaches of measuring beta diversity to sequencing effort, I used the
 same null models used for studying the sensitivity of alpha diversity.
@@
 are unique to either community and does not account for the abundance of
-the OTUs. Jaccard distances were calcualted using the uncorrected OTU
+the OTUs. Jaccard distances were calculated using the uncorrected OTU
 counts, with rarefaction, relative abundances, and following
 normalization using cumulative sum scaling (CSS) and SRS. Only the
-rarefied distances showed a lack of sensitivity to sampling effort
-(Figure 1B). Second, I analyzed the sensitivity of the Bray-Curtis
-distance coefficient, which is a popular metric that incorporates the
-abundance of each OTU. Similar to what I observed with the Jaccard
-coefficient, only the rarefied data showed a lack of sensitivity to
-sampling effort (Figure 1B). Third, I calcualted the Euclidean distance
-on raw OTU counts where the central log-ratio (CLR) was calculated
-(i.e., Aitchison distances) by ignoring OTUs in samples with zero counts
-(Robust CLR), adding a pseudocount of one to all OTU counts prior to
-calculating the CLR (One CLR), adding a pseudocount of one divided by
-the total number of sequences obtained for the community (Nudge CLR),
-and imputing the value of zero counts (Zero CLR). The Aitchison
-distances were all strongly sensitive to sampling effort (Figure 1B).
-Finally, I used the variance-stabilization transformation (VST) from
-DeSeq2 prior to calculating Euclidean distances. Again, there was a
-strong sensitivity to sampling effort (Figure 1B). Although Euclidean
-distances are not typically used on raw or rarefied count data in
-ecology, rarefied Euclidean distances were not sensitive to sampling
-effort. Across each of the beta diversity metrics and approaches used to
-account for uneven sampling effort and sparsity, rarefaction was the
-least sensitive approach to differences in sampling effort.
+distances calculated with rarefaction showed a lack of sensitivity to
+sequencing effort (Figure 1B). Second, I analyzed the sensitivity of the
+Bray-Curtis distance coefficient, which is a popular metric that
+incorporates the abundance of each OTU. Similar to what I observed with
+the Jaccard coefficient, only the data normalized with rarefaction
+showed a lack of sensitivity to sequencing effort (Figure 1B). Third, I
+calculated the Euclidean distance on raw OTU counts where the central
+log-ratio (CLR) was calculated (i.e., Aitchison distances) by ignoring
+OTUs in samples with zero counts (Robust CLR), adding a pseudo-count of
+one to all OTU counts prior to calculating the CLR (One CLR), adding a
+pseudo-count of one divided by the total number of sequences obtained
+for the community (Nudge CLR), and imputing the value of zero counts
+(Zero CLR). The Aitchison distances were all strongly sensitive to
+sequencing effort (Figure 1B). Finally, I used the
+variance-stabilization transformation (VST) from DESeq2 prior to
+calculating Euclidean distances. Again, there was a strong sensitivity
+to sequencing effort (Figure 1B). Although Euclidean distances are not
+typically used on raw or rarefaction-normalized count data in ecology
+(26), Euclidean distances calculated with rarefaction were not sensitive
+to sequencing effort. Across each of the beta diversity metrics and
+approaches used to account for uneven sequencing effort and sparsity,
+rarefaction was the least sensitive approach to differences in
+sequencing effort.
 
 \textbf{\emph{Rarefaction limits the detection of false positives when
-sampling effort and treatment group are confounded.}} Next, I
+sequencing effort and treatment group are confounded.}} Next, I
 investigated the impact of the various strategies and metrics on falsely
@@
 Indeed, for each dataset and alpha and beta diversity metric and
-strategy for accounting for uneven sampling, approximately 5\% of the
+strategy for accounting for uneven sequencing, approximately 5\% of the
 tests yielded a significant result (Figure 2). Second, I employed a
@@
 confounded with the number of sequences in each sample. Under this
-model, only the rarefied data consistently resulted in a 5\% false
-positive rate for alpha and beta diversity metrics (Figure 2). These
-results aligned with the observed sensitivity of alpha and beta
-diversity metrics to sampling effort and underscore the value of
-rarefaction.
+model, only the data normalized with rarefaction consistently resulted
+in a 5\% false positive rate for alpha and beta diversity metrics
+(Figure 2). These results aligned with the observed sensitivity of alpha
+and beta diversity metrics to sequencing effort and underscore the value
+of rarefaction.
 
@@
 differences between treatment groups.}} To assess the impact of
-different approaches to control for uneven sampling effort I performed
+different approaches to control for uneven sequencing effort I performed
 two additional simulations. In the first simulation, I implemented a
 skewed abundance distribution model to create two treatment groups for
-each datasets that were each populated with half of the samples each
-with the same number of sequences as the samples had in the observed
-data. The power to detect differences in richness between the two
-simulated treatment groups by all approaches was low (Figure 4A). This
-was likely because the approach for generating the perturbed community
-did not necessarily change the number of OTUs in each treatment group.
-Regardless, the simulations testing differencse in richness using the
-Rice and Stream datasets had the greatest power when the richness data
-were rarefied. To explore this further, a richness-adjusted community
+each dataset that were each populated with half of the samples each with
+the same number of sequences as the samples had in the observed data.
+The two treatment groups varied in their structure such that one had the
+same abundances as the null distribution above and the other had 10\% of
+its OTUs randomly selected to increase their counts by 5\%. The power to
+detect differences in richness between the two simulated treatment
+groups by all approaches was low (Figure 4A). This was likely because
+the approach for generating the perturbed community did not necessarily
+change the number of OTUs in each treatment group. Regardless, the
+simulations testing differences in richness using the Rice and Stream
+datasets had the greatest power when the richness data were calculated
+with rarefaction. To explore this further, a richness-adjusted community
 model was created by removing 3\% of the OTUs from a null model model.
-As suggested by the first simulation, the rarefied richness data had a
-higher statistical power than the other approaches when measuring
-richness (Figure 5). The simulations testing the power to detect
-differences in Shannon diversity also showed that rarefied data
-performed other methods (Figure 4A). When testing for differences in the
-Inverse Simpson diversity index the the difference between rarefaction
-and the other methods was negligible (Figure 4A). For tests of beta
-diversity I found that rarefaction was the most reliable approach to
-maintain statistical power to detect differences between two
-communities. Among the tests using the Jaccard and Bray-Curtis metrics,
-raw count data and CSS normalized data had little power relative to
-rarefied, relative abundance, and SRS normalized data. The differences
-in power between rarefied, relative abundance, and SRS normalized data
-was small, but if there were differences, the power obtained using
-rarefied data was greater than the other methods. Among the tests using
-Euclidean distances, using raw counts and CLR and DeSeq2 transformed
-data had little power relative to the distances calcualted using
-rarefied and relative abundance data. This power-based analysis of the
+As suggested by the first simulation, the richness data calculated with
+rarefaction had a higher statistical power than the other approaches
+when measuring richness (Figure 5). The simulations testing the power to
+detect differences in Shannon diversity also showed that rarefaction
+performed better than the other methods (Figure 4A). When testing for
+differences in the Inverse Simpson diversity index the the difference
+between rarefaction and the other methods was negligible (Figure 4A).
+For tests of beta diversity I found that rarefaction was the most
+reliable approach to maintain statistical power to detect differences
+between two communities. Among the tests using the Jaccard and
+Bray-Curtis metrics, raw count data and CSS normalized data had little
+power relative to using rarefaction, relative abundance, and SRS to
+normalize the uneven sequencing depths. The differences in power between
+counts normalized with rarefaction, relative abundance, and SRS data was
+small, but if there were differences, the power obtained using
+rarefaction was greater than the other methods. Among the tests using
+Euclidean distances, using raw counts and CLR and DESeq2 transformed
+data had little power relative to the distances calculated using
+rarefaction and relative abundances. This power-based analysis of the
 simulated communities using different methods of handling uneven sample
@@
 \textbf{\emph{Increased rarefaction depth reduces intra-sample variation
-in alpha and beta diversity.}} Once concern with rarefying communities
-is the perceived loss of sequencing information when more a large
-fraction of data appears to be removed when the community with the
-greatest sequencing depth is rarefied to the size of the community with
-the least (e.g., 99\% with the Bioethanol dataset). To assess the
-sensitivity of alpha and beta diversity metrics to rarefaction depth, I
-again used the dataset generated using the null models, but rarefied
-each community to varying sampling depths (Figure 6). The richness
-values increased with sampling effort as rare OTUs would continue be
+in alpha and beta diversity.}} Once concern with using rarefaction is
+the perceived loss of sequencing information when a large fraction of
+data appears to be removed when the community with the greatest
+sequencing depth is sampled to the size of the community with the least
+(e.g., the smallest sample in the Bioethanol dataset had 1.04\% of
+sequences that were in the largest sample). To assess the sensitivity of
+alpha and beta diversity metrics to rarefaction depth, I again used the
+dataset generated using the null models, but used rarefaction with each
+community to varying sequencing depths (Figure 6). The richness values
+increased with sequencing effort as rare OTUs would continue be
 detected. In contrast, the Shannon diversity and Bray-Curtis values
-plateaued with increased sampling effort. This result was expected since
-increased sampling would lead to increased precision in the measured
-abundance of OTUs. Next, I measured the coefficient of variation (i.e.,
-the mean divided by the standard deviation) between samples for
-richness, Shannon diversity, and Bray-Curtis distances. Although the
-richness values appeared to increased unbounded with smapling effort,
-the coefficient of variation for each dataset was relatively stable. In
-general, the coefficient of variation increased slightly with sampling
-depth only to decline once smaller samples were removed from the
-analysis at higher sampling depths. Interestingly, the coefficient of
-variation between Shannon diveristy values decreased towards zero with
-increased sampling effort and the coefficient of variation between
-Bray-Curtis distances tended to increased. Regardless, the coefficients
-of variation were relatively small.
+plateaued with increased sequencing effort. This result was expected
+since increased sequencing would lead to increased precision in the
+measured relative abundance of the OTUs. Next, I measured the
+coefficient of variation (i.e., the mean divided by the standard
+deviation) between samples for richness, Shannon diversity, and
+Bray-Curtis distances. Although the mean richness appeared to increased
+unbounded with sequencing effort, the coefficient of variation for each
+dataset was relatively stable. In general, the coefficient of variation
+increased slightly with sequencing depth only to decline once smaller
+samples were removed from the analysis at higher sequencing depths.
+Interestingly, the coefficient of variation between Shannon diversity
+values decreased towards zero with increased sequencing effort and the
+coefficient of variation between Bray-Curtis distances tended to
+increase. Regardless, the coefficients of variation were relatively
+small. This analysis indicates that there are benefits to increased
+sequencing depths.
 
@@
 coverage for each dataset ranged between 89.4 and 99.8\% for the
-Seagrass and Human datasets, respectively (Figure 7). When I rarefied
-each dataset to the size of the smallest community in the dataset, with
-the exception of the Seagrass, Rice, and Stream datasets, the median
-coverage for the rarefied communities was still greater than 90\%. These
-results suggest that most studies had a level of sequencing coverage
-that aligned with the diversity of the communities. Next, I used the
-null model for each dataset to ask how much sequencing effort was
-required to obtain higher levels of coverage. To obtain 95 and 99\%
-coverage, an average of 2.70 and 101.20-fold more sequence data was
-estimated to be required than was required to obtain 90\% coverage,
+Seagrass and Human datasets, respectively (Figure 7). When I used a
+rarefaction threshold with each dataset at the size of the smallest
+community in the dataset, with the exception of the Seagrass, Rice, and
+Stream datasets, the median coverage with rarefaction was still greater
+than 90\%. These results suggest that most studies had a level of
+sequencing coverage that aligned with the diversity of the communities.
+Next, I used the null model for each dataset to ask how much sequencing
+effort was required to obtain higher levels of coverage. To obtain 95
+and 99\% coverage, an average of 2.70 and 101.20-fold more sequence data
+was estimated to be required than was required to obtain 90\% coverage,
 respectively (Figure 7). As suggested by the simulated coverages curve
 in Figure 7, these estimates are conservative. Regardless, the
-sequencing effort required to acheive higher sequencing depth would
+sequencing effort required to achieve higher sequencing depth would
 likely limit the number of samples that could be sequenced when
-controlling for costs. Although it may be disconcerting to rarefy to a
-sequencing depth that is considerably lower than that obtained for the
-best sequenced community in a dataset, sequencing coverage for many
-environments is probably adequate even at the lower sequencing depth. Of
-course, the results above have demonstrated that rarefaction is
-necessary to avoid problems with making inferences.
+controlling for costs. Although it may be disconcerting to use
+rarefaction to normalize to a sequencing depth that is considerably
+lower than that obtained for the best sequenced community in a dataset,
+sequencing coverage for many environments is probably adequate even at
+the lower sequencing depth. Of course, regardless of the concerns
+surrounding the choice of the rarefaction depth, the results throughout
+this study demonstrate that rarefaction is necessary to avoid reaching
+incorrect inferences.
 
@@
 
-Over the past decade, the question of whether to rarefy microbial
-community sequence data has become controversial. The analyses I
-presented here strongly indicate that rarefaction is necessary to
-control for uneven sampling effort when comparing communities based on
+Over the past decade, the question of whether to use rarefaction with
+microbial community sequence data has become controversial. The analyses
+I presented here strongly indicate that rarefaction is necessary to
+control for uneven sequencing effort when comparing communities based on
 alpha and beta diversity indices. Compared to all other methods,
@@
 97-fold across samples. I showed that this correlation could lead
-falsely detecting differences between treatment groups if sampling depth
-and sequencing effort are confounded (e.g., 25). The correlation with
-sampling effort leads to an artificial increase in the variation between
-samples and a reduced power to detect true differences in alpha and beta
-diversity. For these reasons, rarefaction is a valuable tool to control
-for uneven sampling effort until improved statistical procedures are
-developed or it becomes possible to more evenly distribute sequencing
-effort across samples.
-
-The primary alternative to rarefaction that many advise is to estimate
-alpha diversity has been to use non-parametric or parametric methods
-using raw data and to then compare the estimates (14). My data
-demonstrates that such approaches are limited for several reasons.
-First, it has been widely observed that non-parametric richness esimates
-such as ACE and Chao1 are sensitive to sampling effort. Therefore, these
-strategies do not, in practice, remove the effects of sampling effort.
-Second, parametric approaches, such as those implemented in the
-breakaway R package, generate confidence intervals that are likely to
-include the true richness and theoretically shrink with increased
-sampling effort. For most samples, the confidence intervals around the
-estimates are too wide to be meaningful, again leading to an inability
-to remvoe the effects of sampling effort. Third, it has become an
-increasingly common practice for researchers to remove sequences that
-are rare in a sample (e.g., those that appear once). Although that
-approach was not taken in this study, removing rare sequences would skew
-the distribtion of sequences and OTUs leading to a distortion of the
-measurement of alpha diversity (12). The effects of removing rare
-sequences would vary across samples depending on the number of sequences
-in each sample. One interesting result of this analysis was the
-demonstration that as metrics that depend less on rare taxa are used,
-the effect of uneven sampling effort was reduced. For example, richness
-counts a sequence appearing once as much as sequence appearing 1,000
-times, while Shannon diversityindex places more emphasis on the more
-abundant sequence, and the inverse Simpson index even more. Although
-normalizing communities to a common number of sequences is also
-suggested (e.g., SRS normalization) to control for uneven sampling
-effort, the current analysis indicates that its performance does not
-meet that of rarefaction. For analysis of alpha diversity metrics,
-rarefaction is the most effective and consistent approach to control fo
-uneven sampling effort.
+falsely detecting differences between treatment groups if sequencing
+depth and sequencing effort are confounded. The correlation with
+sequencing effort leads to an artificial increase in the variation
+between samples and a reduced power to detect true differences in alpha
+and beta diversity. For these reasons, rarefaction is a valuable tool to
+control for uneven sequencing effort until improved statistical
+procedures are developed or it becomes possible to more evenly
+distribute sequencing effort across samples.
+
+The primary alternative to rarefaction for measuring alpha diversity is
+to estimate the metric using non-parametric or parametric models with
+raw counts and to then compare the estimates (7, 14). My results
+demonstrate that such approaches are limited for several reasons. First,
+non-parametric richness estimates such as ACE and Chao1 are sensitive to
+sequencing effort. Therefore, these strategies do not, in practice,
+remove the effects of sequencing effort. Second, parametric approaches,
+such as those implemented in the \texttt{breakaway} R package, generate
+confidence intervals that are likely to include the true richness and
+theoretically shrink with increased sequencing effort. Yet for most
+samples, the confidence intervals around the estimates are too wide to
+be meaningful, again leading to an inability to remove the effects of
+sequencing effort. Third, it has become an increasingly common practice
+for researchers to remove sequences that are rare in a sample (e.g.,
+those that appear once). Although that approach was not taken in this
+study, removing rare sequences would skew the distribution of sequences
+and OTUs leading to a distortion of the measurement of alpha diversity
+(27). The effects of removing rare sequences would vary across samples
+depending on the number of sequences in each sample. One interesting
+result of this analysis was the demonstration that as metrics that
+depend less on rare taxa are used, the effect of uneven sequencing
+effort was reduced. For example, richness counts a sequence appearing
+once as much as sequence appearing 1,000 times, while the Shannon
+diversity index places more emphasis on the more abundant sequence, and
+the inverse Simpson index even more. Although normalizing communities to
+a common number of sequences is also suggested (e.g., SRS normalization)
+to control for uneven sequencing effort, the current analysis indicates
+that its performance does not meet that of rarefaction. For analysis of
+alpha diversity metrics, rarefaction is the most effective and
+consistent approach to control for uneven sequencing effort.
 
@@
 more robust alternatives to rarefaction. Again, the only approach that
-consistently removed the effects of uneven sampling effort was
+consistently removed the effects of uneven sequencing effort was
 rarefaction. Most of the recommendations borrow techniques from methods
 for identifying differential gene expression. Unfortunately, there is an
-important but underappreciated difference between gene expression and
+important but under appreciated difference between gene expression and
 community data. This is the interpretation of unobserved data. For gene
 expression analysis in a single organism the lack of any sequencing data
-for a gene would indicate that its expression was below the limit of
-detection. Sequencing the same organism under multiple conditions would
-not lead to a seemingly unbouned number of genes in the organims.
-Rather, the number of genes has a definite limit that is knowable from
-the genome sequence. With microbiome data, an unobserved sequence could
-mean that the organism was present, but below the limit of detection or
-that the organism was missing. Because we have yet to exhaustively
-sample any community in the same way we have sequenced a single genome
-it is unreliable to impute the presence of all organisms. Yet, this is
-exactly what the variance stabilization transformation and most CLR
-techniques do. This analysis has demonstrated a clear correlation
-between distances calculated by these methods and sequencing effort.
-This result is at odds with the claims by others that the distances are
-scale invariant (21, 23). Again, rarefaction is the most effective and
-consistent approach to control for uneven sampling effort when
-calculating beta diversity metrics.
-
-Two common critique of rarefaction is that the approach ommits valid
-data and that the selection of the value to rarefy the sequence data to
-is arbitrary (10). I disagree with both critiques. All of the data are
-used to calculate the mean value of the rarefied metrics; none of it is
-excluded. When the dataset is subsampled, every sequence has a random
-chance of being included in the calculated metric. When that subsampling
-is repeated a large number of times (e.g., 100 or 1000) the risks of
-ignoring or oversampling rare taxa are mititgated. Furthermore, it is
-curious that the study making the original critique removed sequences
-that did not appear in at least three samples or had an abundance of one
-in any sample. A parallel analysis to this study has demonstrated that
-many of these sequences are likely valid and that removal of rare
-sequences can bias alpha and beta diversity metrics and reduce
-statistical power (12). As for the second critism, I would resist the
-claim that the selection of the rarefaction threshold is arbitrary. In
-practice, there is a tradeoff between sampling breadth and depth.
-Greater breadth will increase the statistical power to compare treatment
-groups and greater depth will increase the resolution to describe the
-communities. My process for picking a break involves looking for a
-natural break in the distribution of the number of sequences. For
-example, the Lake dataset used in this study had a clear break around
-10,000 sequences. I would also consider what samples are below any break
-that I select. If there were critical samples below the break I would
-either reduce the break point or I would generate more sequences from
-those samples. Furthermore, regardless of the break point one selects,
-they could repeat an analysis at multiple thresholds to assess the
-tradeoff between statistical power and resolution. As shown in my
-analysis of Good's coverage values, most studies obtain an ample level
-of coverage and would need to increase their sampling depth by 10 fold
-to increase the coverage by several percentage points. I contend that
-for most studies increased sequencing effort should be applied to
-improve sampling breadth and statistical power over depth and
-resolution.
+for a gene would indicate that although the gene was present, its
+expression was below the limit of detection. Sequencing the same
+organism under multiple conditions would not lead to a seemingly
+unbounded number of genes in the organisms. Rather, the number of genes
+has a definite limit that is knowable from the genome sequence. With
+microbiome data, an unobserved sequence could mean that the organism was
+present, but below the limit of detection or that the organism was
+missing. Because we have yet to exhaustively sample any community in the
+same way we have sequenced a single genome, it is unreliable to impute
+the presence of all organisms. Yet, this is exactly what the variance
+stabilization transformation and most CLR techniques do. This analysis
+has demonstrated a clear correlation between distances calculated by
+these methods and sequencing effort. This result is at odds with the
+claims by others that the distances are scale invariant (22, 24). Again,
+rarefaction is the most effective and consistent approach to control for
+uneven sequencing effort when calculating beta diversity metrics.
+
+Two common critiques of rarefaction are that the approach omits valid
+data and that the selection of rarefaction threshold is arbitrary (11).
+I disagree with both critiques. All of the data are used to calculate
+the mean value of the metrics after repeatedly subsampling the data.
+When the dataset is subsampled, every sequence has a random chance of
+being included in the calculated metric. When that subsampling is
+repeated a large number of times (e.g., 100 or 1000) the risks of
+ignoring or oversampling rare taxa are mitigated. Furthermore, it is
+curious that the study making the original critique removed rare
+sequences. A parallel analysis to this study has demonstrated that many
+of these sequences are likely valid and that removal of rare sequences
+can bias alpha and beta diversity metrics and reduce statistical power
+(27). As for the second criticism, I would resist the claim that the
+selection of the rarefaction threshold is arbitrary. In practice, there
+is a tradeoff between sampling breadth and sequencing depth. Greater
+breadth will increase the statistical power to compare treatment groups
+and greater depth will increase the resolution to describe the
+communities. My personal process for rarefaction threshold involves
+looking for a natural break in the distribution of the number of
+sequences. For example, the Lake dataset used in this study had a clear
+break around 10,000 sequences (Figure S1). I would also consider what
+samples are below any break that I select. If there were critical
+samples below the break I would either reduce the threshold or obtain
+more sequences from those samples. As shown in my analysis of Good's
+coverage values, most studies obtain an ample level of coverage and
+would need to increase their sequencing depth by 10-fold to increase the
+coverage by several percentage points. In past work, I have favored
+distributing the total number of sequences to increase sampling breadth
+rather than sequencing depth.
 
 The up to 100-fold difference in sample sizes is an unfortunate
-byproduct of how seqencing libraries are constructed. Researchers
+byproduct of how sequencing libraries are constructed. Researchers
 perform separate PCRs for each sample with unique index (aka, barcode)
-sequencs that all them to later identify which sample each sequence
-belongs. When the PCRs are pooled efforts are taken to pool the
-fragments in equimolar ratios. Researchers use one of two approaches.
-First, they often will quantify the concentration of DNA from each PCR
-and then pool DNA in the desired amounts. Alternatively, they may use
-normalization plates where each well can hybridize a uniform amount of
-DNA that is then eluted and pooled. Clearly, both approaches have
-limitations that reduce the ability to truly achieve equimolar mixture.
-In addition, for some samples it is common to co-amplify non-specific
-DNAs which may add to the challenges of achieving equimolar mixtures of
-the desired amplicons (25). Regardless, it is clear that better
+sequences that allow them to later assign sequences back to the samples
+that they came from. When the PCR products are pooled, efforts are taken
+to pool the fragments in equimolar ratios. Researchers use one of two
+approaches. First, they often will quantify the concentration of DNA
+from each PCR and then pool DNA in the desired amounts. Alternatively,
+they may use normalization plates where each well can hybridize a
+uniform amount of DNA that is then eluted and pooled. Clearly, both
+approaches have limitations that reduce the ability to truly achieve
+equimolar mixture. In addition, for some samples it is common to
+co-amplify non-specific DNAs which may add to the challenges of
+achieving equimolar mixtures of the desired amplicons since those
+sequences will be removed (28). Regardless, it is clear that better
 strategies are needed to reduce the variation in the number of sequences
@@
 All simulations have weaknesses and should be interpreted with caution.
-However, the simulated communites generated and analyzed in this study
+However, the simulated communities generated and analyzed in this study
 had the advantage of being designed with known properties including the
 alpha and beta diversity and the their differences between treatment
-groups. For now, is is perfectly admissible and proper to rarefy
-microbial community data or risk reaching unwarrented conclusions.
+groups. Indeed, is is perfectly admissible and proper to use rarefaction
+with microbial community data. The alternative is to risk reaching
+unwarranted conclusions.
 
@@
 familiarity with the data. The use of paired 250 nt reads to sequence
-the V4 region resulted in a near complete two-fold overap of the V4
+the V4 region resulted in a near complete two-fold overlap of the V4
 region resulting in high quality contigs with a low sequencing error
-rate (1). These data were processed through a standardized sequence
+rate (3). These data were processed through a standardized sequence
 curation pipeline to generate operational taxonomic units (OTUs) using
-the mothur software package (1, 8). OTUs were identified using the
+the mothur software package (3, 9). OTUs were identified using the
 OptiClust algorithm to cluster sequences together that were not more
-than 3\% different from each other (26).
+than 3\% different from each other (29).
 
@@
 the total number of sequences in each OTU. This is a similar approach to
-that of the IS algorithm described by Ulrich and Gotelli (27). Because
+that of the IS algorithm described by Ulrich and Gotelli (30). Because
 the construction of the null models was a stochastic process, 100
-replicates were geneated for each dataset.
+replicates were generated for each dataset.
 
@@
 to one of two treatment groups. In the biased treatment model, samples
-that had more than the median number of seqeunces for a dataset were
+that had more than the median number of sequences for a dataset were
 assigned to one treatment group and the rest were assigned to a second
@@
 treatment groups should have only yielded a significant result in 5\% of
-the simulations when testing under a Type I error (i.e., \(\upalpha\))
-of 0.05.
+the simulations when testing under a Type I error (i.e., \(\alpha\)) of
+0.05.
 
@@
 all samples and using those values as the probability of sampling each
-OTU. This probability distribution was sampled until each sample had the
-same number of sequences that it did in the observed data. Samples in
-the second treatment group were generated by adjusting the relative
-abundances of the OTUs in the firs treatment group by increasing the
-relative abundance of 10\% of the OTUs by 5\%. These values were
-determined after empirically searching for conditions that resulted in a
-large fraction of the randomizations yieleding a significant result
-across most of the studies. Sequences were sampled from the skewed
-community community until each sample had the same number of sequences
-that it did in the observed data. Under the skewed abundance community
-model each sample represented a statistical sampling of two communities
-such that there should not have been a statistically significant
-difference within a treatment group, but there was between the treatment
-groups. Because the construction of the skewed abundance community model
-was a stochastic process, 100 replicates were geneated for each dataset.
+OTU. This probability distribution was sampled, with replacement, until
+each sample had the same number of sequences that it did in the observed
+data. Samples in the second treatment group were generated by adjusting
+the relative abundances of the OTUs in the firs treatment group by
+increasing the relative abundance of 10\% of the OTUs by 5\%. These
+values were determined after empirically searching for conditions that
+resulted in a large fraction of the randomizations yielding a
+significant result across most of the studies. Sequences were sampled
+from the skewed community community until each sample had the same
+number of sequences that it did in the observed data. Under the skewed
+abundance community model, each sample represented a statistical
+sampling of two communities such that there should not have been a
+statistically significant difference within a treatment group, but there
+was between the treatment groups. Because the construction of the skewed
+abundance community model was a stochastic process, 100 replicates were
+generated for each dataset.
 
@@
 remaining OTUs. Sequences were sampled from the richness-adjusted
-community distrubtion until each sample had the same number of sequences
-that it did in the observed data. Under the richness-adjusted community
-model each sample represented a statistical sampling of two communities
-such that there should not have been a statistically significant
-difference within a treatment group, but there was between the treatment
-groups. Because the construction of the richness-adjusted community
-model was a stochastic process, 100 replicates were geneated for each
-dataset.
+community distribtion, with replacement, until each sample had the same
+number of sequences that it did in the observed data. Under the
+richness-adjusted community model, each sample represented a statistical
+sampling of two communities such that there should not have been a
+statistically significant difference within a treatment group, but there
+was between the treatment groups. Because the construction of the
+richness-adjusted community model was a stochastic process, 100
+replicates were generated for each dataset.
 
@@
 performed using the non-parametric two-sample Wilcoxon test as
-implemented in R. This test was selected because the alpha-diversity
-metrics tended to not be normally distributed and each dataset required
-a different transformation to normalize the data. Comparisons of beta
-diversity metrics were performed using the adonis2 function from the
-vegan (v.2.6.2) R package (9). The adonis2 function implements a
-non-parametric multivariate analysis of variance using distances
-matrices as described by McArdle and Anderson (13). Throughout this
-study I used 0.05 as the threshold for asssessing statistical
-significance.
+implemented in \texttt{wilcoxon.test} in the \texttt{stats} base R
+package. This test was selected because the alpha-diversity metrics
+tended to not be normally distributed and each dataset required a
+different transformation to normalize the data. Comparisons of beta
+diversity metrics were performed using the \texttt{adonis2} function
+from the \texttt{vegan} (v.2.6.2) R package (10). The \texttt{adonis2}
+function implements a non-parametric multivariate analysis of variance
+using distances matrices (31). Throughout this study I used 0.05 as the
+threshold for assessing the statistical significance of any P-values.
 
@@
 abundance and richness-adjusted community models were set to impose a
-known effect size when using rarefied community data. The statistical
-power to detect these differences was determined by calculating the
-p-value for each of 100 replicate simulated set of samples from each
-dataset using the various alpha and beta diversity metrics. The
-percentage of tests that yielded a significant P-value was considered
-the statistical power (i.e., 1 minus the Type II error) to detect the
-difference.
+known effect size when using community data normalized by rarefaction.
+The statistical power to detect these differences was determined by
+calculating the p-value for each of 100 replicate simulated set of
+samples from each dataset using the various alpha and beta diversity
+metrics. The percentage of tests that yielded a significant P-value was
+considered the statistical power (i.e., 1 minus the Type II error) to
+detect the difference.
 
 \textbf{Alpha diversity calculations.} Various strategies for handling
-uneven sampling effort were evaluated to identify the best approach for
-calculating community richness and Shannon and inverse Simpson diversity
-indeices. Raw OTU counts were used as input to calculate sample richness
-and Shannon and inverse Simpson diversity using mothur (8, 28). Shannon
-diversity was calculated as
+uneven sequencing effort were evaluated to identify the best approach
+for calculating community richness and Shannon and inverse Simpson
+diversity indices. OTU counts were used as input to calculate sample
+richness and Shannon and inverse Simpson diversity using mothur (9, 32).
+Shannon diversity was calculated as
 
@@
 The inverse Simpson diversity was calculated as \(1/D_{simpson}\). In
-both formulae, \(n_i\) is the number of sequences in OTU \(i\) and \(N\)
-is the number of sequences in the sample. Rarefaction of richness,
-Shannon diversity and Inverse Simpson diversity values were carried out
-in mothur. Briefly, mothur calculates each value on a random draw of the
-same number of sequences from each sample and obtains a mean value based
-on 1,000 random draws. Scaled ranked subsampling (SRS) was used to
-normalize OTU counts to the size of the smallest sample in each dataset
-using the SRS R package (v.0.2.3)(18). Normalized OTU counts were used
-to calculate sample richness and Shannon and inverse Simpson diversity
-values using mothur. Data normalized by cumulative sum scaling (CSS)
-were not reported for alpha-diversity values since the relative
+both formulae, \(n_i\) was the number of sequences in OTU \(i\) and
+\(N\) iwass the number of sequences in the sample. Rarefaction of
+richness, Shannon diversity and Inverse Simpson diversity values were
+carried out in mothur. Briefly, mothur calculates each value on a random
+draw of the same number of sequences from each sample and obtains a mean
+value based on 1,000 random draws. Scaled ranked subsampling (SRS) was
+used to normalize OTU counts to the size of the smallest sample in each
+dataset using the SRS R package (v.0.2.3)(19). Normalized OTU counts
+were used to calculate sample richness and Shannon and inverse Simpson
+diversity values using mothur. Data normalized by cumulative sum scaling
+(CSS) were not reported for alpha-diversity values since the relative
 abundances of the features do not change with the normalization
-procedure (19). The non-parametric bias-corrected Chao1 and ACE richness
+procedure (20). The non-parametric bias-corrected Chao1 and ACE richness
 estimators (16) and a non-parametric estimator of the Shannon diversity
 (15) were calculated using raw OTU counts with mothur. Parametric
-estimates of sample richness were calculated using the breakaway (BA) R
-package (v.4.7.9)(29). The current analysis reports both the results
-from running default model selection procedure and the Poisson model.
-The default model selection returned either the Kemp, Negative Binomial,
-or Poisson models. Relative abundance data were not used to calculate
-alpha diversity metrics since the richness and evenness does not change
-from the raw data when dividing each sample by the total number of
-sequences in the sample.
+estimates of sample richness were calculated using the
+\texttt{breakaway} (BA) R package (v.4.7.9)(17). My analysis reports
+both the results from running default model selection procedure and the
+Poisson model. The default model selection returned either the Kemp,
+Negative Binomial, or Poisson models. Relative abundance data were not
+used to calculate alpha diversity metrics since the richness and
+evenness does not change from the raw data when dividing each sample by
+the total number of sequences in the sample.
 
@@
 calculations, multiple approaches were used to control for uneven
-sampling effort and calculate beta diversity. Raw and OTU counts were
+sequencing effort and calculate beta diversity. Raw and OTU counts were
 used for input to calculate the Jaccard, Bray-Curtis, and Euclidean
-dissimilarity indices using the vegdist function from the vegan R
-package (v.2.6.2)(9). The Jaccard index was calculated as
+dissimilarity indices using the \texttt{vegdist} function from the
+\texttt{vegan} R package (v.2.6.2)(10). The Jaccard index was calculated
+as
 
@@
 
-where \(n_{A,i}\) and \(n_{B,i}\) are the number of sequences observed
+where \(n_{A,i}\) and \(n_{B,i}\) were the number of sequences observed
 in OTU \(i\) from samples \(A\) and \(B\), respectively. \(N_A\) and
-\(N_B\) are the total number of sequences in samples \(A\) and \(B\),
-respectively. \(S_T\) is the total number of OTUs observed between the
+\(N_B\) were the total number of sequences in samples \(A\) and \(B\),
+respectively. \(S_T\) was the total number of OTUs observed between the
 two samples. The Euclidean distance was calculated as
@@
 These metrics were calculated using the relative abundance of each OTU
-using the vegdist function from vegan. The relative abundance was
-calculated as the number of sequences in the OTU (e.g., \(n_{A,i}\))
-divided by the total number of sequences in the sample (e.g., \(N_A\)).
-
-Rarefied beta-diversity values were calculated using the avgdist
-function in vegan. Briefly, vegan's avgdist function calculates each
-pairwise dissimilarity index after obtaining a random draw of the same
-number of sequences from each sample. After obtaining 100 random draws
-it returns the mean value.
+using the \texttt{vegdist} function from the \texttt{vegan} R package.
+The relative abundance was calculated as the number of sequences in the
+OTU (e.g., \(n_{A,i}\)) divided by the total number of sequences in the
+sample (e.g., \(N_A\)).
+
+Beta-diversity values generated with rarefaction were calculated using
+the \texttt{avgdist} function in vegan. Briefly, \texttt{vegan}'s
+\texttt{avgdist} function calculates each pairwise dissimilarity index
+after obtaining a random draw of the same number of sequences from each
+sample. After obtaining 100 random draws it returns the mean value.
 
@@
 samples within a dataset. Scaled ranked subsampling (SRS) and cumulative
-sum scaling (CSS) were used to normalize raw OTU counts using the SRS
-(v.0.2.3) and metagenomeSeq (v.1.36.0) R packages, respectively (18,
-19). The normalized counts were then used to calculate Jaccard and
-Bray-Curtis dissimilarity indices. Finally, the variance-stabilization
-transformation (VST) as implemented in the DESeq2 (v.1.34.0) R package
-was used to normalize the data as described by McMurdie and Holmes (10,
-30). Because the VST approach generates negative values, which are
-incompatible with calculating Jaccard and Bray-Curtis dissimilarity
-values, Euclidean distances were calcualted instead.
-
-Raw OTU counts were used to calculate centered log ratio values for each
-OTU, which were then used to calcualte Euclidean distances; such
-distances are commonly refered to as Aitchison distances. Centered
-log-ratio (CLR) abundances are calculated as:
+sum scaling (CSS) were used to normalize raw OTU counts using the
+\texttt{SRS} (v.0.2.3) and \texttt{metagenomeSeq} (v.1.36.0) R packages,
+respectively (19, 20). The normalized counts were then used to calculate
+Jaccard and Bray-Curtis dissimilarity indices. Finally, the
+variance-stabilization transformation (VST) as implemented in the
+\texttt{DESeq2} (v.1.34.0) R package was used to normalize the data as
+described by McMurdie and Holmes (11, 33). Because the VST approach
+generated negative values, which are incompatible with calculating
+Jaccard and Bray-Curtis dissimilarity values, Euclidean distances were
+calculated instead.
+
+Raw OTU counts were used to calculate centered log ratio (CLR) values
+for each OTU, which were then used to calculate Euclidean distances;
+such distances are commonly referred to as Aitchison distances. CLR
+abundances are calculated as:
 
 \[
-clr\left(n_j\right) = \left[ \ln\frac{x_{ij}}{g(x_j)}, ..., \ln\frac{x_{S_Tj}}{g(x_j)}\right]
+CLR\left(n_j\right) = \left[ \ln\frac{x_{ij}}{g(x_j)}, ..., \ln\frac{x_{S_Tj}}{g(x_j)}\right]
 \]
 
-where \$x\_\{ij\} was the number of sequences observed for OTU \(i\) in
-sample \(j\) and \(g()\) was the geometric mean \(x_j\) was the counts
+where \(x_{ij}\) was the number of sequences observed for OTU \(i\) in
+sample \(j\) and \(g()\) was the geometric mean, \(x_j\) was the count
 of the \(S_T\) OTUs in sample \(j\). Because the geometric mean is zero
@@
 based on the observed data using the zCompositions (v.1.4.0.1) R package
-(31). The second, One CLR, added a pseudocount of 1 to the abundance of
-all OTUs (17, 19). The third, Nudge CLR, added a pseudocount of 1
+(34). The second, One CLR, added a pseudo-count of 1 to the abundance of
+all OTUs (18, 20). The third, Nudge CLR, added a pseudo-count of 1
 divided by the total number of sequences in a sample to each OTU in the
-sample (17, 22). The final approach, Robust CLR, removed unobserved OTUs
-prior to calcualting the CLR (21).
+sample (18, 23). The final approach, Robust CLR, removed unobserved OTUs
+prior to calculating the CLR (22).
 
 \textbf{Analysis of sequencing coverage.} To assess the level of
-sequencing coverage I calcualted Good's coverage (\(C_{Good}\)) using
+sequencing coverage I calculated Good's coverage (\(C_{Good}\)) using
 mothur:
@@
 
-where \(n_1\) is the number of OTUs with only one sequence in the sample
-and \(N_T\) is the total number of sequences in the sample. Good's
-coverage was calculated using (i) the observed OTU counts for each
-sample and dataset, (ii) following rarefaction (1,000 iterations) of the
-observed OTU counts to the size of the smallest sample in each dataset,
-and (iii) after rarefying (1,000 iterations) the null community
-distribution.
+where \(n_1\) was the number of OTUs with only one sequence in the
+sample and \(N_T\) was the total number of sequences in the sample.
+Good's coverage was calculated using (i) the observed OTU counts for
+each sample and dataset, (ii) following rarefaction (1,000 iterations)
+of the observed OTU counts to the size of the smallest sample in each
+dataset, and (iii) after rarefaction (1,000 iterations) of the null
+community distribution.
 
 \textbf{Reproducible data analysis.} A complete reproducible workflow
-written in Snakemake (v.7.15.2) and conda (v.4.12.0) computational
+written in Snakemake (v.7.15.2) and Conda (v.4.12.0) computational
 environment can be obtained from the GitHub hosted git repository for
 this project
-(\url{https://github.com/SchlossLab/Schloss_Rarefaction_XXXXX_2022}).
-This paper was written in R markdown (v.2.16) with the kableExtra
-(v.1.3.4) package. The mothur (v.1.47.0) and R (4.1.3) software packages
-were used for all analyses with extensive use of functions in the
-tidyverse metapackage (v.1.3.1). A preliminary version of this analysis
-was presented as the Rarefaction video series on the Riffomonas Project
-YouTube channel
+(\url{https://github.com/SchlossLab/Schloss_Rarefaction_XXXXX_2023}).
+This paper was written in R-flavored markdown (v.2.16) with the
+\texttt{kableExtra} (v.1.3.4) package. The mothur (v.1.47.0) and R
+(4.1.3) software packages were used for all analyses with extensive use
+of functions in the tidyverse metapackage (v.1.3.1). A preliminary
+version of this analysis was presented as the Rarefaction video series
+on the Riffomonas Project YouTube channel
 (\url{https://www.youtube.com/playlist?list=PLmNrK_nkqBpJuhS93PYC-Xr5oqur7IIWf}).
@@
 \begin{CSLReferences}{0}{1}
-\leavevmode\vadjust pre{\hypertarget{ref-Kozich2013}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-Sogin2006}{}}%
 \CSLLeftMargin{1. }%
-\CSLRightInline{\textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter
-NT}, \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. {Development of
-a dual-index sequencing strategy and curation pipeline for analyzing
-amplicon sequence data on the MiSeq Illumina sequencing platform}.
-Applied and environmental microbiology \textbf{79}:5112--5120.}
+\CSLRightInline{\textbf{Sogin ML}, \textbf{Morrison HG}, \textbf{Huber
+JA}, \textbf{Welch DM}, \textbf{Huse SM}, \textbf{Neal PR},
+\textbf{Arrieta JM}, \textbf{Herndl GJ}. 2006. Microbial diversity in
+the deep sea and the underexplored {``}rare biosphere{''}. Proceedings
+of the National Academy of Sciences \textbf{103}:12115--12120.
+doi:\href{https://doi.org/10.1073/pnas.0605127103}{10.1073/pnas.0605127103}.}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-Kozich2013}{}}%
+\CSLLeftMargin{3. }%
+\CSLRightInline{\textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter
+NT}, \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
+dual-index sequencing strategy and curation pipeline for analyzing
+amplicon sequence data on the {MiSeq} illumina sequencing platform.
+Applied and Environmental Microbiology \textbf{79}:5112--5120.
+doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-Sanders1968}{}}%
-\CSLLeftMargin{3. }%
+\CSLLeftMargin{4. }%
 \CSLRightInline{\textbf{Sanders HL}. 1968. Marine benthic diversity: A
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Hurlbert1971}{}}%
-\CSLLeftMargin{4. }%
+\CSLLeftMargin{5. }%
 \CSLRightInline{\textbf{Hurlbert SH}. 1971. The nonconcept of species
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Dunbar1999}{}}%
-\CSLLeftMargin{5. }%
+\CSLLeftMargin{6. }%
 \CSLRightInline{\textbf{Dunbar J}, \textbf{Takala S}, \textbf{Barns SM},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Hughes2001}{}}%
-\CSLLeftMargin{6. }%
+\CSLLeftMargin{7. }%
 \CSLRightInline{\textbf{Hughes JB}, \textbf{Hellmann JJ},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Moyer1998}{}}%
-\CSLLeftMargin{7. }%
+\CSLLeftMargin{8. }%
 \CSLRightInline{\textbf{Moyer CL}, \textbf{Tiedje JM}, \textbf{Dobbs
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Schloss2009}{}}%
-\CSLLeftMargin{8. }%
+\CSLLeftMargin{9. }%
 \CSLRightInline{\textbf{Schloss PD}, \textbf{Westcott SL},
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Oksanen2022}{}}%
-\CSLLeftMargin{9. }%
-\CSLRightInline{\textbf{Oksanen J}, \textbf{Simpson GL},
-\textbf{Blanchet FG}, \textbf{Kindt R}, \textbf{Legendre P},
-\textbf{Minchin PR}, \textbf{O'Hara RB}, \textbf{Solymos P},
-\textbf{Stevens MHH}, \textbf{Szoecs E}, \textbf{Wagner H},
-\textbf{Barbour M}, \textbf{Bedward M}, \textbf{Bolker B},
-\textbf{Borcard D}, \textbf{Carvalho G}, \textbf{Chirico M}, \textbf{De
-Caceres M}, \textbf{Durand S}, \textbf{Evangelista HBA},
-\textbf{FitzJohn R}, \textbf{Friendly M}, \textbf{Furneaux B},
-\textbf{Hannigan G}, \textbf{Hill MO}, \textbf{Lahti L}, \textbf{McGlinn
-D}, \textbf{Ouellette M-H}, \textbf{Ribeiro Cunha E}, \textbf{Smith T},
-\textbf{Stier A}, \textbf{Ter Braak CJF}, \textbf{Weedon J}. 2022.
-\href{https://CRAN.R-project.org/package=vegan}{Vegan: Community ecology
-package}.}
+\leavevmode\vadjust pre{\hypertarget{ref-Dixon2003}{}}%
+\CSLLeftMargin{10. }%
+\CSLRightInline{\textbf{Dixon P}. 2003. {VEGAN}, a package of r
+functions for community ecology. Journal of Vegetation Science
+\textbf{14}:927--930.
+doi:\href{https://doi.org/10.1111/j.1654-1103.2003.tb02228.x}{10.1111/j.1654-1103.2003.tb02228.x}.}
 
 \leavevmode\vadjust pre{\hypertarget{ref-McMurdie2014}{}}%
-\CSLLeftMargin{10. }%
+\CSLLeftMargin{11. }%
 \CSLRightInline{\textbf{McMurdie PJ}, \textbf{Holmes S}. 2014. Waste
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Bokulich2013}{}}%
-\CSLLeftMargin{11. }%
-\CSLRightInline{\textbf{Bokulich NA}, \textbf{Subramanian S},
-\textbf{Faith JJ}, \textbf{Gevers D}, \textbf{Gordon JI}, \textbf{Knight
-R}, \textbf{Mills DA}, \textbf{Caporaso JG}. 2013. Quality-filtering
-vastly improves diversity estimates from illumina amplicon sequencing.
-Nature Methods \textbf{10}:57--59.
-doi:\href{https://doi.org/10.1038/nmeth.2276}{10.1038/nmeth.2276}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Schloss2020}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-Schloss2023wnwn}{}}%
 \CSLLeftMargin{12. }%
-\CSLRightInline{\textbf{Schloss PD}. 2020. Removal of rare amplicon
-sequence variants from 16S {rRNA} gene sequence surveys biases the
-interpretation of community structure data.
-doi:\href{https://doi.org/10.1101/2020.12.11.422279}{10.1101/2020.12.11.422279}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-McArdle2001}{}}%
+\CSLRightInline{\textbf{Schloss PD}. 2023. Waste not, want not:
+Revisiting the analysis that called into question the practice of
+rarefaction. bio{R}xiv.
+doi:\href{https://doi.org/Pending\%20at\%20bioRxiv}{Pending at
+bioRxiv}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Weiss2017}{}}%
 \CSLLeftMargin{13. }%
-\CSLRightInline{\textbf{McArdle BH}, \textbf{Anderson MJ}. 2001.
-{FITTING} {MULTIVARIATE} {MODELS} {TO} {COMMUNITY} {DATA}: A {COMMENT}
-{ON} {DISTANCE}-{BASED} {REDUNDANCY} {ANALYSIS}. Ecology
-\textbf{82}:290--297.
-doi:\href{https://doi.org/10.1890/0012-9658(2001)082\%5B0290:fmmtcd\%5D2.0.co;2}{10.1890/0012-9658(2001)082{[}0290:fmmtcd{]}2.0.co;2}.}
+\CSLRightInline{\textbf{Weiss S}, \textbf{Xu ZZ}, \textbf{Peddada S},
+\textbf{Amir A}, \textbf{Bittinger K}, \textbf{Gonzalez A},
+\textbf{Lozupone C}, \textbf{Zaneveld JR}, \textbf{Vázquez-Baeza Y},
+\textbf{Birmingham A}, \textbf{Hyde ER}, \textbf{Knight R}. 2017.
+Normalization and microbial differential abundance strategies depend
+upon data characteristics. Microbiome \textbf{5}:27.
+doi:\href{https://doi.org/10.1186/s40168-017-0237-y}{10.1186/s40168-017-0237-y}.}
 
@@
 \CSLRightInline{\textbf{Willis AD}. 2019. Rarefaction, alpha diversity,
-and statistics. Frontiers in Microbiology \textbf{10}.
+and statistics. Frontiers in Microbiology \textbf{10}:2407.
 doi:\href{https://doi.org/10.3389/fmicb.2019.02407}{10.3389/fmicb.2019.02407}.}
@@
 \CSLLeftMargin{15. }%
-\CSLRightInline{\textbf{Chao A}, \textbf{Shen T-J}. 2003. Nonparametric
-estimation of shannon's index of diversity when there are unseen species
-in sample. Environmental and ecological statistics
-\textbf{10}:429--443.}
+\CSLRightInline{\textbf{Chao A}, \textbf{Shen T-J}. 2003. Environmental
+and Ecological Statistics \textbf{10}:429--443.
+doi:\href{https://doi.org/10.1023/a:1026096204727}{10.1023/a:1026096204727}.}
 
@@
 \CSLLeftMargin{16. }%
-\CSLRightInline{\textbf{Chao A}, \textbf{Chiu C-H}. 2016. Species
-richness: Estimation and comparison. Wiley StatsRef: statistics
-reference online \textbf{1}:26.}
+\CSLRightInline{\textbf{Chao A}, \textbf{Chiu C-H}. 2016.
+\href{https://doi.org/10.1002/9781118445112.stat03432.pub2}{Species
+richness: Estimation and comparison}, p. 1--26. \emph{In} Wiley
+StatsRef: Statistics reference online. John Wiley \& Sons, Ltd.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Willis2015}{}}%
+\CSLLeftMargin{17. }%
+\CSLRightInline{\textbf{Willis A}, \textbf{Bunge J}. 2015. Estimating
+diversity via frequency ratios. Biometrics \textbf{71}:1042--1049.
+doi:\href{https://doi.org/10.1111/biom.12332}{10.1111/biom.12332}.}
 
 \leavevmode\vadjust pre{\hypertarget{ref-Lin2020}{}}%
-\CSLLeftMargin{17. }%
+\CSLLeftMargin{18. }%
 \CSLRightInline{\textbf{Lin H}, \textbf{Peddada SD}. 2020. Analysis of
 microbial compositions: A review of normalization and differential
-abundance analysis. npj Biofilms and Microbiomes \textbf{6}.
+abundance analysis. npj Biofilms and Microbiomes \textbf{6}:60.
 doi:\href{https://doi.org/10.1038/s41522-020-00160-w}{10.1038/s41522-020-00160-w}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Beule2020}{}}%
-\CSLLeftMargin{18. }%
-\CSLRightInline{\textbf{Beule L}, \textbf{Karlovsky P}. 2020.
-\href{https://doi.org/10.7717/peerj.9593}{Improved normalization of
-species count data in ecology by scaling with ranked subsampling (SRS):
-Application to microbial communities}. PeerJ \textbf{8}:e9593.}
+\CSLLeftMargin{19. }%
+\CSLRightInline{\textbf{Beule L}, \textbf{Karlovsky P}. 2020. Improved
+normalization of species count data in ecology by scaling with ranked
+subsampling (SRS): Application to microbial communities. PeerJ
+\textbf{8}:e9593.
+doi:\href{https://doi.org/10.7717/peerj.9593}{10.7717/peerj.9593}.}
 
 \leavevmode\vadjust pre{\hypertarget{ref-Paulson2013}{}}%
-\CSLLeftMargin{19. }%
+\CSLLeftMargin{20. }%
 \CSLRightInline{\textbf{Paulson JN}, \textbf{Stine OC}, \textbf{Bravo
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Quinn2019}{}}%
-\CSLLeftMargin{20. }%
+\CSLLeftMargin{21. }%
 \CSLRightInline{\textbf{Quinn TP}, \textbf{Erb I}, \textbf{Gloor G},
@@
 A field guide for the compositional analysis of any-omics data.
-{GigaScience} \textbf{8}.
+{GigaScience} \textbf{8}:giz107.
 doi:\href{https://doi.org/10.1093/gigascience/giz107}{10.1093/gigascience/giz107}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Martino2019}{}}%
-\CSLLeftMargin{21. }%
+\CSLLeftMargin{22. }%
 \CSLRightInline{\textbf{Martino C}, \textbf{Morton JT}, \textbf{Marotz
@@
 \textbf{Zengler K}. 2019. A novel sparse compositional technique reveals
-microbial perturbations. {mSystems} \textbf{4}.
+microbial perturbations. {mSystems} \textbf{4}:00016--19.
 doi:\href{https://doi.org/10.1128/msystems.00016-19}{10.1128/msystems.00016-19}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Costea2014}{}}%
-\CSLLeftMargin{22. }%
+\CSLLeftMargin{23. }%
 \CSLRightInline{\textbf{Costea PI}, \textbf{Zeller G}, \textbf{Sunagawa
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Quinn2018}{}}%
-\CSLLeftMargin{23. }%
+\CSLLeftMargin{24. }%
 \CSLRightInline{\textbf{Quinn TP}, \textbf{Erb I}, \textbf{Richardson
@@
 \leavevmode\vadjust pre{\hypertarget{ref-teBeest2021}{}}%
-\CSLLeftMargin{24. }%
+\CSLLeftMargin{25. }%
 \CSLRightInline{\textbf{Beest DE te}, \textbf{Nijhuis EH},
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-Legendre2012}{}}%
+\CSLLeftMargin{26. }%
+\CSLRightInline{\textbf{Legendre P}, \textbf{Legendre L}. 2012.
+Numerical ecology. Elsevier Science.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Schloss2023sing}{}}%
+\CSLLeftMargin{27. }%
+\CSLRightInline{\textbf{Schloss PD}. 2020. Removal of rare amplicon
+sequence variants from 16S {rRNA} gene sequence surveys biases the
+interpretation of community structure data. bio{R}xiv.
+doi:\href{https://doi.org/10.1101/2020.12.11.422279}{10.1101/2020.12.11.422279}.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-Morris2013}{}}%
-\CSLLeftMargin{25. }%
+\CSLLeftMargin{28. }%
 \CSLRightInline{\textbf{Morris A}, \textbf{Beck JM}, \textbf{Schloss
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Westcott2017}{}}%
-\CSLLeftMargin{26. }%
+\CSLLeftMargin{29. }%
 \CSLRightInline{\textbf{Westcott SL}, \textbf{Schloss PD}. 2017.
 {OptiClust}, an improved method for assigning amplicon-based sequence
-data to operational taxonomic units. {mSphere} \textbf{2}.
+data to operational taxonomic units. {mSphere} \textbf{2}:e00073--17.
 doi:\href{https://doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Ulrich2010}{}}%
-\CSLLeftMargin{27. }%
+\CSLLeftMargin{30. }%
 \CSLRightInline{\textbf{Ulrich W}, \textbf{Gotelli NJ}. 2010. Null model
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-Anderson2001}{}}%
+\CSLLeftMargin{31. }%
+\CSLRightInline{\textbf{Anderson MJ}. 2001. A new method for
+non-parametric multivariate analysis of variance. Austral Ecology
+\textbf{26}:32--46.
+doi:\href{https://doi.org/10.1111/j.1442-9993.2001.01070.pp.x}{10.1111/j.1442-9993.2001.01070.pp.x}.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-Magurran2004}{}}%
-\CSLLeftMargin{28. }%
+\CSLLeftMargin{32. }%
 \CSLRightInline{\textbf{Magurran AE}. 2004.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Willis2015}{}}%
-\CSLLeftMargin{29. }%
-\CSLRightInline{\textbf{Willis A}, \textbf{Bunge J}. 2015. Estimating
-diversity via frequency ratios. Biometrics \textbf{71}:1042--1049.}
-
 \leavevmode\vadjust pre{\hypertarget{ref-Love2014}{}}%
-\CSLLeftMargin{30. }%
+\CSLLeftMargin{33. }%
 \CSLRightInline{\textbf{Love MI}, \textbf{Huber W}, \textbf{Anders S}.
 2014. Moderated estimation of fold change and dispersion for {RNA}-seq
-data with {DESeq}2. Genome Biology \textbf{15}.
+data with {DESeq}2. Genome Biology \textbf{15}:550.
 doi:\href{https://doi.org/10.1186/s13059-014-0550-8}{10.1186/s13059-014-0550-8}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Palarea2015}{}}%
-\CSLLeftMargin{31. }%
+\CSLLeftMargin{34. }%
 \CSLRightInline{\textbf{Palarea-Albaladejo J}, \textbf{Martin-Fernandez
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Li2015}{}}%
-\CSLLeftMargin{32. }%
+\CSLLeftMargin{35. }%
 \CSLRightInline{\textbf{Li Q}, \textbf{Heist EP}, \textbf{Moe LA}. 2015.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Baxter2016}{}}%
-\CSLLeftMargin{33. }%
+\CSLLeftMargin{36. }%
 \CSLRightInline{\textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers
@@
 sensitivity of fecal immunochemical test for detecting colonic lesions.
-Genome Medicine \textbf{8}.
+Genome Medicine \textbf{8}:37.
 doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Beall2015}{}}%
-\CSLLeftMargin{34. }%
+\CSLLeftMargin{37. }%
 \CSLRightInline{\textbf{Beall BFN}, \textbf{Twiss MR}, \textbf{Smith
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Henson2016}{}}%
-\CSLLeftMargin{35. }%
+\CSLLeftMargin{38. }%
 \CSLRightInline{\textbf{Henson MW}, \textbf{Pitre DM}, \textbf{Weckhorst
@@
 Artificial seawater media facilitate cultivating members of the
-microbial majority from the gulf of mexico. {mSphere} \textbf{1}.
+microbial majority from the gulf of mexico. {mSphere}
+\textbf{1}:e00028--16.
 doi:\href{https://doi.org/10.1128/msphere.00028-16}{10.1128/msphere.00028-16}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Baxter2014}{}}%
-\CSLLeftMargin{36. }%
+\CSLLeftMargin{39. }%
 \CSLRightInline{\textbf{Baxter NT}, \textbf{Wan JJ}, \textbf{Schubert
@@
 \leavevmode\vadjust pre{\hypertarget{ref-LevyBooth2018}{}}%
-\CSLLeftMargin{37. }%
+\CSLLeftMargin{40. }%
 \CSLRightInline{\textbf{Levy-Booth DJ}, \textbf{Giesbrecht IJW},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Edwards2015}{}}%
-\CSLLeftMargin{38. }%
+\CSLLeftMargin{41. }%
 \CSLRightInline{\textbf{Edwards J}, \textbf{Johnson C},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Ettinger2017}{}}%
-\CSLLeftMargin{39. }%
+\CSLLeftMargin{42. }%
 \CSLRightInline{\textbf{Ettinger CL}, \textbf{Williams SL},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Graw2018}{}}%
-\CSLLeftMargin{40. }%
+\CSLLeftMargin{43. }%
 \CSLRightInline{\textbf{Graw MF}, \textbf{DAngelo G}, \textbf{Borchers
@@
 microbial communities across sediment horizons in deep marine sediments
-of the south china sea. Frontiers in Microbiology \textbf{9}.
+of the south china sea. Frontiers in Microbiology \textbf{9}:729.
 doi:\href{https://doi.org/10.3389/fmicb.2018.00729}{10.3389/fmicb.2018.00729}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Johnston2016}{}}%
-\CSLLeftMargin{41. }%
+\CSLLeftMargin{44. }%
 \CSLRightInline{\textbf{Johnston ER}, \textbf{Rodriguez-R LM},
@@
 bacterial populations and reduced community diversity across the alaska
-tundra ecosystem. Frontiers in Microbiology \textbf{7}.
+tundra ecosystem. Frontiers in Microbiology \textbf{7}:579.
 doi:\href{https://doi.org/10.3389/fmicb.2016.00579}{10.3389/fmicb.2016.00579}.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Hassell2018}{}}%
-\CSLLeftMargin{42. }%
+\CSLLeftMargin{45. }%
 \CSLRightInline{\textbf{Hassell N}, \textbf{Tinker KA}, \textbf{Moore
@@
 \textbf{Table 1. Summary of studies used in the analysis.} For all
-studies, the number of sequences used from each dataset was rarefied to
-the smallest sample size. A graphical represenation of the distribution
-of sample sizes for each dataset and the samples that were removed from
-each dataset are provided in Figure S1.
+studies, when rarefaction was used the number of sequences used from
+each dataset was the size of the smallest sample. A graphical
+representation of the distribution of sample sizes for each dataset and
+the samples that were removed from each dataset are provided in Figure
+S1. This table is similar to Table 1 from (27).
 
@@
 \endhead
-Bioethanol~(32) & 95 & 3,970,972 & 16,014 & 41,799 &
+Bioethanol~(35) & 95 & 3,970,972 & 16,014 & 41,799 &
 3,690\Hyphdash*356,027 & SRP055545 \\
-Human~(33) & 490 & 20,828,275 & 32,452 & 42,506 &
+Human~(36) & 490 & 20,828,275 & 32,452 & 42,506 &
 10,439\Hyphdash*422,904 & SRP062005 \\
-Lake~(34) & 52 & 3,145,486 & 69,205 & 60,490 & 15,135\Hyphdash*110,993 &
+Lake~(37) & 52 & 3,145,486 & 69,205 & 60,490 & 15,135\Hyphdash*110,993 &
 SRP050963 \\
-Marine~(35) & 7 & 1,484,068 & 213,091 & 212,009 &
+Marine~(38) & 7 & 1,484,068 & 213,091 & 212,009 &
 132,895\Hyphdash*256,758 & SRP068101 \\
-Mice~(1) & 348 & 2,785,641 & 6,426 & 8,004 & 1,804\Hyphdash*30,311 &
+Mice~(3) & 348 & 2,785,641 & 6,426 & 8,004 & 1,804\Hyphdash*30,311 &
 SRP192323 \\
-Peromyscus~(36) & 111 & 1,545,288 & 12,393 & 13,921 &
+Peromyscus~(39) & 111 & 1,545,288 & 12,393 & 13,921 &
 4,454\Hyphdash*33,502 & SRP044050 \\
-Rainforest~(37) & 69 & 936,666 & 11,464 & 13,574 & 4,880\Hyphdash*37,403
+Rainforest~(40) & 69 & 936,666 & 11,464 & 13,574 & 4,880\Hyphdash*37,403
 & ERP023747 \\
-Rice~(38) & 490 & 22,623,937 & 43,399 & 46,171 & 2,777\Hyphdash*192,200
+Rice~(41) & 490 & 22,623,937 & 43,399 & 46,171 & 2,777\Hyphdash*192,200
 & SRP044745 \\
-Seagrass~(39) & 286 & 4,135,440 & 13,538 & 14,459 &
+Seagrass~(42) & 286 & 4,135,440 & 13,538 & 14,459 &
 1,830\Hyphdash*45,076 & SRP092441 \\
-Sediment~(40) & 58 & 1,151,389 & 17,606 & 19,851 & 7,686\Hyphdash*67,763
+Sediment~(43) & 58 & 1,151,389 & 17,606 & 19,851 & 7,686\Hyphdash*67,763
 & SRP097192 \\
-Soil~(41) & 18 & 932,563 & 50,487 & 51,809 & 46,622\Hyphdash*58,935 &
+Soil~(44) & 18 & 932,563 & 50,487 & 51,809 & 46,622\Hyphdash*58,935 &
 ERP012016 \\
-Stream~(42) & 201 & 21,017,610 & 90,621 & 104,565 &
+Stream~(45) & 201 & 21,017,610 & 90,621 & 104,565 &
 8,931\Hyphdash*394,419 & SRP075852 \\
@@
 sequencing depth and alpha diversity (A) and between differences in
-sampling depth and beta (B) diversity metrics when using null community
-models.} Examples of the relationship between different metrics and
-methods for controlling for uneven sequencing effort are provided in
-Figures S2 and S3 for alpha and beta diversity metrics, respectively.
-Each point represents the mean of 100 random null community models; the
-standard deviation was smaller than the size of the plotting symbol.
+sequencing depth and beta (B) diversity metrics when using null
+community models.} Examples of the relationship between different
+metrics and methods for controlling for uneven sequencing effort are
+provided in Figures S2 and S3 for alpha and beta diversity metrics,
+respectively. Each point represents the mean of 100 random null
+community models; the standard deviation was smaller than the size of
+the plotting symbol.
 
@@
 5\% threshold. The false positive rate was the number of times a dataset
-yeilded a significant result.
+yielded a significant result.
 
@@
 treatment groups drawn from a null model does not meaningfully vary from
-5\% when data are rarefied when sequencing depth is confounded with
-treatement group}. Samples were assigned to different treatment groups
-based on whether they were above the median number of sequences for each
-dataset. To calculate the false detection rate, datasets were
-regenerated 100 times and differences in alpha diversity were tested
-using a Wilcoxon test (A) and differences in beta diversity were tested
-using PERMANOVA (B) at a 5\% threshold. The false positive rate was the
-number of times a dataset yeilded a significant result.
+5\% when data are normalized by rarefaction when sequencing depth is
+confounded with treatment group}. Samples were assigned to different
+treatment groups based on whether they were above the median number of
+sequences for each dataset. To calculate the false detection rate,
+datasets were regenerated 100 times and differences in alpha diversity
+were tested using a Wilcoxon test (A) and differences in beta diversity
+were tested using PERMANOVA (B) at a 5\% threshold. The false positive
+rate was the number of times a dataset yielded a significant result.
 
@@
 communities differing in the relative abundance of their OTUs are
-rarefied.} For each dataset samples were randomly assigned to one of two
-community distributions where the abundance of OTUs differed. To
-calculate the power for each datasets, datasets were regenerated 100
-times and differences in alpha diversity were tested using a Wilcoxon
-test (A) and differences in beta diversity were tested using PERMANOVA
-(B) at a 5\% threshold. The power was the number of times a dataset
-yielded a significant result.
+normalized by rarefaction.} For each dataset samples were randomly
+assigned to one of two community distributions where the abundance of
+OTUs differed. To calculate the power for each datasets, datasets were
+regenerated 100 times and differences in alpha diversity were tested
+using a Wilcoxon test (A) and differences in beta diversity were tested
+using PERMANOVA (B) at a 5\% threshold. The power was the number of
+times a dataset yielded a significant result.
 
@@
 \textbf{Figure 5. The ability to detect true differences in treatment
-groups for alpha diversity metrics is greatest when communities
-differing in richness are rarefied.} For each dataset samples were
-randomly assigned to one of two community distributions where one
-distribution contained a subset of OTUs found in the other. To calculate
-the power for each dataset, datasets were regenerated 100 times and
-differences in alpha diversity were tested using a Wilcoxon test (A) and
-differences in beta diversity were tested using PERMANOVA (B) at a 5\%
-threshold. The power was the number of times a dataset yielded a
-significant result.
+groups for alpha diversity metrics is greatest when sequencing depths
+from communities differing in richness are normalized by rarefaction.}
+For each dataset samples were randomly assigned to one of two community
+distributions where one distribution contained a subset of OTUs found in
+the other. To calculate the power for each dataset, datasets were
+regenerated 100 times and differences in alpha diversity were tested
+using a Wilcoxon test (A) and differences in beta diversity were tested
+using PERMANOVA (B) at a 5\% threshold. The power was the number of
+times a dataset yielded a significant result.
 
@@
 
-\textbf{Figure 6. The mean and coefficient of variation for rarefied
-richness, shannon diversity, and Bray-Curtis dissimilarity vary with
-sequencing depth.} For each dataset, a null community distribution was
-created and samples were created to have the same sequencing depth as
-they did originally. The placement of the plotting symbol indicates the
-size of the smallest sample. Results are only shown for sequencing
-depths where a dataset had 5 or more samples.
+\textbf{Figure 6. The mean and coefficient of variation for richness,
+Shannon diversity, and Bray-Curtis dissimilarity values calculated by
+rarefaction vary with sequencing depth.} For each dataset, a null
+community distribution was created and samples were created to have the
+same sequencing depth as they did originally. The placement of the
+plotting symbol indicates the size of the smallest sample. Results are
+only shown for sequencing depths where a dataset had 5 or more samples.
 
@@
 Good's coverage for a different sample in each dataset. The smoothed
-line indicates the simulated coverage for varying levels of sampling
+line indicates the simulated coverage for varying levels of sequencing
 effort when a null community is generated from the observed data. The
 box and whisker plot indicates the range of coverage values when the
-observed commmunity data were rarefied to the size of the least
-sequenced sample.
+observed community data were normalized by rarefaction to the size of
+the least sequenced sample.
 
@@
 sample varied by dataset and was determined based on identifying natural
-breaks in the data.
+breaks in the data. This figure is similar to Figure S1 of (27)
 
@@
 human dataset.} The x-axis indicates the number of sequences in each of
-the samples prior to each method's appraoch of controlling for uneven
-sampling effort. The Spearman correlation coefficient (\(\uprho\)) and
+the samples prior to each method's approach of controlling for uneven
+sequencing effort. The Spearman correlation coefficient (\(\rho\)) and
 test of whether the coefficient was significantly different from zero
@@
 the number of sequences in each of the samples that went into
-calcualting the pairwise distance prior to each method's appraoch of
-controlling for uneven sampling effort. The Spearman correlation
-coefficient (\(\uprho\)) and test of whether the coefficient was
+calculating the pairwise distance prior to each method's approach of
+controlling for uneven sequencing effort. The Spearman correlation
+coefficient (\(\rho\)) and test of whether the coefficient was
 significantly different from zero are indicated for each panel.

### 2023-06-22 code/plot_alpha_beta_depth_correlation.R

@@ -64,7 +64,7 @@ alpha_classes <- c(
 )
 
 alpha_labels <- c(
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "SRS Normalized",
   "ACE Estimate",
@@ -72,12 +72,12 @@ alpha_labels <- c(
   "BA Default",
   "BA Poisson",
   
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "SRS Normalized",
   "Estimate",
   
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "SRS Normalized"
 )
@@ -159,19 +159,19 @@ beta_metrics <- c(
 # )
 
 beta_labels <- c(
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "Rel. abundance",
   "SRS Normalized",
   "CSS Normalized",
 
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "Rel. abundance",
   "SRS Normalized",
   "CSS Normalized",
 
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "Rel. abundance",
   "Robust CLR",

### 2023-06-22 code/plot_cluster_analysis.R

@@ -43,19 +43,19 @@ beta_metrics <- c(
 )
 
 beta_labels <- c(
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "Rel. abundance",
   "SRS Normalized",
   "CSS Normalized",
 
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "Rel. abundance",
   "SRS Normalized",
   "CSS Normalized",
 
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "Rel. abundance",
   "Robust CLR",

### 2023-06-22 code/plot_example_beta_cor.R

@@ -92,7 +92,7 @@ get_rho <- function(p, m, df = correlations) {
 
 
 j_beta_labels <- c(
-  rare = glue("Rarefied\n(\u03C1 = {get_rho('rare', 'jaccard')})"),
+  rare = glue("Rarefaction\n(\u03C1 = {get_rho('rare', 'jaccard')})"),
   raw = glue("Raw\n(\u03C1 = {get_rho('raw', 'jaccard')})"),
   relabund = glue("Rel. abundance\n(\u03C1 = {get_rho('relabund', 'jaccard')})"),
   srs = glue("SRS Normalized\n(\u03C1 = {get_rho('srs', 'jaccard')})"),
@@ -100,7 +100,7 @@ j_beta_labels <- c(
 )
 
 b_beta_labels <- c(
-  rare = glue("Rarefied\n(\u03C1 = {get_rho('rare', 'bray')})"),
+  rare = glue("Rarefaction\n(\u03C1 = {get_rho('rare', 'bray')})"),
   raw = glue("Raw\n(\u03C1 = {get_rho('raw', 'bray')})"),
   relabund = glue("Rel. abundance\n(\u03C1 = {get_rho('relabund', 'bray')})"),
   srs = glue("SRS Normalized\n(\u03C1 = {get_rho('srs', 'bray')})"),

### 2023-06-22 code/plot_false_positive_null_model.R

@@ -60,7 +60,7 @@ alpha_classes <- c(
 )
 
 alpha_labels <- c(
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "SRS Normalized",
   "ACE Estimate",
@@ -68,12 +68,12 @@ alpha_labels <- c(
   "BA Default",
   "BA Poisson",
 
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "SRS Normalized",
   "Estimate",
 
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "SRS Normalized"
 )
@@ -160,19 +160,19 @@ beta_metrics <- c(
 # )
 
 beta_labels <- c(
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "Rel. abundance",
   "SRS Normalized",
   "CSS Normalized",
 
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "Rel. abundance",
   "SRS Normalized",
   "CSS Normalized",
 
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "Rel. abundance",
   "Robust CLR",

### 2023-06-22 code/plot_false_positive_null_model_size.R

@@ -63,7 +63,7 @@ alpha_classes <- c(
 )
 
 alpha_labels <- c(
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "SRS Normalized",
   "ACE Estimate",
@@ -71,12 +71,12 @@ alpha_labels <- c(
   "BA Default",
   "BA Poisson",
 
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "SRS Normalized",
   "Estimate",
 
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "SRS Normalized"
 )
@@ -162,19 +162,19 @@ beta_metrics <- c(
 # )
 
 beta_labels <- c(
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "Rel. abundance",
   "SRS Normalized",
   "CSS Normalized",
 
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "Rel. abundance",
   "SRS Normalized",
   "CSS Normalized",
 
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "Rel. abundance",
   "Robust CLR",

### 2023-06-22 code/plot_power_cffect_model.R

@@ -59,7 +59,7 @@ alpha_classes <- c(
 )
 
 alpha_labels <- c(
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "SRS Normalized",
   "ACE Estimate",
@@ -67,12 +67,12 @@ alpha_labels <- c(
   "BA Default",
   "BA Poisson",
 
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "SRS Normalized",
   "Estimate",
 
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "SRS Normalized"
 )

### 2023-06-22 code/plot_power_effect_model.R

@@ -59,7 +59,7 @@ alpha_classes <- c(
 )
 
 alpha_labels <- c(
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "SRS Normalized",
   "ACE Estimate",
@@ -67,12 +67,12 @@ alpha_labels <- c(
   "BA Default",
   "BA Poisson",
 
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "SRS Normalized",
   "Estimate",
 
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "SRS Normalized"
 )
@@ -161,19 +161,19 @@ beta_metrics <- c(
 # )
 
 beta_labels <- c(
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "Rel. abundance",
   "SRS Normalized",
   "CSS Normalized",
 
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "Rel. abundance",
   "SRS Normalized",
   "CSS Normalized",
 
-  "Rarefied",
+  "Rarefaction",
   "Raw",
   "Rel. abundance",
   "Robust CLR",

### 2023-07-12 submission/manuscript.Rmd
Use gemelli to perform rclr/rpca

@@
  - \modulolinenumbers
- - \usepackage{helvet} % Helvetica font
- - \renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
+ - \usepackage{helvet}
+ - \renewcommand*\familydefault{\sfdefault}
  - \usepackage[T1]{fontenc}
@@
 ---
+
+\raggedright
 
@@
 
-${\dagger}$ To whom corresponsdence should be addressed:
+${\dagger}$ To whom correspondence should be addressed:
 
@@
 How to do this has become a contentious question.
-Some have argued that rarefaction is "inadmissible" because it omits valid data.
+Some have argued that rarefying or rarefaction is "inadmissible" because it omits valid data.
 A number of alternative approaches have been developed to normalize and rescale the data that purport to be invariant to the number of observations.
@@
 
-The ability to generate millions sequence reads from 16S rRNA gene amplicons and metagenomic datasets has allowed researchers to multiplex multiple samples on the same sequencing run by pooling separate PCRs that can be deconvoluted later based on index (aka barcode) sequences that are embedded into the sequence of the PCR primers [@Sogin2006; @Caporaso2010; @Kozich2013].
+The ability to generate millions of 16S rRNA gene amplicon and metagenomic sequence reads has allowed researchers to multiplex multiple samples on the same sequencing run by pooling separate PCRs that can be deconvoluted later based on index (aka barcode) sequences that are embedded into the sequence of the PCR primers [@Sogin2006; @Caporaso2010; @Kozich2013].
 Unfortunately, it is common to observe variation in the number of sequence reads from each sample vary by as much as 100-fold (e.g, see Figure S1).
@@
 With rarefaction the investigator selects a desired threshold of sequencing effort and removes any samples below that threshold.
-They then randomly select that many sequences, with replacement from each sample.
+They then randomly select that many sequences, without replacement from each sample.
 Based on the observed sequence counts, the researcher can then calculate alpha diversity metrics including richness and diversity indices or beta diversity metrics such as a Jaccard or Bray-Curtis dissimilarity index.
@@
 
-[@Schloss2023wnwn]
-Confusion between rarefy and rarefaction
-***In 2014, McMurdie and Holmes [@McMurdie2014] announced that rarefaction of microbial community data is "statistically inadmissible" because it omits valid data.
-In their simulations, they observed that the use of rarefaction reduced the statistical power to compare communities based on beta diversity metrics.
-The communities used in this study were simulated by mixing operational taxonomic unit (OTU) frequencies from marine and human fecal communities, which were then sampled to generate 40 replicate communities each sampled to different depths.
-The simulation with the greatest depth of coverage obtained an average of 10,000 sequences per sample.
-Analysis of the R code that they used indicates that the simulations were actually only based on a single subsample and not traditional rarefaction.
-Furthermore, the authors removed any OTU from the simulated communities that was not observed in three or more samples.
-This is a frequently used approach for analyzing microbiome data [@Bokulich2013].
-Unfortunately, it removes valid data, skews the distribution of the community, and removes true patchiness from the representation of the communities [@Schloss2023sing].
-Finally, the analysis did not include analysis of alpha diversity metrics and chose to assess beta diversity metrics based on clustering quality of samples rather than using the more frequently used approach of employing non-parametric multivariate analysis of variance [@McArdle2001].
-Regardless of these caveats, McMurdie and Holmes recommended the use of variance stabilizing transformations such as those applied for gene expression analysis such as that found in the DESeq and DESeq2 R packages.***
+In 2014, McMurdie and Holmes [@McMurdie2014] announced that "rarefying" of microbial community data was "statistically inadmissible" because it omits valid data.
+In their simulations, they observed that rarefying reduced the statistical power to correctly cluster samples into the same treatment groups based on beta diversity metrics.
+Although the detail was lost on many subsequent researchers, they did not describe rarefaction, but rarefying [@Schloss2023wnwn].
+According to McMurdie and Holmes, rarefying is a single subsampling of an OTU abundance table whereas rarefaction repeats the subsampling step many times.
+Furthermore, their simulations penalized the rarefied data by removing 15% of the samples and evaluated clustering quality using a clustering algorithm that performed worse for rarefied data.
+According to my reanalysis of rarefaction using the original simulation framework, rarefaction outperformed other normalization method being used in 2014 for both alpha and beta diversity metrics [@Schloss2023wnwn].
+Others have also critiqued the original work [@Weiss2017].
+Regardless, there is a general sense in the community that rarefying and rarefaction should not be used with microbiome data.
 
-**Since McMurdie and Holmes published their simulations, others have developed alternative approaches to controlling for uneven sequencing effort in amplicon sequencing studies.
+Since McMurdie and Holmes published their simulations, others have developed alternative approaches to control for uneven sequencing effort in amplicon sequencing studies.
 For alpha diversity metrics, Willis used toy datasets to demonstrate that one could estimate the richness for each sample in a dataset and use those values for statistical comparisons [@Willis2019].
-Non-parametric estimators of richness and diversity [@Chao2003; @Chao2016] and parametric estimators of richness [@Willis2015] that have been used in microbial ecology studies.
+Non-parametric estimators of richness and diversity [@Chao2003; @Chao2016] and parametric estimators of richness [@Willis2015] have been used in microbial ecology studies.
 For beta diversity metrics at least four approaches have been pursued.
@@
 This strategy is purported to control for uneven sequencing effort [@Martino2019; @Quinn2018]; however, some have noticed that this feature breaks down under certain conditions [@teBeest2021].
-Finally, as mentioned above, variance stabilization transformations have been recommended to generate values that can be used to calculate Euclidean distances [@McMurdie2014].**
+Finally, as mentioned above, variance stabilization transformations have been recommended to generate values that can be used to calculate Euclidean distances [@McMurdie2014].
 
-The ongoing controversy over the use of rarefaction and the recent development of alternative strategies to control for uneven sequencing effort caused me to question how these methods compared to each other using a simulation framework that overcame the issues with the original study.
+The ongoing controversy over the use of rarefaction and the recent development of alternative strategies to control for uneven sequencing effort caused me to question how these methods compared to each other using a simulation framework that overcame the issues with the McMurdie and Holmes study.
 My analysis included 16S rRNA gene sequence data from from 12 studies that characterized the variation in bacterial communities from diverse environments (Table 1 and Figure S1).
@@
 Similar to what I observed with the Jaccard coefficient, only the data normalized with rarefaction showed a lack of sensitivity to sequencing effort (Figure 1B).
-Third, I calculated the Euclidean distance on raw OTU counts where the central log-ratio (CLR) was calculated (i.e., Aitchison distances) by ignoring OTUs in samples with zero counts (Robust CLR), adding a pseudo-count of one to all OTU counts prior to calculating the CLR (One CLR), adding a pseudo-count of one divided by the total number of sequences obtained for the community (Nudge CLR), and imputing the value of zero counts (Zero CLR).
+Third, I calculated the Euclidean distance on raw OTU counts where the central log-ratio (CLR) was calculated (i.e., Aitchison distances) using matrix completion to replace zero counts (Robust CLR), adding a pseudo-count of one to all OTU counts prior to calculating the CLR (One CLR), adding a pseudo-count of one divided by the total number of sequences obtained for the community (Nudge CLR), and imputing the value of zero counts (Zero CLR).
 The Aitchison distances were all strongly sensitive to sequencing effort (Figure 1B).
@@
 loss <- table_1 %>%
-  select(nice_name, fold_difference) %>%
-  mutate(loss = 100 - round(100* (1 - 1/fold_difference), 1)) %>%
-  slice_max(loss)
+  mutate(min_max = 100 * min / max) %>%
+  select(nice_name, min_max) %>%
+  slice_min(min_max)
 
@@
 
-***Increased rarefaction depth reduces intra-sample variation in alpha and beta diversity.*** Once concern with using rarefaction is the perceived loss of sequencing information when a large fraction of data appears to be removed when the community with the greatest sequencing depth is sampled to the size of the community with the least (e.g., the smallest sample in the `r loss$nice_name` dataset had `r loss$loss`% of sequences that were in the largest sample).
+***Increased rarefaction depth reduces intra-sample variation in alpha and beta diversity.*** Once concern with using rarefaction is the perceived loss of sequencing information when a large fraction of data appears to be removed when the community with the greatest sequencing depth is sampled to the size of the community with the least (e.g., the smallest sample in the `r loss$nice_name` dataset had `r loss$min_max`% of sequences that were in the largest sample).
 To assess the sensitivity of alpha and beta diversity metrics to rarefaction depth, I again used the dataset generated using the null models, but used rarefaction with each community to varying sequencing depths (Figure 6).
@@
 Indeed, is is perfectly admissible and proper to use rarefaction with microbial community data.
-The alternatie is to risk reaching unwarranted conclusions.
+The alternative is to risk reaching unwarranted conclusions.
 
@@
 In the biased treatment model, samples that had more than the median number of sequences for a dataset were assigned to one treatment group and the rest were assigned to a second treatment group.
-Comparison of any diversity metric across the two treatment groups should have only yielded a significant result in 5% of the simulations when testing under a Type I error (i.e., $\upalpha$) of 0.05.
+Comparison of any diversity metric across the two treatment groups should have only yielded a significant result in 5% of the simulations when testing under a Type I error (i.e., $\alpha$) of 0.05.
 
@@
 Samples in the second treatment group were generated by removing 3% of the OTUs from the dataset and recalculating the relative abundance of the remaining OTUs.
-Sequences were sampled from the richness-adjusted community distribtion, with replacement, until each sample had the same number of sequences that it did in the observed data.
+Sequences were sampled from the richness-adjusted community distribution, with replacement, until each sample had the same number of sequences that it did in the observed data.
 Under the richness-adjusted community model, each sample represented a statistical sampling of two communities such that there should not have been a statistically significant difference within a treatment group, but there was between the treatment groups.
@@
 The third, Nudge CLR, added a pseudo-count of 1 divided by the total number of sequences in a sample to each OTU in the sample [@Costea2014; @Lin2020].
-The final approach, Robust CLR, removed unobserved OTUs prior to calculating the CLR [@Martino2019].
+The final approach, Robust CLR, used matrix completion to replace zero counts for unobserved OTUs prior to calculating the CLR [@Martino2019].
 
@@
 A graphical representation of the distribution of sample sizes for each dataset and the samples that were removed from each dataset are provided in Figure S1.
+This table is similar to Table 1 from [@Schloss2023sing].
 
@@
 **Figure 1.
-Rarefaction eliminates the correlation between sequencing depth and alpha diversity (A) and between differences in squencing depth and beta (B) diversity metrics when using null community models.** Examples of the relationship between different metrics and methods for controlling for uneven sequencing effort are provided in Figures S2 and S3 for alpha and beta diversity metrics, respectively.
+Rarefaction eliminates the correlation between sequencing depth and alpha diversity (A) and between differences in sequencing depth and beta (B) diversity metrics when using null community models.** Examples of the relationship between different metrics and methods for controlling for uneven sequencing effort are provided in Figures S2 and S3 for alpha and beta diversity metrics, respectively.
 Each point represents the mean of 100 random null community models; the standard deviation was smaller than the size of the plotting symbol.
@@
 To calculate the false detection rate, datasets were regenerated 100 times and differences in alpha diversity were tested using a Wilcoxon test (A) and differences in beta diversity were tested using PERMANOVA (B) at a 5% threshold.
-The false positive rate was the number of times a dataset yeilded a significant result.
+The false positive rate was the number of times a dataset yielded a significant result.
 
@@
 The number of sequences observed in each sample for each dataset included in this analysis generally varied by 10 to 100-fold.** The threshold for specifying the number of sequences per sample varied by dataset and was determined based on identifying natural breaks in the data.
+This figure is similar to Figure S1 of [@Schloss2023sing]
 
@@
 Examples of the richness in each of the 490 samples that were generated for one randomization of the null model using the human dataset.** The x-axis indicates the number of sequences in each of the samples prior to each method's approach of controlling for uneven sequencing effort.
-The Spearman correlation coefficient ($\uprho$) and test of whether the coefficient was significantly different from zero are indicated for each panel.
+The Spearman correlation coefficient ($\rho$) and test of whether the coefficient was significantly different from zero are indicated for each panel.
 
@@
 Examples of differences in beta diversity in each of the 490 samples that were generated for one randomization of the null model using the human dataset.** The x-axis indicates the difference in the number of sequences in each of the samples that went into calculating the pairwise distance prior to each method's approach of controlling for uneven sequencing effort.
-The Spearman correlation coefficient ($\uprho$) and test of whether the coefficient was significantly different from zero are indicated for each panel.
+The Spearman correlation coefficient ($\rho$) and test of whether the coefficient was significantly different from zero are indicated for each panel.

### 2023-07-12 submission/manuscript.tex
Submitted to journal
Use gemelli to perform rclr/rpca

@@
 effort and removes any samples below that threshold. They then randomly
-select that many sequences, with replacement from each sample. Based on
-the observed sequence counts, the researcher can then calculate alpha
+select that many sequences, without replacement from each sample. Based
+on the observed sequence counts, the researcher can then calculate alpha
 diversity metrics including richness and diversity indices or beta
@@
 calculated the Euclidean distance on raw OTU counts where the central
-log-ratio (CLR) was calculated (i.e., Aitchison distances) by ignoring
-OTUs in samples with zero counts (Robust CLR), adding a pseudo-count of
+log-ratio (CLR) was calculated (i.e., Aitchison distances) using matrix
+completion to replace zero counts (Robust CLR), adding a pseudo-count of
 one to all OTU counts prior to calculating the CLR (One CLR), adding a
@@
 remaining OTUs. Sequences were sampled from the richness-adjusted
-community distribtion, with replacement, until each sample had the same
+community distribution, with replacement, until each sample had the same
 number of sequences that it did in the observed data. Under the
@@
 divided by the total number of sequences in a sample to each OTU in the
-sample (18, 23). The final approach, Robust CLR, removed unobserved OTUs
-prior to calculating the CLR (22).
+sample (18, 23). The final approach, Robust CLR, used matrix completion
+to replace zero counts for unobserved OTUs prior to calculating the CLR
+(22).
 
@@
 rarefaction. bio{R}xiv.
-doi:\href{https://doi.org/Pending\%20at\%20bioRxiv}{Pending at
-bioRxiv}.}
+doi:\href{https://doi.org/10.1101/2023.06.23.546312}{10.1101/2023.06.23.546312}.}
 

### 2023-07-12 code/get_rclr_beta.R

@@ -2,36 +2,41 @@
 
 library(tidyverse)
 library(data.table)
-library(vegan)
+library(glue)
 
 input <- commandArgs(trailingOnly = TRUE)
-shared_file <- input[1] # shared_file <- "data/soil/data.otu.1.rshared"
-dist_file <- input[2] # dist_file <- "data/soil/data.otu.1.r_rclr_euclidean.dist"
+shared_file <- input[1]
+dist_file <- input[2]
+# shared_file <- "data/soil/data.otu.1.rshared"
+# dist_file <- "data/soil/data.otu.1.r_rclr_euclidean.dist"
+
+temp_otu_file <- str_replace(shared_file, "shared", "shared_gemelli")
+temp_dist_file <- str_replace(shared_file, "shared", "dist_gemelli")
 
 method <- str_replace(dist_file, ".*_(\\w*).dist", "\\1")
+dataset <- str_replace(dist_file, "data/(.*)/data.*", "\\1")
 
-geometric_mean <- function(x) {
+# couldn't get stream, rice, seagrass to reliably get through run_gemelli even
+# with 750 GB of RAM. Skipping...
 
-  exp(mean(log(x[x > 0])))
-
-}
-
-if (method == "euclidean") {
+if (method == "euclidean" && !(dataset %in% c("stream", "rice", "seagrass"))) {
 
   fread(shared_file, colClasses = c(Group = "character")) %>%
     select(Group, starts_with("Otu")) %>%
     pivot_longer(-Group) %>%
-    filter(value > 0) %>%
-    group_by(Group) %>%
-    mutate(rclr = log(value / geometric_mean(value))) %>%
-    ungroup() %>%
-    select(-value) %>%
-    pivot_wider(names_from = name, values_from = rclr, values_fill = 0) %>%
-    column_to_rownames("Group") %>%
-    vegdist(method = method) %>%
-    as.matrix() %>%
-    as_tibble(rownames = "Group") %>%
+    pivot_wider(names_from = Group, values_from = value, values_fill = 0) %>%
+    column_to_rownames("name") %>%
+    write.table(temp_otu_file, sep = "\t", quote = FALSE)
+
+  system(glue("./code/run_gemelli.py {temp_otu_file} {temp_dist_file}"))
+
+  fread(temp_dist_file, header = TRUE) %>%
+    as_tibble() %>%
+    rename(Group = V1) %>%
     write_tsv(dist_file)
+
[…]

### 2023-07-12 code/plot_example_beta_cor.R

@@ -74,7 +74,8 @@ j_distances <- distances %>% filter(calculator == "jaccard")
 
 b_distances <- distances %>% filter(calculator == "bray")
 
-e_distances <- distances %>% filter(calculator == "euclidean")
+e_distances <- distances %>% filter(calculator == "euclidean") %>%
+  mutate(dist = if_else(process == "rclr", 20 * dist, dist))
 
 get_rho <- function(p, m, df = correlations) {
 
@@ -108,7 +109,7 @@ b_beta_labels <- c(
 )
 
 e_beta_labels <- c(
-  rclr = glue("Robust CLR\n(\u03C1 = {get_rho('rclr', 'euclidean')})"),
+  rclr = glue("Robust CLR (x20)\n(\u03C1 = {get_rho('rclr', 'euclidean')})"),
   oclr = glue("One CLR\n(\u03C1 = {get_rho('oclr', 'euclidean')})"),
   nclr = glue("Nudge CLR\n(\u03C1 = {get_rho('nclr', 'euclidean')})"),
   zclr = glue("Zero CLR\n(\u03C1 = {get_rho('zclr', 'euclidean')})"),
@@ -168,7 +169,7 @@ beta_plot <- j_plot / b_plot / e_plot &
     panel.background = element_blank(),
     axis.title.x = element_markdown(),
     strip.text.y = element_blank(),
-    strip.text.x = element_text(hjust = 0, size = 8)
+    strip.text.x = element_text(hjust = 0, size = 7)
   )
   
 ggsave("figures/example_beta_cor.tiff", beta_plot,

### 2023-07-12 code/run_gemelli.py

@@ -1,0 +1,30 @@
+#!/usr/bin/env python3
+
+import sys
+import numpy as np
+from biom import Table
+import pandas as pd
+from gemelli import __version__ as gversion
+from gemelli.preprocessing import rclr_transformation
+from gemelli.rpca import auto_rpca
+
+np.seterr(divide = 'ignore') # for rclr
+
+shared_file = sys.argv[1]
+dist_file = sys.argv[2]
+
+shared_transposed_df = pd.read_csv(shared_file,
+                                   sep = "\t")
+#shared_transposed_df.index.name = 'otu' # format for R 
+
+# convert to biom-formatted file
+shared_transposed_biom = Table(shared_transposed_df.to_numpy(),
+                               shared_transposed_df.index,
+                               shared_transposed_df.columns) 
+
+rclr_gemelli = rclr_transformation(shared_transposed_biom) # compute RCLR
+
+# Run RPCA
+ordination, distance_matrix = auto_rpca(shared_transposed_biom) 
+
+distance_matrix.to_data_frame().to_csv(dist_file, sep='\t') # save just the distances since that is what is output from the R version

### 2023-10-20 code/plot_example_beta_cor.R
Fix superscript in x-axis title

@@ -117,7 +117,6 @@ e_beta_labels <- c(
 )
 
 j_plot <- j_distances %>%
-# slice_sample(n = 1000) %>% #remove this
   ggplot(aes(x = diff, y = dist)) +
   geom_hline(yintercept = 0) +
   geom_point(size = 0.25) +
@@ -132,7 +131,6 @@ j_plot <- j_distances %>%
   )
   
 b_plot <- b_distances %>%
-#  slice_sample(n = 1000) %>% #remove this
   ggplot(aes(x = diff, y = dist)) +
   geom_hline(yintercept = 0) +
   geom_point(size = 0.25) +
@@ -146,9 +144,7 @@ b_plot <- b_distances %>%
     axis.line.y = element_line()
   )
 
-
 e_plot <- e_distances %>%
- # slice_sample(n = 1000) %>% #remove this
   ggplot(aes(x = diff, y = dist)) +
   geom_hline(yintercept = 0) +
   geom_point(size = 0.25) +
@@ -157,21 +153,21 @@ e_plot <- e_distances %>%
   scale_x_continuous(breaks = c(0, 1e5, 2e5, 3e5, 4e5),
                      labels = c(0, 1, 2, 3, 4)) +
   coord_cartesian(clip = "off") +
-  labs(x = "Difference in the number of sequences between pairs of samples (x10^5)",
+  labs(x = "Difference in the number of sequences between pairs of samples (x10<sup>5</sup>)",
        y = "Euclidean distances") +
   theme(
-    axis.line.y = element_line()
+    axis.line.y = element_line(),
+    axis.title.x.bottom = element_markdown()
   )
 
 
 beta_plot <- j_plot / b_plot / e_plot &
   theme(
     panel.background = element_blank(),
-    axis.title.x = element_markdown(),
     strip.text.y = element_blank(),
     strip.text.x = element_text(hjust = 0, size = 7)
   )
-  
+
 ggsave("figures/example_beta_cor.tiff", beta_plot,
       width = 7, height = 7,
       compression = "lzw+p")

### 2023-10-20 submission/manuscript.Rmd
Replace dot dot with here in relative paths

@@
 
-obs_coverage_files <- glue("../data/{datasets}/data.otu.obs_coverage")
+obs_coverage_files <- here(glue("data/{datasets}/data.otu.obs_coverage"))
 names(obs_coverage_files) <- datasets
@@
 
-rare_coverage_files <- glue("../data/{datasets}/data.otu.rarefy_coverage")
+rare_coverage_files <- here(glue("data/{datasets}/data.otu.rarefy_coverage"))
 names(rare_coverage_files) <- datasets

### 2023-10-20 code/get_matrix_rank.R
Calculate ranks for shared files

@@ -1,0 +1,35 @@
+#!/usr/bin/env Rscript
+
+library(data.table)
+library(Matrix)
+library(tidyverse)
+library(glue)
+
+input <- commandArgs(trailingOnly = TRUE)
+dataset <- input[1]
+
+get_matrix_stats <- function(file_name) {
+
+  fread(file_name, colClasses = c(Group = "character")) %>%
+      select(Group, starts_with("Otu")) %>%
+      column_to_rownames("Group") %>% 
+      summarize(rank = rankMatrix(as.matrix(.)),
+                n_samples = nrow(.),
+                n_otus = ncol(.))
+
+}
+
+shared_file <- glue("data/{dataset}/data.otu.1.eshared")
+
+r_seeds <- rep(1:100, 3)
+r_simulations <- rep(c("rshared", "eshared", "cshared"), each = 100)
+
+shared_files <- c(glue("data/{dataset}/data.otu.{r_seeds}.{r_simulations}"),
+                  glue("data/{dataset}/data.otu.shared"))
+
+map_dfr(shared_files, get_matrix_stats) %>%
+  mutate(dataset = dataset,
+        seed = c(r_seeds, 0),
+        simulation = c(r_simulations, "observed"),
+        .before = rank) %>%
+  write_tsv(glue("data/{dataset}/data.rank"))

### 2023-10-20 code/pool_ranks.R

@@ -1,0 +1,12 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(glue)
+
+datasets <- c("bioethanol", "human", "lake", "marine", "mice", "peromyscus",
+            "rainforest", "sediment", "soil", "stream", "rice", "seagrass")
+
+read_tsv(glue("data/{datasets}/data.rank")) %>%
+  group_by(dataset, simulation) %>%
+  summarize(frac_full_rank = mean(rank == n_samples), .groups = "drop") %>%
+  write_tsv("data/process/rank_fractions.tsv")

### 2023-10-23 code/alpha_correlation.R
Calculate alpha diversity with iNEXT

@@ -32,6 +32,19 @@ read_rarefy <- function(x) {
 }
 
 
+read_inext <- function(x) {
+
+# x <- "data/mice/data.otu.1.r_inext_alpha" 
+
+  read_tsv(x, col_types = cols(group = col_character())) %>%
+    select(group, inext_nseqs = value,
+           chao_sobs, chao_shannon, chao_invsimpson,
+           coverage_sobs, coverage_shannon, coverage_invsimpson,
+           size_sobs, size_shannon, size_invsimpson
+          )
+
+}
+
 
 read_srs <- function(x) {
 
@@ -74,19 +87,24 @@ raw <- map_dfr(glue("data/{dataset}/data.otu.{seeds}.r_raw_alpha"),
 rarefy <- map_dfr(glue("data/{dataset}/data.otu.{seeds}.r_rarefy_alpha"),
                   read_rarefy, .id = "seed")
 
+inext <- map_dfr(glue("data/{dataset}/data.otu.{seeds}.r_inext_alpha"),
+               read_inext, .id = "seed")
+
 srs <- map_dfr(glue("data/{dataset}/data.otu.{seeds}.r_srs_alpha"),
                read_srs, .id = "seed")
 
 breakaway <- map_dfr(glue("data/{dataset}/data.otu.{seeds}.r_breakaway_alpha"),
                read_breakaway, .id = "seed")
 
+stopifnot(sum(raw$nseqs != inext$inext_nseqs) == 0)
 stopifnot(sum(raw$nseqs != breakaway$ba_nseqs) == 0)
 stopifnot(sum(rarefy$rare_nseqs != srs$srs_nseqs) == 0)
 
 inner_join(raw, rarefy, by = c("seed", "group")) %>%
+  inner_join(., inext, by = c("seed", "group")) %>%
   inner_join(., srs, by = c("seed", "group")) %>%
   inner_join(., breakaway, by = c("seed", "group")) %>%
-  select(-rare_nseqs, -srs_nseqs, -ba_nseqs) %>%
+  select(-rare_nseqs, -srs_nseqs, -ba_nseqs, -inext_nseqs) %>%
   pivot_longer(cols = -c(seed, group, nseqs),
                names_to = "metric", values_to = "value") %>%
   nest(data = -c(metric, seed)) %>%

### 2023-10-23 code/get_inext_alpha.R

@@ -1,0 +1,80 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(data.table)
+library(iNEXT)
+
+set.seed(19760620)
+
+input <- commandArgs(trailingOnly = TRUE)
+shared_file <- input[1] # e.g. shared_file <- "data/mice/data.otu.1.rshared"
+
+inext_output_file <- str_replace(shared_file,
+                                "shared",
+                                "_inext_alpha")
+
+shared_df <- fread(shared_file,
+      colClasses = c(Group = "character")) %>%
+  select(-label, -numOtus) %>%
+  melt(measure = patterns("^Otu"),
+      variable.name = "otu", value.name = "count") %>%
+  dcast(otu ~ Group, value.var = "count") %>%
+  column_to_rownames("otu")
+
+nseqs <- apply(shared_df, 2, sum) %>%
+  as_tibble(rownames = "Group")
+
+asy_richness <- ChaoRichness(shared_df) %>%
+  as_tibble(rownames = "Group") %>%
+  select(Group, chao_sobs = Estimator)
+
+asy_shannon <- ChaoShannon(shared_df, B = 1) %>%
+  as_tibble(rownames = "Group") %>%
+  select(Group, chao_shannon = Estimator)
+
+asy_simpson <- ChaoSimpson(shared_df, B = 1, transform = TRUE) %>%
+  as_tibble(rownames = "Group") %>%
+  select(Group, chao_invsimpson = Estimator)
+
+asy <- inner_join(asy_richness, asy_shannon, by = "Group") %>%
+  inner_join(., asy_simpson, by = "Group")
+
+coverage_qd <- estimateD(shared_df,
+                         q = c(0, 1, 2),
+                         datatype = "abundance",
+                         base = "coverage",
+                         nboot = 0) %>%
+  mutate(Diversity = recode(Order.q,
+                            "0" = "Species richness",
+                            "1" = "Shannon diversity",
+                            "2" = "Simpson diversity")) %>%
+  select(Assemblage, coverage_method = Method, Diversity, Coverage = qD) %>%
+  pivot_wider(names_from = Diversity, values_from = Coverage) %>%
+  select(Group = Assemblage,
+        coverage_method,
+        coverage_sobs = `Species richness`,
+        coverage_shannon = `Shannon diversity`,
+        coverage_invsimpson = `Simpson diversity`)
+
+size_qd <- estimateD(shared_df,
+                     q = c(0, 1, 2),
+                     datatype = "abundance",
[…]

### 2023-10-23 code/plot_alpha_beta_depth_correlation.R

@@ -19,29 +19,30 @@ pretty_datasets <- tibble(
 )
 
 ## Alpha diversity panel...
-
 alpha_metrics <- c(
   "rare_sobs",
   "raw_sobs",
   "srs_sobs",
   "raw_ace",
-  "raw_chao",
+  "chao_sobs",
+  "size_sobs",
+  "coverage_sobs",
   "ba_default",
   "ba_poisson",
-  
+
   "rare_shannon",
   "raw_shannon",
   "srs_shannon",
-  "raw_npshannon",
-  
+  "chao_shannon",
+  "size_shannon",
+  "coverage_shannon",
+
   "rare_simpson",
   "raw_simpson",
-  "srs_simpson"
-  
-  # "ba_sobs", "rare_ace", "rare_chao", "rare_coverage", "rare_npshannon",
-  # "raw_coverage", 
-  # "srs_ace", "srs_chao", "srs_coverage",
-  # "srs_npshannon", 
+  "srs_simpson",
+  "chao_invsimpson",
+  "size_invsimpson",
+  "coverage_invsimpson"
 )
 
 alpha_classes <- c(
@@ -52,14 +53,21 @@ alpha_classes <- c(
   "richness",
   "richness",
   "richness",
-  
+  "richness",
+  "richness",
+
   "shannon",
   "shannon",
   "shannon",
   "shannon",
-  
+  "shannon",
+  "shannon",
+
   "inv_simpson",
-  "inv_simpson", 
+  "inv_simpson",
+  "inv_simpson",
+  "inv_simpson",
+  "inv_simpson",
   "inv_simpson"
 )
 
@@ -69,18 +77,26 @@ alpha_labels <- c(
   "SRS Normalized",
   "ACE Estimate",
   "Chao1 Estimate",
+  "iNEXT (Size)",
+  "iNEXT (Coverage)",
   "BA Default",
   "BA Poisson",
-  
+
   "Rarefaction",
   "Raw",
   "SRS Normalized",
   "Estimate",
-  
+  "iNEXT (Size)",
+  "iNEXT (Coverage)",
+
   "Rarefaction",
   "Raw",
-  "SRS Normalized"
+  "SRS Normalized",
+  "Estimate",
+  "iNEXT (Size)",
+  "iNEXT (Coverage)"
 )
+
 
 pretty_alpha_classes <- c(
   richness = "Richness",
[…]

### 2023-10-23 code/plot_example_alpha_cor.R

@@ -5,37 +5,62 @@ library(glue)
 library(broom)
 library(ggtext)
 
-raw_sobs <- "data/human/data.otu.1.r_raw_alpha" %>%
+raw <- "data/human/data.otu.1.r_raw_alpha" %>%
   read_tsv(col_types = cols(group = col_character())) %>%
-    select(group, nseqs,
-           raw = sobs)
+  select(group, nseqs,
+        raw = sobs,
+        ace)
 
-rare_sobs <- "data/human/data.otu.1.r_rarefy_alpha" %>%
+rare <- "data/human/data.otu.1.r_rarefy_alpha" %>%
   read_tsv(col_types = cols(group = col_character())) %>%
-    filter(method == "ave") %>%
-    select(group,
-           rare = sobs)
+  filter(method == "ave") %>%
+  select(group,
+          rare = sobs)
 
-srs_sobs <- "data/human/data.otu.1.r_srs_alpha" %>%
+srs <- "data/human/data.otu.1.r_srs_alpha" %>%
   read_tsv(col_types = cols(group = col_character())) %>%
     select(group,
            srs = sobs)
 
-ba_sobs <- "data/human/data.otu.1.r_breakaway_alpha" %>%
+ba <- "data/human/data.otu.1.r_breakaway_alpha" %>%
   read_tsv(col_types = cols(Group = col_character())) %>%
-    select(group = Group,
-           ba_default = ba_est,
-           ba_poisson = p_est)
+  select(group = Group,
+        ba_default = ba_est,
+        ba_poisson = p_est)
+
+inext <- "data/human/data.otu.1.r_inext_alpha" %>%
+  read_tsv(col_types = cols(group = col_character())) %>%
+  select(group,
+        chao = chao_sobs,
+        size = size_sobs,
+        coverage = coverage_sobs
+        )
+
+inext_methods <- "data/human/data.otu.1.r_inext_alpha" %>%
+  read_tsv(col_types = cols(group = col_character())) %>%
+  select(group,
+        coverage = coverage_method,
+        size = size_method) %>%
+  pivot_longer(-group, names_to = "metric", values_to = "method") %>%
+  mutate(method = tolower(method))
 
 
 
-tidy_composite <- inner_join(raw_sobs, rare_sobs, by = "group") %>%
-  inner_join(., srs_sobs, by = "group") %>%
-  inner_join(., ba_sobs, by = "group") %>%
-  select(-ba_poisson) %>%
+
[…]

### 2023-10-23 code/plot_false_positive_null_model.R

@@ -25,19 +25,25 @@ alpha_metrics <- c(
   "sobs_raw",
   "sobs_srs",
   "ace_raw",
-  "chao_raw",
+  "chao_sobs_inext",
+  "size_sobs_inext",
+  "coverage_sobs_inext",
   "default_breakaway",
   "poisson_breakaway",
 
   "shannon_rarefy",
   "shannon_raw",
   "shannon_srs",
-  "npshannon_raw",
+  "chao_shannon_inext",
+  "size_shannon_inext",
+  "coverage_shannon_inext",
 
   "simpson_rarefy",
   "simpson_raw",
-  "simpson_srs"
-
+  "simpson_srs",
+  "chao_invsimpson_inext",
+  "size_invsimpson_inext",
+  "coverage_invsimpson_inext"
 )
 
 alpha_classes <- c(
@@ -48,12 +54,19 @@ alpha_classes <- c(
   "richness",
   "richness",
   "richness",
+  "richness",
+  "richness",
 
   "shannon",
   "shannon",
   "shannon",
   "shannon",
+  "shannon",
+  "shannon",
 
+  "inv_simpson",
+  "inv_simpson",
+  "inv_simpson",
   "inv_simpson",
   "inv_simpson",
   "inv_simpson"
@@ -65,6 +78,8 @@ alpha_labels <- c(
   "SRS Normalized",
   "ACE Estimate",
   "Chao1 Estimate",
+  "iNEXT (Size)",
+  "iNEXT (Coverage)",
   "BA Default",
   "BA Poisson",
 
@@ -72,10 +87,15 @@ alpha_labels <- c(
   "Raw",
   "SRS Normalized",
   "Estimate",
+  "iNEXT (Size)",
+  "iNEXT (Coverage)",
 
   "Rarefaction",
   "Raw",
-  "SRS Normalized"
+  "SRS Normalized",
+  "Estimate",
+  "iNEXT (Size)",
+  "iNEXT (Coverage)"
 )
 
 pretty_alpha_classes <- c(
@@ -126,7 +146,6 @@ alpha <- alpha_composite %>%
       legend.key.size = unit(12, "pt"),
       axis.line.x = element_line()
     )
-
 
 ## Beta diversity panel...
 
@@ -232,13 +251,8 @@ beta <- beta_composite %>%
       axis.line.x = element_line()
       )
 
[…]

### 2023-10-23 code/plot_false_positive_null_model_size.R

@@ -19,28 +19,30 @@ pretty_datasets <- tibble(
 )
 
 ## Alpha diversity panel...
-
 alpha_metrics <- c(
   "sobs_rarefy",
   "sobs_raw",
   "sobs_srs",
   "ace_raw",
-  "chao_raw",
+  "chao_sobs_inext",
+  "size_sobs_inext",
+  "coverage_sobs_inext",
   "default_breakaway",
   "poisson_breakaway",
 
   "shannon_rarefy",
   "shannon_raw",
   "shannon_srs",
-  "npshannon_raw",
+  "chao_shannon_inext",
+  "size_shannon_inext",
+  "coverage_shannon_inext",
 
   "simpson_rarefy",
   "simpson_raw",
-  "simpson_srs"
-
-#"ace_rarefy", "chao_rarefy", "coverage_rarefy", "npshannon_rarefy",
-#"nseqs_rarefy", "npshannon_srs", "coverage_raw", "coverage_srs", "nseqs_srs",
-#"nseqs_raw", "ace_srs", "chao_srs",
+  "simpson_srs",
+  "chao_invsimpson_inext",
+  "size_invsimpson_inext",
+  "coverage_invsimpson_inext"
 )
 
 alpha_classes <- c(
@@ -51,12 +53,19 @@ alpha_classes <- c(
   "richness",
   "richness",
   "richness",
+  "richness",
+  "richness",
 
   "shannon",
   "shannon",
   "shannon",
   "shannon",
+  "shannon",
+  "shannon",
 
+  "inv_simpson",
+  "inv_simpson",
+  "inv_simpson",
   "inv_simpson",
   "inv_simpson",
   "inv_simpson"
@@ -68,6 +77,8 @@ alpha_labels <- c(
   "SRS Normalized",
   "ACE Estimate",
   "Chao1 Estimate",
+  "iNEXT (Size)",
+  "iNEXT (Coverage)",
   "BA Default",
   "BA Poisson",
 
@@ -75,11 +86,17 @@ alpha_labels <- c(
   "Raw",
   "SRS Normalized",
   "Estimate",
+  "iNEXT (Size)",
+  "iNEXT (Coverage)",
 
   "Rarefaction",
   "Raw",
-  "SRS Normalized"
+  "SRS Normalized",
+  "Estimate",
+  "iNEXT (Size)",
+  "iNEXT (Coverage)"
 )
+
 
[…]

### 2023-10-23 code/plot_power_cffect_model.R

@@ -19,24 +19,30 @@ pretty_datasets <- tibble(
 )
 
 ## Alpha diversity panel...
-
 alpha_metrics <- c(
   "sobs_rarefy",
   "sobs_raw",
   "sobs_srs",
   "ace_raw",
-  "chao_raw",
+  "chao_sobs_inext",
+  "size_sobs_inext",
+  "coverage_sobs_inext",
   "default_breakaway",
   "poisson_breakaway",
 
   "shannon_rarefy",
   "shannon_raw",
   "shannon_srs",
-  "npshannon_raw",
+  "chao_shannon_inext",
+  "size_shannon_inext",
+  "coverage_shannon_inext",
 
   "simpson_rarefy",
   "simpson_raw",
-  "simpson_srs"
+  "simpson_srs",
+  "chao_invsimpson_inext",
+  "size_invsimpson_inext",
+  "coverage_invsimpson_inext"
 )
 
 alpha_classes <- c(
@@ -47,12 +53,19 @@ alpha_classes <- c(
   "richness",
   "richness",
   "richness",
+  "richness",
+  "richness",
 
   "shannon",
   "shannon",
   "shannon",
   "shannon",
+  "shannon",
+  "shannon",
 
+  "inv_simpson",
+  "inv_simpson",
+  "inv_simpson",
   "inv_simpson",
   "inv_simpson",
   "inv_simpson"
@@ -64,6 +77,8 @@ alpha_labels <- c(
   "SRS Normalized",
   "ACE Estimate",
   "Chao1 Estimate",
+  "iNEXT (Size)",
+  "iNEXT (Coverage)",
   "BA Default",
   "BA Poisson",
 
@@ -71,11 +86,17 @@ alpha_labels <- c(
   "Raw",
   "SRS Normalized",
   "Estimate",
+  "iNEXT (Size)",
+  "iNEXT (Coverage)",
 
   "Rarefaction",
   "Raw",
-  "SRS Normalized"
+  "SRS Normalized",
+  "Estimate",
+  "iNEXT (Size)",
+  "iNEXT (Coverage)"
 )
+
 
 pretty_alpha_classes <- c(
   richness = "Richness",
@@ -129,5 +150,4 @@ alpha <- alpha_composite %>%
     )
 
 ggsave("figures/power_cffect_model.tiff", alpha, compression = "lzw+p",
-      width = 6, height = 3)
-
+      width = 7, height = 3)

### 2023-10-23 code/plot_power_effect_model.R

@@ -19,24 +19,30 @@ pretty_datasets <- tibble(
 )
 
 ## Alpha diversity panel...
-
 alpha_metrics <- c(
   "sobs_rarefy",
   "sobs_raw",
   "sobs_srs",
   "ace_raw",
-  "chao_raw",
+  "chao_sobs_inext",
+  "size_sobs_inext",
+  "coverage_sobs_inext",
   "default_breakaway",
   "poisson_breakaway",
 
   "shannon_rarefy",
   "shannon_raw",
   "shannon_srs",
-  "npshannon_raw",
+  "chao_shannon_inext",
+  "size_shannon_inext",
+  "coverage_shannon_inext",
 
-  "simpson_raw",
   "simpson_rarefy",
-  "simpson_srs"
+  "simpson_raw",
+  "simpson_srs",
+  "chao_invsimpson_inext",
+  "size_invsimpson_inext",
+  "coverage_invsimpson_inext"
 )
 
 alpha_classes <- c(
@@ -47,12 +53,19 @@ alpha_classes <- c(
   "richness",
   "richness",
   "richness",
+  "richness",
+  "richness",
 
   "shannon",
   "shannon",
   "shannon",
   "shannon",
+  "shannon",
+  "shannon",
 
+  "inv_simpson",
+  "inv_simpson",
+  "inv_simpson",
   "inv_simpson",
   "inv_simpson",
   "inv_simpson"
@@ -64,6 +77,8 @@ alpha_labels <- c(
   "SRS Normalized",
   "ACE Estimate",
   "Chao1 Estimate",
+  "iNEXT (Size)",
+  "iNEXT (Coverage)",
   "BA Default",
   "BA Poisson",
 
@@ -71,10 +86,15 @@ alpha_labels <- c(
   "Raw",
   "SRS Normalized",
   "Estimate",
+  "iNEXT (Size)",
+  "iNEXT (Coverage)",
 
   "Rarefaction",
   "Raw",
-  "SRS Normalized"
+  "SRS Normalized",
+  "Estimate",
+  "iNEXT (Size)",
+  "iNEXT (Coverage)"
 )
 
 pretty_alpha_classes <- c(
@@ -236,19 +256,16 @@ beta <- beta_composite %>%
       axis.line.x = element_line()
       )
 
-layout <- "
-AAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAB
[…]

### 2023-10-23 code/run_alpha_kw.R

@@ -54,6 +54,23 @@ get_raw_results <- function(rawfile, dfile) {
 
 }
 
+get_inext_results <- function(inextfile, dfile) {
+
+  inner_join(read_tsv(inextfile, show_col_types = FALSE),
+             read_tsv(dfile, show_col_types = FALSE),
+             by = c("group" = "Group")) %>%
+    select(group, treatment, chao_sobs, chao_shannon, chao_invsimpson,
+          coverage_sobs, coverage_shannon, coverage_invsimpson, size_sobs,
+          size_shannon, size_invsimpson) %>%
+    pivot_longer(cols = -c(group, treatment), names_to = "metric") %>%
+    nest(data = -metric) %>%
+    mutate(kw = map(data, ~tidy(kruskal.test(.x$value, g = .x$treatment)))) %>%
+    unnest(kw) %>%
+    select(metric, p.value) %>%
+    mutate(method = "inext")
+
+}
+
 get_srs_results <- function(srsfile, dfile) {
 
   inner_join(read_tsv(srsfile, show_col_types = FALSE),
@@ -78,15 +95,15 @@ run_kw_tests <- function(dataset, method, model) {
       alpha_file = list.files(path = dataset,
                 pattern = glue(".*{model}_{method}_alpha"),
                 include.dirs = TRUE, full.names = TRUE)) %>%
-      mutate(seed = str_replace(alpha_file, ".*\\.(\\d+)\\..*", "\\1")) %>%
-      inner_join(., design_files, by = "seed") %>%
-      nest(data = -seed) %>%
-      mutate(results = map(data,
-                          ~do.call(f, list(.x$alpha_file, .x$design_file)))) %>%
-      select(results) %>%
-      unnest(results) %>%
-      group_by(metric, method) %>%
-      summarize(frac = sum(p.value < 0.05) / n(), .groups = "drop")
+    mutate(seed = str_replace(alpha_file, ".*\\.(\\d+)\\..*", "\\1")) %>%
+    inner_join(., design_files, by = "seed") %>%
+    nest(data = -seed) %>%
+    mutate(results = map(data,
+                        ~do.call(f, list(.x$alpha_file, .x$design_file)))) %>%
+    select(results) %>%
+    unnest(results) %>%
+    group_by(metric, method) %>%
+    summarize(frac = sum(p.value < 0.05) / n(), .groups = "drop")
 }
 
 #########
[…]

### 2023-10-26 submission/response_to_reviewers.Rmd
Rough in response to reviewers

---
output: pdf_document
geometry: margin=1.0in
header-includes:
 - \usepackage{setspace}
 - \doublespacing
 - \usepackage{helvet} % Helvetica font
 - \renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
 - \usepackage[T1]{fontenc}
 - \usepackage[shortcuts]{extdash}
 - \usepackage[document]{ragged2e}
---

\raggedright

**Reviewer #1 (Comments for the Author):**

**Within this manuscript Pat Schloss reports on the use of rarefaction to effectively control unequal observation efforts in 16S rRNA gene sequencing due to differences in sequencing depth. This paper is in response to a controversial idea within the microbiome community about whether rarefaction or other methods of controlling sequencing depth during diversity analysis are most appropriate. Overall, this paper shows convincing results that rarefaction is the most appropriate method for controlling unequal sampling depths at the present time. It was a pleasure to read and overall, I was happy with the manuscript. However, I do have a few comments that should be addressed by the author.**

**Major comments:**

**1. I found the result of increasing CoV in Bray-Curtis distances with increasing sequencing depth to be an interesting result given that increased sequencing depth should generally result in better confidence of what is within the underlying population. I'm wondering if the CoV results are difficult to understand due to the changing number of samples included at each sampling depth. Could this analysis be redone using only samples with sequencing depths that are above the highest rarefaction value chosen so that sample number does not vary across depth (this could be done in a subset of the samples/datasets presented). If changing sample number is not to blame are there any further speculations as to what is going on here?**

**2. It is clear based on the results shown that treatment groups that are highly associated with variance in sequencing depth will result in high false positive rates during diversity analysis when rarefaction is not used. However, I would argue that the simulation used does not in general reflect many natural datasets. It would be uncommon for most natural datasets to vary by treatment group by splitting sequencing depth via the median. I am wondering if this analysis could be done with varying levels of association between treatment group and sequencing depth. This will give the readers a better idea of how large this issue is in actual practice as the true association for most datasets may lay somewhere between figures 2 and 3. This would also help address how much of a concern this may be for previously published papers that do not use rarefaction during diversity analysis.**

**3. Based on the results of this paper it seems the key difference that is being demonstrated in these simulations as opposed to other works is the use of true rarefaction rather than subsampling. Further comments on this in the discussion would be useful to highlight why these results differ so significantly from other work such as the paper by Martino et al., in 2019 (PMID: 30801021) that introduced using robust CLR measures for microbiome data (and is quickly becoming one of the preferred choices of microbiome scientists). Furthermore, it would be nice to see whether the use of true rarefaction on the datasets presented in Martino et al., 2019 resolves the conflicting views between this report and Matino et al's findings on robust CLR measures. Overall, a clarification on how these two publications come to differing conclusions about best practices when comparing beta diversity would be helpful to a large audience within the microbiome field.**

**4. I am wondering whether any of the differences in the results derived from studies such as Martino et al., 2019 (PMID: 30801021), can be attributed to sequence processing pipeline. While the author makes convincing arguments that removal of low abundance sequences may not always be appropriate, I am wondering how it would impact the results shown. Especially given the widespread use of this practice in the current literature as indicated by the author in the discussion. This would be especially important given the strong language around this practice in the discussion without primary data to back it up within this manuscript.**

**5. One line 164 the report indicates finding differences in richness between samples that should have the same OTU counts given the way the enrichment sampling was done (increasing sequencing counts not raw number of OTUS in the sample). It would make sense that metrics that consider abundances to change but not richness. Comments on this would be appreciated. Furthermore, this section jumps around between two different models of enrichment (abundance enrichment vs. raw count enrichment). A clearer separation of these two analyses would be appreciated to make parsing the text easier as in some cases it is unclear which enrichment strategy was used for the result that is being talked about.**

**Minor comments:**

**1. Well, the current consensus is that sequence depth is mostly attributed to noise (Gloor et al., 2017. PMID: 29187837). It would be nice to comment on some recent papers that indicate that through certain methods of library preparation sequencing depth may be correlated with sample biomass and how this reflects with the current thoughts in this manuscript (Cruz et al., 2021. PMID: 33717032, Munich et al., 2022. PMID: 36396943).**

**2. It would be nice if confidence intervals could be included in the correlation plots in Figure 1. Furthermore, if understood correctly Figures 2-3 y-axis should be the mean false positive rate over 100 dataset simulations? If so a measure of data spread would be appropriate for these values.**

**3. While I agree that the rarefaction value chosen is not arbitrary and is usually a balance between keeping samples and having the highest sequencing depth possible (as indicated in the discussion of this manuscript). It would be nice to give stronger guidelines around choosing an appropriate value to highlight more robust ways of choosing rarefaction depths. One thought that came across my mind would be the use of CoV calculations from repeated subsampling (rarefaction) to determine appropriate sequencing depths. Or perhaps just a simple calculation of good's coverage at various sequence depths (similar to rarefaction plots used now)?**

**4. Based on these results and various studies on microbiome biases, alpha diversity metrics such as richness do not indicate the true richness of a sample due to a variety of issues included sequencing depth. A comment on this in the discussion would be helpful, for those that are less seasoned in the microbiome field.**

**Reviewer #2 (Comments for the Author):**

**This study addresses a controversy that has been playing out in the literature regarding the processing of 16S rRNA gene amplicon data. Some researchers contend that rarefaction (the random subsampling with replacement to account for differences in sequencing depth among samples) is undesireable because it "wastes data". Schloss does a comprehensive analysis of how rarefaction performs compared to other ways of processing the samples (including using raw counts). He compares both alpha and beta diversity metrics using both a mock dataset and a collection of published datasets from a range of ecosystems. In nearly all cases, rarefaction**

**My only "major" critique relates to the core purpose of the article: to convince readers that rarefaction is a better way to process data than those proposed by other researchers. After reading the article multiple times, I still don't yet comprehend why the other researchers came to their conclusions. At some level, they must have convinced themselves that the use of raw read count (or other methods) out-perform rarefaction. Did they really not do the same comparisons that are done here? Without reading those carefully, I can't tell. Schloss does mention that they removed rare OTUs in their analysis. Is this the main reason why they did not detect the superior performance of rarefaction? I think just a brief explanation for Schloss' interpretation of why his conclusions are different from McMurdle/Holmes would be even more convincing for readers who don't want to dig too deeply into other papers. There is some discussion in the introduction, but I'd like to see it in the actual discussion section, after the full presentation of the new analysis.**

**Kudos. Love the running title. Tell us how you really feel!**

**Minor comments**
**Line 69 - This implies that McMurdle/Holmes used an incorrect interpretation of differences between rarefying and rarefaction. This is an opportunity to clarify! Do you agree with their interpretation (but want to promote the use of rarefaction specifically)?**

**Line 72 - Do you mean that the reanalysis did not remove 15% of the samples?**

**Line 72 - Change "method" to "methods"**

**Line 78 - Did they use the term "toy" in their work? This comes off as unnecessarily snarky. Would it be appropriate to use "simulated" instead?**

**Line 94 - Delete one of the "from"s**

**Line 113 - I would say that the Raw and SRS normalized are completely insensitive to seq depth when using inverse Simpson, based on Fig 1A? Not sure this matters in the end, but be transparent in the explanation of the results.**

**Line 167 - Were the 3% of OTUs selected at random, or did you remove the rarest 3%?**

**Line 194 - Change "increased" to "increase"**

**Line 210 - Drop your excessive sig figs .... Use "101" fold**

**Line 281 - I think this should be "my personal process for selecting the rarefaction threshold"?**

### 2023-10-26 code/render_manuscript.R

@@ -1,11 +1,0 @@
-#!/usr/bin/env Rscript
-
-library('rmarkdown')
-
-if(!tinytex::is_tinytex()) {
-  tinytex::install_tinytex()
-} 
-
-render('submission/manuscript.Rmd', clean=FALSE)
-
-file.rename("submission/manuscript.knit.md", "submission/manuscript.md")

### 2023-10-26 code/render_rmd.R

@@ -1,0 +1,14 @@
+#!/usr/bin/env Rscript
+
+library("rmarkdown")
+
+if (!tinytex::is_tinytex()) {
+  tinytex::install_tinytex(force = TRUE)
+}
+
+args <- commandArgs(trailingOnly = TRUE)
+rmd <- args[1]
+
+render(rmd, output_format = "all", clean = FALSE)
+
+file.remove(gsub("Rmd", "log", rmd))

### 2023-11-16 code/defunct/get_observed_depth_alpha.R
Remove old code no longer in use

@@ -1,45 +1,0 @@
-#!/usr/bin/env Rscript
-
-library(tidyverse)
-library(glue)
-
-input <- commandArgs(trailingOnly = TRUE)
-
-shared_file <- input[1] # shared_file <- "data/human/data.otu.shared"
-depth <- as.numeric(input[2]) # depth <- 1000
-
-mothur_output_file <-
-      str_replace(shared_file,
-                  "shared",
-                  glue("temp_obs_{depth}_alpha.groups.ave-std.summary"))
-
-script_output_file <- str_replace(shared_file,
-                                  "shared",
-                                  glue("obs_{depth}_alpha.summary"))
-
-temp_shared_file <- str_replace(shared_file, "shared",
-                                glue("temp_obs_{depth}_alpha.shared"))
-
-if (file.exists(temp_shared_file)) {
-  unlink(temp_shared_file)
-}
-
-file.copy(shared_file, temp_shared_file, overwrite = TRUE)
-
-
-
-
-mothur <- glue("mothur '#summary.single(shared = {temp_shared_file},
-                        calc=sobs-shannon-simpson-coverage,
-                        subsample={depth})'")
-
-system(mothur)
-
-file.rename(mothur_output_file, script_output_file)
-
-temp_files <- list.files(path = dirname(temp_shared_file),
-                         pattern = str_replace(basename(temp_shared_file),
-                                               ".shared", ""),
-                         full.names = TRUE)
-
-unlink(temp_files)

### 2023-11-16 code/defunct/get_observed_depth_alpha_kw.R

@@ -1,71 +1,0 @@
-#!/usr/bin/env Rscript
-
-library(tidyverse)
-library(glue)
-library(effectsize)
-library(broom)
-
-get_rarefy_results <- function(rfile, design) {
-
-  alpha_data <- read_tsv(rfile, col_types = cols(group = col_character(),
-                                                 method = col_character(),
-                                                 .default = col_double()))
-
-  inner_join(alpha_data,
-             design,
-             by = "group") %>% 
-    filter(method == "ave") %>%
-    select(Group = group, treatment, sobs, shannon, simpson, coverage) %>%
-    pivot_longer(cols = -c(Group, treatment), names_to = "metric") %>%
-    nest(data = -metric) %>%
-    mutate(kw = map(data, ~tidy(kruskal.test(.x$value, g = .x$treatment))),
-           biserial = map(data, ~rank_biserial(.x$value~as.factor(.x$treatment),
-                                                    ci = NULL)[[1]][[1]]),
-           median = map(data,
-                        ~group_by(.x, treatment) %>%
-                          summarize(median = median(value),
-                                    n = n())
-                          )) %>%
-    unnest(c(kw, median, biserial)) %>%
-    select(metric, p.value, biserial, treatment, median, n) %>%
-    pivot_wider(names_from = "treatment", values_from = c("median", "n"))
-
-}
-
-
-run_kw_tests <- function(alpha_file, treatment) {
-
-  nseqs <- as.numeric(str_replace(alpha_file, ".*_(\\d*)_.*", "\\1"))
-
-  design_file <- "data/human/data.odesign"
-
-  design <- read_tsv(design_file,
-                     col_types = cols(.default = col_character())) %>%
-    select(group = Group, treatment = all_of(treatment))
-
-  get_rarefy_results(alpha_file, design) %>%
-    mutate("treatment" = treatment) %>%
-    mutate(nseqs = nseqs, .after = "metric")
-
-}
-
-#########
-
-design_file <- "data/human/data.odesign"
-output_file <- "data/human/data.otu.obs_depth_alpha_kw.sumary"
-
-summary_files <- list.files(path = "data/human",
-                            pattern = "data.otu.obs_\\d*_alpha.summary",
-                            full.names = TRUE)
-
-metadata <- c("srn", "obese", "female", "smoker", "nsaid", "prev_crc",
-              "prev_polyps")
-
-
[…]

### 2023-11-16 code/defunct/get_observed_depth_beta.R

@@ -1,26 +1,0 @@
-#!/usr/bin/env Rscript
-
-library(tidyverse)
-library(glue)
-library(vegan)
-library(data.table)
-
-
-input <- commandArgs(trailingOnly = TRUE)
-
-shared_file <- input[1] # shared_file <- "data/human/data.otu.shared"
-depth <- as.numeric(input[2]) # depth <- 1000
-distance <- input[3] # distance <- "bray"
-
-dist_file <- str_replace(shared_file,
-                         "shared",
-                         glue("obs_{depth}_{distance}.dist"))
-
-
-fread(shared_file, colClasses = c(Group = "character")) %>%
-  select(Group, starts_with("Otu")) %>%
-  column_to_rownames("Group") %>%
-  avgdist(sample = depth, dmethod = distance)  %>%
-      as.matrix() %>%
-      as_tibble(rownames = "Group") %>%
-      write_tsv(dist_file)

### 2023-11-16 code/defunct/get_observed_depth_beta_amova.R

@@ -1,77 +1,0 @@
-#!/usr/bin/env Rscript
-
-library(tidyverse)
-library(glue)
-library(vegan)
-
-run_amova <- function(distance_file, treatment) {
-
-  design <- read_tsv("data/human/data.odesign",
-                     col_types = cols(.default = col_character())) %>%
-    select(Group, "treatment" = all_of(treatment)) %>%
-    distinct()
-
-  distance_design <- read_tsv(distance_file,
-                              col_types = cols(Group = col_character(),
-                                               .default = col_double()
-                                               )
-                              ) %>%
-    inner_join(design, ., by = "Group")
-
-  w_na <- nrow(distance_design)
-  wo_na <- distance_design %>%
-    drop_na() %>%
-    nrow()
-
-  result <- tibble(p.value = NA_real_,
-                   r_squared = NA_real_,
-                   n_FALSE = NA_real_,
-                   n_TRUE = NA_real_
-                   )
-
-  if (w_na == wo_na) {
-    distance_matrix <- distance_design %>%
-      select(Group, distance_design$Group) %>%
-      column_to_rownames("Group") %>%
-      as.dist()
-
-    test <- vegan::adonis2(distance_matrix~distance_design$treatment)
-
-    tf_count <- distance_design %>% count(treatment) %>%
-      pivot_wider(names_from = treatment, values_from = "n")
-
-    result <- tibble(p.value = test$`Pr(>F)`[1],
-                     r_squared = test$R2[1],
-                     n_FALSE = tf_count$`FALSE`,
-                     n_TRUE = tf_count$`TRUE`)
-  }
-
-  return(result)
-
-}
-
-
-design_file <- read_tsv("data/human/data.odesign")
-treatments <- colnames(design_file)[-1]
-
-output_file <- "data/human/data.otu.obs_depth_beta_amova.summary"
-
-distance_design <- expand_grid(
-    distance_file = list.files(path = "data/human",
-                               pattern = "data.otu.obs_\\d*_\\w*.dist",
-                               full.names = TRUE),
-    treatment = treatments
-    )
-
-
-distance_design %>%
-  rowwise() %>%
-  mutate(result = run_amova(distance_file, treatment)) %>%
-  unnest(result) %>%
-  ungroup() %>%
-  separate(distance_file,
-           into = c("dummy", "nseqs", "beta_calculator"),
-           sep = "_") %>%
[…]

### 2023-11-16 code/defunct/get_observed_shared_design.R

@@ -1,35 +1,0 @@
-#!/usr/bin/env Rscript
-
-library(tidyverse)
-library(data.table)
-
-shared_file <- "data/human/data.otu.shared"
-remove_file <- "data/human/data.remove_accnos"
-
-obs_design_file <- "data/human/data.odesign"
-obs_shared_file <- "data/human/data.otu.oshared"
-
-removal <- scan(remove_file, quiet = TRUE, what = character()) %>%
-  str_split(., "-") %>%
-  unlist()
-
-# some sample names appear twice, but their diagnoses are the same. this was
-# because some samples were sequenced twice
-read_tsv("data/human/sra_info.tsv",
-         col_type = cols(Sample_Name = col_character())) %>% 
-  mutate(srn = diagnosis == "Cancer" | diagnosis == "adv Adenoma",
-         obese = BMI >= 30,
-         smoker = as.logical(Smoke),
-         nsaid = as.logical(NSAID),
-         prev_crc = as.logical(Hx_Prev),
-         prev_polyps = as.logical(Hx_of_Polyps),
-         female = Gender == 'female'
-         ) %>%
-  select(Group = Sample_Name, srn, obese, female, smoker, nsaid, 
-         prev_crc, prev_polyps) %>%
-  filter(! Group %in% removal) %>%
-  write_tsv(obs_design_file)
-
-rand_shared <- fread(shared_file, colClasses = c(Group = "character")) %>%
-  filter(! Group %in% removal) %>%
-  write_tsv(obs_shared_file)

### 2023-11-16 code/defunct/run_oalpha_kw.R

@@ -1,144 +1,0 @@
-#!/usr/bin/env Rscript
-
-library(tidyverse)
-library(glue)
-library(broom)
-
-get_rarefy_results <- function(rfile, design) {
-
-  alpha_data <- read_tsv(rfile, col_types = cols(group = col_character(),
-                                                 method = col_character(),
-                                                 .default = col_double()))
-
-  inner_join(alpha_data,
-             design,
-             by = "group") %>%
-    filter(method == "ave") %>%
-    select(Group = group, treatment, nseqs, sobs, shannon,
-          simpson, chao, ace, npshannon, coverage) %>%
-    pivot_longer(cols = -c(Group, treatment), names_to = "metric") %>%
-    nest(data = -metric) %>%
-    mutate(kw = map(data, ~tidy(kruskal.test(.x$value, g = .x$treatment)))) %>%
-    unnest(kw) %>%
-    mutate(median = map(data,
-                        ~group_by(.x, treatment) %>%
-                          summarize(median = median(value))
-                          )) %>%
-    unnest(median) %>%
-    select(metric, p.value, statistic, treatment, median) %>%
-    pivot_wider(names_from = "treatment", values_from = "median") %>%
-    mutate(method = "rarefy")
-
-}
-
-get_breakaway_results <- function(bafile, design) {
-
-  alpha_data <- read_tsv(bafile, col_types = cols(Group = col_character(),
-                                                 ba_model = col_character(),
-                                                 p_model = col_character(),
-                                                 .default = col_double()))
-
-  inner_join(alpha_data,
-             design,
-             by = c("Group" = "group")) %>%
-    select(Group, treatment, ba_est, ba_model, p_est, p_model) %>%
-    pivot_longer(cols = -c(Group, treatment), names_to = c("set", ".value"),
-                 names_pattern = "(.*)_(.*)") %>%
-    nest(data = -set) %>%
-    mutate(kw = map(data, ~tidy(kruskal.test(.x$est, g = .x$treatment)))) %>%
-    unnest(kw) %>%
-    mutate(median = map(data,
-                        ~group_by(.x, treatment) %>%
-                          summarize(median = median(est))
-                          )) %>%
-    unnest(median) %>%
-    select(metric = set, p.value, statistic, treatment, median) %>%
-    pivot_wider(names_from = "treatment", values_from = "median") %>%
[…]

### 2023-11-16 code/defunct/run_obeta_analysis.R

@@ -1,72 +1,0 @@
-#!/usr/bin/env Rscript
-
-library(tidyverse)
-library(glue)
-library(vegan)
-
-run_amova <- function(distance_file, treatment) {
-
-  design <- read_tsv("data/human/data.odesign",
-                     col_types = cols(.default = col_character())) %>%
-    select(Group, "treatment" = all_of(treatment)) %>%
-    distinct()
-
-  distance_design <- read_tsv(distance_file,
-                              col_types = cols(Group = col_character(),
-                                               .default = col_double()
-                                               )
-                              ) %>%
-    inner_join(design, ., by = "Group")
-
-  w_na <- nrow(distance_design)
-  wo_na <- distance_design %>%
-    drop_na() %>%
-    nrow()
-
-  result <- tibble(p.value = NA_real_,
-                   r_squared = NA_real_)
-
-  if (w_na == wo_na) {
-    distance_matrix <- distance_design %>%
-      select(Group, distance_design$Group) %>%
-      column_to_rownames("Group") %>%
-      as.dist()
-
-    test <- vegan::adonis2(distance_matrix~distance_design$treatment)
-
-    result <- tibble(p.value = test$`Pr(>F)`[1],
-                     r_squared = test$R2[1])
-  }
-
-  return(result)
-
-}
-
-dataset <- "data/human"
-output_file <- "data/human/data.o_oamova"
-
-model <- str_replace(output_file, ".*\\.(.)_.amova", "\\1")
-design <- str_replace(output_file, ".*\\.._(.)amova", "\\1")
-
-design_file <- read_tsv("data/human/data.odesign")
-treatments <- colnames(design_file)[-1]
-
-distance_design <- expand_grid(
-    distance_file = list.files(path = dataset,
-                               pattern = glue(".*\\.{model}_.*.dist"),
-                               full.names = TRUE),
-    treatment = treatments
-    )
-
-
-distance_design %>%
-  rowwise() %>%
-  mutate(result = run_amova(distance_file, treatment)) %>%
-  unnest(result) %>%
-  ungroup() %>%
-  separate(distance_file,
-           into = c("model", "beta_process", "beta_calculator"),
-           sep = "_") %>%
-  mutate(beta_calculator = str_replace(beta_calculator, ".dist", "")) %>%
-  select(-model) %>%
-  write_tsv(output_file)

### 2023-11-16 submission/response_to_reviewers.Rmd
Comment on increasing BC COV values

@@
 
+> I have added a set of panels to Figure 6 indicating the standard deviation across the samples.
+This hopefully makes it clear that the standard deviation between samples is decreasing, but not as fast as the mean distance between samples.
+On the whole, the COV values are fairly small.
+I have added text to this effect at LXXX.
+
 **2.
@@
 
+> XXXUse overlapping IQR instead of perfect confounding.
+> Currently have code/get_intraquantile_design.R set at 10%/90%; needs to be run
+> on great lakes
+
 **3.
@@
 Further comments on this in the discussion would be useful to highlight why these results differ so significantly from other work such as the paper by Martino et al., in 2019 (PMID: 30801021) that introduced using robust CLR measures for microbiome data (and is quickly becoming one of the preferred choices of microbiome scientists).
-Furthermore, it would be nice to see whether the use of true rarefaction on the datasets presented in Martino et al., 2019 resolves the conflicting views between this report and Matino et al's findings on robust CLR measures.
+Furthermore, it would be nice to see whether the use of true rarefaction on the datasets presented in Martino et al., 2019 resolves the conflicting views between this report and Martino et al's findings on robust CLR measures.
 Overall, a clarification on how these two publications come to differing conclusions about best practices when comparing beta diversity would be helpful to a large audience within the microbiome field.**
+
+> XXXDo this.
 
@@
 This would be especially important given the strong language around this practice in the discussion without primary data to back it up within this manuscript.**
+
+> XXXResponse to previous point will help to address this.
+But the data processing pipeline won't really factor in since all methods used the same data for a dataset.
+The methods don't know what type of processing was done upstream.
+Could have used purely synthetic distributions.
 
@@
 A clearer separation of these two analyses would be appreciated to make parsing the text easier as in some cases it is unclear which enrichment strategy was used for the result that is being talked about.**
+
+> XXXRevise text
 
@@
 
+> Look at suggested papers; cite Gloor
+
 **2.
@@
 If so a measure of data spread would be appropriate for these values.**
+
+> The caption for Figure 1 indicates that the size of the confidence interval was smaller than the plotting symbol.
+As stated in the captions for Figures 2 and 3, "The false positive rate was the number of times a dataset yielded a significant result."
 
@@
 
+> I'm reluctant to add more than what I already have at L281-289. In that text, I mention using Good's coverage values as a check to make sure that the level of coverage is acceptable.
+
 **4.
@@
 A comment on this in the discussion would be helpful, for those that are less seasoned in the microbiome field.**
+
+> I have added a sentence to this effect in the Discussion at **LXXX**
 

### 2023-11-16 submission/manuscript.Rmd
Comment on increasing BC COV values

@@
 In general, the coefficient of variation increased slightly with sequencing depth only to decline once smaller samples were removed from the analysis at higher sequencing depths.
-Interestingly, the coefficient of variation between Shannon diversity values decreased towards zero with increased sequencing effort and the coefficient of variation between Bray-Curtis distances tended to increase.
+Interestingly, the coefficient of variation between Shannon diversity values decreased towards zero with increased sequencing effort and the coefficient of variation between Bray-Curtis distances tended to increase slightly for samples where the standard deviation did not decrease as quickly as the mean distance between samples.
 Regardless, the coefficients of variation were relatively small.
@@
 Regardless, it is clear that better strategies are needed to reduce the variation in the number of sequences generated for each sample.
+Furthermore, it is important to keep in mind that because of residual sequencing errors, PCR artifacts, and incomplete sequencing, it is not possible to obtain absolute quantification of alpha or beta diversity values.
+Instead, comparisons of these values need to be made on a relative basis.
 

### 2023-11-16 code/plot_intrasample_variation.R

@@ -20,10 +20,10 @@ pretty_datasets <- tibble(
 
 
 get_samples_to_remove <- function(file) {
-  
+
   scan(file, what = character(), quiet = TRUE) %>%
     as_tibble_col(column_name = "sample")
-  
+
 }
 
 remove_files <- glue("data/{datasets}/data.remove_accnos")
@@ -44,7 +44,7 @@ smallest_sample_size <-
   group_by(dataset) %>%
   summarize(
     n_seqs = 2000 * ceiling(min(n_seqs) / 2000))
-  
+
 
 
 alpha_summary_files <- glue("data/{datasets}/data.otu.alpha_depth.summary")
@@ -58,22 +58,17 @@ alpha_composite <- map_dfr(alpha_summary_files, read_tsv, .id = "dataset") %>%
             sd = mean(sd),
             n_samples = mean(n_samples),
             .groups = "drop") %>%
-  filter(n_samples > 5) %>%
-  pivot_longer(cols = c(mean, cov),
+  filter(n_samples >= 5) %>%
+  pivot_longer(cols = c(mean, cov, sd),
                names_to = "statistic",
                values_to = "value") %>%
   mutate(dataset = factor(dataset, level = datasets),
          method = factor(method, levels = c("sobs", "shannon")),
-         statistic = factor(statistic, levels = c("mean", "cov"))
+         statistic = factor(statistic, levels = c("mean", "sd", "cov"))
          )
 
 
 
-datasets <- c("bioethanol", "human", "lake", "marine", "mice", "peromyscus",
-              "rainforest", "rice",
-              "seagrass", "sediment", "soil",
-              "stream"
-              )
 
 beta_summary_files <- glue("data/{datasets}/data.otu.beta_depth.summary")
 names(beta_summary_files) <- datasets
@@ -85,18 +80,18 @@ beta_composite <- map_dfr(beta_summary_files, read_tsv, .id = "dataset") %>%
             cov = 100 * mean(sd / mean),
             sd = mean(sd),
             n_samples = mean(n_samples),
-            .groups = "drop") %>% 
-  filter(n_samples > 5) %>% 
-  pivot_longer(cols = c(mean, cov),
+            .groups = "drop") %>%
+  filter(n_samples >= 5) %>%
+  pivot_longer(cols = c(mean, sd, cov),
                names_to = "statistic",
                values_to = "value") %>%
[…]

### 2023-11-16 code/run_cluster_analysis.R
Remove vestigial cluster analysis now in WNWN paper

@@ -1,64 +1,0 @@
-#!/usr/bin/env Rscript
-
-library(tidyverse)
-library(cluster)
-library(glue)
-
-get_accuracy <- function(seed, model, design, dataset, correct, dist) {
-
-  data <- read_tsv(glue("data/{dataset}/data.otu.{seed}.{model}_{correct}_{dist}.dist"),
-          col_types = cols(Group = col_character())) %>%
-    column_to_rownames("Group")
-  
-  if(is.na(data[1,2])) {
-    return(NA_real_)
-  }
-  
-  data %>%
-    pam(k = 2, diss = T) %>%
-    pluck("clustering") %>%
-    as_tibble(rownames = "Group") %>%
-    rename(cluster = value) %>%
-    inner_join(., read_tsv(glue("data/{dataset}/data.{seed}.{design}design"),
-                          col_types = cols(Group = col_character())),
-              by = "Group") %>%
-    count(cluster, treatment) %>%
-    pivot_wider(names_from = "treatment", values_from = "n", values_fill = 0) %>%
-    summarize(a = .[["A"]][1] + .[["B"]][2], b = .[["A"]][2] + .[["B"]][1]) %>%
-    pivot_longer(cols = c("a", "b")) %>%
-    mutate(accuracy = value / sum(value)) %>%
-    slice_max(accuracy, with_ties = FALSE) %>%
-    pull(accuracy)
-
-}
-
-input <- commandArgs(trailingOnly = TRUE)
-dataset <- input[1]
-
-output_file <- glue("data/{dataset}/cluster_analysis.tsv")
-
-corrections <- c("raw", "relabund", "rare",
-                 "deseq2", "metagenomeseq", "srs",
-                 "nclr", "oclr", "rclr", "zclr")
-
-distances <- c("bray",
-               "euclidean",
-               "jaccard")
-
-model <- c("r", "e")
-
-design <- c("r", "s", "e")
-
-seeds <- 1:100
-
-expand_grid(model, design, corrections, distances, seeds) %>% 
-  filter((model == "r" & design != "e") | (model == "e" & design == "e")) %>%
-  mutate(
-      dataset = dataset
-  ) %>%
-  rowwise() %>%
-  mutate(
-    accuracies = get_accuracy(seeds, model, design, dataset, corrections, distances)
-  ) %>%
-  write_tsv(output_file)
-

### 2023-11-20 submission/response_to_reviewers.Rmd
Implement less extreme confounding analysis

@@
 
-> XXXUse overlapping IQR instead of perfect confounding.
-> Currently have code/get_intraquantile_design.R set at 10%/90%; needs to be run
-> on great lakes
+> I appreciated this suggestion and repeated the analysis with a less severe level of confounding.
+In addition to the perfect confounding of treatment groups by sequencing effort, I created an additional simulation where the smallest 5% of samples were assigned to one treatment group and the largest were assigned to another; the remaining samples were randomly assigned between the groups.
+The new results are in Figure 3 and the previous simulation are now in Figure S4.
+Although the results are less severe in the new simulation, they still affirm the effects of sampling effort on the propensity to get false positive results.
 

### 2023-11-20 submission/manuscript.Rmd
Implement less extreme confounding analysis

@@
 Indeed, for each dataset and alpha and beta diversity metric and strategy for accounting for uneven sequencing, approximately 5% of the tests yielded a significant result (Figure 2).
-Second, I employed a biased null treatment model where the treatment group was completely confounded with the number of sequences in each sample.
-Under this model, only the data normalized with rarefaction consistently resulted in a 5% false positive rate for alpha and beta diversity metrics (Figure 2).
+Second, I employed a biased null treatment model where the treatment group was completely confounded with the number of sequences in each sample (Figure S4).
+Under this model, only the data normalized with rarefaction consistently resulted in a 5% false positive rate for alpha and beta diversity metrics.
+I then used a less severe bias where the smallest 5% of samples were assigned to one treatment group and the largest 5% were assigned to another; the other samples were randomly assigned to both treatment groups.
+Although the sensitivity of the false positive rate to the level of confounding varied across methods of quantifying alpha and beta diversity, rarefaction was the only method to consistently control the false positive rate (Figure 3).
 These results aligned with the observed sensitivity of alpha and beta diversity metrics to sequencing effort and underscore the value of rarefaction.
@@
 In the unbiased model, samples were randomly assigned to one of two treatment groups.
-In the biased treatment model, samples that had more than the median number of sequences for a dataset were assigned to one treatment group and the rest were assigned to a second treatment group.
+I created two biased treatment models.
+In the first, samples that had more than the median number of sequences for a dataset were assigned to one treatment group and the rest were assigned to a second treatment group (see Figure S4).
+In the second, samples whose sequencing depth where in the 5th percentile or lower were assigned to one treatment group while those in the 95% percentile or higher were assigned to another; the remaining samples were randomly assigned to either treatment group (see Figure 3).
 Comparison of any diversity metric across the two treatment groups should have only yielded a significant result in 5% of the simulations when testing under a Type I error (i.e., $\alpha$) of 0.05.
@@
 **Figure 3.The risk of falsely detecting a difference between treatment groups drawn from a null model does not meaningfully vary from 5% when data are normalized by rarefaction when sequencing depth is confounded with treatment group**.
-Samples were assigned to different treatment groups based on whether they were above the median number of sequences for each dataset.
+Samples were assigned to different treatment groups so that the smallest 5% of samples were assigned to one treatment group and the largest 5% to another; the other 90% of samples were randomly assigned to both treatment groups.
 To calculate the false detection rate, datasets were regenerated 100 times and differences in alpha diversity were tested using a Wilcoxon test (A) and differences in beta diversity were tested using PERMANOVA (B) at a 5% threshold.
@@
 The Spearman correlation coefficient ($\rho$) and test of whether the coefficient was significantly different from zero are indicated for each panel.
+
+\newpage
+
+\includegraphics[height=17cm]{figure_s4.png}
+
+**Figure S4.The risk of falsely detecting a difference between treatment groups drawn from a null model does not meaningfully vary from 5% when data are normalized by rarefaction when sequencing depth is confounded with treatment group**.
+Samples were assigned to different treatment groups based on whether they were above the median number of sequences for each dataset.
+To calculate the false detection rate, datasets were regenerated 100 times and differences in alpha diversity were tested using a Wilcoxon test (A) and differences in beta diversity were tested using PERMANOVA (B) at a 5% threshold.
+The false positive rate was the number of times a dataset yielded a significant result.

### 2023-11-20 submission/manuscript.tex
Implement less extreme confounding analysis

@@
 biased null treatment model where the treatment group was completely
-confounded with the number of sequences in each sample. Under this
-model, only the data normalized with rarefaction consistently resulted
-in a 5\% false positive rate for alpha and beta diversity metrics
-(Figure 2). These results aligned with the observed sensitivity of alpha
-and beta diversity metrics to sequencing effort and underscore the value
-of rarefaction.
+confounded with the number of sequences in each sample (Figure S4).
+Under this model, only the data normalized with rarefaction consistently
+resulted in a 5\% false positive rate for alpha and beta diversity
+metrics. I then used a less severe bias where the smallest 5\% of
+samples were assigned to one treatment group and the largest 5\% were
+assigned to another; the other samples were randomly assigned to both
+treatment groups. Although the sensitivity of the false positive rate to
+the level of confounding varied across methods of quantifying alpha and
+beta diversity, rarefaction was the only method to consistently control
+the false positive rate (Figure 3). These results aligned with the
+observed sensitivity of alpha and beta diversity metrics to sequencing
+effort and underscore the value of rarefaction.
 
@@
 coefficient of variation between Bray-Curtis distances tended to
-increase. Regardless, the coefficients of variation were relatively
-small. This analysis indicates that there are benefits to increased
-sequencing depths.
+increase slightly for samples where the standard deviation did not
+decrease as quickly as the mean distance between samples. Regardless,
+the coefficients of variation were relatively small. This analysis
+indicates that there are benefits to increased sequencing depths.
 
@@
 strategies are needed to reduce the variation in the number of sequences
-generated for each sample.
+generated for each sample. Furthermore, it is important to keep in mind
+that because of residual sequencing errors, PCR artifacts, and
+incomplete sequencing, it is not possible to obtain absolute
+quantification of alpha or beta diversity values. Instead, comparisons
+of these values need to be made on a relative basis.
 
@@
 treatment model. In the unbiased model, samples were randomly assigned
-to one of two treatment groups. In the biased treatment model, samples
-that had more than the median number of sequences for a dataset were
-assigned to one treatment group and the rest were assigned to a second
-treatment group. Comparison of any diversity metric across the two
-treatment groups should have only yielded a significant result in 5\% of
-the simulations when testing under a Type I error (i.e., \(\alpha\)) of
-0.05.
+to one of two treatment groups. I created two biased treatment models.
+In the first, samples that had more than the median number of sequences
+for a dataset were assigned to one treatment group and the rest were
+assigned to a second treatment group (see Figure S4). In the second,
+samples whose sequencing depth where in the 5th percentile or lower were
+assigned to one treatment group while those in the 95\% percentile or
+higher were assigned to another; the remaining samples were randomly
+assigned to either treatment group (see Figure 3). Comparison of any
+diversity metric across the two treatment groups should have only
+yielded a significant result in 5\% of the simulations when testing
+under a Type I error (i.e., \(\alpha\)) of 0.05.
 
@@
 confounded with treatment group}. Samples were assigned to different
-treatment groups based on whether they were above the median number of
-sequences for each dataset. To calculate the false detection rate,
-datasets were regenerated 100 times and differences in alpha diversity
-were tested using a Wilcoxon test (A) and differences in beta diversity
-were tested using PERMANOVA (B) at a 5\% threshold. The false positive
-rate was the number of times a dataset yielded a significant result.
+treatment groups so that the smallest 5\% of samples were assigned to
+one treatment group and the largest 5\% to another; the other 90\% of
+samples were randomly assigned to both treatment groups. To calculate
+the false detection rate, datasets were regenerated 100 times and
+differences in alpha diversity were tested using a Wilcoxon test (A) and
+differences in beta diversity were tested using PERMANOVA (B) at a 5\%
+threshold. The false positive rate was the number of times a dataset
+yielded a significant result.
 
@@
 
+\newpage
+
+\includegraphics[height=17cm]{figure_s4.png}
+
+\textbf{Figure S4.The risk of falsely detecting a difference between
+treatment groups drawn from a null model does not meaningfully vary from
+5\% when data are normalized by rarefaction when sequencing depth is
+confounded with treatment group}. Samples were assigned to different
+treatment groups based on whether they were above the median number of
+sequences for each dataset. To calculate the false detection rate,
+datasets were regenerated 100 times and differences in alpha diversity
+were tested using a Wilcoxon test (A) and differences in beta diversity
+were tested using PERMANOVA (B) at a 5\% threshold. The false positive
+rate was the number of times a dataset yielded a significant result.
+
 \end{document}

### 2023-11-20 code/get_intraquantile_design.R

@@ -1,0 +1,49 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(glue)
+
+input <- commandArgs(trailingOnly = TRUE)
+
+group_count_file <- input[1] # e.g. group_count_file <- "data/human/data.group_count"
+remove_accnos_file <- input[2] #e.g. remove_accnos_file <- "data/human/data.remove_accnos"
+seed <- as.numeric(input[3]) # e.g. seed <- 87
+
+set.seed(seed)
+
+samplesize_design_file <- str_replace(group_count_file,
+                             "group_count",
+                             glue("{seed}.idesign"))
+
+remove_accnos <- scan(remove_accnos_file, sep = "\n",
+                      what = character(), quiet = TRUE)
+
+group_count <- read_tsv(group_count_file,
+         col_names = c("Group", "n_seqs"),
+         col_types = cols(Group = col_character(), n_seqs = col_double())
+         ) %>%
+  filter(! Group %in% remove_accnos)
+
+n_samples <- nrow(group_count)
+
+l_quantile <- quantile(group_count$n_seqs, prob = 0.10)
+u_quantile <- quantile(group_count$n_seqs, prob = 0.90)
+
+group_count %>%
+  mutate(treatment = case_when(n_seqs < l_quantile ~ "A",
+                              n_seqs > u_quantile ~ "B")) %>%
+  nest(data = -treatment) %>%
+  mutate(rand_treat = map2(data, treatment,
+                          function(.x, .y) {
+
+                            if (is.na(.y)) {
+                              return(sample(rep(c("A", "B"),
+                                                length.out = nrow(.x))))
+                            }
+
+                            return(.y)
+
+                          })) %>%
+  unnest(c(data, rand_treat)) %>%
+  select(Group, treatment = rand_treat) %>%
+  write_tsv(., samplesize_design_file)

### 2023-11-20 code/plot_false_positive_null_model_size.R

@@ -4,6 +4,9 @@ library(tidyverse)
 library(glue)
 library(patchwork)
 
+args <- commandArgs(trailingOnly = TRUE)
+design <- args[1]
+
 capitalize <- function(string) {
   substr(string, 1, 1) <- toupper(substr(string, 1, 1))
   string
@@ -104,7 +107,7 @@ pretty_alpha_classes <- c(
   inv_simpson = "Inverse\nSimpson"
 )
 
-alpha_summary_files <- glue("data/{datasets}/data.r_salpha_kw")
+alpha_summary_files <- glue("data/{datasets}/data.r_{design}alpha_kw")
 names(alpha_summary_files) <- datasets
 
 alpha_composite <- map_dfr(alpha_summary_files, read_tsv, .id = "dataset") %>%
@@ -206,7 +209,7 @@ pretty_beta_calcs <- c(
   euclidean = "Euclidean"
 )
 
-beta_summary_files <- glue("data/{datasets}/data.r_samova")
+beta_summary_files <- glue("data/{datasets}/data.r_{design}amova")
 names(beta_summary_files) <- datasets
 
 beta_composite <- map_dfr(beta_summary_files, read_tsv, .id = "dataset") %>%
@@ -259,7 +262,7 @@ combo <- alpha + beta +
                             )
     )
 
-ggsave("figures/false_positive_null_model_size.tiff",
+ggsave(glue("figures/false_positive_null_model_{design}size.tiff"),
        combo,
        width = 6, height = 7,
        compression = "lzw+p")

### 2023-12-14 code/get_rclr_beta.R
Use 3 components in RPCA rather than auto

@@ -17,7 +17,7 @@ method <- str_replace(dist_file, ".*_(\\w*).dist", "\\1")
 dataset <- str_replace(dist_file, "data/(.*)/data.*", "\\1")
 
 # couldn't get stream, rice, seagrass to reliably get through run_gemelli even
-# with 750 GB of RAM. Skipping...
+# with 100 GB of RAM
 
 if (method == "euclidean" && !(dataset %in% c("stream", "rice", "seagrass"))) {
 

### 2023-12-14 code/run_gemelli.py

@@ -4,9 +4,8 @@ import sys
 import numpy as np
 from biom import Table
 import pandas as pd
-from gemelli import __version__ as gversion
 from gemelli.preprocessing import rclr_transformation
-from gemelli.rpca import auto_rpca
+from gemelli.rpca import rpca
 
 np.seterr(divide = 'ignore') # for rclr
 
@@ -25,6 +24,6 @@ shared_transposed_biom = Table(shared_transposed_df.to_numpy(),
 rclr_gemelli = rclr_transformation(shared_transposed_biom) # compute RCLR
 
 # Run RPCA
-ordination, distance_matrix = auto_rpca(shared_transposed_biom) 
+ordination, distance_matrix = rpca(shared_transposed_biom, n_components = 3) 
 
 distance_matrix.to_data_frame().to_csv(dist_file, sep='\t') # save just the distances since that is what is output from the R version

### 2023-12-15 submission/response_to_reviewers.Rmd
Editing / submitted

@@
 \raggedright
+
+I thank the editor and reviewers for their helpful comments on my manuscript.
+I have thoroughly addressed their comments.
+In addition, the revised manuscript has included alpha diversity estimates derived using the iNEXT R package, which combines non-parametric estimation and rarefaction to control for uneven effect sizes.
+I have also replaced the robust centered log ratio method, with the full Robust PCA method, which combines the robus CLR transformation with matrix completion.
+Regardless, the results are the same as robust CLR alone.
+When I mention line numbers in my responses to the reviewers' comments below, the line numbers indicated refer to the marked up version of the manuscript.
 
@@
 On the whole, the COV values are fairly small.
-I have added text to this effect at LXXX.
+I have added text to this effect at L221-L224.
 
@@
 Although the results are less severe in the new simulation, they still affirm the effects of sampling effort on the propensity to get false positive results.
+Additional text describing the new analysis has been inserted in the Results section at L163-L171 and the Materials & Methods section at L375-L382.
 
@@
 
-> XXXDo this.
+> I appreciate the reviewer's suggestion to look closer at the Martino et al study and their RPCA method.
+In the revised manuscript I used RPCA rather than the simple RCLR transformation.
+In the process of my work, I found a number of problems with the design and analysis of the benchmarking experiments performed in their study.
+While I mention the differences between my analysis and the Martino et al.
+and McMurdie & Holmes studies in the revised manuscript (L345-L355), a more comprehensive analysis is underway and is beyond the scope of the current study.
 
@@
 
-> XXXResponse to previous point will help to address this.
-But the data processing pipeline won't really factor in since all methods used the same data for a dataset.
+> My responses to the previous points partially address this question.
+The data processing pipeline do not factor in for interpreting the results of this analysis since all of my benchmarking analyses used the same data for a dataset and each dataset was processed with the same pipeline.
 The methods don't know what type of processing was done upstream.
-Could have used purely synthetic distributions.
+Part of the value of using 12 different datasets was to provide diversity in community shapes.
+I could have used purely synthetic distributions as others have done (e.g.
+Martino et al., McMurdie & Holmes) and obtained similar results.
 
 **5.
-One line 164 the report indicates finding differences in richness between samples that should have the same OTU counts given the way the enrichment sampling was done (increasing sequencing counts not raw number of OTUS in the sample).
+On line 164 the report indicates finding differences in richness between samples that should have the same OTU counts given the way the enrichment sampling was done (increasing sequencing counts not raw number of OTUS in the sample).
 It would make sense that metrics that consider abundances to change but not richness.
@@
 
-> XXXRevise text
+> The text in this paragraph has been edited to make it more clear what simulation is being used for each analysis.
+There is also text in the Discussion highlighting the value of using the Shannon or inverse Simpson's diversity indices over richness because they depend less on rare taxa (see paragraph starting at L255).
 
@@
 
-> Look at suggested papers; cite Gloor
+> I have added the citations the reviewer suggested to the text at L355-L340 and fleshed out the discussion in this paragraph.
 
@@
 > The caption for Figure 1 indicates that the size of the confidence interval was smaller than the plotting symbol.
-As stated in the captions for Figures 2 and 3, "The false positive rate was the number of times a dataset yielded a significant result."
+As stated in the captions for Figures 2 and 3, "The false positive rate was the number of times a dataset yielded a significant result." Therefore, it does not have a confidence interval.
 
@@
 
-> I'm reluctant to add more than what I already have at L281-289. In that text, I mention using Good's coverage values as a check to make sure that the level of coverage is acceptable.
+> I'm reluctant to add more than what I already have at L308-L315.
+In that text, I mention using Good's coverage values as a check to make sure that the level of coverage is acceptable.
+In the revised manuscript I did add alpha diversity metrics that incorporate sampling depth and coverage to extrapolate (iNEXT) (L316-L327).
+Although these methods perform better than their "raw" counterparts, there are still effects of sampling effort.
 
@@
 
-> I have added a sentence to this effect in the Discussion at **LXXX**
+> I have added a sentence to this effect in the Discussion at L274-L277.
 
@@
 
+> My pleasure :).
+I have added text to the closing paragraph of the Discussion (L345-L355) to elaborate on why the results of my benchmarking studies differ so strikingly with those that have preceded mine.
+
 **Minor comments**
+
 **Line 69 - This implies that McMurdle/Holmes used an incorrect interpretation of differences between rarefying and rarefaction.
@@
 
+> I have clarified this paragraph of the Introduction (L64-L78).
+My paper describing the re-analysis of WNWN has recently been accepted for publication at *mSphere* and goes into far greater details regarding the problems with the original study.
+
 **Line 72 - Change "method" to "methods"**
+
+> This has been corrected
 
@@
 
+> No snark was intended. "Toy" is jargon in the field for a simplistic example desined to make a point. To avoid confusion I have used "overly simplistic" instead.
+
 **Line 94 - Delete one of the "from"s**
+
+> This has been edited
 
@@
 
+> This text has been clarified
+
 **Line 167 - Were the 3% of OTUs selected at random, or did you remove the rarest 3%?**
+
+> This text has been clarified
 
@@
 
+> This text has been corrected
+
 **Line 210 - Drop your excessive sig figs .... Use "101" fold**
 
+> The number of significant figures has been corrected
+
 **Line 281 - I think this should be "my personal process for selecting the rarefaction threshold"?**
+
+> This text has been corrected

### 2023-12-15 submission/manuscript.Rmd
Editing / submitted

@@
     keep_tex: true
+  word_document: default
 csl: asm.csl
@@
 Unfortunately, it is common to observe variation in the number of sequence reads from each sample vary by as much as 100-fold (e.g, see Figure S1).
-This occurs because pooling of DNA from multiple PCRs is fraught with numerous opportunities for technical errors to compound leading to a skewed distribution.
+This occurs because pooling of DNA from multiple PCRs is fraught with numerous opportunities for random sampling errors to compound leading to a skewed distribution.
 Aside from developing better methods of pooling DNA, the question of how to control for uneven sequencing effort in microbial ecology studies has become controversial.
@@
 In their simulations, they observed that rarefying reduced the statistical power to correctly cluster samples into the same treatment groups based on beta diversity metrics.
-Although the detail was lost on many subsequent researchers, they did not describe rarefaction, but rarefying [@Schloss2023wnwn].
-According to McMurdie and Holmes, rarefying is a single subsampling of an OTU abundance table whereas rarefaction repeats the subsampling step many times.
-Furthermore, their simulations penalized the rarefied data by removing 15% of the samples and evaluated clustering quality using a clustering algorithm that performed worse for rarefied data.
-According to my reanalysis of rarefaction using the original simulation framework, rarefaction outperformed other normalization method being used in 2014 for both alpha and beta diversity metrics [@Schloss2023wnwn].
+I have recently shown that their analysis had a number of important problems [@Schloss2023wnwn].
+First, although the detail was lost on many subsequent researchers, they did not describe rarefaction.
+Instead they described rarefying as a single subsampling of an OTU abundance table whereas rarefaction repeats the subsampling step many times.
+Furthermore, their simulations penalized the rarefied data by removing 15% of the samples but did not remove those samples when considernig their other methods.
+They also evaluated the accuracy of clustering samples using a clustering algorithm that performed worse for rarefied data.
+According to my reanalysis of rarefaction using their original simulation framework, but using all of the samples, rarefaction outperformed the other normalization methods they described for both alpha and beta diversity metrics [@Schloss2023wnwn].
 Others have also critiqued the original work [@Weiss2017].
-Regardless, there is a general sense in the community that rarefying and rarefaction should not be used with microbiome data.
 
 Since McMurdie and Holmes published their simulations, others have developed alternative approaches to control for uneven sequencing effort in amplicon sequencing studies.
-For alpha diversity metrics, Willis used toy datasets to demonstrate that one could estimate the richness for each sample in a dataset and use those values for statistical comparisons [@Willis2019].
-Non-parametric estimators of richness and diversity [@Chao2003; @Chao2016] and parametric estimators of richness [@Willis2015] have been used in microbial ecology studies.
-For beta diversity metrics at least four approaches have been pursued.
+For alpha diversity metrics, Willis used overly simplistic datasets to demonstrate that one could estimate the richness for each sample in a dataset and use those values for statistical comparisons [@Willis2019].
+Non-parametric estimators of the minimum richness in a community and diversity [@Chao2003; @Chao2016; @Chao2014] and parametric estimators of richness [@Willis2015] have been used in microbial ecology studies.
+One non-parametric approach, iNEXT, combines rarefaction on samples more than a threshold number of sequences and extrapoliation on samples below the threshold [@Chao2014].
+This method has become popular in traditional ecology studies, but has not been widely used amongst microbial ecologists.
+For beta diversity metrics at least four approaches have been pursued as alternatives to rarefaction.
 First, one could use relative abundance values where the observed number of sequences in an OTU is divided by the total number of sequences in the sample [@Lin2020].
@@
 This strategy is purported to control for uneven sequencing effort [@Martino2019; @Quinn2018]; however, some have noticed that this feature breaks down under certain conditions [@teBeest2021].
-Finally, as mentioned above, variance stabilization transformations have been recommended to generate values that can be used to calculate Euclidean distances [@McMurdie2014].
+Finally variance stabilization transformations have been recommended to generate values that can be used to calculate Euclidean distances [@McMurdie2014].
 
-The ongoing controversy over the use of rarefaction and the recent development of alternative strategies to control for uneven sequencing effort caused me to question how these methods compared to each other using a simulation framework that overcame the issues with the McMurdie and Holmes study.
-My analysis included 16S rRNA gene sequence data from from 12 studies that characterized the variation in bacterial communities from diverse environments (Table 1 and Figure S1).
+The ongoing controversy over the use of rarefaction and the recent development of alternative strategies to control for uneven sequencing effort caused me to question how these methods compared to each other using a simulation framework that overcame the issues with the McMurdie and Holmes study [@McMurdie2014].
+My analysis included 16S rRNA gene sequence data from 12 diverse environments (Table 1 and Figure S1).
 The sequences were assigned to OTUs using a standard pipeline and their frequencies and the number of sequences found in each sample were used to generate simulated communities and treatment effects.
@@
 Figure S2).
-For each dataset, all of the approaches, except for rarefaction, showed a strong correlation between richness and the number of sequences in the sample (Figure 1A).
-Next, I assessed diversity using the Shannon diversity index and the inverse Simpson diversity index without any correction, using normalized OTU counts, and rarefaction; I also used a non-parametric estimator of Shannon diversity.
-The correlation between sequencing depth and the diversity metric was not as strong as it was for richness and the inverse Simpson diversity values were less sensitive than the Shannon diversity values; however, the correlation to the diversity metrics calculated with rarefaction were the lowest for all of the metrics and studies (Figure 1A).
+For each dataset, all of the approaches, except for rarefaction and possibly the size-based iNEXT estimates, showed a strong correlation between richness and the number of sequences in the sample (Figure 1A).
+The size-based iNEXT estimates were correlated with sampling effort for the Marine and Soil datasts, which had fewer samples and more uniform sequencing depth distributions than the other datasets.
+Next, I assessed diversity using the Shannon diversity index and the inverse Simpson diversity index without any correction, using normalized OTU counts, non-parametric estimation, and rarefaction.
+The correlation between sequencing depth and the Shannon diversity metric was not as strong as it was for richness.
+The inverse Simpson diversity values was not sensitive to sampling effort (Figure 1A).
 The alpha-diversity metrics calculated with rarefaction consistently demonstrated a lack of sensitivity to sequencing depth.
@@
 Similar to what I observed with the Jaccard coefficient, only the data normalized with rarefaction showed a lack of sensitivity to sequencing effort (Figure 1B).
-Third, I calculated the Euclidean distance on raw OTU counts where the central log-ratio (CLR) was calculated (i.e., Aitchison distances) using matrix completion to replace zero counts (Robust CLR), adding a pseudo-count of one to all OTU counts prior to calculating the CLR (One CLR), adding a pseudo-count of one divided by the total number of sequences obtained for the community (Nudge CLR), and imputing the value of zero counts (Zero CLR).
+Third, I calculated the Euclidean distance on raw OTU counts where the central log-ratio (CLR) was calculated (i.e., Aitchison distances) by adding a pseudo-count of one to all OTU counts prior to calculating the CLR (One CLR), adding a pseudo-count of one divided by the total number of sequences obtained for the community (Nudge CLR), imputing the value of zero counts (Zero CLR), and using CLR on non-zero counts followed by matrix completion on zero counts (Robust Principle Components Analysis [PCA]).
 The Aitchison distances were all strongly sensitive to sequencing effort (Figure 1B).
@@
 Again, there was a strong sensitivity to sequencing effort (Figure 1B).
-Although Euclidean distances are not typically used on raw or rarefaction-normalized count data in ecology [@Legendre2012], Euclidean distances calculated with rarefaction were not sensitive to sequencing effort.
+Although use of Euclidean distances is not advised with raw or rarefaction-normalized count data in ecology [@Legendre2012], Euclidean distances calculated with rarefaction were not sensitive to sequencing effort.
 Across each of the beta diversity metrics and approaches used to account for uneven sequencing effort and sparsity, rarefaction was the least sensitive approach to differences in sequencing effort.
@@
 Indeed, for each dataset and alpha and beta diversity metric and strategy for accounting for uneven sequencing, approximately 5% of the tests yielded a significant result (Figure 2).
-Second, I employed a biased null treatment model where the treatment group was completely confounded with the number of sequences in each sample (Figure S4).
+Second, I employed a biased null treatment model where the treatment group was confounded with the number of sequences in each sample such that the 50% of samples with the least sequencing effort were in one treatment group and the others were in another treatment group (Figure S4).
 Under this model, only the data normalized with rarefaction consistently resulted in a 5% false positive rate for alpha and beta diversity metrics.
 I then used a less severe bias where the smallest 5% of samples were assigned to one treatment group and the largest 5% were assigned to another; the other samples were randomly assigned to both treatment groups.
-Although the sensitivity of the false positive rate to the level of confounding varied across methods of quantifying alpha and beta diversity, rarefaction was the only method to consistently control the false positive rate (Figure 3).
+Although the sensitivity of the false positive rate to the level of confounding varied across methods of quantifying alpha and beta diversity, rarefaction and the iNEXT size-based estimates for alpha diveristy were the only methods to consistently control the false positive rate (Figure 3).
 These results aligned with the observed sensitivity of alpha and beta diversity metrics to sequencing effort and underscore the value of rarefaction.
@@
 The power to detect differences in richness between the two simulated treatment groups by all approaches was low (Figure 4A).
-This was likely because the approach for generating the perturbed community did not necessarily change the number of OTUs in each treatment group.
+This was likely because the approach for generating the perturbed community did not necessarily change the number of OTUs in each treatment group; the differences primarily affected the distribution of the sequences across OTUs.
 Regardless, the simulations testing differences in richness using the Rice and Stream datasets had the greatest power when the richness data were calculated with rarefaction.
-To explore this further, a richness-adjusted community model was created by removing 3% of the OTUs from a null model model.
-As suggested by the first simulation, the richness data  calculated with rarefaction had a higher statistical power than the other approaches when measuring richness (Figure 5).
-The simulations testing the power to detect differences in Shannon diversity also showed that rarefaction performed better than the other methods (Figure 4A).
-When testing for differences in the Inverse Simpson diversity index the the difference between rarefaction and the other methods was negligible (Figure 4A).
-For tests of beta diversity I found that rarefaction was the most reliable approach to maintain statistical power to detect differences between two communities.
+To explore this further, a richness-adjusted community model was created by randomly selecting and removing 3% of the OTUs from a null model.
+As suggested by the first simulation, the richness data calculated with rarefaction consistently had the highest statistical power among the approaches used to measure richness followed by the iNEXT size-based estimates (Figure 5).
+Returning to simulations where treatment groups varied in their structure, the power to detect differences in Shannon diversity was the lowest when using raw relative abundances than when using the other methods (Figures 4A and 5).
+When testing for differences in the Inverse Simpson diversity index the the difference between rarefaction and the other methods was negligible (Figures 4A and 5).
+For tests of beta diversity, I found that rarefaction was the most reliable approach to maintain statistical power to detect differences between two communities (Figure 4B).
 Among the tests using the Jaccard and Bray-Curtis metrics, raw count data and CSS normalized data had little power relative to using rarefaction, relative abundance, and SRS to normalize the uneven sequencing depths.
@@
          ninetynine = ninetynine / ninety - 1) %>%
-  summarize(m = round(mean(ninetyfive), 1),
-            n = round(mean(ninetynine), 1)
+  summarize(m = round(mean(ninetyfive), 0),
+            n = round(mean(ninetynine), 0)
             )
@@
 
-***Increased rarefaction depth reduces intra-sample variation in alpha and beta diversity.*** Once concern with using rarefaction is the perceived loss of sequencing information when a large fraction of data appears to be removed when the community with the greatest sequencing depth is sampled to the size of the community with the least (e.g., the smallest sample in the `r loss$nice_name` dataset had `r loss$min_max`% of sequences that were in the largest sample).
+***Increased rarefaction depth reduces intra-sample variation in alpha and beta diversity.*** One concern with using rarefaction is the perceived loss of sequencing information when a large fraction of data appears to be removed when the community with the greatest sequencing depth is sampled to the size of the community with the least (e.g., the smallest sample in the `r loss$nice_name` dataset had `r loss$min_max`% of sequences that were in the largest sample).
 To assess the sensitivity of alpha and beta diversity metrics to rarefaction depth, I again used the dataset generated using the null models, but used rarefaction with each community to varying sequencing depths (Figure 6).
@@
 Next, I measured the coefficient of variation (i.e., the mean divided by the standard deviation) between samples for richness, Shannon diversity, and Bray-Curtis distances.
-Although the mean richness appeared to increased unbounded with sequencing effort, the coefficient of variation for each dataset was relatively stable.
+Although the mean richness appeared to increase unbounded with sequencing effort, the coefficient of variation for each dataset was relatively stable.
 In general, the coefficient of variation increased slightly with sequencing depth only to decline once smaller samples were removed from the analysis at higher sequencing depths.
@@
 The median coverage for each dataset ranged between `r oxford_comma(obs_median$med)`% for the `r oxford_comma(capitalize(obs_median$dataset))` datasets, respectively (Figure 7).
-When I used a rarefaction threshold with each dataset at the size of the smallest community in the dataset, with the exception of the `r oxford_comma(capitalize(rare_median$dataset))` datasets, the median coverage with rarefaction was still greater than 90%.
+When I used a rarefaction threshold with each dataset at the size of the smallest community in the dataset, with the exception of the `r oxford_comma(capitalize(rare_median$dataset))` datasets, the median coverage with rarefaction was still greater than 90% (see the box plots in Figure 7).
 These results suggest that most studies had a level of sequencing coverage that aligned with the diversity of the communities.
 Next, I used the null model for each dataset to ask how much sequencing effort was required to obtain higher levels of coverage.
-To obtain 95 and 99% coverage, an average of `r extrapolate$m` and `r extrapolate$n`-fold more sequence data was estimated to be required than was required to obtain 90% coverage, respectively (Figure 7).
+To obtain 95 and 99% coverage, an average of `r inline_hook(extrapolate$m, 0)` and `r inline_hook(extrapolate$n, 0)`-fold more sequence data was estimated to be required than was required to obtain 90% coverage, respectively (Figure 7).
 As suggested by the simulated coverages curve in Figure 7, these estimates are conservative.
@@
 The analyses I presented here strongly indicate that rarefaction is necessary to control for uneven sequencing effort when comparing communities based on alpha and beta diversity indices.
-Compared to all other methods, rarefaction removed the correlation between sequencing depth and alpha or beta diversity metrics when the sequencing depth varied by as much as 97-fold across samples.
+Compared to all other methods, rarefaction was the most consistent method for removing the correlation between sequencing depth and alpha or beta diversity metrics even when the sequencing depth varied by as much as 97-fold across samples.
 I showed that this correlation could lead falsely detecting differences between treatment groups if sequencing depth and sequencing effort are confounded.
@@
 
-The primary alternative to rarefaction for measuring alpha diversity is to estimate the metric using non-parametric or parametric models with raw counts and to then compare the estimates [@Hughes2001; @Willis2019].
+Historically, the primary alternative to rarefaction for measuring alpha diversity is to estimate the metric using non-parametric or parametric models with raw counts and to then compare the estimates [@Hughes2001; @Willis2019].
 My results demonstrate that such approaches are limited for several reasons.
 First, non-parametric richness estimates such as ACE and Chao1 are sensitive to sequencing effort.
-Therefore, these strategies do not, in practice, remove the effects of sequencing effort.
+Therefore, these estimators do not, in practice, remove the effects of sequencing effort.
 Second, parametric approaches, such as those implemented in the `breakaway` R package, generate confidence intervals that are likely to include the true richness and theoretically shrink with increased sequencing effort.
@@
 Although normalizing communities to a common number of sequences is also suggested (e.g., SRS normalization) to control for uneven sequencing effort, the current analysis indicates that its performance does not meet that of rarefaction.
-For analysis of alpha diversity metrics, rarefaction is the most effective and consistent approach to control for uneven sequencing effort.
+Because alpha diversity metrics are so sensitive to sampling effort and residual PCR and sequencing artifacts, it is important to treat their values on a relative (e.g., 5% difference) rather than absolute basis (e.g., difference of 48 OTUs).
+Rarefaction is an effective and consistent approach to control for uneven sequencing effort.
 
 Use of relative abundances, normalized counts, variance stabilizing transformations, and centered log ratios have each been recommended as more robust alternatives to rarefaction.
-Again, the only approach that consistently removed the effects of uneven sequencing effort was rarefaction.
-Most of the recommendations borrow techniques from methods for identifying differential gene expression.
+Again, the only approach that consistently removed the effects of uneven sequencing effort across alpha and beta diversity analyses was rarefaction.
+Many of the alternative methods have been borrowed from differential gene expression analysis.
 Unfortunately, there is an important but under appreciated difference between gene expression and community data.
@@
 When that subsampling is repeated a large number of times (e.g., 100 or 1000) the risks of ignoring or oversampling rare taxa are mitigated.
-Furthermore, it is curious that the study making the original critique removed rare sequences.
-A parallel analysis to this study has demonstrated that many of these sequences are likely valid and that removal of rare sequences can bias alpha and beta diversity metrics and reduce statistical power [@Schloss2023sing].
 As for the second criticism, I would resist the claim that the selection of the rarefaction threshold is arbitrary.
@@
 Greater breadth will increase the statistical power to compare treatment groups and greater depth will increase the resolution to describe the communities.
-My personal process for rarefaction threshold involves looking for a natural break in the distribution of the number of sequences.
+My personal process for selecting a rarefaction threshold involves looking for a natural break in the distribution of the number of sequences.
 For example, the Lake dataset used in this study had a clear break around 10,000 sequences (Figure S1).
-I would also consider what samples are below any break that I select.
-If there were critical samples below the break I would either reduce the threshold or obtain more sequences from those samples.
+I would also consider what samples are below any threshold that I select.
+If there were critical samples below the threshold I would either reduce the threshold or obtain more sequences from those samples.
 As shown in my analysis of Good's coverage values, most studies obtain an ample level of coverage and would need to increase their sequencing depth by 10-fold to increase the coverage by several percentage points.
-In past work, I have favored distributing the total number of sequences to increase sampling breadth rather than sequencing depth.
+In past work, I have favored increasing sampling breadth over sequencing depth.
+
+An additional critique of rarefaction is that the variation between alpha diversity metrics shrinks as the rarefaction threshold decreases, which may compromise the power to detect differences between treatment groups.
+The iNEXT methods attempt to minimize this effect by estimating alpha diversity metrics for samples whose sequencing depth is below a threshold and using rarefaction on samples with a sequencing depth above that threshold [@Chao2014].
+This threshold is set at either twice the size of the smallest sample (i.e., size-based) or by controlling for the coverage when the number of sequences in each sample is doubled.
+An example of this can be seen in Figure S2 where richness values obtained by estimation are shown in gray and those by rarefaction are in black.
+While my data suggest that the size-based approach has promise for estimating alpha diversity, I found that it was not as consistent as rarefaction for smaller datasets where there was a narrow distribution in sequencing depths.
+This is likely because in these situatons the algorithm relies more on estimation than rarefaction.
 
@@
 Researchers perform separate PCRs for each sample with unique index (aka, barcode) sequences that allow them to later assign sequences back to the samples that they came from.
-When the PCR products are pooled, efforts are taken to pool the fragments in equimolar ratios.
-Researchers use one of two approaches.
+When the PCR products are pooled, efforts are often taken to pool the fragments in equimolar ratios using one of two approaches.
 First, they often will quantify the concentration of DNA from each PCR and then pool DNA in the desired amounts.
@@
 Clearly, both approaches have limitations that reduce the ability to truly achieve equimolar mixture.
-In addition, for some samples it is common to co-amplify non-specific DNAs which may add to the challenges of achieving equimolar mixtures of the desired amplicons since those sequences will be removed [@Morris2013].
-Regardless, it is clear that better strategies are needed to reduce the variation in the number of sequences generated for each sample.
+Although normalization procedures are expected to remove the relationship between sequencing depth and sample biomass [@Gloor2017], several recent studies have found a positive correlation [@Minich2022; @Cruz2021].
+Even in cases where the is no correlation, for some samples it is common to co-amplify non-specific DNAs which are later removed during sequence processing [@Morris2013].
+It is clear that better strategies are needed to reduce the variation in the number of sequences generated for each sample.
 Furthermore, it is important to keep in mind that because of residual sequencing errors, PCR artifacts, and incomplete sequencing, it is not possible to obtain absolute quantification of alpha or beta diversity values.
@@
 All simulations have weaknesses and should be interpreted with caution.
-However, the simulated communities generated and analyzed in this study had the advantage of being designed with known properties including the alpha and beta diversity and the their differences between treatment groups.
-Indeed, is is perfectly admissible and proper to use rarefaction with microbial community data.
+Previous studies have reached different conclusions than I have drawn here and elsewhere [@Schloss2023wnwn].
+I believe that the differences are largely due to confusion over the differences in definitions of rarefaction, the types of analysis being used to test the effects of rarefaction, and the use of simulated and actual communities that represent extremes, which are rarely found in real studies.
+It is worth noting that others have also observed sequencing depth correlations with CLR-based distances, but indicated that the correlation was less than what was observed without rarefaction (e.g., see Figure S3 from [@Martino2019]).
+The simulated communities generated and analyzed in this study had the advantage of being designed with known properties including the alpha and beta diversity and the their differences between treatment groups.
+Furthermore, I have used a wide array of alpha and beta diversity metrics and tests of significance that are commonly used in microbial ecology studies.
+The result is the conclusion that it is perfectly admissible and proper to use rarefaction with microbial community data.
 The alternative is to risk reaching unwarranted conclusions.
@@
 
-**Null treatment models.** I created an unbiased and biased treatment model.
+**Null treatment models.** I created unbiased and biased treatment models.
 In the unbiased model, samples were randomly assigned to one of two treatment groups.
@@
 In the second, samples whose sequencing depth where in the 5th percentile or lower were assigned to one treatment group while those in the 95% percentile or higher were assigned to another; the remaining samples were randomly assigned to either treatment group (see Figure 3).
-Comparison of any diversity metric across the two treatment groups should have only yielded a significant result in 5% of the simulations when testing under a Type I error (i.e., $\alpha$) of 0.05.
+Regardless of how the null treatment models were constructed, comparison of any diversity metric across the two treatment groups should have only yielded a significant result in 5% of the simulations when testing under a Type I error (i.e., $\alpha$) of 0.05.
 
@@
 **Power analysis.** The parameters used to design the skewed abundance and richness-adjusted community models were set to impose a known effect size when using community data normalized by rarefaction.
-The statistical power to detect these differences was determined by calculating the p-value for each of 100 replicate simulated set of samples from each dataset using the various alpha and beta diversity metrics.
+The statistical power to detect these differences was determined by calculating the P-value for each of 100 replicate simulated set of samples from each dataset using the various alpha and beta diversity metrics.
 The percentage of tests that yielded a significant P-value was considered the statistical power (i.e., 1 minus the Type II error) to detect the difference.
@@
 The inverse Simpson diversity was calculated as $1/D_{simpson}$.
-In both formulae, $n_i$ was the number of sequences in OTU $i$ and $N$ iwass the number of sequences in the sample.
+In both formulae, $n_i$ was the number of sequences in OTU $i$ and $N$ was the number of sequences in the sample.
 Rarefaction of richness, Shannon diversity and Inverse Simpson diversity values were carried out in mothur.
@@
 The non-parametric bias-corrected Chao1 and ACE richness estimators [@Chao2016] and a non-parametric estimator of the Shannon diversity [@Chao2003] were calculated using raw OTU counts with mothur.
+The iNEXT R package was used to estimate richness and Shannon and inverse Simpson diversity metrics with default parameters in the estimateD function (v.
+3.0.0)[@Chao2014].
+In the size-based approach, estimateD estimates alpha diversity at a level that is twice the size of the smallest sample and in the coverage-based approach it does so at a level that matches the smallest predicted coverage when the number of sequences in each sample is doubled.
 Parametric estimates of sample richness were calculated using the `breakaway` (BA) R package (v.`r package_version("breakaway")`)[@Willis2015].
@@
 Raw OTU counts were used to calculate centered log ratio (CLR) values for each OTU, which were then used to calculate Euclidean distances; such distances are commonly referred to as Aitchison distances.
-CLR abundances are calculated as:
+CLR abundances were calculated as:
 
@@
 To overcome this problem, I attempted a four approaches.
-The first, Zero CLR, imputed the value of the zeroes based on the observed data using the zCompositions (v.`r package_version("zCompositions")`) R package [@Palarea2015].
-The second, One CLR, added a pseudo-count of 1 to the abundance of all OTUs [@Paulson2013; @Lin2020].
-The third, Nudge CLR, added a pseudo-count of 1 divided by the total number of sequences in a sample to each OTU in the sample [@Costea2014; @Lin2020].
-The final approach, Robust CLR, used matrix completion to replace zero counts for unobserved OTUs prior to calculating the CLR [@Martino2019].
+The first, One CLR, added a pseudo-count of 1 to the abundance of all OTUs [@Paulson2013; @Lin2020].
+The second, Nudge CLR, added a pseudo-count of 1 divided by the total number of sequences in a sample to each OTU in the sample [@Costea2014; @Lin2020].
+The third, Zero CLR, imputed the value of the zeroes based on the observed data using the zCompositions (v.`r package_version("zCompositions")`) R package [@Palarea2015].
+The final approach, Robust PCA, calculated the CLR on non-zero counts and then used matrix completion to replace the zero counts; this was implemented using the rpca function in the Gemelli Python package (v.0.0.9) with default parameters [@Martino2019].
+It was not possible to perform Robust PCA on the Rice, Seagrass, or Stream datasets since more than 250 GB of RAM was required; the other datasets required less than 16 GB of RAM.
 
@@
 
-**Reproducible data analysis.** A complete reproducible workflow written in Snakemake (v.7.15.2) and Conda (v.4.12.0) computational environment can be obtained from the GitHub hosted git repository for this project (https://github.com/SchlossLab/Schloss_Rarefaction_XXXXX_2023).
+**Reproducible data analysis.** A complete reproducible workflow written in Snakemake (v.7.15.2) and Conda (v.4.12.0) computational environment can be obtained from the GitHub hosted git repository for this project (https://github.com/SchlossLab/Schloss_Rarefaction_mSphere_2024).
 This paper was written in R-flavored markdown (v.`r package_version("rmarkdown")`) with the `kableExtra` (v.`r package_version("kableExtra")`) package.
@@
 
-\includegraphics[height=17cm]{figure_1.png}
-
 **Figure 1.
@@
 Each point represents the mean of 100 random null community models; the standard deviation was smaller than the size of the plotting symbol.
-
-\newpage
-
-\includegraphics[height=17cm]{figure_2.png}
 
@@
 
-\newpage
-
-\includegraphics[height=17cm]{figure_3.png}
-
-**Figure 3.The risk of falsely detecting a difference between treatment groups drawn from a null model does not meaningfully vary from 5% when data are normalized by rarefaction when sequencing depth is confounded with treatment group**.
+**Figure 3.The risk of falsely detecting a difference between treatment groups drawn from a null model did not meaningfully vary from 5% when data are normalized by rarefaction when sequencing depth was confounded with treatment group**.
 Samples were assigned to different treatment groups so that the smallest 5% of samples were assigned to one treatment group and the largest 5% to another; the other 90% of samples were randomly assigned to both treatment groups.
@@
 The false positive rate was the number of times a dataset yielded a significant result.
-
-\newpage
-
-\includegraphics[height=17cm]{figure_4.png}
 
@@
 
-\newpage
-
-\includegraphics{figure_5.png}
-
 **Figure 5.
@@
 The power was the number of times a dataset yielded a significant result.
-
-\newpage
-
-\includegraphics{figure_6.png}
 
@@
 
-\newpage
-
-\includegraphics{figure_7.png}
-
 **Figure 7.
 Most datasets are sequenced to a level that provides a high level of coverage.** Each plotting symbol represents the observed Good's coverage for a different sample in each dataset.
+The box and whisker plot indicates the range of coverage values when the observed community data were normalized by rarefaction to the size of the least sequenced sample.
 The smoothed line indicates the simulated coverage for varying levels of sequencing effort when a null community is generated from the observed data.
-The box and whisker plot indicates the range of coverage values when the observed community data were normalized by rarefaction to the size of the least sequenced sample.
 
 \newpage
-
-\includegraphics{figure_s1.png}
 
@@
 
-\newpage
-
-\includegraphics{figure_s2.png}
-
 **Figure S2.
@@
 The Spearman correlation coefficient ($\rho$) and test of whether the coefficient was significantly different from zero are indicated for each panel.
-
-\newpage
-
-\includegraphics{figure_s3.png}
+The gray points for the iNEXT estimates indicate samples where richness values were estimated while the black points are those were rarefaction was used.
 
@@
 Examples of differences in beta diversity in each of the 490 samples that were generated for one randomization of the null model using the human dataset.** The x-axis indicates the difference in the number of sequences in each of the samples that went into calculating the pairwise distance prior to each method's approach of controlling for uneven sequencing effort.
+The values on the y-axis for the Robust PCA distances were multiplied by 20 to fit them on the same scale as the other methods.
 The Spearman correlation coefficient ($\rho$) and test of whether the coefficient was significantly different from zero are indicated for each panel.
 
-\newpage
-
-\includegraphics[height=17cm]{figure_s4.png}
-
-**Figure S4.The risk of falsely detecting a difference between treatment groups drawn from a null model does not meaningfully vary from 5% when data are normalized by rarefaction when sequencing depth is confounded with treatment group**.
+**Figure S4.The risk of falsely detecting a difference between treatment groups drawn from a null model did not meaningfully vary from 5% when data are normalized by rarefaction when sequencing depth was completely confounded with treatment group**.
 Samples were assigned to different treatment groups based on whether they were above the median number of sequences for each dataset.

### 2023-12-15 submission/manuscript.tex
Editing / submitted

@@
 because pooling of DNA from multiple PCRs is fraught with numerous
-opportunities for technical errors to compound leading to a skewed
+opportunities for random sampling errors to compound leading to a skewed
 distribution. Aside from developing better methods of pooling DNA, the
@@
 reduced the statistical power to correctly cluster samples into the same
-treatment groups based on beta diversity metrics. Although the detail
-was lost on many subsequent researchers, they did not describe
-rarefaction, but rarefying (12). According to McMurdie and Holmes,
-rarefying is a single subsampling of an OTU abundance table whereas
-rarefaction repeats the subsampling step many times. Furthermore, their
-simulations penalized the rarefied data by removing 15\% of the samples
-and evaluated clustering quality using a clustering algorithm that
+treatment groups based on beta diversity metrics. I have recently shown
+that their analysis had a number of important problems (12). First,
+although the detail was lost on many subsequent researchers, they did
+not describe rarefaction. Instead they described rarefying as a single
+subsampling of an OTU abundance table whereas rarefaction repeats the
+subsampling step many times. Furthermore, their simulations penalized
+the rarefied data by removing 15\% of the samples but did not remove
+those samples when considernig their other methods. They also evaluated
+the accuracy of clustering samples using a clustering algorithm that
 performed worse for rarefied data. According to my reanalysis of
-rarefaction using the original simulation framework, rarefaction
-outperformed other normalization method being used in 2014 for both
-alpha and beta diversity metrics (12). Others have also critiqued the
-original work (13). Regardless, there is a general sense in the
-community that rarefying and rarefaction should not be used with
-microbiome data.
+rarefaction using their original simulation framework, but using all of
+the samples, rarefaction outperformed the other normalization methods
+they described for both alpha and beta diversity metrics (12). Others
+have also critiqued the original work (13).
 
@@
 in amplicon sequencing studies. For alpha diversity metrics, Willis used
-toy datasets to demonstrate that one could estimate the richness for
-each sample in a dataset and use those values for statistical
-comparisons (14). Non-parametric estimators of richness and diversity
-(15, 16) and parametric estimators of richness (17) have been used in
-microbial ecology studies. For beta diversity metrics at least four
-approaches have been pursued. First, one could use relative abundance
-values where the observed number of sequences in an OTU is divided by
-the total number of sequences in the sample (18). Second, normalization
-strategies have been developed where the relative abundance is
-multiplied by the size of the minimum desired sequencing effort and
-fractional values are reapportioned to the OTUs to obtain integer values
-(19, 20). Third, a variety of center log-ratio methods have been
-developed where the compositional nature of the OTU counts is removed
-and used to calculated Euclidean distances (aka Aitchison distances)
-(18, 20--23). This strategy is purported to control for uneven
-sequencing effort (22, 24); however, some have noticed that this feature
-breaks down under certain conditions (25). Finally, as mentioned above,
-variance stabilization transformations have been recommended to generate
-values that can be used to calculate Euclidean distances (11).
+overly simplistic datasets to demonstrate that one could estimate the
+richness for each sample in a dataset and use those values for
+statistical comparisons (14). Non-parametric estimators of the minimum
+richness in a community and diversity (15--17) and parametric estimators
+of richness (18) have been used in microbial ecology studies. One
+non-parametric approach, iNEXT, combines rarefaction on samples more
+than a threshold number of sequences and extrapoliation on samples below
+the threshold (17). This method has become popular in traditional
+ecology studies, but has not been widely used amongst microbial
+ecologists. For beta diversity metrics at least four approaches have
+been pursued as alternatives to rarefaction. First, one could use
+relative abundance values where the observed number of sequences in an
+OTU is divided by the total number of sequences in the sample (19).
+Second, normalization strategies have been developed where the relative
+abundance is multiplied by the size of the minimum desired sequencing
+effort and fractional values are reapportioned to the OTUs to obtain
+integer values (20, 21). Third, a variety of center log-ratio methods
+have been developed where the compositional nature of the OTU counts is
+removed and used to calculated Euclidean distances (aka Aitchison
+distances) (19, 21--24). This strategy is purported to control for
+uneven sequencing effort (23, 25); however, some have noticed that this
+feature breaks down under certain conditions (26). Finally variance
+stabilization transformations have been recommended to generate values
+that can be used to calculate Euclidean distances (11).
 
@@
 using a simulation framework that overcame the issues with the McMurdie
-and Holmes study. My analysis included 16S rRNA gene sequence data from
-from 12 studies that characterized the variation in bacterial
-communities from diverse environments (Table 1 and Figure S1). The
-sequences were assigned to OTUs using a standard pipeline and their
-frequencies and the number of sequences found in each sample were used
-to generate simulated communities and treatment effects. For each
-dataset and simulation, 100 replicate datasets were generated and used
-as input to each of the strategies for controlling for uneven sequencing
-effort. My overall conclusion was that rarefaction outperformed the
-alternative strategies.
+and Holmes study (11). My analysis included 16S rRNA gene sequence data
+from 12 diverse environments (Table 1 and Figure S1). The sequences were
+assigned to OTUs using a standard pipeline and their frequencies and the
+number of sequences found in each sample were used to generate simulated
+communities and treatment effects. For each dataset and simulation, 100
+replicate datasets were generated and used as input to each of the
+strategies for controlling for uneven sequencing effort. My overall
+conclusion was that rarefaction outperformed the alternative strategies.
 
@@
 parametric approaches, and using rarefaction (e.g.~Figure S2). For each
-dataset, all of the approaches, except for rarefaction, showed a strong
-correlation between richness and the number of sequences in the sample
-(Figure 1A). Next, I assessed diversity using the Shannon diversity
-index and the inverse Simpson diversity index without any correction,
-using normalized OTU counts, and rarefaction; I also used a
-non-parametric estimator of Shannon diversity. The correlation between
-sequencing depth and the diversity metric was not as strong as it was
-for richness and the inverse Simpson diversity values were less
-sensitive than the Shannon diversity values; however, the correlation to
-the diversity metrics calculated with rarefaction were the lowest for
-all of the metrics and studies (Figure 1A). The alpha-diversity metrics
-calculated with rarefaction consistently demonstrated a lack of
-sensitivity to sequencing depth.
+dataset, all of the approaches, except for rarefaction and possibly the
+size-based iNEXT estimates, showed a strong correlation between richness
+and the number of sequences in the sample (Figure 1A). The size-based
+iNEXT estimates were correlated with sampling effort for the Marine and
+Soil datasts, which had fewer samples and more uniform sequencing depth
+distributions than the other datasets. Next, I assessed diversity using
+the Shannon diversity index and the inverse Simpson diversity index
+without any correction, using normalized OTU counts, non-parametric
+estimation, and rarefaction. The correlation between sequencing depth
+and the Shannon diversity metric was not as strong as it was for
+richness. The inverse Simpson diversity values was not sensitive to
+sampling effort (Figure 1A). The alpha-diversity metrics calculated with
+rarefaction consistently demonstrated a lack of sensitivity to
+sequencing depth.
 
@@
 calculated the Euclidean distance on raw OTU counts where the central
-log-ratio (CLR) was calculated (i.e., Aitchison distances) using matrix
-completion to replace zero counts (Robust CLR), adding a pseudo-count of
-one to all OTU counts prior to calculating the CLR (One CLR), adding a
-pseudo-count of one divided by the total number of sequences obtained
-for the community (Nudge CLR), and imputing the value of zero counts
-(Zero CLR). The Aitchison distances were all strongly sensitive to
+log-ratio (CLR) was calculated (i.e., Aitchison distances) by adding a
+pseudo-count of one to all OTU counts prior to calculating the CLR (One
+CLR), adding a pseudo-count of one divided by the total number of
+sequences obtained for the community (Nudge CLR), imputing the value of
+zero counts (Zero CLR), and using CLR on non-zero counts followed by
+matrix completion on zero counts (Robust Principle Components Analysis
+{[}PCA{]}). The Aitchison distances were all strongly sensitive to
 sequencing effort (Figure 1B). Finally, I used the
@@
 calculating Euclidean distances. Again, there was a strong sensitivity
-to sequencing effort (Figure 1B). Although Euclidean distances are not
-typically used on raw or rarefaction-normalized count data in ecology
-(26), Euclidean distances calculated with rarefaction were not sensitive
+to sequencing effort (Figure 1B). Although use of Euclidean distances is
+not advised with raw or rarefaction-normalized count data in ecology
+(27), Euclidean distances calculated with rarefaction were not sensitive
 to sequencing effort. Across each of the beta diversity metrics and
@@
 tests yielded a significant result (Figure 2). Second, I employed a
-biased null treatment model where the treatment group was completely
-confounded with the number of sequences in each sample (Figure S4).
-Under this model, only the data normalized with rarefaction consistently
-resulted in a 5\% false positive rate for alpha and beta diversity
-metrics. I then used a less severe bias where the smallest 5\% of
-samples were assigned to one treatment group and the largest 5\% were
-assigned to another; the other samples were randomly assigned to both
-treatment groups. Although the sensitivity of the false positive rate to
-the level of confounding varied across methods of quantifying alpha and
-beta diversity, rarefaction was the only method to consistently control
-the false positive rate (Figure 3). These results aligned with the
-observed sensitivity of alpha and beta diversity metrics to sequencing
-effort and underscore the value of rarefaction.
+biased null treatment model where the treatment group was confounded
+with the number of sequences in each sample such that the 50\% of
+samples with the least sequencing effort were in one treatment group and
+the others were in another treatment group (Figure S4). Under this
+model, only the data normalized with rarefaction consistently resulted
+in a 5\% false positive rate for alpha and beta diversity metrics. I
+then used a less severe bias where the smallest 5\% of samples were
+assigned to one treatment group and the largest 5\% were assigned to
+another; the other samples were randomly assigned to both treatment
+groups. Although the sensitivity of the false positive rate to the level
+of confounding varied across methods of quantifying alpha and beta
+diversity, rarefaction and the iNEXT size-based estimates for alpha
+diveristy were the only methods to consistently control the false
+positive rate (Figure 3). These results aligned with the observed
+sensitivity of alpha and beta diversity metrics to sequencing effort and
+underscore the value of rarefaction.
 
@@
 the approach for generating the perturbed community did not necessarily
-change the number of OTUs in each treatment group. Regardless, the
-simulations testing differences in richness using the Rice and Stream
-datasets had the greatest power when the richness data were calculated
-with rarefaction. To explore this further, a richness-adjusted community
-model was created by removing 3\% of the OTUs from a null model model.
-As suggested by the first simulation, the richness data calculated with
-rarefaction had a higher statistical power than the other approaches
-when measuring richness (Figure 5). The simulations testing the power to
-detect differences in Shannon diversity also showed that rarefaction
-performed better than the other methods (Figure 4A). When testing for
-differences in the Inverse Simpson diversity index the the difference
-between rarefaction and the other methods was negligible (Figure 4A).
-For tests of beta diversity I found that rarefaction was the most
-reliable approach to maintain statistical power to detect differences
-between two communities. Among the tests using the Jaccard and
-Bray-Curtis metrics, raw count data and CSS normalized data had little
-power relative to using rarefaction, relative abundance, and SRS to
-normalize the uneven sequencing depths. The differences in power between
-counts normalized with rarefaction, relative abundance, and SRS data was
-small, but if there were differences, the power obtained using
-rarefaction was greater than the other methods. Among the tests using
-Euclidean distances, using raw counts and CLR and DESeq2 transformed
-data had little power relative to the distances calculated using
-rarefaction and relative abundances. This power-based analysis of the
-simulated communities using different methods of handling uneven sample
-sizes demonstrated the value of rarefaction for preserving the
-statistical power to detect differences between treatment groups for
-measures of alpha and beta diversity.
+change the number of OTUs in each treatment group; the differences
+primarily affected the distribution of the sequences across OTUs.
+Regardless, the simulations testing differences in richness using the
+Rice and Stream datasets had the greatest power when the richness data
+were calculated with rarefaction. To explore this further, a
+richness-adjusted community model was created by randomly selecting and
+removing 3\% of the OTUs from a null model. As suggested by the first
+simulation, the richness data calculated with rarefaction consistently
+had the highest statistical power among the approaches used to measure
+richness followed by the iNEXT size-based estimates (Figure 5).
+Returning to simulations where treatment groups varied in their
+structure, the power to detect differences in Shannon diversity was the
+lowest when using raw relative abundances than when using the other
+methods (Figures 4A and 5). When testing for differences in the Inverse
+Simpson diversity index the the difference between rarefaction and the
+other methods was negligible (Figures 4A and 5). For tests of beta
+diversity, I found that rarefaction was the most reliable approach to
+maintain statistical power to detect differences between two communities
+(Figure 4B). Among the tests using the Jaccard and Bray-Curtis metrics,
+raw count data and CSS normalized data had little power relative to
+using rarefaction, relative abundance, and SRS to normalize the uneven
+sequencing depths. The differences in power between counts normalized
+with rarefaction, relative abundance, and SRS data was small, but if
+there were differences, the power obtained using rarefaction was greater
+than the other methods. Among the tests using Euclidean distances, using
+raw counts and CLR and DESeq2 transformed data had little power relative
+to the distances calculated using rarefaction and relative abundances.
+This power-based analysis of the simulated communities using different
+methods of handling uneven sample sizes demonstrated the value of
+rarefaction for preserving the statistical power to detect differences
+between treatment groups for measures of alpha and beta diversity.
 
 \textbf{\emph{Increased rarefaction depth reduces intra-sample variation
-in alpha and beta diversity.}} Once concern with using rarefaction is
-the perceived loss of sequencing information when a large fraction of
-data appears to be removed when the community with the greatest
-sequencing depth is sampled to the size of the community with the least
-(e.g., the smallest sample in the Bioethanol dataset had 1.04\% of
-sequences that were in the largest sample). To assess the sensitivity of
-alpha and beta diversity metrics to rarefaction depth, I again used the
-dataset generated using the null models, but used rarefaction with each
+in alpha and beta diversity.}} One concern with using rarefaction is the
+perceived loss of sequencing information when a large fraction of data
+appears to be removed when the community with the greatest sequencing
+depth is sampled to the size of the community with the least (e.g., the
+smallest sample in the Bioethanol dataset had 1.04\% of sequences that
+were in the largest sample). To assess the sensitivity of alpha and beta
+diversity metrics to rarefaction depth, I again used the dataset
+generated using the null models, but used rarefaction with each
 community to varying sequencing depths (Figure 6). The richness values
@@
 deviation) between samples for richness, Shannon diversity, and
-Bray-Curtis distances. Although the mean richness appeared to increased
+Bray-Curtis distances. Although the mean richness appeared to increase
 unbounded with sequencing effort, the coefficient of variation for each
@@
 Stream datasets, the median coverage with rarefaction was still greater
-than 90\%. These results suggest that most studies had a level of
-sequencing coverage that aligned with the diversity of the communities.
-Next, I used the null model for each dataset to ask how much sequencing
-effort was required to obtain higher levels of coverage. To obtain 95
-and 99\% coverage, an average of 2.70 and 101.20-fold more sequence data
-was estimated to be required than was required to obtain 90\% coverage,
-respectively (Figure 7). As suggested by the simulated coverages curve
-in Figure 7, these estimates are conservative. Regardless, the
-sequencing effort required to achieve higher sequencing depth would
-likely limit the number of samples that could be sequenced when
-controlling for costs. Although it may be disconcerting to use
-rarefaction to normalize to a sequencing depth that is considerably
-lower than that obtained for the best sequenced community in a dataset,
-sequencing coverage for many environments is probably adequate even at
-the lower sequencing depth. Of course, regardless of the concerns
-surrounding the choice of the rarefaction depth, the results throughout
-this study demonstrate that rarefaction is necessary to avoid reaching
-incorrect inferences.
+than 90\% (see the box plots in Figure 7). These results suggest that
+most studies had a level of sequencing coverage that aligned with the
+diversity of the communities. Next, I used the null model for each
+dataset to ask how much sequencing effort was required to obtain higher
+levels of coverage. To obtain 95 and 99\% coverage, an average of 3 and
+101-fold more sequence data was estimated to be required than was
+required to obtain 90\% coverage, respectively (Figure 7). As suggested
+by the simulated coverages curve in Figure 7, these estimates are
+conservative. Regardless, the sequencing effort required to achieve
+higher sequencing depth would likely limit the number of samples that
+could be sequenced when controlling for costs. Although it may be
+disconcerting to use rarefaction to normalize to a sequencing depth that
+is considerably lower than that obtained for the best sequenced
+community in a dataset, sequencing coverage for many environments is
+probably adequate even at the lower sequencing depth. Of course,
+regardless of the concerns surrounding the choice of the rarefaction
+depth, the results throughout this study demonstrate that rarefaction is
+necessary to avoid reaching incorrect inferences.
 
@@
 alpha and beta diversity indices. Compared to all other methods,
-rarefaction removed the correlation between sequencing depth and alpha
-or beta diversity metrics when the sequencing depth varied by as much as
-97-fold across samples. I showed that this correlation could lead
-falsely detecting differences between treatment groups if sequencing
-depth and sequencing effort are confounded. The correlation with
-sequencing effort leads to an artificial increase in the variation
-between samples and a reduced power to detect true differences in alpha
-and beta diversity. For these reasons, rarefaction is a valuable tool to
-control for uneven sequencing effort until improved statistical
-procedures are developed or it becomes possible to more evenly
-distribute sequencing effort across samples.
-
-The primary alternative to rarefaction for measuring alpha diversity is
-to estimate the metric using non-parametric or parametric models with
-raw counts and to then compare the estimates (7, 14). My results
-demonstrate that such approaches are limited for several reasons. First,
-non-parametric richness estimates such as ACE and Chao1 are sensitive to
-sequencing effort. Therefore, these strategies do not, in practice,
-remove the effects of sequencing effort. Second, parametric approaches,
-such as those implemented in the \texttt{breakaway} R package, generate
-confidence intervals that are likely to include the true richness and
-theoretically shrink with increased sequencing effort. Yet for most
-samples, the confidence intervals around the estimates are too wide to
-be meaningful, again leading to an inability to remove the effects of
-sequencing effort. Third, it has become an increasingly common practice
-for researchers to remove sequences that are rare in a sample (e.g.,
-those that appear once). Although that approach was not taken in this
-study, removing rare sequences would skew the distribution of sequences
-and OTUs leading to a distortion of the measurement of alpha diversity
-(27). The effects of removing rare sequences would vary across samples
-depending on the number of sequences in each sample. One interesting
-result of this analysis was the demonstration that as metrics that
-depend less on rare taxa are used, the effect of uneven sequencing
-effort was reduced. For example, richness counts a sequence appearing
-once as much as sequence appearing 1,000 times, while the Shannon
-diversity index places more emphasis on the more abundant sequence, and
-the inverse Simpson index even more. Although normalizing communities to
-a common number of sequences is also suggested (e.g., SRS normalization)
-to control for uneven sequencing effort, the current analysis indicates
-that its performance does not meet that of rarefaction. For analysis of
-alpha diversity metrics, rarefaction is the most effective and
-consistent approach to control for uneven sequencing effort.
+rarefaction was the most consistent method for removing the correlation
+between sequencing depth and alpha or beta diversity metrics even when
+the sequencing depth varied by as much as 97-fold across samples. I
+showed that this correlation could lead falsely detecting differences
+between treatment groups if sequencing depth and sequencing effort are
+confounded. The correlation with sequencing effort leads to an
+artificial increase in the variation between samples and a reduced power
+to detect true differences in alpha and beta diversity. For these
+reasons, rarefaction is a valuable tool to control for uneven sequencing
+effort until improved statistical procedures are developed or it becomes
+possible to more evenly distribute sequencing effort across samples.
+
+Historically, the primary alternative to rarefaction for measuring alpha
+diversity is to estimate the metric using non-parametric or parametric
+models with raw counts and to then compare the estimates (7, 14). My
+results demonstrate that such approaches are limited for several
+reasons. First, non-parametric richness estimates such as ACE and Chao1
+are sensitive to sequencing effort. Therefore, these estimators do not,
+in practice, remove the effects of sequencing effort. Second, parametric
+approaches, such as those implemented in the \texttt{breakaway} R
+package, generate confidence intervals that are likely to include the
+true richness and theoretically shrink with increased sequencing effort.
+Yet for most samples, the confidence intervals around the estimates are
+too wide to be meaningful, again leading to an inability to remove the
+effects of sequencing effort. Third, it has become an increasingly
+common practice for researchers to remove sequences that are rare in a
+sample (e.g., those that appear once). Although that approach was not
+taken in this study, removing rare sequences would skew the distribution
+of sequences and OTUs leading to a distortion of the measurement of
+alpha diversity (28). The effects of removing rare sequences would vary
+across samples depending on the number of sequences in each sample. One
+interesting result of this analysis was the demonstration that as
+metrics that depend less on rare taxa are used, the effect of uneven
+sequencing effort was reduced. For example, richness counts a sequence
+appearing once as much as sequence appearing 1,000 times, while the
+Shannon diversity index places more emphasis on the more abundant
+sequence, and the inverse Simpson index even more. Although normalizing
+communities to a common number of sequences is also suggested (e.g., SRS
+normalization) to control for uneven sequencing effort, the current
+analysis indicates that its performance does not meet that of
+rarefaction. Because alpha diversity metrics are so sensitive to
+sampling effort and residual PCR and sequencing artifacts, it is
+important to treat their values on a relative (e.g., 5\% difference)
+rather than absolute basis (e.g., difference of 48 OTUs). Rarefaction is
+an effective and consistent approach to control for uneven sequencing
+effort.
 
@@
 more robust alternatives to rarefaction. Again, the only approach that
-consistently removed the effects of uneven sequencing effort was
-rarefaction. Most of the recommendations borrow techniques from methods
-for identifying differential gene expression. Unfortunately, there is an
-important but under appreciated difference between gene expression and
-community data. This is the interpretation of unobserved data. For gene
-expression analysis in a single organism the lack of any sequencing data
-for a gene would indicate that although the gene was present, its
-expression was below the limit of detection. Sequencing the same
-organism under multiple conditions would not lead to a seemingly
-unbounded number of genes in the organisms. Rather, the number of genes
-has a definite limit that is knowable from the genome sequence. With
-microbiome data, an unobserved sequence could mean that the organism was
-present, but below the limit of detection or that the organism was
-missing. Because we have yet to exhaustively sample any community in the
-same way we have sequenced a single genome, it is unreliable to impute
-the presence of all organisms. Yet, this is exactly what the variance
-stabilization transformation and most CLR techniques do. This analysis
-has demonstrated a clear correlation between distances calculated by
-these methods and sequencing effort. This result is at odds with the
-claims by others that the distances are scale invariant (22, 24). Again,
-rarefaction is the most effective and consistent approach to control for
-uneven sequencing effort when calculating beta diversity metrics.
+consistently removed the effects of uneven sequencing effort across
+alpha and beta diversity analyses was rarefaction. Many of the
+alternative methods have been borrowed from differential gene expression
+analysis. Unfortunately, there is an important but under appreciated
+difference between gene expression and community data. This is the
+interpretation of unobserved data. For gene expression analysis in a
+single organism the lack of any sequencing data for a gene would
+indicate that although the gene was present, its expression was below
+the limit of detection. Sequencing the same organism under multiple
+conditions would not lead to a seemingly unbounded number of genes in
+the organisms. Rather, the number of genes has a definite limit that is
+knowable from the genome sequence. With microbiome data, an unobserved
+sequence could mean that the organism was present, but below the limit
+of detection or that the organism was missing. Because we have yet to
+exhaustively sample any community in the same way we have sequenced a
+single genome, it is unreliable to impute the presence of all organisms.
+Yet, this is exactly what the variance stabilization transformation and
+most CLR techniques do. This analysis has demonstrated a clear
+correlation between distances calculated by these methods and sequencing
+effort. This result is at odds with the claims by others that the
+distances are scale invariant (23, 25). Again, rarefaction is the most
+effective and consistent approach to control for uneven sequencing
+effort when calculating beta diversity metrics.
 
@@
 repeated a large number of times (e.g., 100 or 1000) the risks of
-ignoring or oversampling rare taxa are mitigated. Furthermore, it is
-curious that the study making the original critique removed rare
-sequences. A parallel analysis to this study has demonstrated that many
-of these sequences are likely valid and that removal of rare sequences
-can bias alpha and beta diversity metrics and reduce statistical power
-(27). As for the second criticism, I would resist the claim that the
-selection of the rarefaction threshold is arbitrary. In practice, there
-is a tradeoff between sampling breadth and sequencing depth. Greater
-breadth will increase the statistical power to compare treatment groups
-and greater depth will increase the resolution to describe the
-communities. My personal process for rarefaction threshold involves
-looking for a natural break in the distribution of the number of
-sequences. For example, the Lake dataset used in this study had a clear
-break around 10,000 sequences (Figure S1). I would also consider what
-samples are below any break that I select. If there were critical
-samples below the break I would either reduce the threshold or obtain
-more sequences from those samples. As shown in my analysis of Good's
-coverage values, most studies obtain an ample level of coverage and
-would need to increase their sequencing depth by 10-fold to increase the
-coverage by several percentage points. In past work, I have favored
-distributing the total number of sequences to increase sampling breadth
-rather than sequencing depth.
+ignoring or oversampling rare taxa are mitigated. As for the second
+criticism, I would resist the claim that the selection of the
+rarefaction threshold is arbitrary. In practice, there is a tradeoff
+between sampling breadth and sequencing depth. Greater breadth will
+increase the statistical power to compare treatment groups and greater
+depth will increase the resolution to describe the communities. My
+personal process for selecting a rarefaction threshold involves looking
+for a natural break in the distribution of the number of sequences. For
+example, the Lake dataset used in this study had a clear break around
+10,000 sequences (Figure S1). I would also consider what samples are
+below any threshold that I select. If there were critical samples below
+the threshold I would either reduce the threshold or obtain more
+sequences from those samples. As shown in my analysis of Good's coverage
+values, most studies obtain an ample level of coverage and would need to
+increase their sequencing depth by 10-fold to increase the coverage by
+several percentage points. In past work, I have favored increasing
+sampling breadth over sequencing depth.
+
+An additional critique of rarefaction is that the variation between
+alpha diversity metrics shrinks as the rarefaction threshold decreases,
+which may compromise the power to detect differences between treatment
+groups. The iNEXT methods attempt to minimize this effect by estimating
+alpha diversity metrics for samples whose sequencing depth is below a
+threshold and using rarefaction on samples with a sequencing depth above
+that threshold (17). This threshold is set at either twice the size of
+the smallest sample (i.e., size-based) or by controlling for the
+coverage when the number of sequences in each sample is doubled. An
+example of this can be seen in Figure S2 where richness values obtained
+by estimation are shown in gray and those by rarefaction are in black.
+While my data suggest that the size-based approach has promise for
+estimating alpha diversity, I found that it was not as consistent as
+rarefaction for smaller datasets where there was a narrow distribution
+in sequencing depths. This is likely because in these situatons the
+algorithm relies more on estimation than rarefaction.
 
@@
 sequences that allow them to later assign sequences back to the samples
-that they came from. When the PCR products are pooled, efforts are taken
-to pool the fragments in equimolar ratios. Researchers use one of two
+that they came from. When the PCR products are pooled, efforts are often
+taken to pool the fragments in equimolar ratios using one of two
 approaches. First, they often will quantify the concentration of DNA
@@
 approaches have limitations that reduce the ability to truly achieve
-equimolar mixture. In addition, for some samples it is common to
-co-amplify non-specific DNAs which may add to the challenges of
-achieving equimolar mixtures of the desired amplicons since those
-sequences will be removed (28). Regardless, it is clear that better
-strategies are needed to reduce the variation in the number of sequences
-generated for each sample. Furthermore, it is important to keep in mind
-that because of residual sequencing errors, PCR artifacts, and
-incomplete sequencing, it is not possible to obtain absolute
-quantification of alpha or beta diversity values. Instead, comparisons
-of these values need to be made on a relative basis.
+equimolar mixture. Although normalization procedures are expected to
+remove the relationship between sequencing depth and sample biomass
+(29), several recent studies have found a positive correlation (30, 31).
+Even in cases where the is no correlation, for some samples it is common
+to co-amplify non-specific DNAs which are later removed during sequence
+processing (32). It is clear that better strategies are needed to reduce
+the variation in the number of sequences generated for each sample.
+Furthermore, it is important to keep in mind that because of residual
+sequencing errors, PCR artifacts, and incomplete sequencing, it is not
+possible to obtain absolute quantification of alpha or beta diversity
+values. Instead, comparisons of these values need to be made on a
+relative basis.
 
 All simulations have weaknesses and should be interpreted with caution.
-However, the simulated communities generated and analyzed in this study
-had the advantage of being designed with known properties including the
-alpha and beta diversity and the their differences between treatment
-groups. Indeed, is is perfectly admissible and proper to use rarefaction
-with microbial community data. The alternative is to risk reaching
-unwarranted conclusions.
+Previous studies have reached different conclusions than I have drawn
+here and elsewhere (12). I believe that the differences are largely due
+to confusion over the differences in definitions of rarefaction, the
+types of analysis being used to test the effects of rarefaction, and the
+use of simulated and actual communities that represent extremes, which
+are rarely found in real studies. It is worth noting that others have
+also observed sequencing depth correlations with CLR-based distances,
+but indicated that the correlation was less than what was observed
+without rarefaction (e.g., see Figure S3 from (23)). The simulated
+communities generated and analyzed in this study had the advantage of
+being designed with known properties including the alpha and beta
+diversity and the their differences between treatment groups.
+Furthermore, I have used a wide array of alpha and beta diversity
+metrics and tests of significance that are commonly used in microbial
+ecology studies. The result is the conclusion that it is perfectly
+admissible and proper to use rarefaction with microbial community data.
+The alternative is to risk reaching unwarranted conclusions.
 
@@
 OptiClust algorithm to cluster sequences together that were not more
-than 3\% different from each other (29).
+than 3\% different from each other (33).
 
@@
 the total number of sequences in each OTU. This is a similar approach to
-that of the IS algorithm described by Ulrich and Gotelli (30). Because
+that of the IS algorithm described by Ulrich and Gotelli (34). Because
 the construction of the null models was a stochastic process, 100
@@
 
-\textbf{Null treatment models.} I created an unbiased and biased
-treatment model. In the unbiased model, samples were randomly assigned
-to one of two treatment groups. I created two biased treatment models.
-In the first, samples that had more than the median number of sequences
-for a dataset were assigned to one treatment group and the rest were
-assigned to a second treatment group (see Figure S4). In the second,
-samples whose sequencing depth where in the 5th percentile or lower were
+\textbf{Null treatment models.} I created unbiased and biased treatment
+models. In the unbiased model, samples were randomly assigned to one of
+two treatment groups. I created two biased treatment models. In the
+first, samples that had more than the median number of sequences for a
+dataset were assigned to one treatment group and the rest were assigned
+to a second treatment group (see Figure S4). In the second, samples
+whose sequencing depth where in the 5th percentile or lower were
 assigned to one treatment group while those in the 95\% percentile or
 higher were assigned to another; the remaining samples were randomly
-assigned to either treatment group (see Figure 3). Comparison of any
-diversity metric across the two treatment groups should have only
-yielded a significant result in 5\% of the simulations when testing
-under a Type I error (i.e., \(\alpha\)) of 0.05.
+assigned to either treatment group (see Figure 3). Regardless of how the
+null treatment models were constructed, comparison of any diversity
+metric across the two treatment groups should have only yielded a
+significant result in 5\% of the simulations when testing under a Type I
+error (i.e., \(\alpha\)) of 0.05.
 
@@
 function implements a non-parametric multivariate analysis of variance
-using distances matrices (31). Throughout this study I used 0.05 as the
+using distances matrices (35). Throughout this study I used 0.05 as the
 threshold for assessing the statistical significance of any P-values.
@@
 The statistical power to detect these differences was determined by
-calculating the p-value for each of 100 replicate simulated set of
+calculating the P-value for each of 100 replicate simulated set of
 samples from each dataset using the various alpha and beta diversity
@@
 diversity indices. OTU counts were used as input to calculate sample
-richness and Shannon and inverse Simpson diversity using mothur (9, 32).
+richness and Shannon and inverse Simpson diversity using mothur (9, 36).
 Shannon diversity was calculated as
@@
 both formulae, \(n_i\) was the number of sequences in OTU \(i\) and
-\(N\) iwass the number of sequences in the sample. Rarefaction of
+\(N\) was the number of sequences in the sample. Rarefaction of
 richness, Shannon diversity and Inverse Simpson diversity values were
@@
 used to normalize OTU counts to the size of the smallest sample in each
-dataset using the SRS R package (v.0.2.3)(19). Normalized OTU counts
+dataset using the SRS R package (v.0.2.3)(20). Normalized OTU counts
 were used to calculate sample richness and Shannon and inverse Simpson
@@
 abundances of the features do not change with the normalization
-procedure (20). The non-parametric bias-corrected Chao1 and ACE richness
+procedure (21). The non-parametric bias-corrected Chao1 and ACE richness
 estimators (16) and a non-parametric estimator of the Shannon diversity
-(15) were calculated using raw OTU counts with mothur. Parametric
-estimates of sample richness were calculated using the
-\texttt{breakaway} (BA) R package (v.4.7.9)(17). My analysis reports
-both the results from running default model selection procedure and the
-Poisson model. The default model selection returned either the Kemp,
-Negative Binomial, or Poisson models. Relative abundance data were not
-used to calculate alpha diversity metrics since the richness and
+(15) were calculated using raw OTU counts with mothur. The iNEXT R
+package was used to estimate richness and Shannon and inverse Simpson
+diversity metrics with default parameters in the estimateD function (v.
+3.0.0)(17). In the size-based approach, estimateD estimates alpha
+diversity at a level that is twice the size of the smallest sample and
+in the coverage-based approach it does so at a level that matches the
+smallest predicted coverage when the number of sequences in each sample
+is doubled. Parametric estimates of sample richness were calculated
+using the \texttt{breakaway} (BA) R package (v.4.7.9)(18). My analysis
+reports both the results from running default model selection procedure
+and the Poisson model. The default model selection returned either the
+Kemp, Negative Binomial, or Poisson models. Relative abundance data were
+not used to calculate alpha diversity metrics since the richness and
 evenness does not change from the raw data when dividing each sample by
@@
 \texttt{SRS} (v.0.2.3) and \texttt{metagenomeSeq} (v.1.36.0) R packages,
-respectively (19, 20). The normalized counts were then used to calculate
+respectively (20, 21). The normalized counts were then used to calculate
 Jaccard and Bray-Curtis dissimilarity indices. Finally, the
@@
 \texttt{DESeq2} (v.1.34.0) R package was used to normalize the data as
-described by McMurdie and Holmes (11, 33). Because the VST approach
+described by McMurdie and Holmes (11, 37). Because the VST approach
 generated negative values, which are incompatible with calculating
@@
 such distances are commonly referred to as Aitchison distances. CLR
-abundances are calculated as:
+abundances were calculated as:
 
@@
 unobserved OTUs in a sample. To overcome this problem, I attempted a
-four approaches. The first, Zero CLR, imputed the value of the zeroes
-based on the observed data using the zCompositions (v.1.4.0.1) R package
-(34). The second, One CLR, added a pseudo-count of 1 to the abundance of
-all OTUs (18, 20). The third, Nudge CLR, added a pseudo-count of 1
-divided by the total number of sequences in a sample to each OTU in the
-sample (18, 23). The final approach, Robust CLR, used matrix completion
-to replace zero counts for unobserved OTUs prior to calculating the CLR
-(22).
+four approaches. The first, One CLR, added a pseudo-count of 1 to the
+abundance of all OTUs (19, 21). The second, Nudge CLR, added a
+pseudo-count of 1 divided by the total number of sequences in a sample
+to each OTU in the sample (19, 24). The third, Zero CLR, imputed the
+value of the zeroes based on the observed data using the zCompositions
+(v.1.4.0.1) R package (38). The final approach, Robust PCA, calculated
+the CLR on non-zero counts and then used matrix completion to replace
+the zero counts; this was implemented using the rpca function in the
+Gemelli Python package (v.0.0.9) with default parameters (23). It was
+not possible to perform Robust PCA on the Rice, Seagrass, or Stream
+datasets since more than 250 GB of RAM was required; the other datasets
+required less than 16 GB of RAM.
 
@@
 this project
-(\url{https://github.com/SchlossLab/Schloss_Rarefaction_XXXXX_2023}).
+(\url{https://github.com/SchlossLab/Schloss_Rarefaction_mSphere_2024}).
 This paper was written in R-flavored markdown (v.2.16) with the
@@
 Revisiting the analysis that called into question the practice of
-rarefaction. bio{R}xiv.
-doi:\href{https://doi.org/10.1101/2023.06.23.546312}{10.1101/2023.06.23.546312}.}
+rarefaction. mSphere.
+doi:\href{https://doi.org/10.1128/msphere.00355-23}{10.1128/msphere.00355-23}.}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-Chao2014}{}}%
+\CSLLeftMargin{17. }%
+\CSLRightInline{\textbf{Chao A}, \textbf{Gotelli NJ}, \textbf{Hsieh TC},
+\textbf{Sander EL}, \textbf{Ma KH}, \textbf{Colwell RK}, \textbf{Ellison
+AM}. 2014. Rarefaction and extrapolation with hill numbers: A framework
+for sampling and estimation in species diversity studies. Ecological
+Monographs \textbf{84}:45--67.
+doi:\href{https://doi.org/10.1890/13-0133.1}{10.1890/13-0133.1}.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-Willis2015}{}}%
-\CSLLeftMargin{17. }%
+\CSLLeftMargin{18. }%
 \CSLRightInline{\textbf{Willis A}, \textbf{Bunge J}. 2015. Estimating
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Lin2020}{}}%
-\CSLLeftMargin{18. }%
+\CSLLeftMargin{19. }%
 \CSLRightInline{\textbf{Lin H}, \textbf{Peddada SD}. 2020. Analysis of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Beule2020}{}}%
-\CSLLeftMargin{19. }%
+\CSLLeftMargin{20. }%
 \CSLRightInline{\textbf{Beule L}, \textbf{Karlovsky P}. 2020. Improved
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Paulson2013}{}}%
-\CSLLeftMargin{20. }%
+\CSLLeftMargin{21. }%
 \CSLRightInline{\textbf{Paulson JN}, \textbf{Stine OC}, \textbf{Bravo
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Quinn2019}{}}%
-\CSLLeftMargin{21. }%
+\CSLLeftMargin{22. }%
 \CSLRightInline{\textbf{Quinn TP}, \textbf{Erb I}, \textbf{Gloor G},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Martino2019}{}}%
-\CSLLeftMargin{22. }%
+\CSLLeftMargin{23. }%
 \CSLRightInline{\textbf{Martino C}, \textbf{Morton JT}, \textbf{Marotz
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Costea2014}{}}%
-\CSLLeftMargin{23. }%
+\CSLLeftMargin{24. }%
 \CSLRightInline{\textbf{Costea PI}, \textbf{Zeller G}, \textbf{Sunagawa
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Quinn2018}{}}%
-\CSLLeftMargin{24. }%
+\CSLLeftMargin{25. }%
 \CSLRightInline{\textbf{Quinn TP}, \textbf{Erb I}, \textbf{Richardson
@@
 \leavevmode\vadjust pre{\hypertarget{ref-teBeest2021}{}}%
-\CSLLeftMargin{25. }%
+\CSLLeftMargin{26. }%
 \CSLRightInline{\textbf{Beest DE te}, \textbf{Nijhuis EH},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Legendre2012}{}}%
-\CSLLeftMargin{26. }%
+\CSLLeftMargin{27. }%
 \CSLRightInline{\textbf{Legendre P}, \textbf{Legendre L}. 2012.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Schloss2023sing}{}}%
-\CSLLeftMargin{27. }%
+\CSLLeftMargin{28. }%
 \CSLRightInline{\textbf{Schloss PD}. 2020. Removal of rare amplicon
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-Gloor2017}{}}%
+\CSLLeftMargin{29. }%
+\CSLRightInline{\textbf{Gloor GB}, \textbf{Macklaim JM},
+\textbf{Pawlowsky-Glahn V}, \textbf{Egozcue JJ}. 2017. Microbiome
+datasets are compositional: And this is not optional. Frontiers in
+Microbiology \textbf{8}.
+doi:\href{https://doi.org/10.3389/fmicb.2017.02224}{10.3389/fmicb.2017.02224}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Minich2022}{}}%
+\CSLLeftMargin{30. }%
+\CSLRightInline{\textbf{Minich JJ}, \textbf{Härer A}, \textbf{Vechinski
+J}, \textbf{Frable BW}, \textbf{Skelton ZR}, \textbf{Kunselman E},
+\textbf{Shane MA}, \textbf{Perry DS}, \textbf{Gonzalez A},
+\textbf{McDonald D}, \textbf{Knight R}, \textbf{Michael TP},
+\textbf{Allen EE}. 2022. Host biology, ecology and the environment
+influence microbial biomass and diversity in 101 marine fish species.
+Nature Communications \textbf{13}.
+doi:\href{https://doi.org/10.1038/s41467-022-34557-2}{10.1038/s41467-022-34557-2}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Cruz2021}{}}%
+\CSLLeftMargin{31. }%
+\CSLRightInline{\textbf{Cruz GNF}, \textbf{Christoff AP},
+\textbf{Oliveira LFV de}. 2021. Equivolumetric protocol generates
+library sizes proportional to total microbial load in 16S amplicon
+sequencing. Frontiers in Microbiology \textbf{12}.
+doi:\href{https://doi.org/10.3389/fmicb.2021.638231}{10.3389/fmicb.2021.638231}.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-Morris2013}{}}%
-\CSLLeftMargin{28. }%
+\CSLLeftMargin{32. }%
 \CSLRightInline{\textbf{Morris A}, \textbf{Beck JM}, \textbf{Schloss
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Westcott2017}{}}%
-\CSLLeftMargin{29. }%
+\CSLLeftMargin{33. }%
 \CSLRightInline{\textbf{Westcott SL}, \textbf{Schloss PD}. 2017.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Ulrich2010}{}}%
-\CSLLeftMargin{30. }%
+\CSLLeftMargin{34. }%
 \CSLRightInline{\textbf{Ulrich W}, \textbf{Gotelli NJ}. 2010. Null model
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Anderson2001}{}}%
-\CSLLeftMargin{31. }%
+\CSLLeftMargin{35. }%
 \CSLRightInline{\textbf{Anderson MJ}. 2001. A new method for
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Magurran2004}{}}%
-\CSLLeftMargin{32. }%
+\CSLLeftMargin{36. }%
 \CSLRightInline{\textbf{Magurran AE}. 2004.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Love2014}{}}%
-\CSLLeftMargin{33. }%
+\CSLLeftMargin{37. }%
 \CSLRightInline{\textbf{Love MI}, \textbf{Huber W}, \textbf{Anders S}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Palarea2015}{}}%
-\CSLLeftMargin{34. }%
+\CSLLeftMargin{38. }%
 \CSLRightInline{\textbf{Palarea-Albaladejo J}, \textbf{Martin-Fernandez
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Li2015}{}}%
-\CSLLeftMargin{35. }%
+\CSLLeftMargin{39. }%
 \CSLRightInline{\textbf{Li Q}, \textbf{Heist EP}, \textbf{Moe LA}. 2015.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Baxter2016}{}}%
-\CSLLeftMargin{36. }%
+\CSLLeftMargin{40. }%
 \CSLRightInline{\textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Beall2015}{}}%
-\CSLLeftMargin{37. }%
+\CSLLeftMargin{41. }%
 \CSLRightInline{\textbf{Beall BFN}, \textbf{Twiss MR}, \textbf{Smith
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Henson2016}{}}%
-\CSLLeftMargin{38. }%
+\CSLLeftMargin{42. }%
 \CSLRightInline{\textbf{Henson MW}, \textbf{Pitre DM}, \textbf{Weckhorst
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Baxter2014}{}}%
-\CSLLeftMargin{39. }%
+\CSLLeftMargin{43. }%
 \CSLRightInline{\textbf{Baxter NT}, \textbf{Wan JJ}, \textbf{Schubert
@@
 \leavevmode\vadjust pre{\hypertarget{ref-LevyBooth2018}{}}%
-\CSLLeftMargin{40. }%
+\CSLLeftMargin{44. }%
 \CSLRightInline{\textbf{Levy-Booth DJ}, \textbf{Giesbrecht IJW},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Edwards2015}{}}%
-\CSLLeftMargin{41. }%
+\CSLLeftMargin{45. }%
 \CSLRightInline{\textbf{Edwards J}, \textbf{Johnson C},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Ettinger2017}{}}%
-\CSLLeftMargin{42. }%
+\CSLLeftMargin{46. }%
 \CSLRightInline{\textbf{Ettinger CL}, \textbf{Williams SL},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Graw2018}{}}%
-\CSLLeftMargin{43. }%
+\CSLLeftMargin{47. }%
 \CSLRightInline{\textbf{Graw MF}, \textbf{DAngelo G}, \textbf{Borchers
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Johnston2016}{}}%
-\CSLLeftMargin{44. }%
+\CSLLeftMargin{48. }%
 \CSLRightInline{\textbf{Johnston ER}, \textbf{Rodriguez-R LM},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Hassell2018}{}}%
-\CSLLeftMargin{45. }%
+\CSLLeftMargin{49. }%
 \CSLRightInline{\textbf{Hassell N}, \textbf{Tinker KA}, \textbf{Moore
@@
 the samples that were removed from each dataset are provided in Figure
-S1. This table is similar to Table 1 from (27).
+S1. This table is similar to Table 1 from (28).
 
@@
 \endhead
-Bioethanol~(35) & 95 & 3,970,972 & 16,014 & 41,799 &
+Bioethanol~(39) & 95 & 3,970,972 & 16,014 & 41,799 &
 3,690\Hyphdash*356,027 & SRP055545 \\
-Human~(36) & 490 & 20,828,275 & 32,452 & 42,506 &
+Human~(40) & 490 & 20,828,275 & 32,452 & 42,506 &
 10,439\Hyphdash*422,904 & SRP062005 \\
-Lake~(37) & 52 & 3,145,486 & 69,205 & 60,490 & 15,135\Hyphdash*110,993 &
+Lake~(41) & 52 & 3,145,486 & 69,205 & 60,490 & 15,135\Hyphdash*110,993 &
 SRP050963 \\
-Marine~(38) & 7 & 1,484,068 & 213,091 & 212,009 &
+Marine~(42) & 7 & 1,484,068 & 213,091 & 212,009 &
 132,895\Hyphdash*256,758 & SRP068101 \\
@@
 SRP192323 \\
-Peromyscus~(39) & 111 & 1,545,288 & 12,393 & 13,921 &
+Peromyscus~(43) & 111 & 1,545,288 & 12,393 & 13,921 &
 4,454\Hyphdash*33,502 & SRP044050 \\
-Rainforest~(40) & 69 & 936,666 & 11,464 & 13,574 & 4,880\Hyphdash*37,403
+Rainforest~(44) & 69 & 936,666 & 11,464 & 13,574 & 4,880\Hyphdash*37,403
 & ERP023747 \\
-Rice~(41) & 490 & 22,623,937 & 43,399 & 46,171 & 2,777\Hyphdash*192,200
+Rice~(45) & 490 & 22,623,937 & 43,399 & 46,171 & 2,777\Hyphdash*192,200
 & SRP044745 \\
-Seagrass~(42) & 286 & 4,135,440 & 13,538 & 14,459 &
+Seagrass~(46) & 286 & 4,135,440 & 13,538 & 14,459 &
 1,830\Hyphdash*45,076 & SRP092441 \\
-Sediment~(43) & 58 & 1,151,389 & 17,606 & 19,851 & 7,686\Hyphdash*67,763
+Sediment~(47) & 58 & 1,151,389 & 17,606 & 19,851 & 7,686\Hyphdash*67,763
 & SRP097192 \\
-Soil~(44) & 18 & 932,563 & 50,487 & 51,809 & 46,622\Hyphdash*58,935 &
+Soil~(48) & 18 & 932,563 & 50,487 & 51,809 & 46,622\Hyphdash*58,935 &
 ERP012016 \\
-Stream~(45) & 201 & 21,017,610 & 90,621 & 104,565 &
+Stream~(49) & 201 & 21,017,610 & 90,621 & 104,565 &
 8,931\Hyphdash*394,419 & SRP075852 \\
@@
 \subsection{Figures}\label{figures}}
-
-\includegraphics[height=17cm]{figure_1.png}
 
@@
 
-\newpage
-
-\includegraphics[height=17cm]{figure_2.png}
-
 \textbf{Figure 2. The risk of falsely detecting a difference between
@@
 
-\newpage
-
-\includegraphics[height=17cm]{figure_3.png}
-
 \textbf{Figure 3.The risk of falsely detecting a difference between
-treatment groups drawn from a null model does not meaningfully vary from
-5\% when data are normalized by rarefaction when sequencing depth is
+treatment groups drawn from a null model did not meaningfully vary from
+5\% when data are normalized by rarefaction when sequencing depth was
 confounded with treatment group}. Samples were assigned to different
@@
 yielded a significant result.
-
-\newpage
-
-\includegraphics[height=17cm]{figure_4.png}
 
@@
 
-\newpage
-
-\includegraphics{figure_5.png}
-
 \textbf{Figure 5. The ability to detect true differences in treatment
@@
 
-\newpage
-
-\includegraphics{figure_6.png}
-
 \textbf{Figure 6. The mean and coefficient of variation for richness,
@@
 
-\newpage
-
-\includegraphics{figure_7.png}
-
 \textbf{Figure 7. Most datasets are sequenced to a level that provides a
 high level of coverage.} Each plotting symbol represents the observed
-Good's coverage for a different sample in each dataset. The smoothed
-line indicates the simulated coverage for varying levels of sequencing
-effort when a null community is generated from the observed data. The
-box and whisker plot indicates the range of coverage values when the
-observed community data were normalized by rarefaction to the size of
-the least sequenced sample.
+Good's coverage for a different sample in each dataset. The box and
+whisker plot indicates the range of coverage values when the observed
+community data were normalized by rarefaction to the size of the least
+sequenced sample. The smoothed line indicates the simulated coverage for
+varying levels of sequencing effort when a null community is generated
+from the observed data.
 
 \newpage
-
-\includegraphics{figure_s1.png}
 
@@
 sample varied by dataset and was determined based on identifying natural
-breaks in the data. This figure is similar to Figure S1 of (27)
-
-\newpage
-
-\includegraphics{figure_s2.png}
+breaks in the data. This figure is similar to Figure S1 of (28)
 
@@
 test of whether the coefficient was significantly different from zero
-are indicated for each panel.
-
-\newpage
-
-\includegraphics{figure_s3.png}
+are indicated for each panel. The gray points for the iNEXT estimates
+indicate samples where richness values were estimated while the black
+points are those were rarefaction was used.
 
@@
 calculating the pairwise distance prior to each method's approach of
-controlling for uneven sequencing effort. The Spearman correlation
-coefficient (\(\rho\)) and test of whether the coefficient was
-significantly different from zero are indicated for each panel.
-
-\newpage
-
-\includegraphics[height=17cm]{figure_s4.png}
+controlling for uneven sequencing effort. The values on the y-axis for
+the Robust PCA distances were multiplied by 20 to fit them on the same
+scale as the other methods. The Spearman correlation coefficient
+(\(\rho\)) and test of whether the coefficient was significantly
+different from zero are indicated for each panel.
 
 \textbf{Figure S4.The risk of falsely detecting a difference between
-treatment groups drawn from a null model does not meaningfully vary from
-5\% when data are normalized by rarefaction when sequencing depth is
-confounded with treatment group}. Samples were assigned to different
-treatment groups based on whether they were above the median number of
-sequences for each dataset. To calculate the false detection rate,
-datasets were regenerated 100 times and differences in alpha diversity
-were tested using a Wilcoxon test (A) and differences in beta diversity
-were tested using PERMANOVA (B) at a 5\% threshold. The false positive
-rate was the number of times a dataset yielded a significant result.
+treatment groups drawn from a null model did not meaningfully vary from
+5\% when data are normalized by rarefaction when sequencing depth was
+completely confounded with treatment group}. Samples were assigned to
+different treatment groups based on whether they were above the median
+number of sequences for each dataset. To calculate the false detection
+rate, datasets were regenerated 100 times and differences in alpha
+diversity were tested using a Wilcoxon test (A) and differences in beta
+diversity were tested using PERMANOVA (B) at a 5\% threshold. The false
+positive rate was the number of times a dataset yielded a significant
+result.
 

### 2023-12-15 code/plot_alpha_beta_depth_correlation.R

@@ -162,16 +162,13 @@ beta_metrics <- c(
   "rare_euclidean",
   "raw_euclidean",
   "relabund_euclidean",
-  "rclr_euclidean",
   "oclr_euclidean",
   "nclr_euclidean",
   "zclr_euclidean",
+  "rclr_euclidean",
   "deseq2_euclidean"
 )
 
-# beta_classes <- c(
-#
-# )
 
 beta_labels <- c(
   "Rarefaction",
@@ -189,10 +186,10 @@ beta_labels <- c(
   "Rarefaction",
   "Raw",
   "Rel. abundance",
-  "Robust CLR",
   "One CLR",
   "Nudge CLR",
   "Zero CLR",
+  "Robust PCA",
   "DeSeq2"
 )
 
@@ -248,7 +245,7 @@ combo <- alpha + beta +
   plot_annotation(tag_levels = "A") &
   theme(
     plot.tag = element_text(size = 18, face = "bold",
-                            margin = margin(t = 0, r = -15, b = -15, l = 0)
+                            margin = margin(t = 0, r = -40, b = -15, l = 0)
                             )
     )
 

### 2023-12-15 code/plot_example_beta_cor.R

@@ -58,8 +58,8 @@ get_dists <- function(x) {
 distances <- map_dfr(dist_files, get_dists) %>%
   mutate(process = factor(process,
                           levels = c("rare", "raw", "relabund", "srs",
-                                     "metagenomeseq", "rclr", "oclr", "nclr",
-                                     "zclr", "deseq2")))
+                                     "metagenomeseq", "oclr", "nclr",
+                                     "zclr", "rclr", "deseq2")))
 
 correlations <- distances %>%
   nest(data = -c(process, calculator)) %>%
@@ -109,10 +109,10 @@ b_beta_labels <- c(
 )
 
 e_beta_labels <- c(
-  rclr = glue("Robust CLR (x20)\n(\u03C1 = {get_rho('rclr', 'euclidean')})"),
   oclr = glue("One CLR\n(\u03C1 = {get_rho('oclr', 'euclidean')})"),
   nclr = glue("Nudge CLR\n(\u03C1 = {get_rho('nclr', 'euclidean')})"),
   zclr = glue("Zero CLR\n(\u03C1 = {get_rho('zclr', 'euclidean')})"),
+  rclr = glue("Robust PCA (x20)\n(\u03C1 = {get_rho('rclr', 'euclidean')})"),
   deseq2 = glue("DeSeq2\n(\u03C1 = {get_rho('deseq2', 'euclidean')})")
 )
 

### 2023-12-15 code/plot_false_positive_null_model.R

@@ -116,8 +116,9 @@ alpha_composite <- map_dfr(alpha_summary_files, read_tsv, .id = "dataset") %>%
          )
 
 alpha <- alpha_composite %>%
-  ggplot(aes(x = metric_method, y = frac, color = dataset, shape = dataset)) +
-    geom_hline(yintercept = 0.05, size = 0.25, color = "gray") +
+  mutate(percentage = 100 * frac) %>%
+  ggplot(aes(x = metric_method, y = percentage, color = dataset, shape = dataset)) +
+    geom_hline(yintercept = 5, size = 0.25, color = "gray") +
     geom_point(position = position_dodge(width = 0.3), fill = "white") +
     facet_grid(. ~ class,
                scales = "free_x", space = "free_x",
@@ -134,8 +135,8 @@ alpha <- alpha_composite %>%
                        labels = pretty_datasets$pretty
                        ) +
     scale_x_discrete(breaks = alpha_metrics, labels = alpha_labels) +
-    scale_y_continuous(limits = c(0, 0.15), breaks = seq(0, 0.15, 0.05)) +
-    labs(x = NULL, y = "False positive rate") +
+    scale_y_continuous(limits = c(0, 15), breaks = seq(0, 15, 5)) +
+    labs(x = NULL, y = "False positive rate (%)") +
     theme(
       axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
       panel.background = element_rect(fill = "white"),
@@ -165,18 +166,12 @@ beta_metrics <- c(
   "rare_euclidean",
   "raw_euclidean",
   "relabund_euclidean",
-  "rclr_euclidean",
   "oclr_euclidean",
   "nclr_euclidean",
   "zclr_euclidean",
+  "rclr_euclidean",
   "deseq2_euclidean"
-  # "srs_euclidean",
-  # "metagenomeseq_euclidean"
 )
-
-# beta_classes <- c(
-#
-# )
 
 beta_labels <- c(
   "Rarefaction",
@@ -194,10 +189,10 @@ beta_labels <- c(
   "Rarefaction",
   "Raw",
   "Rel. abundance",
-  "Robust CLR",
   "One CLR",
   "Nudge CLR",
   "Zero CLR",
+  "Robust PCA",
   "DeSeq2"
 )
 
@@ -220,9 +215,10 @@ beta_composite <- map_dfr(beta_summary_files, read_tsv, .id = "dataset") %>%
 
 beta <- beta_composite %>%
   drop_na(significant) %>%
[…]

### 2023-12-15 code/plot_false_positive_null_model_size.R

@@ -119,8 +119,9 @@ alpha_composite <- map_dfr(alpha_summary_files, read_tsv, .id = "dataset") %>%
          )
 
 alpha <- alpha_composite %>%
-  ggplot(aes(x = metric_method, y = frac, color = dataset, shape = dataset)) +
-    geom_hline(yintercept = 0.05, size = 0.25, color = "gray") +
+  mutate(percentage = 100 * frac) %>%
+  ggplot(aes(x = metric_method, y = percentage, color = dataset, shape = dataset)) +
+    geom_hline(yintercept = 5, size = 0.25, color = "gray") +
     geom_point(position = position_dodge(width = 0.3), fill = "white") +
     facet_grid(. ~ class,
                scales = "free_x", space = "free_x",
@@ -138,7 +139,7 @@ alpha <- alpha_composite %>%
                        ) +
     scale_x_discrete(breaks = alpha_metrics, labels = alpha_labels) +
     # scale_y_continuous(limits = c(0, NA), breaks = seq(0, 0.15, 0.05)) +
-    labs(x = NULL, y = "False positive rate") +
+    labs(x = NULL, y = "False positive rate (%)") +
     theme(
       axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
       panel.background = element_rect(fill = "white"),
@@ -168,16 +169,13 @@ beta_metrics <- c(
   "rare_euclidean",
   "raw_euclidean",
   "relabund_euclidean",
-  "rclr_euclidean",
   "oclr_euclidean",
   "nclr_euclidean",
   "zclr_euclidean",
+  "rclr_euclidean",
   "deseq2_euclidean"
 )
 
-# beta_classes <- c(
-#
-# )
 
 beta_labels <- c(
   "Rarefaction",
@@ -195,10 +193,10 @@ beta_labels <- c(
   "Rarefaction",
   "Raw",
   "Rel. abundance",
-  "Robust CLR",
   "One CLR",
   "Nudge CLR",
   "Zero CLR",
+  "Robust PCA",
   "DeSeq2"
 
 )
@@ -222,9 +220,10 @@ beta_composite <- map_dfr(beta_summary_files, read_tsv, .id = "dataset") %>%
 
 beta <- beta_composite %>%
   drop_na(significant) %>%
+  mutate(significant = 100 * significant) %>%
   ggplot(aes(x = process_calc, y = significant,
              color = dataset, shape = dataset)) +
[…]

### 2023-12-15 code/plot_intrasample_variation.R

@@ -109,7 +109,7 @@ plot_method_statistic <- function(m, s, data = composite,
     ggplot(aes(x = n_seqs, y = value,
                shape = dataset, group = dataset, color = dataset)) +
     geom_line() +
-    geom_point(data = filtered_point, fill = "white") +
+    geom_point(data = filtered_point, fill = "white", size = 2) +
     scale_y_continuous(limits = c(0, NA)) +
     scale_x_log10(breaks = c(1e3, 1e4, 1e5),
                   labels = c("10<sup>3</sup>", "10<sup>4</sup>",

### 2023-12-15 code/plot_power_cffect_model.R

@@ -116,10 +116,8 @@ alpha_composite <- map_dfr(alpha_summary_files, read_tsv, .id = "dataset") %>%
          )
 
 alpha <- alpha_composite %>%
-  ggplot(aes(x = metric_method, y = frac, color = dataset, shape = dataset)) +
-    # geom_hline(yintercept = 0.05, size = 0.25, color = "gray") +
-    # geom_line(aes(group = dataset), position = position_dodge(width = 0.3),
-    #           color = "gray", size = 0.1) +
+  mutate(percent = 100 * frac) %>%
+  ggplot(aes(x = metric_method, y = percent, color = dataset, shape = dataset)) +
     geom_point(position = position_dodge(width = 0.3), fill = "white") +
     facet_grid(. ~ class,
                scales = "free_x", space = "free_x",
@@ -136,8 +134,8 @@ alpha <- alpha_composite %>%
                        labels = pretty_datasets$pretty
                        ) +
     scale_x_discrete(breaks = alpha_metrics, labels = alpha_labels) +
-    # scale_y_continuous(limits = c(0, NA), breaks = seq(0, 0.15, 0.05)) +
-    labs(x = NULL, y = "Power") +
+    scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 25)) +
+    labs(x = NULL, y = "Power (%)") +
     theme(
       axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
       panel.background = element_rect(fill = "white"),

### 2023-12-15 code/plot_power_effect_model.R

@@ -115,10 +115,8 @@ alpha_composite <- map_dfr(alpha_summary_files, read_tsv, .id = "dataset") %>%
          )
 
 alpha <- alpha_composite %>%
-  ggplot(aes(x = metric_method, y = frac, color = dataset, shape = dataset)) +
-    # geom_hline(yintercept = 0.05, size = 0.25, color = "gray") +
-    # geom_line(aes(group = dataset), position = position_dodge(width = 0.3),
-    #           color = "gray", size = 0.1) +
+  mutate(percentage = 100 * frac) %>%
+  ggplot(aes(x = metric_method, y = percentage, color = dataset, shape = dataset)) +
     geom_point(position = position_dodge(width = 0.3), fill = "white") +
     facet_grid(. ~ class,
                scales = "free_x", space = "free_x",
@@ -135,15 +133,14 @@ alpha <- alpha_composite %>%
                        labels = pretty_datasets$pretty
                        ) +
     scale_x_discrete(breaks = alpha_metrics, labels = alpha_labels) +
-    # scale_y_continuous(limits = c(0, NA), breaks = seq(0, 0.15, 0.05)) +
-    labs(x = NULL, y = "Power") +
+    labs(x = NULL, y = "Power (%)") +
     theme(
       axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
       panel.background = element_rect(fill = "white"),
       panel.grid = element_blank(),
       axis.ticks = element_blank(),
       legend.key = element_blank(),
-      legend.margin = margin(0,0,0,0),
+      legend.margin = margin(0, 0, 0, 0),
       legend.key.size = unit(12, "pt"),
       axis.line.x = element_line()
     )
@@ -169,10 +166,10 @@ beta_metrics <- c(
   "rare_euclidean",
   "raw_euclidean",
   "relabund_euclidean",
-  "rclr_euclidean",
   "oclr_euclidean",
   "nclr_euclidean",
   "zclr_euclidean",
+  "rclr_euclidean",
   "deseq2_euclidean"
 )
 
@@ -196,10 +193,10 @@ beta_labels <- c(
   "Rarefaction",
   "Raw",
   "Rel. abundance",
-  "Robust CLR",
   "One CLR",
   "Nudge CLR",
   "Zero CLR",
+  "Robust PCA",
   "DeSeq2"
 )
 
@@ -223,9 +220,9 @@ beta_composite <- map_dfr(beta_summary_files, read_tsv, .id = "dataset") %>%
 
[…]

### 2023-12-15 submission/manuscript.Rmd
Final polish before submitting

@@
 
-Considering it is common to find as much as 100-fold variation in the number of 16S rRNA gene sequences found across samples in a study, researchers need to control for the effect of uneven sequencing effort.
+Considering it is common to find as much as 100-fold variation in the number of 16S rRNA gene sequences across samples in a study, researchers need to control for the effect of uneven sequencing effort.
 How to do this has become a contentious question.

### 2023-12-15 submission/manuscript.tex
Final polish before submitting

@@
 Considering it is common to find as much as 100-fold variation in the
-number of 16S rRNA gene sequences found across samples in a study,
-researchers need to control for the effect of uneven sequencing effort.
-How to do this has become a contentious question. Some have argued that
-rarefying or rarefaction is ``inadmissible'' because it omits valid
-data. A number of alternative approaches have been developed to
-normalize and rescale the data that purport to be invariant to the
-number of observations. I generated community distributions based on 12
-published datasets where I was able to assess the ability of multiple
-methods to control for uneven sequencing effort. Rarefaction was the
-only method that could control for variation in uneven sequencing effort
-when measuring commonly used alpha and beta diversity metrics. Next, I
-compared the false detection rate and power to detect true differences
-between simulated communities with a known effect size using various
-alpha and beta diversity metrics. Although all methods of controlling
-for uneven sequencing effort had an acceptable false detection rate when
-samples were randomly assigned to two treatment groups, rarefaction was
-consistently able to control for differences in sequencing effort when
-sequencing depth was confounded with treatment group. Finally, the
-statistical power to detect differences in alpha and beta diversity
-metrics was consistently the highest when using rarefaction. These
-simulations underscore the importance of using rarefaction to normalize
-the number of sequences across samples in amplicon sequencing analyses.
+number of 16S rRNA gene sequences across samples in a study, researchers
+need to control for the effect of uneven sequencing effort. How to do
+this has become a contentious question. Some have argued that rarefying
+or rarefaction is ``inadmissible'' because it omits valid data. A number
+of alternative approaches have been developed to normalize and rescale
+the data that purport to be invariant to the number of observations. I
+generated community distributions based on 12 published datasets where I
+was able to assess the ability of multiple methods to control for uneven
+sequencing effort. Rarefaction was the only method that could control
+for variation in uneven sequencing effort when measuring commonly used
+alpha and beta diversity metrics. Next, I compared the false detection
+rate and power to detect true differences between simulated communities
+with a known effect size using various alpha and beta diversity metrics.
+Although all methods of controlling for uneven sequencing effort had an
+acceptable false detection rate when samples were randomly assigned to
+two treatment groups, rarefaction was consistently able to control for
+differences in sequencing effort when sequencing depth was confounded
+with treatment group. Finally, the statistical power to detect
+differences in alpha and beta diversity metrics was consistently the
+highest when using rarefaction. These simulations underscore the
+importance of using rarefaction to normalize the number of sequences
+across samples in amplicon sequencing analyses.
 
